### 2015-10-01 Manuscript/carkeek_grid.tex
first commit

%--------------------------------------||--------------------------------------%

\documentclass[11pt,letterpaper]{article} %  font size

%-----------------------------------PACKAGES-----------------------------------%

\usepackage[T1]{fontenc} % Choose an output font encoding (T1) that has support for the accented characters used by the most widespread European languages
\usepackage[utf8]{inputenc} % Allow input of accented characters (and more...)
\usepackage{graphicx} %  figures
\usepackage[round]{natbib} % names in citations
\usepackage{lineno} % line numbers
\usepackage{authblk} % allows more intuitive formatting for multiple authors/affiliations
\usepackage[margin=1in]{geometry} % make the margins 1 inch on all sides of the document
%\usepackage{amsmath} % useful for formatting math stuff, especially complex equations
%\usepackage{pdflscape} % rotate table into landscape mode
%\usepackage{subfigure} % side by side figures
%\usepackage{longtable} % for tables that span multiple pages.
\usepackage{setspace} % double spacing

%----------------------------------FORMATTING-----------------------------------

%\topmargin -1cm %0.0cm
%\textwidth 16cm % what does this do?
%\textheight 21cm % what does this do?
%\footskip 1.0cm % what does this do?
%\oddsidemargin 0.0cm

\date{\today}
\doublespacing % initiate double spacing (package setspace)
%\linespread{2} % alternate method of double spacing

%------------------------------------TITLE--------------------------------------

\title{There once was a grid at ol' Carkeek}

%-----------------------------------AUTHORS-------------------------------------

% using package 'authblk':
\author[1]{First Author\thanks{first.author@funstuff.com}}

\author[1,2]{Second Author}

\author[2]{Third Author}

\affil[1]{Department of Computer Science, \LaTeX\ University}
\affil[2]{Department of Mechanical Engineering, Superfabulous University}

%----------------------------------FORMATTING-----------------------------------
\begin{document}
\maketitle
\linenumbers % start line numbers
\def\linenumberfont{\normalfont\small\rmfamily} % change line number font

%----------------------------------KEYWORDS-------------------------------------
\section*{Keywords}
Stuff, things, neat, cool, wow, instafun, tags4likes, etc

%----------------------------------ABSTRACT-------------------------------------
\section*{Abstract}
This is the text of the abstract.

%---------------------------------INTRODUCTION----------------------------------
\section*{Introduction}
For centuries, humankind has wondered: If I have two apples, and someone gives me another two apples, how many apples do I have? Some people did this \citep{Darwin1859}.

%-----------------------------------METHODS-------------------------------------
\section*{Methods}
We use the general framework outlined by Shelton et al (CITE).
That study outlined the structure for estimation of the proportional biomass of a taxon ($B_i$) given the proportional counts of sequences recovered from a parallel sequencing run ($Z_i$).
We constructed the following mathematical model (\ref{my_equation_label}) to better understand the concept:
% using '\ref{something}' in the text will refer to any object (e.g. figure, equation, table) which contains the corresponding '\label{something}'
% Note that you need to run latex a few times to get it to register numbers correctly

\begin{equation}\label{my_equation_label}
	Y = 2a + 2a
\end{equation}

Where $a$ represents an apple.

%-----------------------------------RESULTS-------------------------------------
\section*{Results}
We found that if you have two apples, and someone gives you another two apples, you have four apples.

%----------------------------------DISCUSSION-----------------------------------
\section*{Discussion}
Boy those results sure are neat. Now, the pressing question becomes: How do you like them apples?

%-------------------------------ACKNOWLEDGEMENTS--------------------------------
\section*{Acknowledgements}
We wish to thank all of the little people.

%-----------------------------------FUNDING-------------------------------------
\section*{Funding}
This study was funded by our super-rich uncle.

%--------------------------------CONTRIBUTIONS----------------------------------
\section*{Author Contributions}
Conceived and designed the experiments: .
Collected the data: .
Conducted the analyses: .
Wrote the first draft: .
Edited the manuscript: .

%-------------------------------------DATA-------------------------------------%
\section*{Data Availablity}
% Are the data and code available in a permanent, publicly accessible data archive or repository?
The data and code used to generate our results can be found at the following url:

%----------------------------------REFERENCES-----------------------------------
%\section*{References} % commented out because the section title is automatically inserted if using an automatically-generated bibliography

\bibliographystyle{apalike} % or: plain,unsrt,alpha,abbrv,acm,apalike,ieeetr
\bibliography{/Users/threeprime/Documents/Publications/bibtex/library} % path to your .bib file excluding .bib extension (e.g. /Users/threeprime/Documents/Publications/bibtex/library)

%-----------------------------------FIGURES-------------------------------------
\section*{Figures}

%\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
%  \centering
%    \includegraphics[width=1\textwidth]{figure_filename.pdf}
%    \caption{This is the figure caption.}
%  \label{myfigure} % use this to refer to your figure in the text, so that numbering updates automatically
%\end{figure}

%----------------------------------SUPPLEMENT-----------------------------------
%\section*{Supplemental Material}

\end{document}

### 2015-10-01 Analysis/Ole/eDNA_grid_model.R

@@ -1,0 +1,236 @@
+#!/usr/bin/env Rscript
+
+# setwd("")
+library(R2jags)
+
+# This is code to estimate some model parameters for proportional abundance of DNA in a sample given reads in the file from a sequencer
+
+
+# Let:
+# i = taxon i (1:I); Carkeek grid dataset I = ? (limit to 10 for tests)
+# j = PCR replicate j (1:J); Carkeek grid dataset J = 4
+# k = location k (1:K); Carkeek grid dataset K = 24
+
+
+# Requires as input:
+# 1. otu file
+# 2. metadata file
+
+# a table of counts of sequences (Z, length = ...)
+counts_table <- read.csv("OTUs_top10_4000m.csv", row.names = 1)
+# rows/rownames = samples, columns/colnames = taxa, cells = integer counts
+# if table is incorrectly oriented, transpose it:
+# counts_table <- as.data.frame(t(counts_table))
+
+# load metadata
+metadata <- read.csv("metadata_spatial.csv", stringsAsFactors = FALSE)
+
+# what is the name of the column containing the sample id (SEQUENCING samples)
+colname_sampleid <- "sample_id"
+
+# what is the name of the column containing PCR replicate levels
+colname_pcr <- "PCR_replicate"
+
+# what is the name of the column containing relevant site replicate levels
+colname_site <- "transect"
+
+# of each taxon (length = I)
+taxa <- colnames(counts_table)
+
+# from each PCR replicate (length = J)
+pcr <- unique(metadata[, colname_pcr])
+
+# at each site (length = K)
+sites <- unique(metadata[, colname_site])
+
+
+
+
+
+
+
+
+# JAGS won't do some basic R functions, so we need to give it the lengths explicitly
+N_taxa <- length(taxa)
+N_pcr <- length(pcr)
+N_sites <- length(sites)
+
+
+# make a vector of the site for each sample name in the counts table
+site_by_rowname <- metadata[ , colname_site][
+  match(rownames(counts_table), metadata[ , colname_sampleid])
+]
+
+# unused for now
+# sampleid_by_site <- split(x = metadata[, colname_sampleid], f = metadata[, colname_site], drop = TRUE)
+
+counts_by_site <- split(x = counts_table, f = site_by_rowname)
+
+
+
+
+
+
+# create an array for JAGS called "counts"
[…]

### 2015-10-01 Analysis/data_prep.R

@@ -1,0 +1,52 @@
+# Manipulate data in preparation for modeling in JAGS
+
+filename_orig <- "/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Data/OTUs_fam_w30_top10.csv"
+
+filename_out <- file.path(dirname(filename_orig), "top10_OTUs.csv")
+
+metadata <- read.csv("/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Data/metadata_spatial.csv")
+
+column_name <- "sample_id"
+
+table_orig <- read.csv(filename_orig, row.names = 1)
+
+
+# restrict to only samples in the metadata file
+# it will really screw things up if you don't convert the sample id column to a character vector!
+# table_new <- table_orig[,metadata$sample_id]
+table_new <- table_orig[,as.character(metadata$sample_id)]
+
+
+# write the file
+write.csv(x = table_new, file = filename_out, quote = FALSE)
+
+
+# --------------- DATA TRIMMING ------------------------------------------
+# a table of counts of sequences (Z, length = ...)
+counts_table <- read.csv("/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Data/OTUs_top10.csv", row.names = 1)
+# rows/rownames = samples, columns/colnames = taxa, cells = integer counts
+# if table is incorrectly oriented, transpose it:
+counts_table <- as.data.frame(t(counts_table))
+
+# load metadata
+metadata <- read.csv("/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Data/metadata_spatial.csv", stringsAsFactors = FALSE)
+
+# what is the name of the column containing the sample id (SEQUENCING samples)
+colname_sampleid <- "sample_id"
+
+# what is the name of the column containing distance from shore data 
+colname_dist <- "dist_from_shore"
+
+# what distance should we look at?
+dist_of_interest <- 4000
+
+# subset to only include samples taken at 4000m from shore
+metadata <- droplevels(metadata[ metadata[ , colname_dist] == dist_of_interest , ])
+
+# the ids of the sequenced samples we care about are given by
+relevant_sampleids <- metadata[ , colname_sampleid]
+
+# the counts that we care about are given by:
+counts_table <- counts_table[relevant_sampleids , ]
[…]

### 2015-10-01 Analysis/eDNA_grid_model.R

@@ -1,0 +1,235 @@
+#!/usr/bin/env Rscript
+
+library(R2jags)
+
+# This is code to estimate some model parameters for proportional abundance of DNA in a sample given reads in the file from a sequencer
+
+
+# Let:
+# i = taxon i (1:I); Carkeek grid dataset I = ? (limit to 10 for tests)
+# j = PCR replicate j (1:J); Carkeek grid dataset J = 4
+# k = location k (1:K); Carkeek grid dataset K = 24
+
+
+# Requires as input:
+# 1. otu file
+# 2. metadata file
+
+# a table of counts of sequences (Z, length = ...)
+counts_table <- read.csv("/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Data/OTUs_top10_4000m.csv", row.names = 1)
+# rows/rownames = samples, columns/colnames = taxa, cells = integer counts
+# if table is incorrectly oriented, transpose it:
+# counts_table <- as.data.frame(t(counts_table))
+
+# load metadata
+metadata <- read.csv("/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Data/metadata_spatial.csv", stringsAsFactors = FALSE)
+
+# what is the name of the column containing the sample id (SEQUENCING samples)
+colname_sampleid <- "sample_id"
+
+# what is the name of the column containing PCR replicate levels
+colname_pcr <- "PCR_replicate"
+
+# what is the name of the column containing relevant site replicate levels
+colname_site <- "transect"
+
+# of each taxon (length = I)
+taxa <- colnames(counts_table)
+
+# from each PCR replicate (length = J)
+pcr <- unique(metadata[, colname_pcr])
+
+# at each site (length = K)
+sites <- unique(metadata[, colname_site])
+
+
+
+
+
+
+
+
+# JAGS won't do some basic R functions, so we need to give it the lengths explicitly
+N_taxa <- length(taxa)
+N_pcr <- length(pcr)
+N_sites <- length(sites)
+
+
+# make a vector of the site for each sample name in the counts table
+site_by_rowname <- metadata[ , colname_site][
+  match(rownames(counts_table), metadata[ , colname_sampleid])
+]
+
+# unused for now
[…]

### 2015-10-01 Analysis/eDNA_model_single_env_sample.R

@@ -1,0 +1,219 @@
+#!/usr/bin/env Rscript
+
+library(R2jags)
+
+# This is code to estimate some model parameters for proportional abundance of DNA in a sample given reads in the file from a sequencer
+
+
+# Let:
+# i = taxon i (1:I); Carkeek grid dataset I = ? (limit to 10 for tests)
+# j = PCR replicate j (1:J); Carkeek grid dataset J = 4
+# k = location k (1:K); Carkeek grid dataset K = 24
+
+
+# Requires as input:
+# 1. otu file
+# 2. metadata file
+
+# a table of counts of sequences (Z, length = ...)
+counts_table <- read.csv("/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Data/OTUs_top10_4000m.csv", row.names = 1)
+# rows/rownames = samples, columns/colnames = taxa, cells = integer counts
+# if table is incorrectly oriented, transpose it:
+# counts_table <- as.data.frame(t(counts_table))
+
+# load metadata
+metadata <- read.csv("/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Data/metadata_spatial.csv", stringsAsFactors = FALSE)
+
+# what is the name of the column containing the sample id (SEQUENCING samples)
+colname_sampleid <- "sample_id"
+
+# what is the name of the column containing PCR replicate levels
+colname_pcr <- "PCR_replicate"
+
+# what is the name of the column containing relevant site replicate levels
+colname_site <- "transect"
+
+# of each taxon (length = I)
+taxa <- colnames(counts_table)
+
+# from each PCR replicate (length = J)
+pcr <- unique(metadata[, colname_pcr])
+
+# at each site (length = K)
+sites <- unique(metadata[, colname_site])
+
+
+
+
+
+
+
+
+# JAGS won't do some basic R functions, so we need to give it the lengths explicitly
+N_taxa <- length(taxa)
+N_pcr <- length(pcr)
+N_sites <- length(sites)
+
+
+# make a vector of the site for each sample name in the counts table
+site_by_rowname <- metadata[ , colname_site][
[…]

### 2015-10-01 Analysis/general_framework/Analysis/BHM_eDNA.R

@@ -1,0 +1,196 @@
+# Estimation of some underlying/unobserved quantity from some other observed quantity
+# (e.g. relative abundance from eDNA counts)
+
+# Set this to the analysis directory of this project
+analysis_dir <- "~/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Analysis"
+
+
+
+setwd(analysis_dir)
+data_dir <- file.path("..", "Data")
+fig_dir <- file.path("..", "Figures")
+
+
+############################################################################
+# The following requires as input:
+# 1. A matrix of counts (integers) of DNA sequences, where:
+# 		rows = samples
+# 		columns = taxa whose names are given in column names
+# eg: dput(mydata) 
+# structure(c(18094L, 18136L, 21198L, 13643L, 14198L, 12219L, 12703L, 
+# 13873L, 11773L, 12022L, 11865L, 13554L, 13840L, 9871L, 10999L, 
+# 10070L, 10827L, 8920L, 9708L, 10033L, 9303L, 6802L, 7853L, 7927L, 
+# 3118L, 3344L, 4107L), .Dim = c(3L, 9L), .Dimnames = list(NULL, 
+    # c("taxon_A", "taxon_B", "taxon_C", "taxon_D", "taxon_E", 
+    # "taxon_F", "taxon_G", "taxon_H", "taxon_I")))
+
+
+# If loading from a file
+# counts <- as.matrix(read.csv(file = file.path(data_dir, "counts.csv")))
+# taxon_names <- colnames(counts)
+
+# If reading from a currently existing object (e.g. you simulated some eDNA data)
+# counts <- as.matrix(mydata)
+
+# make into a vector
+counts_vec <- as.vector(counts)
+names(counts_vec) <- rep(x = colnames(counts), each = nrow(counts))
+
+
+
+
+############
+# Create Design Matrix for the categorical effects
+############
+
+# Should have nrow = length of the input 
+
+cat_mat <- matrix(data = 0, nrow = length(counts_vec), ncol = ncol(counts))
+
+# Not sure why, but you set the first species to equal 1...
+# I think this might be because everything else is expressed relative to this?
+cat_mat[,1]	<- 1
+
+# Changed Ole's original version because it will break if you don't use the "right" taxon names (i.e. with an integer corresponding to their index in the third position == A.3)
+for(i in 1:length(taxon_names)){
[…]

### 2015-10-01 Analysis/general_framework/Analysis/original OS - eDNA Pipers v4_JO.r

@@ -1,0 +1,1801 @@
+rm(list=ls())
+### GO GET THE NECESSARY LIBRARIES
+
+##### TILAPIA IS OTU 4 ####
+
+library(gtools)
+library(R2jags)
+
+# Set this to the subdirectory Analysis within the project folder (which should also contain subdirectories called Data and Figures)
+# analysis_dir <- "/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Analysis"
+
+
+# data_dir <- file.path("..", "Data")
+# fig_dir <- file.path("..", "Figures")
+
+setwd("/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Analysis/general_framework/Data")
+
+
+
+
+
+
+
+#------------------------------------------
+## Read in the clusters data + sort to make long form
+# this file is a matrix of counts (integers) where
+# rows = taxa
+# columns = samples
+# AND
+# columns = distance measure from taxon's primer binding site and each of the tags (primer indexes)
+# for original modeling paper, dim =  9,151; only top 10 OTUs
+dat		<-	read.csv("abundance_distance_matrix+OLE.csv")
+dat		<-	dat[dat$X!='OTU_4',] # remove Tilapia # no need once tables are separated appropriately
+
+# OTU_table.csv
+# primer_mismatch_both.csv
+# primer_mismatch_F.csv
+# primer_mismatch_R.csv
+#------------------------------------------
+
+
+
+
+
+
+
+
+
+#------------------------------------------
+# Read in the "duplicate table":
+# this file is a matrix of counts (integers)
+# rows = unique DNA sequences ("DUP")
+# columns = samples
+# dim = (27973,75); 75 samples, 27973 unique sequences
+dat.total.DNA	<-	read.csv("all_clusters.csv", header=TRUE, row.names = 1)
+
+# duplicate_table.csv
+#------------------------------------------
+
+# calculate the total number of sequence counts per sample
+dat.DNA			<-	colSums(dat.total.DNA)
+
+
+
+
+
+
+
+#------------------------------------------
+# Read in the sequencing metadata
+# contains primer index data, environmental sample name, etc
+dat.surv	<-	read.csv("sample_data.csv")
+# sequencing_metadata.csv
+#------------------------------------------
+
+
+# isolate the primer index sequence ("tag") and the sample name
[…]

### 2015-10-01 Analysis/simulate_eDNA_counts.R

@@ -1,0 +1,158 @@
+############################################################################
+# This script simulates some data that resembles eDNA sequence data
+############################################################################
+
+# Original code by Ole Shelton
+# Hacked up by Jimmy O'Donnell
+# Comments mostly reflect Jimmy's imperfect understanding of the underlying concepts.
+
+library(gtools) # gtools::rdirichlet
+library(R2jags) # R2jags::jags
+
+
+
+# Suppose we have an aquarium which contains N_tax taxa
+N_tax <- 9
+
+
+
+# Give them names
+taxon_names <- paste("taxon_", LETTERS[1:N_tax], sep = "")
+
+
+# The distribution of DNA molecules derived from each of the taxa might look like the following
+# Ole originally called these alpha because they eventually become the alpha parameter of the Dirichlet distribution, which sorta determines how likely each component is to be drawn
+# I call these "true" because they represent the "true" abundance of DNA from each taxon in the sample.
+# Set seed to be able to reproduce draws from a probability distribution (gets reset each time one of these functions is used)
+# Instead of runif, you might try using rexp or random from the Pareto distribution to more accurately resemble "real" eDNA results
+set.seed(407744)
+DNA_truth		<- sort(round(runif(N_tax,1,100)), decreasing = TRUE)
+names(DNA_truth) <- taxon_names
+
+# those same values expressed as proportions
+DNA_truth_prop <- DNA_truth/sum(DNA_truth)
+
+
+
+
+
+############################################################################
+# The actual simulation 
+############################################################################
+
+# This simulates data that might be expected from high throughput sequencing of PCR amplicons generated from environmental samples: proportional abundance of sequences from each of the taxa/OTUs/dups
+# draw samples from the Dirichlet distribution, using alpha given by DNA_truth to approximate the sampling of DNA molecules from the environment
+# Set seed to be able to reproduce draws from a probability distribution (gets reset each time one of these functions is used)
+
+
+# If you were to take N_samples samples, what kind of answer would you expect?
+N_samples_small	<-	3
+set.seed(407744)
+eDNA_counts_small <- rdirichlet(N_samples_small, DNA_truth)
+colnames(eDNA_counts_small) <- taxon_names
+
+
+############
+# OR...
[…]

### 2015-10-06 Analysis/eDNA_grid_model.R
many changes to attemp to get model to run

@@ -1,9 +1,13 @@
 #!/usr/bin/env Rscript
 
-library(R2jags)
-
 # This is code to estimate some model parameters for proportional abundance of DNA in a sample given reads in the file from a sequencer
 
+library(R2jags)
+
+setwd("~/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Analysis")
+
+data_dir <- file.path("..", "Data")
+fig_dir <- file.path("..", "Figures")
 
 # Let:
 # i = taxon i (1:I); Carkeek grid dataset I = ? (limit to 10 for tests)
@@ -16,13 +20,15 @@ library(R2jags)
 # 2. metadata file
 
 # a table of counts of sequences (Z, length = ...)
-counts_table <- read.csv("/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Data/OTUs_top10_4000m.csv", row.names = 1)
+counts_file_path <- file.path(data_dir, "OTUs_top10_4000m.csv")
+counts_table <- read.csv(file = counts_file_path, row.names = 1)
 # rows/rownames = samples, columns/colnames = taxa, cells = integer counts
 # if table is incorrectly oriented, transpose it:
 # counts_table <- as.data.frame(t(counts_table))
 
 # load metadata
-metadata <- read.csv("/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Data/metadata_spatial.csv", stringsAsFactors = FALSE)
+metadata_file_path <- file.path(data_dir, "metadata_spatial.csv")
+metadata <- read.csv(file = metadata_file_path, stringsAsFactors = FALSE)
 
 # what is the name of the column containing the sample id (SEQUENCING samples)
 colname_sampleid <- "sample_id"
@@ -88,31 +94,24 @@ counts <- array(
 # }
 
 # thus, you should be able to reference stuff like so:
-counts[        ,         , sites[2] ]
 counts[ pcr[4] ,         ,          ]
 counts[        , taxa[8] ,          ]
-
-
-
+counts[        ,         , sites[2] ]
 
 # lambda = mean of the Poisson dist that describes variation in counts
 # beta = intercept
-# eta = random effect for i,j
+# eta = random effect for i,j,k
[…]

### 2015-10-06 Analysis/eDNA_model_single_env_sample.R

@@ -1,9 +1,13 @@
 #!/usr/bin/env Rscript
 
-library(R2jags)
-
 # This is code to estimate some model parameters for proportional abundance of DNA in a sample given reads in the file from a sequencer
 
+library(R2jags)
+
+setwd("~/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Analysis")
+
+data_dir <- file.path("..", "Data")
+fig_dir <- file.path("..", "Figures")
 
 # Let:
 # i = taxon i (1:I); Carkeek grid dataset I = ? (limit to 10 for tests)
@@ -16,7 +20,8 @@ library(R2jags)
 # 2. metadata file
 
 # a table of counts of sequences (Z, length = ...)
-counts_table <- read.csv("/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Data/OTUs_top10_4000m.csv", row.names = 1)
+counts_file_path <- file.path(data_dir, "OTUs_top10_4000m.csv")
+counts_table <- read.csv(file = counts_file_path, row.names = 1)
 # rows/rownames = samples, columns/colnames = taxa, cells = integer counts
 # if table is incorrectly oriented, transpose it:
 # counts_table <- as.data.frame(t(counts_table))
@@ -100,16 +105,10 @@ counts[        ,         , sites[2] ]
 # epsilon = random effect for i,k
 # tau2 = variance of normal distribution describing epsilon (attributable to location - k)
 
-# gamma = coefficient of something, I think -- maybe the covariate matrix?
-# phi =
-
 # pi = proportional abundance of each taxon
 
-# counts <- counts_table
+model_loc <- "eDNA_model_single_env_sample.txt"
 
-model_loc <- c("jags_model_multi_location.txt")
-
-jags.model()
 jagsscript <- cat(
 "
   model {
@@ -124,14 +123,18 @@ jagsscript <- cat(
 
       # GLM
       lambda[j,i,k] <- beta_0 + beta[i] + eta[j,i,k] + epsilon[j,i,k]
-      # alt:
+      # single site: lambda[i,j] <- beta_0 + beta[i] + eta[j,i]
+      # alt format:
       # fixed[i,j] <- beta_0 + beta[i]
       # lambda[i,j] <- fixed[i] + eta[i,j]
 
[…]

### 2015-10-06 Manuscript/carkeek_grid.tex
added description of basics of model

@@
 That study outlined the structure for estimation of the proportional biomass of a taxon ($B_i$) given the proportional counts of sequences recovered from a parallel sequencing run ($Z_i$).
-We constructed the following mathematical model (\ref{my_equation_label}) to better understand the concept:
+
+We modeled the counts of DNA sequences ($Z$) from each of a given taxon $i$, in each replicate PCR $j$, from each replicate of a given location $k$ (hence, $Z_{ijk}$), as though they are ?(proportional to/drawn from)?
+a Poisson distribution.
+Poisson distributions are described by one and only one parameter, their mean; in this case the mean is given by an exponential function $e^\lambda$.
+% WHY IS IT EXPONENTIATED?
+COUNTS?
+Thus,
+\begin{equation}\label{some_cool_eqn_name}
+	Z_{ijk} \sim Poisson(e^{\lambda_{ijk}})
+\end{equation}
+
+In turn, we further assume this parameter $\lambda$ is linearly proportional to a suite of taxon-, pcr-, and site- specific parameters describing the variance associated with each sub-process linking the amount of DNA ($Y$) of a given taxon $i$ at a given location $k$ in a DNA extract (hence $Y_{ik}$):
+
+\begin{equation}\label{GLM}
+	\lambda_{ijk} = \beta_0 + \beta_i + \eta_{ijk} + \epsilon_{ijk}
+\end{equation}
+
+Where $\beta_0$ is a general intercept across all taxa, $\beta_i$ is a fixed effect accounting for the variance associated with taxon $i$, and $\eta_{ijk}$ and $\epsilon_{ijk}$ are random effects of variance resulting from the processes associated with PCR and spatial location, respectively.
+
+%We constructed the following mathematical model (\ref{my_equation_label}) to better understand the concept:
 % using '\ref{something}' in the text will refer to any object (e.g. figure, equation, table) which contains the corresponding '\label{something}'
@@
 
-\begin{equation}\label{my_equation_label}
-	Y = 2a + 2a
-\end{equation}
-
-Where $a$ represents an apple.
+%\begin{equation}\label{my_equation_label}
+%	Y = 2a + 2a
+%\end{equation}
+%
+%Where $a$ represents an apple.
 

### 2015-10-07 Manuscript/carkeek_grid.tex
small changes to wording of manuscript; increased MCMC iterations

@@
 a Poisson distribution.
-Poisson distributions are described by one and only one parameter, their mean; in this case the mean is given by an exponential function $e^\lambda$.
+A Poisson distribution is described by one and only one parameter, $\lambda$, which is equal to both the mean and variance.
+Because in this case our modeled values are discrete counts, we use the natural exponent, $e^\lambda$.
 % WHY IS IT EXPONENTIATED?
@@
 \section*{Author Contributions}
-Conceived and designed the experiments: .
-Collected the data: .
+Conceived and designed the experiments: James L. O'Donnell, Ryan P. Kelly, A. Ole Shelton.
+Collected the data: James L. O'Donnell, Greg Williams, Natalie C. Lowell, Ryan P. Kelly, A. Ole Shelton, .
 Conducted the analyses: .

### 2015-10-07 Analysis/eDNA_model_grid.R

@@ -110,8 +110,8 @@ counts[        ,         , sites[2] ]
 
 model_loc <- "eDNA_model_grid.jags"
 
-jagsscript <- cat(
-"
+# jagsscript <- cat("
+jagsscript <- "
 model {
   
     ## MODEL STRUCTURE
@@ -178,8 +178,9 @@ model {
     }
 
 }
-",
-file = model_loc)
+"
+# ,
+# file = model_loc)
 
 jags_data <- list(
                 "counts",
@@ -196,15 +197,16 @@ jags_params <- c(
                 "beta_0"
 )
 
-N_burn <- 1000
-N_iter <- 1000
+N_burn <- 0
+N_iter <- 100000
 N_chain <- 3
 
 my_jags <- jags(
 				data = jags_data,
 				inits = NULL,
 				parameters.to.save = jags_params,
-				model.file = model_loc,
+				model.file = textConnection(jagsscript), 
+				# model.file = model_loc,
 				n.chains = N_chain,
 				n.iter = N_iter + N_burn,
 				n.burnin = N_burn, #floor(n.iter/2)

### 2015-10-27 Analysis/eDNA_model_grid.R
added trace plots

@@ -178,7 +178,8 @@ model {
     }
 
 }
-# "
+"
+"
 
 
 # ,
@@ -234,17 +235,104 @@ attach.jags(my_jags, overwrite = TRUE)
 # general plot
 plot(my_jags)
 
-str(my_jags)
+# F this S. Keep getting the error "Error in coda.samples(my_jags, jags_params, N_iter = 1000) : attempt to apply non-function"
+# coda.samples(my_jags, jags_params, N_iter = 1000)
+# jags.samples(my_jags, jags_params, N_iter = 1000)
 
+str(my_jags)
 names(my_jags) # "model" "BUGSoutput" "parameters.to.save" "model.file" "n.iter" "DIC"
 
+str(my_jags$BUGSoutput)
 names(my_jags$BUGSoutput)
 
-someoutput <- my_jags$BUGSoutput$summary
+# simulations list
+names(my_jags$BUGSoutput$sims.list)
 
-class(someoutput)
+# simulations array
+str(my_jags$BUGSoutput$sims.array)
 
-str(my_jags$BUGSoutput)
+
+dim(my_jags$BUGSoutput$sims.array)
+dimnames(my_jags$BUGSoutput$sims.array)
+dimnames(my_jags$BUGSoutput$sims.array)[[3]] # parameter names
+
+dim(my_jags$BUGSoutput$sims.array[,,])
+
+
+
+
+mcmc_summary <- my_jags$BUGSoutput$summary
+
+
+
+
+
+
+
+
+################################################################################################
+# PLOT TRACES
+################################################################################################
+
+mcmc_array <- my_jags$BUGSoutput$sims.array
+
+dim(mcmc_array)
+
+mcmc_iter <- dim(mcmc_array)[1]
+mcmc_chains <- dim(mcmc_array)[3]
+mcmc_params <- dimnames(mcmc_array)[[3]]
+thin_factor <- 10
+
+trace_layout <- layout(
+
+		mat = matrix(
+					data = 1:length(mcmc_params), 
+					nrow = 6, 
+					byrow = TRUE
+					)
+	)
+
+layout.show(trace_layout)
+
+for(param in mcmc_params){
+
+	mcmc_samples <- mcmc_array[,,param]
+
+	plot(
+		x= 0, 
+		pch = '', 
[…]

### 2016-03-04 Manuscript/carkeek_grid.tex
added trace plots
add to introduction

@@
 \section*{Introduction}
-For centuries, humankind has wondered: If I have two apples, and someone gives me another two apples, how many apples do I have? Some people did this \citep{Darwin1859}.
+
+Biodiversity surveillance is being revolutionized by DNA-based detection of organisms from environmental samples. ?(specifically speed and scope of ecological studies).
+Many researchers are justifiably cautious about the ?(adoption) of this new form of data.
+Their apprehension is rooted in the premise that traditional survey approaches are more accurate because the chain of inference between observation and ecological data is usually short: A researcher sees two swans in Lake Hopatcong and infers the lake is occupied by at least 2 swans.
+DNA based surveys, on the other hand, consist of a longer chain of inference ?(including several links at which the links are complex or unknown)?:
+Dna sequences are observed from a sequencer, the sequencer sequences products of a polymerase chain reaction (PCR), PCR amplified sequences from a purified DNA sample, DNA is purified (extracted) from an environmental sample, environmental samples contain DNA from organisms present, the organisms present are representative of the biological community about which we wish to make inference.
+(Could tie this to the swans of Lake Hopatcong).
+Clearly, this process is more complex than visual surveys. ?(and includes several steps at which the relationship between stages is complex or unknown)
 
@@
 Conceived and designed the experiments: James L. O'Donnell, Ryan P. Kelly, A. Ole Shelton.
-Collected the data: James L. O'Donnell, Greg Williams, Natalie C. Lowell, Ryan P. Kelly, A. Ole Shelton, .
+Collected the data: James L. O'Donnell, Greg Williams, Natalie C. Lowell, Ryan P. Kelly, A. Ole Shelton, Jameal F. Samhouri.
 Conducted the analyses: .

### 2016-03-11 Manuscript/carkeek_grid.tex
add introduction

@@
 Their apprehension is rooted in the premise that traditional survey approaches are more accurate because the chain of inference between observation and ecological data is usually short: A researcher sees two swans in Lake Hopatcong and infers the lake is occupied by at least 2 swans.
-DNA based surveys, on the other hand, consist of a longer chain of inference ?(including several links at which the links are complex or unknown)?:
-Dna sequences are observed from a sequencer, the sequencer sequences products of a polymerase chain reaction (PCR), PCR amplified sequences from a purified DNA sample, DNA is purified (extracted) from an environmental sample, environmental samples contain DNA from organisms present, the organisms present are representative of the biological community about which we wish to make inference.
-(Could tie this to the swans of Lake Hopatcong).
-Clearly, this process is more complex than visual surveys. ?(and includes several steps at which the relationship between stages is complex or unknown)
+DNA based surveys, on the other hand, consist of a longer chain of inference:
+DNA sequences are reported by a sequencing machine, the machine identifies the sequence of products of a polymerase chain reaction (PCR), PCR amplifies pieces of DNA from a purified genomic DNA sample, DNA is purified (extracted) from an environmental sample, environmental samples contain DNA from organisms present, the organisms present are representative of the biological community about which we wish to make inference.
+?( reverse order?
+tie to concrete example (swans of Lake Hopatcong)).
+Clearly, this process is more complex than visual surveys, as the relationship between several steps is complex or unknown.
+But consider that the processes ?(behind | underlying) other more widely-used ecological survey techniques are similarly complex, such as bird surveys based on song, or visual identification of fungal spores.
+When alternate survey approaches are impossible or inefficient, we are more willing to accept any available survey data, regardless of the complexity or uncertainty underlying it.
+(microbiologists have enthusiastically relied on DNA-based surveys for years for this reason, (though yes, they also do not have the problem of disconnect between individual and cell)).
+
+The ability of DNA surveys to make quantitative inference about communities has been touted by some (CITE new fish quantitation paper) and doubted by others (CITE european eelgrass PLOSONE).
+For example, a study linking (blah blah blah) concluded that "metabarcoding is powerful, yet blind" (CITE european eelgrass).
+Conversely, others have reported strong quantitative and intuitive links between DNA-based and traditional survey methods (CITE Port 2016 MOLECO).
+These studies usually rely on simple statistical models to link DNA quantity to some measurable ecosystem property like biomass (but see CITE).
+When confronted with data collected in ?(complex ways/studies/whatever), simple models ?(may | often) fail to detect relationships when they exist, or vice versa ?(they are prone to inflated risk of BOTH type I and type II error) (CITE, see Woltman 2012).
+For example, (CITE, look for that Gelman paper) have demonstrated that when data are structured in a hierarchical fashion (e.g.
+test scores of students in schools belonging to districts belonging to states), a low number of replicates at the first level of hierarchy (SEE THE PAPER).
+Similarly, (describe hospital/school problems).
+
+Shelton et al. (CITE Shelton 2016) outlined an approach for structuring statistical models of DNA surveys that address these issues.
+This framework improved on alternative statistical techniques by explicitly accounting for the ?(hierarchical | nested | multilevel) structure of the study design, which allows error and uncertainty at each level to be ?(explicitly accounted for| modeled | propagated throughout the model).
+That study demonstrated an improvement in the estimate of higher-level (e.g. ecological community) quantities when the processes linking them to the data are specified.
+As an example, it was shown that incorporation of data about the mismatch between primer and template DNA sequence can improve the estimate of the relative abundance of unique DNA templates input to a PCR.
+
+Here, we apply this framework to a DNA survey of ?(nearshore | coastal) marine habitat.
+(TODO add commentary on current dogma surrounding distribution of DNA in well-mixed (marine) habitats).
+We document the variability associated with lab based ?(procedures | replication | treatment; i.e. filter+DNA+PCR+seq), and the spatial scale over which DNA communities vary in this habitat.
+We ?(show that | tested whether) a taxon's spatial distribution predicts ( the slope of the relationship between distance from shore and DNA abundance ~or~ to what degree DNA abundance is explained by distance from shore for each taxon).
+We focus partly on species with known life histories that define their spatial distribution (e.g.
+shallow water livebearing fishes or sessile intertidal organisms with ?(motile/planktonic/pelagic) larvae or gametes).
+For these taxa whose spatial distribution is well-documented and restricted, we calculate the rate of change in space and compare this rate among taxa with similar spatial distributions.
+In turn, the distribution of rate of change serves as an estimate of the spatial distribution of DNA in this habitat.
 

### 2016-03-15 Manuscript/carkeek_grid.tex
add to intro and methods

@@
 
+We would love to estimate the minimum distance over which eDNA community differences can be detected.
+
 %-----------------------------------METHODS-------------------------------------
@@
 
+%(\ref{my_equation_label})
 Where $\beta_0$ is a general intercept across all taxa, $\beta_i$ is a fixed effect accounting for the variance associated with taxon $i$, and $\eta_{ijk}$ and $\epsilon_{ijk}$ are random effects of variance resulting from the processes associated with PCR and spatial location, respectively.
+
+Sidenote justification for k-means clustering: While variation among sampled ecological is indeed continuous, ecologists often apply categorical groupings to ecological communities.
+For example, we may sample along a transect that intersects rainforest and desert.
+As our samples become smaller and closer together, we may have increasing difficulty assigning samples to a distinct habitat type.
 
@@
 
+\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
+  \centering
+    \includegraphics[width=1\textwidth]{site_map.pdf}
+    \caption{Geographic position of collected samples. Lines give 10m isobaths.}
+  \label{site_map} % use this to refer to your figure in the text, so that numbering updates automatically
+\end{figure}
+
+\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
+  \centering
+    \includegraphics[width=1\textwidth]{clustering_fig.pdf}
+    \caption{Plot of ?(non-metric multi-dimensional scaling | principal components ) analysis. Points are colored by their membership to clusters of k-means clustering analysis.}
+  \label{clustering_fig} % use this to refer to your figure in the text, so that numbering updates automatically
+\end{figure}
+
+\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
+  \centering
+    \includegraphics[width=1\textwidth]{dissim_by_distance.pdf}
+    \caption{Pairwise dissimilarity ?(BRAY CURTIS/ETC) of eDNA communities plotted against pairwise spatial distance.}
+  \label{dissim_by_distance} % use this to refer to your figure in the text, so that numbering updates automatically
+\end{figure}
+
+\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
+  \centering
+    \includegraphics[width=1\textwidth]{slope_plots.pdf}
+    \caption{Fit lines of DNA sequence counts as a function of distance from shore for a selection of taxa for which we have strong preconceived expectations (left).
+Box plots of the estimates of the slopes for taxa ?(100 most abundant)?, grouped by life history traits (right).}
+  \label{slope_plots} % use this to refer to your figure in the text, so that numbering updates automatically
+\end{figure}
+
+\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
+  \centering
+    \includegraphics[width=1\textwidth]{var_boxplots.pdf}
+    \caption{Box plots of estimates of variance associated with each level of the multilevel model, corresponding to stages of the eDNA sampling protocol.}
+  \label{variance_boxplots} % use this to refer to your figure in the text, so that numbering updates automatically
+\end{figure}
+
+\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
+  \centering
+    \includegraphics[width=1\textwidth]{clustering_map.pdf}
+    \caption{Geographic position of collected samples, colored by membership to clusters identified by k-means clustering analysis. Lines give 10m isobaths.}
+  \label{clustering_map} % use this to refer to your figure in the text, so that numbering updates automatically
+\end{figure}
+
 %\begin{figure}[h!] % [h!] forces the figure to be placed roughly here

### 2016-03-16 Manuscript/carkeek_grid.tex
add methods subsections

@@
 \section*{Methods}
+\subsection*{Environmental Sampling}
+We collected samples from 8 points along three parallel transects, separated by 1000m.
+The first sample was collected over a lower-intertidal patch of \textit{Zostera marina}, with samples taken at 75, 125, 250, 500, 1000, 2000, and 4000 meters along the transect.
+
+\subsection*{Laboratory Methods}
+
+\subsection*{Data Preparation (Bioinformatics)}
+
+\subsection*{Model Formulation}
 We use the general framework outlined by Shelton et al (CITE).

### 2016-03-17 Manuscript/carkeek_grid.tex
add site map placeholder

@@
 
+\subsection*{Community Analysis}
+% Used k-medoids (PAM) rather than k-means because it does not require euclidean distances can be computed and that the observations follow a gaussian distribution
+% because of the nature of our data (lots of zeros, with some very high counts), the mean is not very informative
+We simultaneously assessed the existence of distinct community types and the membership of samples to those community types using a partitioning around mediods algorithm (CITE PAM, sometimes referred to as k-mediods clustering), as implemented in the R package fpc (CITE fpc).
+The classification of samples to communities was made on the basis of their pairwise Bray-Curtis dissimilarity, calculated using the function vegdist in the R package vegan (CITE VEGAN).
+% consider Gower distance
+
 \subsection*{Model Formulation}
@@
 Where $\beta_0$ is a general intercept across all taxa, $\beta_i$ is a fixed effect accounting for the variance associated with taxon $i$, and $\eta_{ijk}$ and $\epsilon_{ijk}$ are random effects of variance resulting from the processes associated with PCR and spatial location, respectively.
-
-Sidenote justification for k-means clustering: While variation among sampled ecological is indeed continuous, ecologists often apply categorical groupings to ecological communities.
-For example, we may sample along a transect that intersects rainforest and desert.
-As our samples become smaller and closer together, we may have increasing difficulty assigning samples to a distinct habitat type.
 
@@
   \centering
-    \includegraphics[width=1\textwidth]{site_map.pdf}
-    \caption{Geographic position of collected samples. Lines give 10m isobaths.}
+    \includegraphics[width=1\textwidth]{../Figures/site_map.pdf}
+    \caption{Geographic position of collected samples. Lines give XXX meter isobaths.}
   \label{site_map} % use this to refer to your figure in the text, so that numbering updates automatically

### 2016-03-17 Manuscript/carkeek_grid.tex
add to methods

@@
 \subsection*{Laboratory Methods}
+The sequencing run consisted of 14 samples ('libraries') prepared using different index sequences ligated during library preparation.
+Of these libraries, ten comprised of amplicons prepared using the 16S protocol reported above, and four comprised of amplicons prepared using a 12S protocol similar to that reported by (CITE PORT 2015).
+
+Pooled libraries were sequenced on an Illumina NextSeq at the Stanford Center for Functional Genomics (machine ID: NS50061; run ID: 115; flowcell ID: H3LFLAFXX).
 
 \subsection*{Data Preparation (Bioinformatics)}
+We calculated rates of cross-library contamination by counting occurrences of primer sequences: 12S primer sequences appearing in a 16S library (and vice versa) indicate an error in the preparation or sequencing procedures.
 

### 2016-03-18 Manuscript/carkeek_grid.tex
add to manuscript

@@
 \subsection*{Environmental Sampling}
-We collected samples from 8 points along three parallel transects, separated by 1000m.
+We collected samples from 8 points along three parallel transects separated by 1000 meters.
 The first sample was collected over a lower-intertidal patch of \textit{Zostera marina}, with samples taken at 75, 125, 250, 500, 1000, 2000, and 4000 meters along the transect.
@@
 \subsection*{Laboratory Methods}
+
+Samples were randomly assigned to PCR primer and library adapter index sequences. 
 The sequencing run consisted of 14 samples ('libraries') prepared using different index sequences ligated during library preparation.
@@
 
-\subsection*{Model Formulation}
+% Calculate Moran's I on ?(relative abundance | raw count) of ?(taxa | OTU | uniq_sequences) to report dispersion of DNA in space?
+
+\subsection*{Spatial Model Formulation}
 We use the general framework outlined by Shelton et al (CITE).
@@
 \section*{Results}
-We found that if you have two apples, and someone gives you another two apples, you have four apples.
+\subsection*{Data Quality (Bioinformatics)}
+All value ranges are reported as (mean $\pm$ standard deviation).\\
+There was a very low frequency of cross-contamination from other libraries into those reported here (5e-05$\pm$8e-05; max 0.00034)
+
+\subsection*{Community Analysis}
+
+\subsection*{Spatial Model Output}
 

### 2016-03-18 Analysis/primer_summary.R
add primer summary library contamination analysis

@@ -1,0 +1,28 @@
+primer_summary <- read.table("/Users/jimmy.odonnell/Kelly_Lab_Big/Illumina_Data_Raw/20150717/libraries/all/primer_summary_re.txt", header = TRUE, stringsAsFactors = FALSE)
+
+plot(primer_summary$X12SF/primer_summary$total_seq)
+
+lib_PhiX <- grep("Undetermined", primer_summary$filename)
+
+primer_proportion <- primer_summary[-lib_PhiX,3:6]/primer_summary[-lib_PhiX,2]
+
+stripchart(primer_proportion, las = 1, pch = 1, method = "jitter")
+
+contam <- 
+
+primer_proportion[,]
+
+which(primer_proportion[,c("X16SF", "X16SR")] < 0.5)
+p12S_in_16S <- primer_proportion[,c("X12SF", "X12SR")][primer_proportion[,c("X12SF", "X12SR")] < 0.2]
+
+range(p12S_in_16S)
+round(mean(p12S_in_16S), digits = 5); round(sd(p12S_in_16S), digits = 5); round(max(p12S_in_16S), digits = 5)
+primer_summary_long <- melt(primer_summary[,-2], id = c("filename"))
+pdf(file = "primer_summary.pdf", width = 7, height = 3)
+p <- ggplot(aes(x = filename, weight=value/1000000, fill=variable), data = primer_summary_long)
+p + geom_bar() + 
+  scale_fill_discrete("Primer") + 
+  scale_x_discrete(breaks = primer_summary$filename, labels = 1:nrow(primer_summary)) + 
+  labs(x="Libraries", y="Millions of reads", title="Primer Sequence Abundance")
+dev.off()
+

### 2016-03-18 Analysis/metadata_assembly.R
add coordinates to metadata

@@ -1,0 +1,19 @@
+data_dir <- file.path("..", "Data")
+fig_dir <- file.path("..", "Figures")
+# assemble metadata
+metadata_file <- "metadata_spatial.csv"
+
+coords_file <- "transects.csv"
+
+coords_df <- read.csv(file.path(data_dir, coords_file), stringsAsFactors = FALSE)
+
+sample_name <- paste("PCT", substr(coords_df$trans, 1, 1), sprintf("%04d", coords_df$point), sep = "-")
+coords_df <- data.frame(coords_df, sample_name, stringsAsFactors = FALSE)[,c(1,2,3,4,5,7,6)]
+
+
+metadata <- read.csv(file.path(data_dir, metadata_file), stringsAsFactors = FALSE)
+
+coords_of_metadata <- match(metadata$env_sample_name, coords_df$sample_name)
+
+metadata_with_coords <- cbind.data.frame(metadata, coords_df[coords_of_metadata,c("lon", "lat")], stringsAsFactors = FALSE)
+write.csv(metadata_with_coords, file = "metadata_spatial.csv", quote = TRUE, row.names = FALSE)

### 2016-03-18 Analysis/read_data.R
start final data process

@@ -1,0 +1,18 @@
+
+library(fpc)
+library(vegan)
+
+
+data_dir <- file.path("..", "Data")
+fig_dir <- file.path("..", "Figures")
+
+
+otu_table_filename <- "otu_table_filtered_per_samp.csv"
+metadata_filename <- "metadata_spatial.csv"
+
+otu_table <- read.csv(file.path(data_dir, otu_table_filename), row.names = 1)
+metadata <- read.csv(file.path(data_dir, metadata_filename))
+
+
+head(otu_table)[,1:10]
+

### 2016-03-19 Analysis/cluster.R
add cluster analysis

@@ -1,0 +1,68 @@
+
+# INPUT COMES FROM SCRIPT "load_data.r"
+
+library(fpc) #pamk()
+library(cluster) #pam
+library(vegan) # vegdist()
+
+# function for plot colors (sorta like ggplot)
+gghue <- function(n){
+	hues = seq(15, 375, length = n+1)
+	hcl(h = hues, l = 65, c = 100)[1:n]
+}
+
+# choose between using proportional or raw counts
+# DATA <- otu_table
+DATA <- otu_table_prop[,]
+
+mydist <- vegdist(DATA)
+
+mypam <- pamk(data = mydist) # to restrict range of Ks considered: , krange = 2:4
+pam_out <- mypam$pamobject
+mypam$nc # number of clusters
+
+plot(pam(mydist, k = mypam$nc), which.plots = 1)
+?pam
+
+mycolors <- gghue(mypam$nc)
+
+# see ?clusplot.default
+pdf(file = file.path(fig_dir, "pam_plot.pdf"))
+	par(mar = c(5,5,1,1))
+	clusplot(
+		x = as.matrix(mydist), 
+		clus = pam_out$clustering, 
+		sub = NULL, 
+		lines = 0, 
+		labels = 4, 
+		color = FALSE, 
+		col.p = mycolors[pam_out$clustering], 
+		col.txt = mycolors, 
+		col.clus = mycolors, 
+		main = NA
+	)
+dev.off()
+
+
+set.seed(2) # to control jitter
+# pdf(file = file.path(fig_dir, "medoids_in_space.pdf"), width = 7, height = 7)
+	plot(
+		jitter(x = metadata$dist_along_shore, factor = 0.4), 
+		xaxt = "n", 
+		xlim = c(-500, 2500), 
+		y = jitter(log(metadata$dist_from_shore + 10), factor = 0.2), 
+		yaxt = "n",
+		# log = "y",
+		col = mycolors[mypam$pamobject$clustering[metadata$sample_id]], 
+		# bg = mycolors[mypam$pamobject$clustering[metadata$sample_id]], 
+		pch = as.character(mypam$pamobject$clustering[metadata$sample_id]-1), 
+		cex = 1, 
+		# cex = 1.5, 
+		main = "membership to PAM classifications", 
+		las = 1, 
+		xlab = "Position along shore (meters)", 
+		ylab = "Position from 0 (meters)"
+	)
[…]

### 2016-03-19 Manuscript/carkeek_grid.tex
changes

@@
 \subsection*{Environmental Sampling}
-We collected samples from 8 points along three parallel transects separated by 1000 meters.
-The first sample was collected over a lower-intertidal patch of \textit{Zostera marina}, with samples taken at 75, 125, 250, 500, 1000, 2000, and 4000 meters along the transect.
+Starting from lower-intertidal patches of \textit{Zostera marina}, we collected water samples at 1 meter depth from 8 points (0, 75, 125, 250, 500, 1000, 2000, and 4000 meters) along three parallel transects separated by 1000 meters (Figure \ref{site_map}).
 
@@
 
+We calculated the great circle distance between points using the Haversine method as implemented by the R package geosphere (CITE geosphere).
 % Calculate Moran's I on ?(relative abundance | raw count) of ?(taxa | OTU | uniq_sequences) to report dispersion of DNA in space?
@@
     \includegraphics[width=1\textwidth]{../Figures/site_map.pdf}
-    \caption{Geographic position of collected samples. Lines give XXX meter isobaths.}
+    \caption{TODO: Plot with GEBCO 30-second data or remove grid coloring and color by isobath.
+Looking into filling by contour.
+Geographic position of collected samples.
+Lines give XXX meter isobaths.}
   \label{site_map} % use this to refer to your figure in the text, so that numbering updates automatically
@@
   \centering
-    \includegraphics[width=1\textwidth]{clustering_fig.pdf}
-    \caption{Plot of ?(non-metric multi-dimensional scaling | principal components ) analysis. Points are colored by their membership to clusters of k-means clustering analysis.}
-  \label{clustering_fig} % use this to refer to your figure in the text, so that numbering updates automatically
+    \includegraphics[width=1\textwidth]{../Figures/pam_plot.pdf}
+    \caption{TODO figure out color of ellipses; I can't even plot them gray without Plot of partitioning around medoids (PAM) analysis of OTU sequence abundance from 4 replicate PCRs at each of 24 sampling points.
+Points represent communities of OTUs; color and shape indicate cluster membership as determined by PAM analysis.
+Ellipses indicate the smallest area of a cluster that contains all of its members.}
+  \label{pam_plot.pdf} % use this to refer to your figure in the text, so that numbering updates automatically
 \end{figure}
@@
   \centering
-    \includegraphics[width=1\textwidth]{dissim_by_distance.pdf}
-    \caption{Pairwise dissimilarity ?(BRAY CURTIS/ETC) of eDNA communities plotted against pairwise spatial distance.}
-  \label{dissim_by_distance} % use this to refer to your figure in the text, so that numbering updates automatically
+    \includegraphics[width=1\textwidth]{../Figures/comm_diss_by_geo_dist.pdf}
+    \caption{Pairwise Bray-Curtis dissimilarity of eDNA communities plotted against pairwise spatial distance.}
+  \label{comm_diss_by_geo_dist} % use this to refer to your figure in the text, so that numbering updates automatically
 \end{figure}
@@
   \centering
-    \includegraphics[width=1\textwidth]{clustering_map.pdf}
-    \caption{Geographic position of collected samples, colored by membership to clusters identified by k-means clustering analysis. Lines give 10m isobaths.}
+    \includegraphics[width=1\textwidth]{../Figures/medoids_in_space.pdf}
+    \caption{Geographic position of collected samples, colored by membership to clusters identified by partitioning around medoids algorithm.
+Points are jittered in both horizontal and vertical dimension to distinguish among four replicate PCR products sequenced from each environmental sample.}
   \label{clustering_map} % use this to refer to your figure in the text, so that numbering updates automatically
@@
 
-%\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
-%  \centering
-%    \includegraphics[width=1\textwidth]{figure_filename.pdf}
-%    \caption{This is the figure caption.}
-%  \label{myfigure} % use this to refer to your figure in the text, so that numbering updates automatically
-%\end{figure}
+\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
+  \centering
+    \includegraphics[width=1\textwidth]{../Figures/otu_in_space.pdf}
+    \caption{Example of a DNA sequence's spatial distribution.
+This sequence is annotated to SPECIES X, which is found only in shallow, structured habitats such as patches of \textit{Zostera marina}.
+Point size and color transparency indicates abundance relative to other DNA sequences from that sample, scaled to the maximum value for this sequence (no fill = 0, full fill = 1).
+Samples from which this sequence was not recovered are indicated by an "x".}
+  \label{otu_in_space} % use this to refer to your figure in the text, so that numbering updates automatically
+\end{figure}
 

### 2016-03-19 Analysis/comm_diss_by_geo_dist.R
add code for assessing community dissimilarity wrt geographic distance

@@ -1,0 +1,41 @@
+# FIRST RUN 'load_data.r'
+
+library(geosphere) # distm()
+
+# calculate pairwise great circle distance between sampling locations using Haversine method
+geo_dist <- distm(x = metadata[,c("lon", "lat")])
+dimnames(geo_dist) <- list(metadata$sample_id, metadata$sample_id)
+
+# test:
+for(i in 1:length(metadata$sample_id)){
+	if(geo_dist[metadata$sample_id[i],metadata$sample_id[i]] != 0){
+		warning("distance between a point and itself is not zero, this indicates a problem")
+	}
+}
+
+
+for(i in 1:length(metadata$sample_id)){
+	
+}
+
+
+comm_dist <- as.matrix(vegdist(otu_table, method = "bray", diag = TRUE, upper = TRUE))
+if(!(identical(dimnames(comm_dist), dimnames(geo_dist)))){
+	warning("Whoa there! the row/column names of the two distance matrices do not seem to add up. This is bad.")
+}
+
+pdf(file = file.path(fig_dir, "comm_diss_by_geo_dist.pdf"), width = 8, height = 3)
+	par(mar = c(4,5,1,1))
+	plot(
+		x = log(geo_dist + 100), 
+		y = comm_dist, 
+		ylim = c(0,1), 
+		xaxt = "n", 
+		las = 1, 
+		cex = 0.5, 
+		col = rgb(0,0,0,alpha = 0.1), 
+		xlab = "Distance between samples (meters)", 
+		ylab = "Bray-Curtis dissimilarity"
+	)
+	axis(side = 1, at = unique(log(metadata$dist_from_shore + 100)), labels = unique(metadata$dist_from_shore))
+dev.off()

### 2016-03-19 Analysis/load_data.R
separate data loading from other tasks

@@ -1,0 +1,35 @@
+# !!! SET WORKING DIRECTORY TO THIS PROJECT'S SUBDIRECTORY 'Analysis'
+# Set directories from which to read/write data and write figures
+data_dir <- file.path("..", "Data")
+fig_dir <- file.path("..", "Figures")
+
+# path to otu file: rows are samples, columns are OTUs, cells are counts.
+# rownames correspond to column "sample_id" in metadata
+otu_table_filename <- "otu_table_filtered_per_samp.csv"
+
+# path to metadata file.
+# column "sample_id" contains rownames of otu file.
+metadata_filename <- "metadata_spatial.csv"
+
+otu_table <- read.csv(file.path(data_dir, otu_table_filename), row.names = 1, stringsAsFactors = FALSE)
+metadata <- read.csv(file.path(data_dir, metadata_filename), stringsAsFactors = FALSE)
+dist_along_shore <- c(1000,2000,0)[as.numeric(as.factor(metadata$transect))]
+metadata <- cbind.data.frame(metadata, dist_along_shore, stringsAsFactors = FALSE)
+
+# exclude rows from OTU table that are not in metadata
+otu_table <- otu_table[metadata$sample_id,]
+
+otu_table_prop <- otu_table/rowSums(otu_table)
+
+head(otu_table)[,1:10]
+head(metadata)
+metadata$sample_id
+
+
+boxplot(otu_table[,1:20])
+# boxplot(log(otu_table[,1:20]))
+boxplot(otu_table_prop[,1:20])
+boxplot(scale(otu_table[,1:20]))
+
+min(scale(otu_table[,1:20])+2)
+boxplot(scale(otu_table[,1:20])+2)

### 2016-03-19 Analysis/otu_in_space.R
code to plot OTU abundance in space

@@ -1,0 +1,66 @@
+# INPUT COMES FROM SCRIPT "load_data.r"
+
+# plot scaled proportional abundance in space (i.e. darkest color is wherever OTU's proportional abundance is at maximum)
+
+for(i in c(1:5,7:10)){
+	plot(
+		metadata$lon, 
+		metadata$lat, 
+		bg = rgb(r = 1, g = 0, b = 0, alpha = otu_table_prop[metadata$sample_id, i]/max(otu_table_prop[metadata$sample_id, i])), 
+		pch = 21, 
+		main = colnames(otu_table_prop)[i], 
+		xlab = "Longitude", 
+		ylab = "Latitude"
+	)
+}
+
+otu_table[,6]
+
+pdf(file = file.path(fig_dir, "otu_in_space.pdf"), width = 7, height = 7)
+	plot(
+		metadata$lon, 
+		metadata$lat, 
+		bg = rgb(r = 1, g = 0, b = 0, alpha = otu_table_prop[metadata$sample_id, "DUP_3"]/max(otu_table_prop[metadata$sample_id, "DUP_3"])), 
+		pch = 21, 
+		cex = otu_table_prop[metadata$sample_id, "DUP_3"]/max(otu_table_prop[metadata$sample_id, "DUP_3"])*1.5, 
+		# cex = 1.5, 
+		main = "DUP_3", 
+		xlab = "Longitude", 
+		ylab = "Latitude"
+	)
+	points(
+		metadata$lon, 
+		metadata$lat, 		
+		pch = ifelse(otu_table_prop[metadata$sample_id, "DUP_3"]/max(otu_table_prop[metadata$sample_id, "DUP_3"]) == 0, 4, NA_integer_) 
+	)
+dev.off()
+
+set.seed(8) # to control jitter
+pdf(file = file.path(fig_dir, "otu_in_space.pdf"), width = 7, height = 7)
+	plot(
+		jitter(x = metadata$dist_along_shore, factor = 0.2), 
+		xaxt = "n", 
+		xlim = c(-500, 2500), 
+		y = jitter(log(metadata$dist_from_shore + 10), factor = 0.2), 
+		yaxt = "n",
+		# log = "y",
+		bg = rgb(r = 1, g = 0, b = 0, alpha = otu_table_prop[metadata$sample_id, "DUP_3"]/max(otu_table_prop[metadata$sample_id, "DUP_3"])), 
+		pch = 21, 
[…]

### 2016-03-19 Analysis/read_data.R
removed google icon and old data reading file

@@ -1,18 +1,0 @@
-
-library(fpc)
-library(vegan)
-
-
-data_dir <- file.path("..", "Data")
-fig_dir <- file.path("..", "Figures")
-
-
-otu_table_filename <- "otu_table_filtered_per_samp.csv"
-metadata_filename <- "metadata_spatial.csv"
-
-otu_table <- read.csv(file.path(data_dir, otu_table_filename), row.names = 1)
-metadata <- read.csv(file.path(data_dir, metadata_filename))
-
-
-head(otu_table)[,1:10]
-

### 2016-03-29 Manuscript/v0.0/carkeek_grid.tex
add bioinformatic methods
fix references to figures broken by change in directory structure

--- 2016-03-19 Manuscript/carkeek_grid.tex
@@
 
+Detailed bioinformatic methods are provided in the supplement, and scripts used from raw sequencer output onward can be found in the project directory on GitHub.%\href{https://github.com/jimmyodonnell/Carkeek_eDNA_grid}{https://github.com/jimmyodonnell/Carkeek_eDNA_grid}
+%----------------------------------------------------------------------------
+%METHODS
+%Sequences passing were demultiplexed on the basis of the ?(8-bp) sequence ligated during library preparation and read during the index read, and filtered for general quality by the sequencing facility.
+%
+%We used fastqc to assess the fastq files output from the sequencer for low-quality indications of a problematic run.
+%
+%Reads that contain primer index sequences (or combinations thereof) that we did not actually use in laboratory procedures are known to ?(be derived from contamination) be a result of processes that leave the same signature as contamination.
+For example, errors during oligonucleotide synthesis of the indexed primers would cause reads to be erroneously assigned to samples.
+%Because this represents the minimum rate at which we can be certain that sequences from one environmental sample could be erroneously assigned, we considered for further analysis only those reads occurring with greater frequency than this.
+%
+%To infer a general estimate of the rate of cross-sample contamination likely to be associated with lab procedures, we calculated the proportion of reads containing primer index sequences (or combinations thereof?) that we did not actually use
+%
+%RESULTS
+%Reads were of generally high-quality, with ?(96.48)? % of base calls having a 0.001 or lower probability of incorrect base call (Phred q-score of 30 or higher).
+%
+%
+%Total sequencing clusters passing filter per read:
+%
+%reads, reads passing filter, % >= Q30, for each run ( 7,052,307 ; 6,752,829; 97.47 :: 7,052,307 ; 6,752,829 ; 95.50)
+%
+%Cluster density of 358 \UTF{00EF}\UTF{00BF}\UTF{0153}22	clusters per mm^2
+%
+%XXX % of reads aligned to the PhiX genome
+%(31.02 \UTF{00EF}\UTF{00BF}\UTF{0153}0.59	; 29.17 \UTF{00EF}\UTF{00BF}\UTF{0153}0.89)
+%
+%and error rate as determined from alignment of PhiX reads to the PhiX genome was low (Read 1: 1.26 \UTF{00EF}\UTF{00BF}\UTF{0153}0.07; Read 2: 1.38 \UTF{00EF}\UTF{00BF}\UTF{0153}0.07)
+%----------------------------------------------------------------------------
+
 \subsection*{Community Analysis}
@@
   \centering
-    \includegraphics[width=1\textwidth]{../Figures/site_map.pdf}
+    \includegraphics[width=1\textwidth]{../../Figures/site_map.pdf}
     \caption{TODO: Plot with GEBCO 30-second data or remove grid coloring and color by isobath.
@@
   \centering
-    \includegraphics[width=1\textwidth]{../Figures/pam_plot.pdf}
+    \includegraphics[width=1\textwidth]{../../Figures/pam_plot.pdf}
     \caption{TODO figure out color of ellipses; I can't even plot them gray without Plot of partitioning around medoids (PAM) analysis of OTU sequence abundance from 4 replicate PCRs at each of 24 sampling points.
@@
   \centering
-    \includegraphics[width=1\textwidth]{../Figures/comm_diss_by_geo_dist.pdf}
+    \includegraphics[width=1\textwidth]{../../Figures/comm_diss_by_geo_dist.pdf}
     \caption{Pairwise Bray-Curtis dissimilarity of eDNA communities plotted against pairwise spatial distance.}
@@
   \centering
-    \includegraphics[width=1\textwidth]{../Figures/medoids_in_space.pdf}
+    \includegraphics[width=1\textwidth]{../../Figures/medoids_in_space.pdf}
     \caption{Geographic position of collected samples, colored by membership to clusters identified by partitioning around medoids algorithm.
@@
   \centering
-    \includegraphics[width=1\textwidth]{../Figures/otu_in_space.pdf}
+    \includegraphics[width=1\textwidth]{../../Figures/otu_in_space.pdf}
     \caption{Example of a DNA sequence's spatial distribution.

### 2016-03-30 Manuscript/v0.0/carkeek_grid.tex
small changes

@@
 %\linespread{2} % alternate method of double spacing
+%\graphicspath{{../../Figures/}} % I think if you do this it saves typing the path over and over in the figures section
 
@@
 
-Pooled libraries were sequenced on an Illumina NextSeq at the Stanford Center for Functional Genomics (machine ID: NS50061; run ID: 115; flowcell ID: H3LFLAFXX).
+Pooled libraries were sequenced on the Illumina NextSeq platform at the Stanford Center for Functional Genomics (machine ID: NS50061; run ID: 115; flowcell ID: H3LFLAFXX).
+Raw sequence data in fastq format is publicly available (see Data Availability).
 
 \subsection*{Data Preparation (Bioinformatics)}
-We calculated rates of cross-library contamination by counting occurrences of primer sequences: 12S primer sequences appearing in a 16S library (and vice versa) indicate an error in the preparation or sequencing procedures.
+Detailed bioinformatic methods are provided in the supplemental material, and scripts used from raw sequencer output onward can be found in the project directory on GitHub (see Data Availability).
 
-Detailed bioinformatic methods are provided in the supplement, and scripts used from raw sequencer output onward can be found in the project directory on GitHub.%\href{https://github.com/jimmyodonnell/Carkeek_eDNA_grid}{https://github.com/jimmyodonnell/Carkeek_eDNA_grid}
 %----------------------------------------------------------------------------
@@
 %
+We calculated rates of cross-library contamination by counting occurrences of primer sequences: 12S primer sequences appearing in a 16S library (and vice versa) indicate an error in the preparation or sequencing procedures.
+
 %To infer a general estimate of the rate of cross-sample contamination likely to be associated with lab procedures, we calculated the proportion of reads containing primer index sequences (or combinations thereof?) that we did not actually use
@@
 \section*{Data Availablity}
-% Are the data and code available in a permanent, publicly accessible data archive or repository?
-The data and code used to generate our results can be found at the following url:
+% TODO SUBMIT DATA AND METADATA; ADD URL
+All sequence files and metadata are available from EMBL:\\ \verb!http://www.ebi.ac.uk/ena/data/view/XXXXXXXXX! \\
+All analyses were performed using scripts available from the project repository on GitHub:\\  \verb!https://github.com/jimmyodonnell/Carkeek_eDNA_grid!\\
 
@@
 %\section*{References} % commented out because the section title is automatically inserted if using an automatically-generated bibliography
-
 \bibliographystyle{apalike} % or: plain,unsrt,alpha,abbrv,acm,apalike,ieeetr
-\bibliography{/Users/threeprime/Documents/Publications/bibtex/library} % path to your .bib file excluding .bib extension (e.g. /Users/threeprime/Documents/Publications/bibtex/library)
+\bibliography{carkeek_grid_bib} % path to your .bib file excluding .bib extension (e.g. /Users/threeprime/Documents/Publications/bibtex/library)
 
@@
   \centering
-    \includegraphics[width=1\textwidth]{slope_plots.pdf}
+    \includegraphics[width=1\textwidth]{../../Figures/slope_plots.pdf}
     \caption{Fit lines of DNA sequence counts as a function of distance from shore for a selection of taxa for which we have strong preconceived expectations (left).
@@
   \centering
-    \includegraphics[width=1\textwidth]{var_boxplots.pdf}
+    \includegraphics[width=1\textwidth]{../../Figures/var_boxplots.pdf}
     \caption{Box plots of estimates of variance associated with each level of the multilevel model, corresponding to stages of the eDNA sampling protocol.}
@@
 %----------------------------------SUPPLEMENT-----------------------------------
-%\section*{Supplemental Material}
+\section*{Supplemental Material}
+
+\subsection*{Bioinformatic Methods}
 

### 2016-03-30 Analysis/data_prep.R
furthered data prep

@@ -5,11 +5,19 @@ fig_dir <- file.path("..", "Figures")
 
 # generate a file of only the samples in the metadata
 # (assumes the metadata only has samples we're interested in rather than the full sequencing run)
+
+# holy macaroni, we need to decide on a standard orientation.
+# "OTUs_fam_w30_top10.csv" is in rows=OTU, cols=SAMPLES
+# YOU SHOULD PROBABLY DO THIS: start with script 'load_data.R',
+# use the resulting object 'otu_filt' in place of table_transposed
+# OR use t(otu_filt) as object 'table_full'
+
 filename_orig <- file.path(data_dir, "OTUs_fam_w30_top10.csv")
 
 filename_out <- file.path(data_dir, "OTUs_top10.csv")
 
-table_full <- read.csv(filename_orig, row.names = 1)
+# table_full <- read.csv(filename_orig, row.names = 1)
+# maybe ? table_full <- t(otu_filt)
 
 # load metadata
 metadata_file <- file.path(data_dir, "metadata_spatial.csv")
@@ -25,6 +33,8 @@ transect_line <- as.numeric(as.factor(paste(transect_line_rep, transect_position
 # what is the name of the column containing the sample id (SEQUENCING samples)
 colname_sampleid <- "sample_id"
 
+metadata_exp <- cbind.data.frame(metadata, sequenced_sample, transect_position, transect_line)
+
 # restrict to only samples in the metadata file
 # !!! it will really screw things up if you don't convert the sample id column to a character vector!
 table_restricted <- table_full[ , as.character(metadata[,colname_sampleid])]
@@ -34,13 +44,14 @@ write.csv(x = table_restricted, file = filename_out, quote = FALSE)
 
 # COMBINE UP DATA
 table_transposed <- t(table_restricted[,as.character(metadata[,colname_sampleid])])
+
 if(
   identical(
     rownames(table_transposed),
     metadata[,colname_sampleid]
   )
 ){
-  data_full <- cbind.data.frame(metadata, sequenced_sample, transect_position, transect_line, table_transposed)
+  data_full <- cbind.data.frame(metadata_exp, table_transposed)
   if(identical(rownames(data_full), data_full[,colname_sampleid])){
     rownames(data_full) <- NULL
   } else {
@@ -50,8 +61,23 @@ if(
   warning("something does not match up, and you could screw up the data big time by proceeding.")
 }
 
[…]

### 2016-03-30 Analysis/load_data.R

@@ -5,13 +5,20 @@ fig_dir <- file.path("..", "Figures")
 
 # path to otu file: rows are samples, columns are OTUs, cells are counts.
 # rownames correspond to column "sample_id" in metadata
-otu_table_filename <- "otu_table_filtered_per_samp.csv"
+otu_table_filename <- "OTU_table.csv"
 
 # path to metadata file.
 # column "sample_id" contains rownames of otu file.
 metadata_filename <- "metadata_spatial.csv"
 
 otu_table <- read.csv(file.path(data_dir, otu_table_filename), row.names = 1, stringsAsFactors = FALSE)
+dim(otu_table)
+
+if(length(grep("lib_", colnames(otu_table))) > 0){
+	otu_table <- t(otu_table)
+}
+dim(otu_table)
+
 metadata <- read.csv(file.path(data_dir, metadata_filename), stringsAsFactors = FALSE)
 dist_along_shore <- c(1000,2000,0)[as.numeric(as.factor(metadata$transect))]
 metadata <- cbind.data.frame(metadata, dist_along_shore, stringsAsFactors = FALSE)
@@ -19,14 +26,40 @@ metadata <- cbind.data.frame(metadata, dist_along_shore, stringsAsFactors = FALS
 # exclude rows from OTU table that are not in metadata
 otu_table <- otu_table[metadata$sample_id,]
 
+# exclude OTUs not found in these samples
+otu_table <- otu_table[,which(colSums(otu_table) > 0)]
+dim(otu_table)
+
+# calculate the minimum number of reads assigned to these OTUs in these samples
+minreads <- min(rowSums(otu_table))
+
+# calculate proportional abundance of OTUs in each sample
 otu_table_prop <- otu_table/rowSums(otu_table)
 
-head(otu_table)[,1:10]
-head(metadata)
-metadata$sample_id
+# scale the proportional abundance of the OTU in each sample to the minimum number of reads
+otu_scaled <- otu_table_prop * minreads
+
+# ignore counts OTUs found fewer than 0.5 times (anything greater would be rounded to 1)
+otu_scaled[otu_scaled < 0.5] <- 0
+
+# round to whole numbers
+otu_scaled <- round(otu_scaled)
+dim(otu_scaled)
+
+# again exclude OTUs not found in these samples
+otu_scaled <- otu_scaled[,which(colSums(otu_scaled) > 0)]
[…]

### 2016-03-30 Analysis/stan_4.R
add full stan code

@@ -1,0 +1,30 @@
+data_dir <- file.path("..", "Data")
+
+data_file <- file.path(data_dir, "data_full_top10.csv")
+
+data_full <- read.csv(data_file)
+
+mydata <- data_full
+
+colname_level1 <- "sequenced_sample"
+colname_level2 <- "transect_line"
+colname_level3 <- "transect_position"
+colname_predictor <- "dist_from_shore"
+colname_outcome <- "DUP_8"
+
+lookup_3in2 <- unique(mydata[,c(colname_level2, colname_level3)])[,colname_level3]
+
+plot(mydata[,colname_predictor], mydata[,colname_outcome])
+
+stan_data <- list(Ni          = length(unique(mydata[,colname_level1])), 
+                  Nj          = length(unique(mydata[,colname_level2])), 
+                  Nk          = length(unique(mydata[,colname_level3])), 
+                  level2      = mydata[,colname_level2],
+                  level3      = mydata[,colname_level3],
+                  lev3ForLev2 = lookup_3in2,
+                  Y_ijk       = mydata[,colname_outcome],
+                  X_1ijk      = mydata[,colname_predictor]
+                  )
+
+lm_out <- lm(mydata[,colname_outcome] ~ mydata[,colname_predictor] )
+summary(lm_out)

### 2016-03-30 Analysis/stan_4.R
build 4 level model stan code

@@ -1,6 +1,6 @@
 data_dir <- file.path("..", "Data")
 
-data_file <- file.path(data_dir, "data_full_top10.csv")
+data_file <- file.path(data_dir, "data_full_long_20.csv")
 
 data_full <- read.csv(data_file)
 
@@ -9,19 +9,24 @@ mydata <- data_full
 colname_level1 <- "sequenced_sample"
 colname_level2 <- "transect_line"
 colname_level3 <- "transect_position"
+colname_level4 <- "OTU"
 colname_predictor <- "dist_from_shore"
-colname_outcome <- "DUP_8"
+colname_outcome <- "count"
 
 lookup_3in2 <- unique(mydata[,c(colname_level2, colname_level3)])[,colname_level3]
+lookup_4in3 <- as.character(unique(mydata[,c(colname_level3, colname_level4)])[,colname_level4])
 
-plot(mydata[,colname_predictor], mydata[,colname_outcome])
+# plot(mydata[,colname_predictor], mydata[,colname_outcome])
 
 stan_data <- list(Ni          = length(unique(mydata[,colname_level1])), 
                   Nj          = length(unique(mydata[,colname_level2])), 
                   Nk          = length(unique(mydata[,colname_level3])), 
+                  Nl          = length(unique(mydata[,colname_level4])), 
                   level2      = mydata[,colname_level2],
                   level3      = mydata[,colname_level3],
+                  level4      = mydata[,colname_level4],
                   lev3ForLev2 = lookup_3in2,
+                  lev4ForLev2 = lookup_4in3,
                   Y_ijk       = mydata[,colname_outcome],
                   X_1ijk      = mydata[,colname_predictor]
                   )

### 2016-03-30 Analysis/stan_4.R
add in actual stan code

@@ -16,7 +16,7 @@ colname_outcome <- "count"
 lookup_3in2 <- unique(mydata[,c(colname_level2, colname_level3)])[,colname_level3]
 lookup_4in3 <- as.character(unique(mydata[,c(colname_level3, colname_level4)])[,colname_level4])
 
-# plot(mydata[,colname_predictor], mydata[,colname_outcome])
+# plot(mydata[,colname_predictor], mydata[,colname_outcome], xlab = "distance from shore", ylab = "number of reads", las = 1)
 
 stan_data <- list(Ni          = length(unique(mydata[,colname_level1])), 
                   Nj          = length(unique(mydata[,colname_level2])), 
@@ -33,3 +33,125 @@ stan_data <- list(Ni          = length(unique(mydata[,colname_level1])),
 
 lm_out <- lm(mydata[,colname_outcome] ~ mydata[,colname_predictor] )
 summary(lm_out)
+
+#-------------------------------------------------------------------------------
+# MODEL IN STAN CODE
+#-------------------------------------------------------------------------------
+stan_code <- '
+data {
+  // Define variables in data
+  // Number of level-1 observations (an integer)
+  int<lower=0> Ni;
+  // Number of level-2 clusters
+  int<lower=0> Nj;
+  // Number of level-3 clusters
+  int<lower=0> Nk;
+
+  // Cluster IDs
+  int<lower=1> level2[Ni];
+  int<lower=1> level3[Ni];
+  int<lower=1> lev3ForLev2[Nj];
+
+  // Continuous outcome
+  real Y_ijk[Ni];
+  // Continuous predictor
+  real X_1ijk[Ni];
+}
+
+parameters {
+  // Define parameters to estimate
+  // Population intercept (a real number)
+  real beta_0;
+  // Population slope
+  real beta_1;
+
+  // Level-1
+  real<lower=0> sigma_e0;
+
+  // Level-2 random effect
+  real u_0jk[Nj];
+  real<lower=0> sigma_u0jk;
+
+  // Level-3 random effect
+  real u_0k[Nk];
+  real<lower=0> sigma_u0k;
+
+  // overdispersion parameter
+  real<lower=0> tau;
+
+}
+
+transformed parameters  {
+  // Varying intercepts
[…]

### 2016-03-30 Analysis/stan_4.R
add to stan code

@@ -46,11 +46,15 @@ data {
   int<lower=0> Nj;
   // Number of level-3 clusters
   int<lower=0> Nk;
+  // Number of level-4 clusters
+  int<lower=0> Nl;
 
   // Cluster IDs
   int<lower=1> level2[Ni];
   int<lower=1> level3[Ni];
+  int<lower=1> level4[Nl];
   int<lower=1> lev3ForLev2[Nj];
+  int<lower=1> lev4ForLev3[Nk];
 
   // Continuous outcome
   real Y_ijk[Ni];

### 2016-04-05 Analysis/stan_4.R
add 4 level model

@@ -26,9 +26,9 @@ stan_data <- list(Ni          = length(unique(mydata[,colname_level1])),
                   level3      = mydata[,colname_level3],
                   level4      = mydata[,colname_level4],
                   lev3ForLev2 = lookup_3in2,
-                  lev4ForLev2 = lookup_4in3,
-                  Y_ijk       = mydata[,colname_outcome],
-                  X_1ijk      = mydata[,colname_predictor]
+                  lev4ForLev3 = lookup_4in3,
+                  Y_ijkl      = mydata[,colname_outcome],
+                  X_1ijkl     = mydata[,colname_predictor]
                   )
 
 lm_out <- lm(mydata[,colname_outcome] ~ mydata[,colname_predictor] )
@@ -52,14 +52,14 @@ data {
   // Cluster IDs
   int<lower=1> level2[Ni];
   int<lower=1> level3[Ni];
-  int<lower=1> level4[Nl];
+  int<lower=1> level4[Ni];
   int<lower=1> lev3ForLev2[Nj];
   int<lower=1> lev4ForLev3[Nk];
 
-  // Continuous outcome
-  real Y_ijk[Ni];
-  // Continuous predictor
-  real X_1ijk[Ni];
+  // Outcome: counts
+  int<lower=0> Y_ijkl[Ni];
+  // Predictor: continuous?
+  real X_1ijkl[Ni];
 }
 
 parameters {
@@ -73,12 +73,16 @@ parameters {
   real<lower=0> sigma_e0;
 
   // Level-2 random effect
-  real u_0jk[Nj];
-  real<lower=0> sigma_u0jk;
+  real u_0jkl[Nj];
+  real<lower=0> sigma_u0jkl;
 
   // Level-3 random effect
-  real u_0k[Nk];
-  real<lower=0> sigma_u0k;
+  real u_0kl[Nk];
+  real<lower=0> sigma_u0kl;
+
+  // Level-4 random effect
+  real u_0l[Nl];
+  real<lower=0> sigma_u0l;
 
   // overdispersion parameter
   real<lower=0> tau;
@@ -87,17 +91,24 @@ parameters {
 
 transformed parameters  {
   // Varying intercepts
-  real beta_0jk[Nj];
-  real beta_0k[Nk];
+  real beta_0jkl[Nj];
+  real beta_0kl[Nk];
+  real beta_0l[Nl];
 
[…]

### 2016-04-05 Analysis/0_load_data.R
set threshold with variable

@@ -54,10 +54,14 @@ dim(otu_scaled)
 otu_scaled <- otu_scaled[,order(colSums(otu_scaled), decreasing = TRUE)]
 otu_scaled[1:10,1:10]
 
-# exclude OTUs that were counted a total of fewer than 100 times across samples
-otu_filt <- otu_scaled[,which(colSums(otu_scaled) > 100)]
-dim(otu_filt)
+# how many OTUs occur more than a threshold number of times?
+abundance_threshold <- 100
+sum(colSums(otu_scaled) > abundance_threshold)
 
+
+# exclude OTUs that were counted a total of fewer than 100 times across samples
+otu_filt <- otu_scaled[,which(colSums(otu_scaled) > abundance_threshold)]
+dim(otu_filt)
 
 boxplot(otu_filt[,1:20])
 # boxplot(log(otu_table[,1:20]))

### 2016-04-05 Analysis/plot_rank_abundance.R
add rank abundance plotting

@@ -1,0 +1,39 @@
+
+my_otu_table <- otu_scaled
+
+threshold_1 <- 100
+threshold_2 <- 1000
+threshold_3 <- 10000
+
+gt_t1 <- sum(colSums(my_otu_table) > threshold_1)
+gt_t2 <- sum(colSums(my_otu_table) > threshold_2)
+gt_t3 <- sum(colSums(my_otu_table) > threshold_3)
+
+plot_scale <- 1e6
+
+plot_values <- sort(colSums(my_otu_table)/plot_scale, decreasing = TRUE)
+
+output_path <- file.path(fig_dir, "rank_abundance_rescaled.pdf")
+pdf(file = output_path, width = 8, height = 4)
+par(mar = c(5, 5, 1, 1))
+barplot(
+  plot_values, 
+  ylim = c(0,3), 
+  xaxt = "n", 
+  # yaxt = "n", 
+  xlab = paste(length(plot_values), "OTUs sorted by decreasing abundance"), 
+  ylab = "Millions of reads across samples", 
+  las = 1)
+
+# axis(side = 2, at = pretty(range(plot_values)), las = 1) #, lwd = 0, lwd.ticks = 1
+
+text1 <- paste(gt_t1, "sequences >", threshold_1)
+text2 <- paste(gt_t2, "sequences >", threshold_2)
+text3 <- paste(gt_t3, "sequences >", threshold_3)
+
+legend("topright", legend = c(text1, text2, text3), bty = "n")
+
+text(x = length(plot_values), y = 10000/plot_scale, labels = threshold_3, col = "red", pos = 3)
+abline(h = threshold_3/plot_scale, col = "red")
+
+dev.off()

### 2016-04-05 Analysis/dissimilarity.R
add dissimilarity index calculation

@@ -1,0 +1,35 @@
+# calculate dissimilarity
+
+library(vegan) # vegdist
+# rownames must be unique sample IDs. get this from 1_data_prep.R
+otu_table <- otu_filt
+
+
+# vector of groups to which samples belong.
+group_vector <- metadata$env_sample_name
+
+dis_by_sample <- function(contingency_table, grouping_vector)
+{
+  return(
+    lapply(
+      lapply(
+        split(otu_table, group_vector), 
+        matrix, ncol = ncol(otu_table)
+      ), 
+      vegdist, method = "bray", upper = TRUE, diag = TRUE
+    )
+  )
+}
+
+my_dis <- dis_by_sample(otu_table, group_vector)
+
+
+
+par(mar = c(4,6,1,1))
+stripchart(my_dis, method = "jitter", pch = 21, las = 1, 
+           xlab = "Pairwise Bray-Curtis Dissimilarity", bg = rgb(0,0,0,alpha = 0.2))
+
+# note high value for sample "lib_B_tag_GCGCTC" in PCT-C-0500
+
+# my_dis[1]
+# lapply(my_dis, function(x) max())

### 2016-04-05 Analysis/plot_stacked_bar.R
add plotting of stacked bar column chart

@@ -1,0 +1,16 @@
+# plot stacked bar
+
+head(data_full_long)
+
+
+tmp <- data_full_long[order(data_full_long$env_sample_name),]
+head(plot_dat)
+plot_dat <- tmp[tmp$env_sample_name == "PCT-C-0500",]
+plot_dat <- plot_dat[plot_dat$count > 200,]
+library(ggplot2)
+p <- ggplot(aes(x = sample_id, weight = count, fill = OTU), data = plot_dat)
+p + geom_bar() + 
+  scale_fill_discrete("Legend Title") + 
+  labs(x="X Label", y="Y Label", title="An Example Stacked Column Chart")
+
+# note high value for sample "lib_B_tag_GCGCTC" in PCT-C-0500

### 2016-04-05 Analysis/1_data_prep.R
add removal of outlier sample

@@ -12,9 +12,9 @@ fig_dir <- file.path("..", "Figures")
 # use the resulting object 'otu_filt' in place of table_transposed
 # OR use t(otu_filt) as object 'table_full'
 
-filename_orig <- file.path(data_dir, "OTUs_fam_w30_top10.csv")
+# filename_orig <- file.path(data_dir, "OTUs_fam_w30_top10.csv")
 
-filename_out <- file.path(data_dir, "OTUs_top10.csv")
+# filename_out <- file.path(data_dir, "OTUs_top10.csv")
 
 # table_full <- read.csv(filename_orig, row.names = 1)
 # maybe ? table_full <- t(otu_filt)
@@ -40,11 +40,11 @@ metadata_exp <- cbind.data.frame(metadata, sequenced_sample, transect_position, 
 table_restricted <- table_full[ , as.character(metadata[,colname_sampleid])]
 
 # write the file
-write.csv(x = table_restricted, file = filename_out, quote = FALSE)
+# write.csv(x = table_restricted, file = filename_out, quote = FALSE)
 
 # COMBINE UP DATA
 table_transposed <- t(table_restricted[,as.character(metadata[,colname_sampleid])])
-
+# table_transposed <- otu_filt
 if(
   identical(
     rownames(table_transposed),
@@ -63,9 +63,10 @@ if(
 
 # data_full_path <- file.path(data_dir, "data_full_top10.csv")
 data_full_path <- file.path(data_dir, "data_full_filt.csv")
-write.csv(x = data_full, file = data_full_path, row.names = FALSE)
+# write.csv(x = data_full, file = data_full_path, row.names = FALSE)
 
 # LONG FORMAT:
+library(reshape2)
 otu_long <- melt(otu_filt)
 colnames(otu_long) <- c("sample_id", "OTU", "count")
 otu_long$OTU <- as.character(otu_long$OTU)
@@ -76,8 +77,18 @@ row_match_long <- match(otu_long$sample_id , metadata_exp[, colname_sampleid])
 data_full_long <- cbind.data.frame(metadata_exp[row_match_long,], otu_long[,c("OTU", "count")])
 
 data_full_long_path <- file.path(data_dir, "data_full_long.csv")
-write.csv(x = data_full_long, file = data_full_long_path, row.names = FALSE)
+# write.csv(x = data_full_long, file = data_full_long_path, row.names = FALSE)
 
[…]

### 2016-04-05 Manuscript/v0.0/carkeek_grid.tex
describe rarefaction
add total number of reads
note removal of outlier PCr

@@
 
+We assessed PCR contamination by evaluating the dissimilarity of replicate PCRs of the same DNA sample, and removed one sample for which the Bray-Curtis dissimilarities between itseld and the other replicates exceeded 0.1 (lib_B_tag_GCGCTC).
+
 %To infer a general estimate of the rate of cross-sample contamination likely to be associated with lab procedures, we calculated the proportion of reads containing primer index sequences (or combinations thereof?) that we did not actually use
@@
 %
-%Total sequencing clusters passing filter per read:
+%Total sequencing clusters passing filter per read: 371,576,190 on NEXTSEQ
 %
+% THESE BELOW ARE FROM A MISEQ RUN
 %reads, reads passing filter, % >= Q30, for each run ( 7,052,307 ; 6,752,829; 97.47 :: 7,052,307 ; 6,752,829 ; 95.50)
@@
 %----------------------------------------------------------------------------
+
+To scale the OTU counts, we calculated the minimum number of OTU-assigned reads (as opposed to raw number of reads) found in these samples (130402), multiplied this by within-sample proportional abundance of each OTU, and finally rounded these numbers.
 

### 2016-04-05 Analysis/cluster.R
minor tweak

@@ -13,7 +13,7 @@ gghue <- function(n){
 
 # choose between using proportional or raw counts
 # DATA <- otu_table
-DATA <- otu_table_prop[,]
+DATA <- otu_filt[,]
 
 mydist <- vegdist(DATA)
 
@@ -22,12 +22,11 @@ pam_out <- mypam$pamobject
 mypam$nc # number of clusters
 
 plot(pam(mydist, k = mypam$nc), which.plots = 1)
-?pam
 
 mycolors <- gghue(mypam$nc)
 
 # see ?clusplot.default
-pdf(file = file.path(fig_dir, "pam_plot.pdf"))
+# pdf(file = file.path(fig_dir, "pam_plot.pdf"))
 	par(mar = c(5,5,1,1))
 	clusplot(
 		x = as.matrix(mydist), 
@@ -41,7 +40,7 @@ pdf(file = file.path(fig_dir, "pam_plot.pdf"))
 		col.clus = mycolors, 
 		main = NA
 	)
-dev.off()
+# dev.off()
 
 
 set.seed(2) # to control jitter

### 2016-04-12 Manuscript/v0.0/carkeek_grid.tex
finalize outlier removal

@@
 
-We assessed PCR contamination by evaluating the dissimilarity of replicate PCRs of the same DNA sample, and removed one sample for which the Bray-Curtis dissimilarities between itseld and the other replicates exceeded 0.1 (lib_B_tag_GCGCTC).
+We assessed PCR contamination by evaluating the dissimilarity of replicate PCRs of the same DNA sample, and removed one sample for which the Bray-Curtis dissimilarities between itseld and the other replicates exceeded 0.1 (lib\_B\_tag\_GCGCTC).
 
@@
 There was a very low frequency of cross-contamination from other libraries into those reported here (5e-05$\pm$8e-05; max 0.00034)
+
+We assessed the consistency of PCR by conducting 4 replicate PCRs for each environmental sample and calculating the mean pairwise Bray-Curtis dissimilarity of the resulting communities (scaled to minimum read depth per sample).
+92 of the 96 amplicon samples had mean Bray-Curtis dissimilarity $\leq$ 0.052; 1 sample had a value of 0.341, which elevates the value of the other replicates. 
+After removal of this sample, the highest mean Bray-Curtis dissimilarity among replicates within an environmental sample was 0.034.
 

### 2016-04-12 Analysis/1_data_prep.R

@@ -15,6 +15,7 @@
 # write the file
 # write.csv(x = table_restricted, file = filename_out, quote = FALSE)
 
+# RESCALE TO EQUAL SEQUENCING DEPTHS PER SAMPLE
 # calculate the minimum number of reads assigned to these OTUs in these samples
 minreads <- min(rowSums(otu_table))
 
@@ -54,9 +55,11 @@ boxplot(otu_filt[,1:20])
 boxplot(otu_table_prop[,1:20])
 boxplot(scale(otu_table[,1:20]))
 
-min(scale(otu_table[,1:20])+2)
-boxplot(scale(otu_table[,1:20])+2)
 
+# CHECK FOR OUTLIERS
+# ------------------
+# If you'd like to check for and remove replicates that seem inconsistent, go to:
+# 'dissimilarity.R'
 
 # COMBINE UP DATA
 if(
@@ -94,16 +97,6 @@ data_for_cast <- data.frame(
   sample_id = data_full_long$sample_id, 
   OTU = data_full_long$OTU, 
   count = data_full_long$count)
-
-# note high value for sample "lib_B_tag_GCGCTC" in PCT-C-0500
-# remove that sample?
-remove_outlier <- TRUE
-if(remove_outlier == TRUE){
-  metadata <- metadata[metadata$sample_id != "lib_B_tag_GCGCTC",]
-  metadata_exp <- metadata_exp[metadata_exp$sample_id != "lib_B_tag_GCGCTC",]
-  otu_filt <- otu_filt[rownames(otu_filt) != "lib_B_tag_GCGCTC",]
-  print("removed outlier PCR")
-}
 
 
 #-----------------------------------------------------------

### 2016-04-12 Analysis/dissimilarity.R

@@ -25,23 +25,63 @@ dis_by_group <- lapply(
 
 my_dis <- dis_by_sample(my_table, group_vector)
 
+# which sample has the highest BC dissimilarity?
+dis_high <- which.max(sapply(my_dis, max))
+as.dist(my_dis[[dis_high]], upper = FALSE, diag = FALSE)
+mean(as.dist(my_dis[[dis_high]], upper = FALSE, diag = FALSE)[1:3])
+
 # calculate mean dissimilarity
 dis_mean <- lapply(my_dis, function(x) rowMeans(as.matrix(x)))
 
 # Which samples have high dissimilarity among other samples from the same environmental sample?
 dis_mean_v <- unlist(dis_mean)
 names(dis_mean_v) <- do.call(c, lapply(dis_mean, names))
-names(which.max(dis_mean_v))
 
-par(mar = c(4,6,1,1))
+# One "bad" replicate will have a high mean BC dissimilarity to all other replicates, but it will also elevate the BC dissimilarity of the other replicates
+round(sort(dis_mean_v), digits = 3)
+plot(dis_mean_v)
+# pdf(file = file.path(fig_dir, "BC_before_cleaning.pdf"))
+par(mar = c(4,7,1,1))
 stripchart(my_dis, method = "jitter", pch = 21, las = 1, 
            xlab = "Pairwise Bray-Curtis Dissimilarity", bg = rgb(0,0,0,alpha = 0.2))
+# dev.off()
 
 # note high value for sample "lib_B_tag_GCGCTC" in PCT-C-0500
+# in this case, 0.2 seems to be a reasonable cutoff beyond which we think a PCR replicate is "bad"
+names(which(dis_mean_v > 0.2))
+names(which.max(dis_mean_v))
+
+# remove that sample?
+remove_outlier <- TRUE
+if(remove_outlier == TRUE){
+  outlier <- names(which(dis_mean_v > 0.2))
+  metadata <- metadata[metadata$sample_id != outlier,]
+  metadata_exp <- metadata_exp[metadata_exp$sample_id != outlier,]
+  otu_filt <- otu_filt[rownames(otu_filt) != outlier,]
+  print("removed outlier PCR")
+}
+
+my_table <- otu_filt
+if(!(identical(metadata[,colname_sampleid], rownames(my_table)))){
+	warning("hold on, the otu and metadata rows are not in the same order; disaster could ensue")
+}
+# vector of groups to which samples belong.
[…]

### 2016-04-13 Analysis/slice_sample.R
add function to sample columns

@@ -1,0 +1,47 @@
+################################################################################
+# A function to sample columns from a ( matrix | data frame | table )
+# prop_prob : should the columns be sampled...
+# ... randomly ('equal')
+# ... proportional to their sum ('prop')
+# ... inversely proportional to their sum ('invprop')
+# frac : fraction of columns to sample
+################################################################################
+
+slice_sample <- function(dat, prop_prob = "equal", frac = 1) { # threshold
+	
+	if(class(dat) != "matrix" && class(dat) != "data.frame") {
+		stop("'dat' must be a matrix or data frame")
+	}
+	
+	if(prop_prob != "equal" && prop_prob != "prop" && prop_prob != "invprop" ) {
+		stop("'prop_prob' must be 'equal', 'prop' or 'invprop'")		
+	}
+	
+	if(frac > 1 | frac <= 0) {
+		stop("'frac' must be > 0 and < = 1")
+	}
+	
+	sums <- colSums(dat)
+	
+	if       (prop_prob == "equal") {
+		probs <- NULL
+	} else if(prop_prob == "prop") {
+		probs <- sums/sum(sums)	
+	} else if(prop_prob == "invprop" ) {
+		probs <- (sum(sums)/sums)/sum(sum(sums)/sums)
+	}
+
+	samp_cols <- sample(
+		x = 1:ncol(dat), 
+		size = round(frac * ncol(dat)), 
+		replace = FALSE, 
+		prob = probs
+		)
+		
+	return(dat[ , sort(samp_cols)])
+	
+}
+
+# tests
+# DAT <- matrix(data = rpois(100, lambda = 1), nrow = 10)
+slice_sample(dat = DAT, prop_prob = "equal", frac = 0.5)

### 2016-04-13 Analysis/rm_single_rows.R
add function to strip columns with occurrence in a single sample

@@ -1,0 +1,16 @@
+################################################################################
+# A function to remove from a ( matrix | data frame | table )
+# columns which have a value in only one row.
+################################################################################
+
+rm_single_rows <- function(dat) {
+
+	if(class(dat) != "matrix" && class(dat) != "data.frame") {
+		stop("'dat' must be a matrix or data frame")
+	}
+	
+	n_samples <- apply(X = dat, MARGIN = 2, FUN = function(x) sum(x > 0))
+	
+	return(dat[ , n_samples > 1])
+
+}

### 2016-04-13 Analysis/1_data_prep.R
incorporate exclusion of spatially invariant otus

@@ -86,6 +86,12 @@ stripchart(as.data.frame(otu_01[,1:20]), pch = 19, col = rgb(0,0,0, alpha = 0.2)
 #-------------------------------------------------------------------------------
 
 #-------------------------------------------------------------------------------
+# exclude otus that have no spatial variance (occur in only one sample)
+source("rm_single_rows.R")
+otu_spvar <- rm_single_rows(otu_mean)
+#-------------------------------------------------------------------------------
+
+#-------------------------------------------------------------------------------
 # COMBINE UP DATA
 if(
   identical(

### 2016-04-13 Analysis/diss_by_dist.R
omit within sample comparisons

@@ -4,20 +4,20 @@ library(geosphere) # distm()
 
 my_otu <- otu_mean
 rownames(otu_mean) # PCT-C-0000 etc, aka "env_sample_name"
-my_metadata <- metadata_exp[!duplicated(metadata_exp[,"env_sample_name"]),]
+my_metadata <- metadata_mean# metadata_exp[!duplicated(metadata_exp[,"env_sample_name"]),]
 
 # calculate pairwise great circle distance between sampling locations using Haversine method
-geo_dist <- distm(x = my_metadata[,c("lon", "lat")])
-dimnames(geo_dist) <- list(my_metadata$env_sample_name, my_metadata$env_sample_name)
+geo_dist <- as.dist(distm(x = my_metadata[,c(colname_lon, colname_lat)], fun = distHaversine))
+# dimnames(geo_dist) <- list(my_metadata$env_sample_name, my_metadata$env_sample_name)
 
 # test:
-for(i in 1:length(my_metadata$env_sample_name)){
-	if(geo_dist[my_metadata$env_sample_name[i], my_metadata$env_sample_name[i]] != 0){
-		warning("distance between a point and itself is not zero, this indicates a problem")
-	}
-}
+# for(i in 1:length(my_metadata$env_sample_name)){
+	# if(geo_dist[my_metadata$env_sample_name[i], my_metadata$env_sample_name[i]] != 0){
+		# warning("distance between a point and itself is not zero, this indicates a problem")
+	# }
+# }
 
-comm_dist <- as.matrix(vegdist(my_otu, method = "bray", diag = TRUE, upper = TRUE))
+comm_dist <- vegdist(my_otu, method = "bray", diag = TRUE, upper = TRUE)
 if(!(identical(dimnames(comm_dist), dimnames(geo_dist)))){
 	warning("Whoa there! the row/column names of the two distance matrices do not seem to add up. This is bad.")
 }

### 2016-04-13 Analysis/diss_by_dist.R
add smoothing line

@@ -22,7 +22,7 @@ if(!(identical(dimnames(comm_dist), dimnames(geo_dist)))){
 	warning("Whoa there! the row/column names of the two distance matrices do not seem to add up. This is bad.")
 }
 
-pdf(file = file.path(fig_dir, "diss_by_dist.pdf"), width = 8, height = 3)
+# pdf(file = file.path(fig_dir, "diss_by_dist.pdf"), width = 8, height = 3)
 	par(mar = c(4,5,1,1))
 	plot(
 		x = log(geo_dist + 100), 
@@ -37,8 +37,16 @@ pdf(file = file.path(fig_dir, "diss_by_dist.pdf"), width = 8, height = 3)
 		ylab = "Bray-Curtis dissimilarity"
 	)
 	axis(side = 1, at = unique(log(metadata$dist_from_shore + 100)), labels = unique(metadata$dist_from_shore))
-dev.off()
+# dev.off()
 # abline(v = unique(log(metadata$dist_from_shore + 100)))
+smoothed <- loess.smooth(
+				x = log(geo_dist + 100), 
+				y = comm_dist, 
+				span = 2/3, 
+				degree = 1, 
+				family = "gaussian"
+				)
+lines(smoothed, col="green", lwd=2)
 
 
 

### 2016-04-13 Analysis/diss_by_dist.R
improve analysis and plotting of community dissimilarity by geographic distance

@@ -3,7 +3,7 @@
 library(geosphere) # distm()
 
 my_otu <- otu_mean
-rownames(otu_mean) # PCT-C-0000 etc, aka "env_sample_name"
+rownames(my_otu) # should be e.g. PCT-C-0000 etc, aka "env_sample_name"
 my_metadata <- metadata_mean# metadata_exp[!duplicated(metadata_exp[,"env_sample_name"]),]
 
 # calculate pairwise great circle distance between sampling locations using Haversine method
@@ -22,54 +22,69 @@ if(!(identical(dimnames(comm_dist), dimnames(geo_dist)))){
 	warning("Whoa there! the row/column names of the two distance matrices do not seem to add up. This is bad.")
 }
 
+geo_dist_scaled <- log(geo_dist + 100)
+plot_x <- geo_dist # geo_dist_scaled
+
 # pdf(file = file.path(fig_dir, "diss_by_dist.pdf"), width = 8, height = 3)
 	par(mar = c(4,5,1,1))
 	plot(
-		x = log(geo_dist + 100), 
+		x = plot_x, 
 		y = comm_dist, 
 		ylim = c(0,1), 
 		xaxt = "n", 
-		pch = 1, 
+		pch = 21, 
 		las = 1, 
 		cex = 1, 
-		col = rgb(0,0,0), #,alpha = 0.1 
+		col = 1, 
+		bg = rgb(0,0,0,alpha = 0.1 ), #,alpha = 0.1 
 		xlab = "Distance between samples (meters)", 
 		ylab = "Bray-Curtis dissimilarity"
 	)
-	axis(side = 1, at = unique(log(metadata$dist_from_shore + 100)), labels = unique(metadata$dist_from_shore))
-# dev.off()
+	axis(side = 1)
+	#, at = unique(log(metadata$dist_from_shore + 100)), labels = unique(metadata$dist_from_shore)
 # abline(v = unique(log(metadata$dist_from_shore + 100)))
 smoothed <- loess.smooth(
-				x = log(geo_dist + 100), 
+				x = plot_x, 
 				y = comm_dist, 
 				span = 2/3, 
 				degree = 1, 
 				family = "gaussian"
 				)
-lines(smoothed, col="green", lwd=2)
+lines(smoothed, col="red", lwd=2)
+# dev.off()
 
 
-
-
-
+#-------------------------------------------------------------------------------
+# SPLIT BY TRANSECT
[…]

### 2016-04-13 Manuscript/v0.0/carkeek_grid.tex
add text on relationship between distance and dissimilarity

@@
 \usepackage[utf8]{inputenc} % Allow input of accented characters (and more...)
+\usepackage{textcomp,marvosym} % additional figures, including textdegree
 \usepackage{graphicx} %  figures
@@
 
+% We measure 'extra-(corporeal|organismal) DNA
+Some authors have cautioned against the use of DNA-based macrobial communities in marine environments because they are subject to dynamic physical forces (CITE).
+If environmental circulation 
+In general, the relationship between community dissimilarity (0 = identical; 1 = completely different) and spatial distance is expected to be asymptotic, because communities nearer to each other tend to be more similar than those farther apart.
+The intercept is expected to be near 0, because samples taken at the same place should be very similar.
+% technically impossible for samples to be taken at exactly 0; they are actually immediately adjacent.
+Deviation in the intercept from 0 indicates heterogeneous community composition/structure over fine scales.
+A flat relationship between dissimilarity and distance indicates that heterogeneity is not assorted spatially, and can be interpreted in different ways, depending on the mean.
+If the mean is near 1, the spatial heterogeneity has overwhelmed the spatial scale of sampling.
+If the mean is near 0, there is no community heterogeneity over the scale sampled.
+
 %-----------------------------------METHODS-------------------------------------
@@
 
-We assessed PCR contamination by evaluating the dissimilarity of replicate PCRs of the same DNA sample, and removed one sample for which the Bray-Curtis dissimilarities between itseld and the other replicates exceeded 0.1 (lib\_B\_tag\_GCGCTC).
+We assessed PCR contamination by evaluating the dissimilarity of replicate PCRs of the same DNA sample, and removed one sample for which the Bray-Curtis dissimilarities between itself and the other replicates exceeded 0.1 (lib\_B\_tag\_GCGCTC).
 
@@
   \centering
-    \includegraphics[width=1\textwidth]{../../Figures/comm_diss_by_geo_dist.pdf}
-    \caption{Pairwise Bray-Curtis dissimilarity of eDNA communities plotted against pairwise spatial distance.}
+    \includegraphics[width=1\textwidth]{../../Figures/diss_by_dist.pdf}
+    \caption{Pairwise Bray-Curtis dissimilarity of eDNA communities plotted against pairwise spatial distance.
+    Line represents prediction of Gaussian LOESS (degree = 1; span = 2/3).
+    Restricting comparison to within-transect has no qualitative difference in the outcome (see 'diss\_by\_dist\_by\_transect.pdf').
+    }
   \label{comm_diss_by_geo_dist} % use this to refer to your figure in the text, so that numbering updates automatically

### 2016-04-13 Analysis/cluster.R
improve PAM output

@@ -13,11 +13,11 @@ gghue <- function(n){
 
 # choose between using proportional or raw counts
 # DATA <- otu_table
-DATA <- otu_filt[,]
+# DATA <- otu_filt[,]
 DATA <- otu_mean
 metadata <- metadata_exp[!duplicated(metadata_exp[,"env_sample_name"]),]
 
-mydist <- vegdist(DATA, binary = TRUE)
+mydist <- vegdist(DATA, method = "bray") # , binary = TRUE
 
 mypam <- pamk(data = mydist) # to restrict range of Ks considered: , krange = 2:4
 pam_out <- mypam$pamobject
@@ -28,7 +28,7 @@ plot(pam(mydist, k = mypam$nc), which.plots = 1)
 mycolors <- gghue(mypam$nc)
 
 # see ?clusplot.default
-# pdf(file = file.path(fig_dir, "pam_plot.pdf"))
+pdf(file = file.path(fig_dir, "pam_plot.pdf"))
 	par(mar = c(5,5,1,1))
 	clusplot(
 		x = as.matrix(mydist), 
@@ -42,7 +42,7 @@ mycolors <- gghue(mypam$nc)
 		col.clus = mycolors, 
 		main = NA
 	)
-# dev.off()
+dev.off()
 
 
 set.seed(2) # to control jitter
@@ -70,7 +70,7 @@ set.seed(2) # to control jitter
 
 
 # USING MEAN DATA
-# pdf(file = file.path(fig_dir, "medoids_in_space_mean.pdf"), width = 7, height = 7)
+pdf(file = file.path(fig_dir, "pam_in_space.pdf"), width = 5, height = 7)
 plot(
 		x = metadata$dist_along_shore, 
 		xaxt = "n", 
@@ -80,7 +80,24 @@ plot(
 		# log = "y",
 		col = mycolors[mypam$pamobject$clustering[metadata$env_sample_name]], 
 		# bg = mycolors[mypam$pamobject$clustering[metadata$sample_id]], 
-		pch = as.character(mypam$pamobject$clustering[metadata$env_sample_name]-1), 
+		pch = 19, #as.character(mypam$pamobject$clustering[metadata$env_sample_name]-1)
+		cex = 4, 
+		# cex = 1.5, 
+		main = "membership to PAM classifications", 
+		las = 1, 
+		xlab = "Position along shore (meters)", 
+		ylab = "Position from 0 (meters)"
+	)
+points(
+		x = metadata$dist_along_shore, 
[…]

### 2016-04-14 Manuscript/v0.0/carkeek_grid.tex
add correct pam plot
improve PAM output
tried clustering on spatially invariant taxa

@@
 % because of the nature of our data (lots of zeros, with some very high counts), the mean is not very informative
-We simultaneously assessed the existence of distinct community types and the membership of samples to those community types using a partitioning around mediods algorithm (CITE PAM, sometimes referred to as k-mediods clustering), as implemented in the R package fpc (CITE fpc).
+We simultaneously assessed the existence of distinct community types and the membership of samples to those community types using a partitioning around mediods algorithm (CITE PAM, sometimes referred to as k-medoids clustering), as implemented in the R package fpc (CITE fpc).
 The classification of samples to communities was made on the basis of their pairwise Bray-Curtis dissimilarity, calculated using the function vegdist in the R package vegan (CITE VEGAN).
@@
 \subsection*{Community Analysis}
+Excluding spatially-invariant taxa ?(taxa which occur in only one spatial location) had no discernible effect on the outcome of the PAM analysis (number of clusters, assignment to clusters).
 
@@
   \centering
+    \includegraphics[width=1\textwidth]{../../Figures/diss_by_dist.pdf}
+    \caption{Pairwise Bray-Curtis dissimilarity of eDNA communities plotted against pairwise spatial distance.
+    Line represents prediction of Gaussian LOESS (degree = 1; span = 2/3).
+    Restricting comparison to within-transect has no qualitative difference in the outcome (see 'diss\_by\_dist\_by\_transect.pdf').
+    }
+  \label{comm_diss_by_geo_dist} % use this to refer to your figure in the text, so that numbering updates automatically
+\end{figure}
+
+\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
+  \centering
     \includegraphics[width=1\textwidth]{../../Figures/pam_plot.pdf}
@@
   \centering
-    \includegraphics[width=1\textwidth]{../../Figures/diss_by_dist.pdf}
-    \caption{Pairwise Bray-Curtis dissimilarity of eDNA communities plotted against pairwise spatial distance.
-    Line represents prediction of Gaussian LOESS (degree = 1; span = 2/3).
-    Restricting comparison to within-transect has no qualitative difference in the outcome (see 'diss\_by\_dist\_by\_transect.pdf').
+    \includegraphics[width=1\textwidth]{../../Figures/pam_in_space.pdf}
+    \caption{Geographic position of collected samples, colored by membership to clusters identified by partitioning around medoids algorithm. 
+%    Points are jittered in both horizontal and vertical dimension to distinguish among four replicate PCR products sequenced from each environmental sample.
 }
-  \label{comm_diss_by_geo_dist} % use this to refer to your figure in the text, so that numbering updates automatically
+  \label{clustering_map} % use this to refer to your figure in the text, so that numbering updates automatically
 \end{figure}
@@
   \centering
-    \includegraphics[width=1\textwidth]{../../Figures/medoids_in_space.pdf}
-    \caption{Geographic position of collected samples, colored by membership to clusters identified by partitioning around medoids algorithm.
-Points are jittered in both horizontal and vertical dimension to distinguish among four replicate PCR products sequenced from each environmental sample.}
-  \label{clustering_map} % use this to refer to your figure in the text, so that numbering updates automatically
-\end{figure}
-
-\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
-  \centering
     \includegraphics[width=1\textwidth]{../../Figures/otu_in_space.pdf}

### 2016-04-14 Analysis/cluster.R

@@ -1,5 +1,7 @@
 
 # INPUT COMES FROM SCRIPT "load_data.r"
+# Watch out: if number of clusters is >9, the plotting will look funky
+# Only single digits can be used as symbols in base plotting
 
 library(fpc) #pamk()
 library(cluster) #pam
@@ -12,12 +14,12 @@ gghue <- function(n){
 }
 
 # choose between using proportional or raw counts
-# DATA <- otu_table
-# DATA <- otu_filt[,]
-DATA <- otu_mean
-metadata <- metadata_exp[!duplicated(metadata_exp[,"env_sample_name"]),]
+# my_table <- otu_table
+# my_table <- otu_filt[,]
+my_table <- otu_mean # otu_mean, otu_spvar
+my_metadata <- metadata_mean #metadata_exp[!duplicated(metadata_exp[,"env_sample_name"]),]
 
-mydist <- vegdist(DATA, method = "bray") # , binary = TRUE
+mydist <- vegdist(my_table, method = "bray") # , binary = TRUE
 
 mypam <- pamk(data = mydist) # to restrict range of Ks considered: , krange = 2:4
 pam_out <- mypam$pamobject
@@ -31,80 +33,83 @@ mycolors <- gghue(mypam$nc)
 pdf(file = file.path(fig_dir, "pam_plot.pdf"))
 	par(mar = c(5,5,1,1))
 	clusplot(
-		x = as.matrix(mydist), 
-		clus = pam_out$clustering, 
-		sub = NULL, 
-		lines = 0, 
-		labels = 4, 
-		color = FALSE, 
-		col.p = mycolors[pam_out$clustering], 
-		col.txt = mycolors, 
-		col.clus = mycolors, 
+		x = as.matrix(mydist),
+		clus = pam_out$clustering,
+		sub = NULL,
+		lines = 0,
+		labels = 4,
+		color = FALSE,
+		col.p = mycolors[pam_out$clustering],
+		col.txt = mycolors,
+		col.clus = mycolors,
 		main = NA
 	)
 dev.off()
 
 
-set.seed(2) # to control jitter
-# pdf(file = file.path(fig_dir, "medoids_in_space.pdf"), width = 7, height = 7)
-	plot(
-		x = metadata$dist_along_shore, 
-		xaxt = "n", 
-		xlim = c(-500, 2500), 
-		y = log(metadata$dist_from_shore + 10), 
-		yaxt = "n",
-		# log = "y",
[…]

### 2016-04-14 Manuscript/v0.0/carkeek_grid.tex
explain data subsetting

@@
 \subsection*{Community Analysis}
+
+We subset the data in a variety of ways and conducted each analysis on all subsets.
+We report the subset used with each analysis, and report results on alternative subsets in the supplemental material. 
+For all analyses beyond the assessment of PCR consistency, we use the mean taxon abundance across PCR replicates from each of the 24 environmental samples.
+Our subsetting methods were (1) exclude rare taxa ?(threshold)?, (2) exclude abundant taxa ?(threshold)?, (3) subsampling of taxa randomly, (4) subsampling of taxa proportional to their abundance, (5) subsampling of taxa inversely proportional to their abundance, (6) exclude taxa found in only one environmental sample (spatially invariant), (7) exclude non-marine taxa (e.g.
+humans, pigs), (8) exclude taxa whose known individual range (including gametes and larvae) exceeds the spatial scale of our study.
+We also tested a variety of transformations of the mean scaled abundance data, including (1) log ($log_{e}x$), and (2) binary ($1 = x > 1$; $0 = x < 1$).
+
 % Used k-medoids (PAM) rather than k-means because it does not require euclidean distances can be computed and that the observations follow a gaussian distribution

### 2016-04-14 Manuscript/v0.0/carkeek_grid.tex
small changes to dissimilarity

@@
 Some authors have cautioned against the use of DNA-based macrobial communities in marine environments because they are subject to dynamic physical forces (CITE).
-If environmental circulation 
+%If environmental circulation 
 In general, the relationship between community dissimilarity (0 = identical; 1 = completely different) and spatial distance is expected to be asymptotic, because communities nearer to each other tend to be more similar than those farther apart.
-The intercept is expected to be near 0, because samples taken at the same place should be very similar.
+The intercept is expected to be 0, because only within-sample comparisons can have 0 spatial separation, and communities have no dissimilarity within a sample.
 % technically impossible for samples to be taken at exactly 0; they are actually immediately adjacent.
@@
 A flat relationship between dissimilarity and distance indicates that heterogeneity is not assorted spatially, and can be interpreted in different ways, depending on the mean.
-If the mean is near 1, the spatial heterogeneity has overwhelmed the spatial scale of sampling.
-If the mean is near 0, there is no community heterogeneity over the scale sampled.
+If the mean is 1, the spatial heterogeneity has overwhelmed the spatial scale of sampling.
+If the mean is 0, all samples are identical, and we infer there is complete community homogeneity over the scale sampled.
 

### 2016-04-14 Analysis/diss_by_dist.R

@@ -10,14 +10,7 @@ my_metadata <- metadata_mean# metadata_exp[!duplicated(metadata_exp[,"env_sample
 geo_dist <- as.dist(distm(x = my_metadata[,c(colname_lon, colname_lat)], fun = distHaversine))
 # dimnames(geo_dist) <- list(my_metadata$env_sample_name, my_metadata$env_sample_name)
 
-# test:
-# for(i in 1:length(my_metadata$env_sample_name)){
-	# if(geo_dist[my_metadata$env_sample_name[i], my_metadata$env_sample_name[i]] != 0){
-		# warning("distance between a point and itself is not zero, this indicates a problem")
-	# }
-# }
-
-comm_dist <- vegdist(my_otu, method = "bray", diag = TRUE, upper = TRUE)
+comm_dist <- vegdist(my_otu, method = "bray") #, diag = TRUE, upper = TRUE
 if(!(identical(dimnames(comm_dist), dimnames(geo_dist)))){
 	warning("Whoa there! the row/column names of the two distance matrices do not seem to add up. This is bad.")
 }

### 2016-04-15 Analysis/variogram.R
add variogram code and plots

@@ -1,0 +1,19 @@
+my_table    <- otu_mean
+my_metadata <- metadata_mean
+
+library(geoR) # variog
+
+pdf(file = file.path(fig_dir, "variograms_taxa.pdf"))
+for(i in 1:20){
+
+	vg <- variog(coords = my_metadata[,c(colname_lon, colname_lat)], data = my_table[,i])
+
+	plot(vg, type = "b", main = colnames(my_table)[i])
+
+}
+dev.off()
+# vg.summary <- cbind(c(1:10), v2$v, v2$n)
+# colnames(v2.summary) <- c("lag", "semi-variance", "# of pairs")
+# v2.summary
+
+# plot(v2, type = "b", main = "Variogram: Av8top")

### 2016-04-15 Analysis/variogram.R
add more variogram plots

@@ -3,7 +3,8 @@ my_metadata <- metadata_mean
 
 library(geoR) # variog
 
-pdf(file = file.path(fig_dir, "variograms_taxa.pdf"))
+# variogram for individual taxon abundance
+pdf(file = file.path(fig_dir, "variogram_taxa.pdf"))
 for(i in 1:20){
 
 	vg <- variog(coords = my_metadata[,c(colname_lon, colname_lat)], data = my_table[,i])
@@ -12,8 +13,22 @@ for(i in 1:20){
 
 }
 dev.off()
-# vg.summary <- cbind(c(1:10), v2$v, v2$n)
-# colnames(v2.summary) <- c("lag", "semi-variance", "# of pairs")
-# v2.summary
 
-# plot(v2, type = "b", main = "Variogram: Av8top")
+
+# variogram for shannon index
+pdf(file = file.path(fig_dir, "variogram_shannon.pdf"))
+shannon <- diversity(my_table, index = "shannon")
+vg <- variog(coords = my_metadata[,c(colname_lon, colname_lat)], data = my_table[,i])
+plot(vg, type = "b", main = "Variogram of Shannon Index")
+dev.off()
+
+
+# variogram for taxon richness (number of taxa w/ mean abundance > 1)
+thresholds <- c(0, 1, 10, 100, 1000, 10000)
+pdf(file = file.path(fig_dir, "variogram_richness.pdf"))
+for(i in 1:length(thresholds)){
+	richness <- rowSums(my_table > i)
+	vg <- variog(coords = my_metadata[,c(colname_lon, colname_lat)], data = richness)
+	plot(vg, type = "b", main = paste("Variogram of Richness, abundance threshold = ", thresholds[i]))
+}
+dev.off()

### 2016-04-15 Analysis/moran.R
calculate morans I

@@ -1,0 +1,33 @@
+# Calculate Moran's I
+
+my_table    <- otu_mean[,1:50]
+my_metadata <- metadata_mean
+
+library(ape) # Moran.I
+
+my_dist <-  distm(my_metadata[,c(colname_lon, colname_lat)])
+my_dist_inv <- 1/my_dist
+diag(my_dist_inv) <- 0
+
+# Calculate Moran's I for the taxa:
+moran_out <- t(
+	apply(
+		X = my_table, 
+		MARGIN = 2, 
+		FUN = function(x) {
+			unlist(Moran.I(x = x, weight = my_dist_inv))
+		}
+		)
+	)
+
+plot(sort(moran_out[,"p.value"]))
+abline(h = 0.05, lty = "dashed")
+
+# is significance correlated with abundance?
+plot(colSums(my_table), moran_out[,"p.value"])
+abline(h = 0.05, lty = "dashed")
+
+# is significance correlated with the number of sites at which a taxon occurs?
+plot(colSums(my_table > 1), moran_out[,"p.value"])
+abline(h = 0.05, lty = "dashed")
+

### 2016-04-15 Analysis/diss_by_dist.R
add some analysis that might be of interest to community dissimilarity decay

@@ -46,6 +46,15 @@ smoothed <- loess.smooth(
 lines(smoothed, col="red", lwd=2)
 # dev.off()
 
+#-------------------------------------------------------------------------------
+# Some Possible analyses
+#-------------------------------------------------------------------------------
+lm_out <- lm(log(1/comm_dist) ~ geo_dist)
+# Then the regression coefficient is usually used in the literature as the descriptor of distance decay, or the distance at which 50% of the maximum similarity is observed.
+summary(lm_out)
+
+mantel(comm_dist, geo_dist, perm = 9999)
+
 
 #-------------------------------------------------------------------------------
 # SPLIT BY TRANSECT

### 2016-04-19 Analysis/add_names.R
misc attempts to add taxon names

@@ -1,4 +1,5 @@
-# add organism names
+# add organism names to an otu table
+the_otu_table <- otu_mean
 
 blast_file <- "query_hit_LCA.csv"
 
@@ -7,17 +8,96 @@ taxa_df <- read.csv(file.path(data_dir, blast_file))
 head(taxa_df)
 
 # remove abundance annotation from query sequence names
-
 taxa_df$query_seq <- gsub(pattern = ";size=.*", replacement = "", taxa_df$query_seq)
 head(taxa_df)
 
 
-taxa_vector <- match(colnames(otu_filt), taxa_df$query_seq)
+taxa_vector <- match(colnames(the_otu_table), taxa_df$query_seq)
+nrow(taxa_df)
 
-as.character(taxa_df$class_beste[taxa_vector])
-taxa_vector
-taxa_df[430,]
-otu_filt[1:5,1:5]
+# which taxa are not in the blast file?
+not_in_blastfile <- colnames(the_otu_table)[!colnames(the_otu_table) %in% taxa_df$query_seq]
+
+# how abundant are they?
+colSums(the_otu_table)[!colnames(the_otu_table) %in% taxa_df$query_seq]
+
+# taxon ID of each otu
+taxa_df$query_seq == 
+colnames(the_otu_table)
+
+colnames(taxa_df)
+head(classification_df)
+
+as.character(classification_df$query)
+
+# columns in the taxa_df with taxon_id
+as.character(taxa_df$LCA_id_all)
+as.character(taxa_df$LCA_id_beste)
+
+# column of the classification_df with taxon_id
+as.character(classification_df$query)
+
+match(as.character(taxa_df$LCA_id_all), as.character(classification_df$query))
+match(as.character(classification_df$query), as.character(taxa_df$LCA_id_beste))
+
+# what is the row of the taxa_df for each of the classification_df taxids
+unique(match(as.character(classification_df$query), as.character(taxa_df$LCA_id_beste)))
+
+classification_list <- split(classification_df, classification_df$query)
+classification_list[["1001339"]]
+some_id <- "1001339"
+classification_list[[some_id]]
+taxid_column <- "LCA_id_beste"
+seqnames <- colnames(otu_mean)
+out <- vector()
+taxids <- vector()
+for(i in 1:length(seqnames)) {
+  print(seqnames[i])
+  # out[i] <- (length(which(taxa_df$query_seq == seqnames[i])))
[…]

### 2016-04-19 Analysis/classification_df.R

@@ -1,0 +1,34 @@
+
+classifications_file <- "classifications20160329.RData"
+
+# I don't know how this works, I am just clicking to load the data
+# classifications <- load(file.path(data_dir, classifications_file))
+
+classifications
+
+
+the_ranks <- lapply(lapply(classifications, function(x) x[2]), "[[", 1)
+the_names <- lapply(classifications, "[[", 1)
+
+# unique(unlist(lapply(lapply(classifications, function(x) x[2]), "[[", 1), use.names = FALSE))
+
+classification_df <- data.frame(
+         query = rep(names(the_names), lapply(the_names, length)),
+         rankname = unlist(the_ranks), 
+         taxname = unlist(the_names), 
+         row.names = NULL
+         )
+
+current_time <- gsub(pattern = " ", replacement = "_", x = Sys.time())
+write.csv(classification_df,file = file.path(data_dir, paste("classification_", current_time, ".csv", sep = "")))
+
+#----------------------------------------------------------------------------------------
+# Get names of taxonomic ranks (e.g. "kingdom", "subphylum", etc)
+#----------------------------------------------------------------------------------------
+ranknames <- getranknames()
+unique_ranks <- sort(as.numeric(unique(ranknames[,"rankid"])))
+all_ranks <- tolower(ranknames[match(as.character(unique_ranks), ranknames[,"rankid"]),"rankname"]) # c(, "no rank")
+all_ranks_full <- c(rbind(all_ranks, paste("below-", all_ranks, sep = "")))
+
+
+

### 2016-04-19 Analysis/cluster.R

@@ -16,7 +16,7 @@ gghue <- function(n){
 # choose between using proportional or raw counts
 # my_table <- otu_table
 # my_table <- otu_filt[,]
-my_table <- otu_mean # otu_mean, otu_spvar
+my_table <- otu_mean # otu_mean, otu_spvar, otu_named
 my_metadata <- metadata_mean #metadata_exp[!duplicated(metadata_exp[,"env_sample_name"]),]
 
 mydist <- vegdist(my_table, method = "bray") # , binary = TRUE

### 2016-04-19 Analysis/otu_in_space.R
further messing around with taxa

@@ -6,8 +6,17 @@
 #    - "lon" and "lat" columns, or x and y coordinates
 #    - data matching rownames of abundance/otu table
 
-my_table <- otu_mean
+my_table <- otu_mean # otu_named
+if(sum(colSums(my_table) < 1) > 0){
+	print("removing otus with total abundance < 1")
+	my_table <- my_table[,colSums(my_table) >= 1]
+}
 my_metadata <- metadata_mean
+
+LOG_TRANSFORM <- FALSE
+if(LOG_TRANSFORM) {
+  my_table <- log(my_table + 1)
+}
 
 if(!identical(rownames(my_table), my_metadata[, colname_env_sample])){
 	warning("Hold your horses: the rownames of the otu table don't match up with the metadata. This will make bad things happen.")

### 2016-04-20 Manuscript/v0.0/carkeek_grid.tex
compile life history data
added michaelis-menten model of community dissimilarity by spatial distance

@@
 %If environmental circulation
-In general, the relationship between community dissimilarity (0 = identical; 1 = completely different) and spatial distance is expected to be asymptotic, because communities nearer to each other tend to be more similar than those farther apart.
+
+Samples collected of ecological communities may vary in dissimilarity from 0 (completely identical) to 1 (completely different). 
+For samples collected from multiple locations, the relationship between their spatial distance and community dissimilarity is of interest because it reflects the amount of community heterogeneity over the spatial scale sampled.
+%In general, the relationship between community dissimilarity (0 = identical; 1 = completely different) and spatial distance is expected to be asymptotic, because communities nearer to each other tend to be more similar than those farther apart.
 The intercept is expected to be 0, because only within-sample comparisons can have 0 spatial separation, and communities have no dissimilarity within a sample.
 % technically impossible for samples to be taken at exactly 0; they are actually immediately adjacent.
-Deviation in the intercept from 0 indicates heterogeneous community composition/structure over fine scales.
+Likewise, dissimilarity cannot exceed 1. 
+Deviation from 0 indicates heterogeneous community composition/structure over fine scales.
 A flat relationship between dissimilarity and distance indicates that heterogeneity is not assorted spatially, and can be interpreted in different ways, depending on the mean.
-If the mean is 1, the spatial heterogeneity has overwhelmed the spatial scale of sampling.
+If the mean is close to 1, there is high spatial heterogeneity over the spatial scale of sampling.
 If the mean is 0, all samples are identical, and we infer there is complete community homogeneity over the scale sampled.
+The rate at which community dissimilarity approaches the mean gives an indication of the rate of community turnover.
 
@@
 
+We compiled life history data for taxonomic groups at the family level and higher.
+For groups with sparse data at the family level (e.g. Nemertea), we used the data from the higher level group.
+
 \subsection*{Community Analysis}
@@
 % Calculate Moran's I on ?(relative abundance | raw count) of ?(taxa | OTU | uniq_sequences) to report dispersion of DNA in space?
+
+To estimate the maximum dissimilarity and the rate of community turnover in space, we modeled community dissimilarity as a function of distance from shore following a Michaelis-Menten relationship:
+\begin{equation}\label{MichaelisMenten}
+	com \sim V_{max}[d]/K_{m}[d]
+\end{equation}
+where $com$ is community dissimilarity, $d$ is spatial distance, and where the asymptote is given by $V_{max}$, and the distance at which half the asymptote has been reached is given by $K_{m}$. 
+Model fit was assessed using the function nls in R (CITE R).
 
@@
 Excluding spatially-invariant taxa ?(taxa which occur in only one spatial location) had no discernible effect on the outcome of the PAM analysis (number of clusters, assignment to clusters).
+
+The estimated asymptote of community dissimilarity as a function of spatial distance ($V_m$) was 0.72 ($p <<< 0.05$), and the distance at which half this dissimilarity was accumulated ($K_m$) is 23.8 kilometers ($p = 0.006$).
+Residual standard error of the fit of the model is 0.1563 on 274 degrees of freedom.
 
@@
     \caption{Pairwise Bray-Curtis dissimilarity of eDNA communities plotted against pairwise spatial distance.
-    Line represents prediction of Gaussian LOESS (degree = 1; span = 2/3).
+    Line represents prediction of the Non-linear Least Squares regression to a Michaelis-Menten model ($V_m = 0.72$, $p <<< 0.05$; $K_m = 23.8$ kilometers, $p = 0.006$; RSE = 0.1563; df = 274.).
     Restricting comparison to within-transect has no qualitative difference in the outcome (see 'diss\_by\_dist\_by\_transect.pdf').
@@
 
-\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
-  \centering
-    \includegraphics[width=1\textwidth]{../../Figures/slope_plots.pdf}
-    \caption{Fit lines of DNA sequence counts as a function of distance from shore for a selection of taxa for which we have strong preconceived expectations (left).
+%\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
+%  \centering
+%    \includegraphics[width=1\textwidth]{../../Figures/slope_plots.pdf}
+%    \caption{Fit lines of DNA sequence counts as a function of distance from shore for a selection of taxa for which we have strong preconceived expectations (left).
 Box plots of the estimates of the slopes for taxa ?(100 most abundant)?, grouped by life history traits (right).}
-  \label{slope_plots} % use this to refer to your figure in the text, so that numbering updates automatically
-\end{figure}
+%  \label{slope_plots} % use this to refer to your figure in the text, so that numbering updates automatically
+%\end{figure}
 
-\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
-  \centering
-    \includegraphics[width=1\textwidth]{../../Figures/var_boxplots.pdf}
-    \caption{Box plots of estimates of variance associated with each level of the multilevel model, corresponding to stages of the eDNA sampling protocol.}
-  \label{variance_boxplots} % use this to refer to your figure in the text, so that numbering updates automatically
-\end{figure}
+%\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
+%  \centering
+%    \includegraphics[width=1\textwidth]{../../Figures/var_boxplots.pdf}
+%    \caption{Box plots of estimates of variance associated with each level of the multilevel model, corresponding to stages of the eDNA sampling protocol.}
+%  \label{variance_boxplots} % use this to refer to your figure in the text, so that numbering updates automatically
+%\end{figure}
 

### 2016-04-20 Analysis/1_data_prep.R

@@ -43,15 +43,21 @@ dim(otu_scaled)
 otu_scaled <- otu_scaled[,order(colSums(otu_scaled), decreasing = TRUE)]
 otu_scaled[1:10,1:10]
 
-# how many OTUs occur more than a threshold number of times?
-abundance_threshold <- 100
-sum(colSums(otu_scaled) > abundance_threshold)
-# round(colSums(otu_scaled)/sum(otu_scaled)*100, digits = 2)
-# plot(colSums(otu_scaled)/sum(otu_scaled), type = "l")
+#-------------------------------------------------------------------------------
+# Should OTUs be excluded that occur fewer than a threshold number of times?
+EXCLUDE_RARE_OTUs <- TRUE
+abundance_threshold <- 1
 
-# exclude OTUs that were counted a total of fewer than 100 times across samples
-otu_filt <- otu_scaled[,which(colSums(otu_scaled) > abundance_threshold)]
-dim(otu_filt)
+if(EXCLUDE_RARE_OTUs) {
+  # how many OTUs will be retained?
+  sum(colSums(otu_scaled) > abundance_threshold)
+  # round(colSums(otu_scaled)/sum(otu_scaled)*100, digits = 2)
+  # plot(colSums(otu_scaled)/sum(otu_scaled), type = "l")
+  # exclude OTUs that were counted a total of fewer than 100 times across samples
+  otu_filt <- otu_scaled[,which(colSums(otu_scaled) > abundance_threshold)]
+  dim(otu_filt)
+}
+
 
 boxplot(otu_filt[,1:20])
 # boxplot(log(otu_table[,1:20]))
@@ -60,9 +66,11 @@ boxplot(scale(otu_table[,1:20]))
 
 
 #-------------------------------------------------------------------------------
-# CHECK FOR OUTLIERS
-# If you'd like to check for and remove replicates that seem inconsistent, go to:
-source('dissimilarity.R')
+CHECK_FOR_OUTLIERS <- TRUE
+# If you'd like to check for and remove inconsistent PCR replicates, go to:
+if(CHECK_FOR_OUTLIERS) {
+  source('dissimilarity.R')
+}
 #-------------------------------------------------------------------------------
 
 

### 2016-04-20 Analysis/diss_by_dist.R

@@ -15,10 +15,22 @@ if(!(identical(dimnames(comm_dist), dimnames(geo_dist)))){
 	warning("Whoa there! the row/column names of the two distance matrices do not seem to add up. This is bad.")
 }
 
+# fit michaelis menten curve
+the_data <- data.frame(comm = as.vector(comm_dist), space = as.vector(geo_dist))
+mm_fit <- nls(
+  formula = comm ~ Vm * space/(Km + space), 
+  data = the_data, 
+  start = list(
+    Vm = max(the_data$comm), # comment out and set Vm to 1 if asymptote should be 1
+    Km = max(the_data$comm)/2
+  )
+)
+mm_prediction <- predict(mm_fit)
+
 geo_dist_scaled <- log(geo_dist + 100)
 plot_x <- geo_dist # geo_dist_scaled
 
-# pdf(file = file.path(fig_dir, "diss_by_dist.pdf"), width = 8, height = 3)
+pdf(file = file.path(fig_dir, "diss_by_dist.pdf")) #, width = 8, height = 3
 	par(mar = c(4,5,1,1))
 	plot(
 		x = plot_x, 
@@ -36,15 +48,20 @@ plot_x <- geo_dist # geo_dist_scaled
 	axis(side = 1)
 	#, at = unique(log(metadata$dist_from_shore + 100)), labels = unique(metadata$dist_from_shore)
 # abline(v = unique(log(metadata$dist_from_shore + 100)))
-smoothed <- loess.smooth(
-				x = plot_x, 
-				y = comm_dist, 
-				span = 2/3, 
-				degree = 1, 
-				family = "gaussian"
-				)
-lines(smoothed, col="red", lwd=2)
-# dev.off()
+
+# Add Michaelis Menten Fit
+lines(x = c(0, sort(geo_dist)), y = c(0, sort(mm_prediction)), col = "red", lwd = 2, lty = 2)
+
+# Add LOESS line
+# smoothed <- loess.smooth(
+				# x = plot_x, 
+				# y = comm_dist, 
+				# span = 2/3, 
+				# degree = 1, 
+				# family = "gaussian"
+				# )
+# lines(smoothed, col="red", lwd=2)
+dev.off()
 
 #-------------------------------------------------------------------------------
 # Some Possible analyses

### 2016-04-22 Analysis/cluster.R
test cluster robustness

@@ -14,22 +14,22 @@ gghue <- function(n){
 }
 
 # choose between using proportional or raw counts
-my_table <- as.binary(otu_mean[,1:100]) # otu_mean, otu_spvar, otu_named, as.binary(otu_mean), 
+my_table <- otu_mean # otu_mean, otu_spvar, otu_named, as.binary(otu_mean), [,1:100]
 my_metadata <- metadata_mean #metadata[!duplicated(metadata[,"env_sample_name"]),]
 
 mydist <- vegdist(my_table, method = "bray") # , binary = TRUE
 
-mypam <- pamk(data = mydist) # to restrict range of Ks considered: , krange = 2:4
+mypam <- pamk(data = mydist, krange = 1:(attributes(mydist)$Size-1)) # to restrict range of Ks considered: , krange = 2:4
 pam_out <- mypam$pamobject
-mypam$nc # number of clusters
+(K_optim <- mypam$nc) # number of clusters
 
 # plot(pam(mydist, k = mypam$nc), which.plots = 1)
 
 mycolors <- gghue(mypam$nc)
 
 # see ?clusplot.default
-pdf(file = file.path(fig_dir, "pam_plot.pdf"))
-	par(mar = c(5,5,1,1))
+# pdf(file = file.path(fig_dir, "pam_plot.pdf"))
+	par(mar = c(5,5,2,1))
 	clusplot(
 		x = as.matrix(mydist),
 		clus = pam_out$clustering,
@@ -42,11 +42,11 @@ pdf(file = file.path(fig_dir, "pam_plot.pdf"))
 		col.clus = mycolors,
 		main = NA
 	)
-dev.off()
+# dev.off()
 
 
 # USING MEAN DATA
-pdf(file = file.path(fig_dir, "pam_in_space.pdf"), width = 5, height = 7)
+# pdf(file = file.path(fig_dir, "pam_in_space.pdf"), width = 5, height = 7)
 plot(
 		x = my_metadata[,colname_xcoord],
 		xaxt = "n",
@@ -59,31 +59,32 @@ plot(
 		pch = 19, #as.character(mypam$pamobject$clustering[my_metadata$env_sample_name]-1)
 		cex = 4,
 		# cex = 1.5,
-		main = "membership to PAM classifications",
+		main = paste("membership to PAM clusters, K =", K_optim),
+		bty = "n", 
 		las = 1,
 		xlab = "Position along shore (meters)",
[…]

### 2016-04-22 Manuscript/v0.0/carkeek_grid.tex
refocus introduction

@@
 
-Biodiversity surveillance is being revolutionized by DNA-based detection of organisms from environmental samples. ?(specifically speed and scope of ecological studies).
+Biodiversity surveillance is being revolutionized by DNA-based detection of organisms from environmental samples.
+?(specifically speed and scope of ecological studies).
+% "The surveillance of biodiversity using DNA-based detection from environmental samples"
+While this approach has been used for decades to survey micro-organisms (CITE VENTER), it has more recently become used as a tool for surveying macro-organisms (CITE EARLY EDNA).
+The technique is founded on the premises that:
+1) organisms shed DNA-containing material into the environment, 
+2) the DNA contained in this material breaks down over time, 
+which leads to the conclusion that:
+(1) the concentration of DNA originating from an organism should decrease with distance from its source.
+%there should be higher concentrations of DNA close to the organism that shed it
+
+%If DNA never broke down, there would be equal concentrations of DNA from two individuals that passed through the same environment many years apart.
+Similarly, there would be no relationship between DNA concentration and distance from a source.
+
+For an effectively immobile organism (e.g. fungal hyphae) 
+in an effectively static environmental medium (e.g. soil), 
+the relationship between DNA concentration and distance appears straightforward (CITE ???).
+One might expect the relationship to break down in a dynamic environment: 
+An immobile tree may shed cells into the air, which may be carried by the wind for some distance.
+This has been a cause for concern for ecologists interested in surveying aquatic organisms, 
+because the robustness of eDNA surveys in aquatic environments depends on the distance that DNA travels from a source. 
+This distance will be affected by the nature of the environment: 
+Material is less likely to be transported (quickly over) great distances in an ice-covered lake 
+than along a turbulent shoreline or a fast-moving stream.
+One expects more DNA downstream of a fish in a river than upstream, and
+
+In marine environments, this relationship is further complicated by the fact 
+that many organisms have dispersive early life history stages (gametes and larvae) 
+that can travel great distances but are difficult to track.
+These individuals will obscure the expected relationship between DNA concentration and adult abundance because they 
+shed genetic material into the environment and 
+they are small enough to be captured directly by environmental sampling protocols.
+For example, while a traditional survey of a sessile, intertidal bivalve would 
+identify the most biomass along the shore, there may be a greater concentration 
+of their genetic material at some distance away during a spawning event, or in 
+an ocean current carrying many larvae.
+
 Many researchers are justifiably cautious about the ?(adoption) of this new form of data.

### 2016-04-27 Manuscript/v0.0/carkeek_grid.tex
add small note
finish life history data and document

@@
 
-%------------------------------------TITLE--------------------------------------
+%-------------------------------------------------------------------------------
 
@@
 
+%-------------------------------------------------------------------------------
 %----------------------------------ABSTRACT-------------------------------------
+%-------------------------------------------------------------------------------
 \section*{Abstract}
@@
 
+%-------------------------------------------------------------------------------
 %---------------------------------INTRODUCTION----------------------------------
+%-------------------------------------------------------------------------------
 \section*{Introduction}
@@
 an ocean current carrying many larvae.
+
+% Over two marine habitats, we may KNOW their benthic communities are different, their pelagic/planktonic communities MAY be different, and the DNA community MAY be different.
 
@@
 
+%-------------------------------------------------------------------------------
 %-----------------------------------METHODS-------------------------------------
+%-------------------------------------------------------------------------------
 \section*{Methods}
@@
 
+\subsection*{Laboratory Methods}
+We collated coarse-scale data on life history characteristic for each of the major taxonomic groups recovered, 
+including dispersal range of the gametes, larvae, and adults, adult habitat type and selectivity, and adult body size. 
+Dispersal range was given as an order-of-magnitude approximation of the scale of dispersal: for example, internally fertilized species were assigned a gamete range of 0 km, while broadcast spawners were assigned a gamete range of 10 km.
+Similarly, adult range size was approximated as 0 km (sessile), 1 km (motile but not pelagic), or 10 km (highly mobile, pelagic). 
+Variables were specified as 'multiple' for groups known to span more than 1 magnitude of range size. 
+For groups to which sequences were annotated with high confidence, but for which life history strategy is diverse or poorly known (e.g.
+families in the phylum Nemertea), we used conservative, coarse approximations at a higher taxonomic rank.
+We assessed whether or not marine invertebrate taxa are thought to be present in Puget Sound by checking for their presence in a comprehensive checklist of invertebrates of Puget Sound CITE KOZLOFF. 
+These data are available as part of the REFERENCE SUPPLEMENTAL DATA.
+
 \subsection*{Spatial Model Formulation}
@@
 
+%-------------------------------------------------------------------------------
 %-----------------------------------RESULTS-------------------------------------
+%-------------------------------------------------------------------------------
 \section*{Results}
@@
 
+%-------------------------------------------------------------------------------
 %----------------------------------DISCUSSION-----------------------------------
+%-------------------------------------------------------------------------------
 \section*{Discussion}
@@
   \centering
-    \includegraphics[width=1\textwidth]{../../Figures/pam_plot.pdf}
+    \includegraphics[height=0.5\textheight]{../../Figures/pam_plot.pdf}
     \caption{TODO figure out color of ellipses; I can't even plot them gray without Plot of partitioning around medoids (PAM) analysis of OTU sequence abundance from 4 replicate PCRs at each of 24 sampling points.
@@
   \centering
-    \includegraphics[width=1\textwidth]{../../Figures/pam_in_space.pdf}
+    \includegraphics[height=0.5\textheight]{../../Figures/pam_in_space.pdf}
     \caption{Geographic position of collected samples, colored by membership to clusters identified by partitioning around medoids algorithm. 

### 2016-04-29 Manuscript/v0.0/carkeek_grid.tex
document final data prep procedure

@@
 
-We assessed PCR contamination by evaluating the dissimilarity of replicate PCRs of the same DNA sample, and removed one sample for which the Bray-Curtis dissimilarities between itself and the other replicates exceeded 0.1 (lib\_B\_tag\_GCGCTC).
+We checked for experimental error by evaluating the Bray-Curtis dissimilarity of proportional read abundance among replicate PCRs of the same DNA sample (0.033$\pm$0.063), and excluded one PCR replicate for which the dissimilarities between itself and the other replicates exceeded 1 SD (lib\_B\_tag\_GCGCTC).
 
@@
 
-To scale the OTU counts, we calculated the minimum number of OTU-assigned reads (as opposed to raw number of reads) found in these samples (130402), multiplied this by within-sample proportional abundance of each OTU, and finally rounded these numbers.
+To account for variation in the number of sequencing reads (sequencing depth) recovered per sample, we multiplied the within-sample proportional abundance of each OTU by the minimum sequencing depth (130402), and rounded to the nearest integer.
+
+Because each step in the massively parallel sequencing workflow is sensitive to contamination, it is possible that some sequences are the result of contamination during field sampling, filtration, DNA extraction, PCR, fragment size selection, quantitation, sequencing adapter ligation, or the sequencing process itself.
+Some authors have argued that these risks could bias sequence abundance, making those data meaningless and prohibiting quantitative estimates, yet convert count data to binary presence absence data on the basis of the sequence abundance greater than some arbitrary threshold).
+%massively parallel metabarcoding,
+Recent work has shown that this binary treatment of data can (?falsely?) overestimate taxon richness and falsely elevate the estimate of taxon turnover among samples (CITE LERAY FORTHCOMING).
+We take the view that it is unlikely that contaminants would manifest as sequences in the final dataset in consistent abundance across replicates; indeed, our data show that the process from PCR onward is remarkably consistent.
+Thus, we calculated from our data the maximum number of sequence counts (after scaling to correct for sequencing depth variation) for which there is turnover in presence-absence among PCR replicates within an environmental sample.
+We use this number to determine a conservative minimum threshold above which we can be confident that counts are consistent among replicates and not of (?spurious | dubious?) origin, and exclude from further analysis observations where the mean abundance across PCR replicates within samples does not reach this threshold.
+% the number is turnover_threshold=178
 

### 2016-04-29 Analysis/1_data_prep.R
finalize data prep

@@ -68,7 +68,7 @@ if(CHECK_FOR_OUTLIERS) {
   	my_metadata = metadata,
   	sample_id_column = colname_sampleid,
   	grouping_column = colname_env_sample, 
-  	threshold_sd = 1.5)
+  	threshold_sd = 1)
   otu_clean <- strip_absent(cleaned[[1]])
   metadata_clean <- cleaned[[2]]
   rm(cleaned)
@@ -95,31 +95,8 @@ if(RESCALE_SEQUENCING_DEPTH) {
 # where it is completely absent from another PCR?
 source("no_turnover.R")
 turnover_thresholds <- no_turnover(otu_scaled, metadata_clean[,colname_env_sample])
-turnover_threshold <- max(turnover_thresholds)
+(turnover_threshold <- max(turnover_thresholds))
 #-------------------------------------------------------------------------------
-
-
-#-------------------------------------------------------------------------------
-# Should OTUs be excluded that occur fewer than a threshold number of times?
-EXCLUDE_RARE_OTUs <- TRUE
-abundance_threshold <- turnover_threshold
-
-if(EXCLUDE_RARE_OTUs) {
-  # how many OTUs will be retained?
-  sum(colSums(otu_scaled) > abundance_threshold)
-  # round(colSums(otu_scaled)/sum(otu_scaled)*100, digits = 2)
-  # plot(colSums(otu_scaled)/sum(otu_scaled), type = "l")
-  # exclude OTUs that were counted a total of fewer than 100 times across samples
-  otu_filt <- otu_scaled[,which(colSums(otu_scaled) > abundance_threshold)]
-  dim(otu_filt)
-}
-
-
-# boxplot(otu_filt[,1:20])
-# boxplot(log(otu_table_in[,1:20]))
-# boxplot(otu_table_prop[,1:20])
-# boxplot(scale(otu_table_in[,1:20]))
-
 
 #-------------------------------------------------------------------------------
 # take mean of OTU abundance across replicate PCRs
@@ -134,8 +111,39 @@ metadata_mean <- metadata_clean[match(rownames(otu_mean), metadata_clean[,colnam
 #-------------------------------------------------------------------------------
 
 #-------------------------------------------------------------------------------
+# Should OTUs be excluded that occur fewer than a threshold number of times?
+EXCLUDE_RARE_OTUs <- TRUE
+abundance_threshold <- turnover_threshold
+
[…]

### 2016-04-29 Analysis/dissimilarity.R
modify reporting

@@ -46,6 +46,13 @@ find_bad_PCR <- function(my_table, my_metadata,
 	# calculate mean dissimilarity
 	dis_mean <- mean(my_dis_v)
 	dis_sd <- sd(my_dis_v)
+	print(
+		paste(
+			"dissimilarity among replicates within sample (mean, sd):", 
+			paste(
+				round(dis_mean, digits = 3), 
+				round(dis_sd, digits = 3), 
+				collapse = " ")))
 	dissimilarity_cutoff <- dis_mean + (dis_sd * THRESHOLD_SD)
 
 	# which sample group contains dissimilarities over the threshold?

### 2016-04-29 Analysis/cluster.R
use filtered mean otu table (otu_filt)

@@ -2,6 +2,7 @@
 # INPUT COMES FROM SCRIPT "load_data.r"
 # Watch out: if number of clusters is >9, the plotting will look funky
 # Only single digits can be used as symbols in base plotting
+#-------------------------------------------------------------------------------
 
 library(fpc) #pamk()
 library(cluster) #pam
@@ -14,10 +15,10 @@ gghue <- function(n){
 }
 
 # choose between using proportional or raw counts
-my_table <- otu_mean # otu_mean, otu_spvar, otu_named, as.binary(otu_mean), [,1:100]
+my_table <- otu_filt # otu_mean, otu_spvar, otu_named, as.binary(otu_mean), [,1:100] , otu_filt
 my_metadata <- metadata_mean #metadata[!duplicated(metadata[,"env_sample_name"]),]
 
-mydist <- vegdist(my_table, method = "bray") # , binary = TRUE
+mydist <- vegdist(my_table, method = "bray", binary = FALSE) # , binary = TRUE
 
 mypam <- pamk(data = mydist, krange = 1:(attributes(mydist)$Size-1)) # to restrict range of Ks considered: , krange = 2:4
 pam_out <- mypam$pamobject

### 2016-05-03 Manuscript/v0.0/carkeek_grid.tex
polish up model fit and plot of distance/dissimilarity relationship

@@
 
-\subsection*{Laboratory Methods}
+\subsection*{Laboratory Procedures}
 
@@
 \subsection*{Data Preparation (Bioinformatics)}
+
 Detailed bioinformatic methods are provided in the supplemental material, and scripts used from raw sequencer output onward can be found in the project directory on GitHub (see Data Availability).
@@
 Recent work has shown that this binary treatment of data can (?falsely?) overestimate taxon richness and falsely elevate the estimate of taxon turnover among samples (CITE LERAY FORTHCOMING).
-We take the view that it is unlikely that contaminants would manifest as sequences in the final dataset in consistent abundance across replicates; indeed, our data show that the process from PCR onward is remarkably consistent.
+We take the view that contaminants are unlikely to manifest as sequences in the final dataset in consistent abundance across replicates; indeed, our data show that the process from PCR onward is remarkably consistent.
 Thus, we calculated from our data the maximum number of sequence counts (after scaling to correct for sequencing depth variation) for which there is turnover in presence-absence among PCR replicates within an environmental sample.
@@
 
-We compiled life history data for taxonomic groups at the family level and higher.
-For groups with sparse data at the family level (e.g. Nemertea), we used the data from the higher level group.
-
-\subsection*{Community Analysis}
+\subsection*{Ecological Analyses}
 

### 2016-05-03 Analysis/diss_by_dist.R

@@ -1,12 +1,13 @@
 # FIRST RUN 'load_data.r'
 
 # dissimilarity by distance for reduced data set (mean abundance within sample for each OTU)
+#-------------------------------------------------------------------------------
 
 # REQUIRES:
 # 1. dataframe called "metadata" with unique sequenced samples given in column "sample_id"
 my_metadata <- metadata_mean # metadata[!duplicated(metadata[,colname_env_sample]),]
 # 2. OTU table with rownames that correspond to aforementioned column "sample_id"
-my_table <- otu_mean # otu_mean, otu_spvar, otu_named, as.binary(otu_mean), log(otu_mean + 1)
+my_table <- otu_filt # otu_mean, otu_spvar, otu_named, as.binary(otu_mean), otu_log, otu_filt
 rownames(my_table) # should be e.g. PCT-C-0000 etc, aka "env_sample_name"
 
 
@@ -15,30 +16,99 @@ library(geosphere) # distm()
 
 # calculate pairwise great circle distance between sampling locations using Haversine method
 geo_dist <- as.dist(distm(x = my_metadata[,c(colname_lon, colname_lat)], fun = distHaversine))
+geo_dist_v <- as.vector(geo_dist)
 # dimnames(geo_dist) <- list(my_metadata$env_sample_name, my_metadata$env_sample_name)
 
 comm_dist <- vegdist(my_table, method = "bray") #, diag = TRUE, upper = TRUE
+comm_dist_v <- as.vector(comm_dist)
 
 if(!(identical(dimnames(comm_dist), dimnames(geo_dist)))){
 	warning("Whoa there! the row/column names of the two distance matrices do not seem to add up. This is bad.")
 }
 
-# fit michaelis menten curve
+# arrange the data for model fitting
 the_data <- data.frame(comm = as.vector(comm_dist), space = as.vector(geo_dist))
+
+#-------------------------------------------------------------------------------
+# Fit some models, estimate some parameters
+model_out <- list()
+model_pred <- list()
+#-------------------------------------------------------------------------------
+
+#-------------------------------------------------------------------------------
+# Michaelis-Menten
+fix_asymptote_1 <- TRUE
+if(fix_asymptote_1){
+	Vm <- 1
+	start_list <- list(Km = max(the_data$comm)/2)
+} else {
[…]

### 2016-05-03 Analysis/1_data_prep.R
add log transform

@@ -139,6 +139,10 @@ if(EXCLUDE_RARE_OTUs) {
 }
 #-------------------------------------------------------------------------------
 
+#-------------------------------------------------------------------------------
+# Log transform
+otu_log <- log(otu_filt + 1)
+#-------------------------------------------------------------------------------
 
 #-------------------------------------------------------------------------------
 # rescale each column to its maximum value, ranging from 0 to 1
@@ -154,4 +158,3 @@ source("rm_single_rows.R")
 otu_spvar <- rm_single_rows(otu_filt)
 dim(otu_spvar)
 #-------------------------------------------------------------------------------
-

### 2016-05-04 Analysis/add_names.R
small change

@@ -1,5 +1,5 @@
 # add organism names to an otu table
-the_otu_table <- otu_mean
+the_otu_table <- otu_filt
 
 blast_file <- "query_hit_LCA.csv"
 
@@ -16,10 +16,17 @@ taxa_vector <- match(colnames(the_otu_table), taxa_df$query_seq)
 nrow(taxa_df)
 
 # which taxa are not in the blast file?
-not_in_blastfile <- colnames(the_otu_table)[!colnames(the_otu_table) %in% taxa_df$query_seq]
+(not_in_blastfile <- colnames(the_otu_table)[!colnames(the_otu_table) %in% taxa_df$query_seq])
 
 # how abundant are they?
-colSums(the_otu_table)[!colnames(the_otu_table) %in% taxa_df$query_seq]
+round(colSums(the_otu_table)[!colnames(the_otu_table) %in% taxa_df$query_seq] / sum(the_otu_table), digits = 3)
+
+# plot
+# plot(
+  # colSums(the_otu_table), 
+  # col = as.numeric(!colnames(the_otu_table) %in% taxa_df$query_seq) + 1, 
+  # log = "y", ylab = "", las = 1
+# )
 
 # taxon ID of each otu
 taxa_df$query_seq == 

### 2016-05-04 Analysis/variogram.R
tried multivariate variogram using geodata object; failed.

@@ -1,7 +1,19 @@
-my_table    <- otu_mean
+my_table    <- otu_filt
 my_metadata <- metadata_mean
 
 library(geoR) # variog
+
+# Attempting multivariate variogram
+# TODO as of now, I think it just calculates individual variograms
+my_geodata <- as.geodata(
+	obj = cbind.data.frame(
+		my_metadata[,c(colname_lon, colname_lat)], #c(colname_xcoord, colname_ycoord)
+		my_table), 
+	coords.col = 1:2, 
+	data.col = 2 + 1:ncol(my_table)
+)
+vg_all <- variog(my_geodata)
+plot(vg_all, type = "b")
 
 # variogram for individual taxon abundance
 pdf(file = file.path(fig_dir, "variogram_taxa.pdf"))

### 2016-05-04 Manuscript/v0.0/carkeek_grid.tex
incorporate Ole figures

@@
 Samples collected of ecological communities may vary in dissimilarity from 0 (completely identical) to 1 (completely different). 
-For samples collected from multiple locations, the relationship between their spatial distance and community dissimilarity is of interest because it reflects the amount of community heterogeneity over the spatial scale sampled.
+For samples collected from multiple locations, the relationship between their spatial distance and community dissimilarity is of interest because it reflects the amount of community heterogeneity---changes in abundance and composition---over the spatial scale sampled.
 %In general, the relationship between community dissimilarity (0 = identical; 1 = completely different) and spatial distance is expected to be asymptotic, because communities nearer to each other tend to be more similar than those farther apart.
-The intercept is expected to be 0, because only within-sample comparisons can have 0 spatial separation, and communities have no dissimilarity within a sample.
+The intercept is expected to be 0, because only within-sample comparisons can have 0 spatial separation, and communities have no dissimilarity within a sample if sampling method is repeatable.
 % technically impossible for samples to be taken at exactly 0; they are actually immediately adjacent.
-Likewise, dissimilarity cannot exceed 1. 
-Deviation from 0 indicates heterogeneous community composition/structure over fine scales.
-A flat relationship between dissimilarity and distance indicates that heterogeneity is not assorted spatially, and can be interpreted in different ways, depending on the mean.
-If the mean is close to 1, there is high spatial heterogeneity over the spatial scale of sampling.
-If the mean is 0, all samples are identical, and we infer there is complete community homogeneity over the scale sampled.
+%Deviation from 0 indicates heterogeneous community composition/structure over fine scales. 
+Likewise, dissimilarity cannot exceed 1, and reaches 1 only when multiple discrete community types are sampled.
+The trend is expected to be asymptotic if communities within a habitat are spatially structured, where the value of the asymptote and the rate of increase provide insight about community turnover. 
+A flat relationship indicates that heterogeneity is not assorted spatially, and can be interpreted in different ways, depending on the asymptote. % ?or no relationship both 
+If the asymptote is close to 1, there is high spatial heterogeneity over the spatial scale of sampling. 
+If the asymptote is close 0, all samples are similar, and we infer there is complete community homogeneity over the scale sampled. 
 The rate at which community dissimilarity approaches the mean gives an indication of the rate of community turnover.
@@
 
-\subsection*{Laboratory Methods}
+\subsection*{Organismal Life History Data}
 We collated coarse-scale data on life history characteristic for each of the major taxonomic groups recovered, 
@@
   \centering
+    \includegraphics[width=1\textwidth]{../../Figures/dissimilarity_from_shore.pdf}
+    \caption{Pairwise dissimilarity (Bray-Curtis) across transects plotted against distance from shore. A linear model determined no effect of distance from shore on dissimilarity.}
+  \label{dissimilarity_from_shore} % use this to refer to your figure in the text, so that numbering updates automatically
+\end{figure}
+
+\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
+  \centering
+    \includegraphics[width=1\textwidth]{../../Figures/variance_from_shore.pdf}
+    \caption{Within-OTU variance across transects plotted against distance from shore.}
+  \label{variance_from_shore} % use this to refer to your figure in the text, so that numbering updates automatically
+\end{figure}
+
+\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
+  \centering
+    \includegraphics[width=1\textwidth]{../../Figures/heatmap_sites.pdf}
+    \caption{Heatmap of pairwise site dissimilarity (Bray-Curtis).}
+  \label{heatmap} % use this to refer to your figure in the text, so that numbering updates automatically
+\end{figure}
+
+\begin{figure}[h!] % [h!] forces the figure to be placed roughly here
+  \centering
     \includegraphics[height=0.5\textheight]{../../Figures/pam_plot.pdf}

### 2016-05-04 Analysis/plot_variance_shore.R

@@ -1,0 +1,125 @@
+#!/usr/bin/env Rscript
+#-------------------------------------------------------------------------------
+
+the_otu_table <- otu_filt
+the_metadata  <- metadata_mean
+
+
+#-------------------------------------------------------------------------------
+data_by_dist <- split(
+	x = as.data.frame(the_otu_table), 
+	f = metadata_mean[,colname_ycoord]
+)
+
+distances <- rep(names(data_by_dist), times = sapply(data_by_dist, nrow))
+
+dis_by_dist <- lapply(data_by_dist, function(x) 
+	as.vector(vegdist(x, method = "bray"))
+)
+
+to_plot <- data.frame(dist = as.numeric(distances), vegdist = unlist(dis_by_dist))
+
+# transform distance to shore 
+lm_out <- lm(to_plot$vegdist ~ log(to_plot$dist + 1))
+summary(lm_out)
+# no significant slope
+
+
+pdf(file = file.path(fig_dir, "dissimilarity_from_shore.pdf"),
+	width = 4, 
+	height = 4
+)
+par(mar = c(4,4,1,1))
+plot(
+	x = to_plot$dist,
+	xlab = "Distance from shore", 
+	y = to_plot$vegdist, 
+	ylab = "Bray-Curtis Dissimilarity", 
+	ylim = c(0,1), 
+	pch = 21, 
+	col = hsv(1,1,0), 
+	bg  = hsv(1,1,0,0.2), 
+	# log = "x"
+	las = 1
+)
+points(
+	y = sapply(dis_by_dist, mean), 
+	x = as.numeric(names(dis_by_dist)), 
+	pch = 23, 
+	col = hsv(1,1,1), 
+	bg  = hsv(1,1,1,0.2)
+)
+legend(
+	"bottomright", 
+	bty = "n", 
+	legend = "mean", 
+	pch = 23, 
+	col = hsv(1,1,1), 
+	pt.bg  = hsv(1,1,1,0.2)
+)
+# abline(lm_out, col = hsv(0.6, 1, 1), lwd = 2, lty = 2)
+dev.off()
+
+#-------------------------------------------------------------------------------
+# individual otu abundance variance
+var_by_dist <- lapply(data_by_dist, function(x)
[…]

### 2016-05-13 Analysis/classification_df.R
correct formatting of date in filename

@@ -4,7 +4,7 @@ classifications_file <- "classifications20160329.RData"
 # I don't know how this works, I am just clicking to load the data
 # classifications <- load(file.path(data_dir, classifications_file))
 
-classifications
+classifications[[1]]
 
 
 the_ranks <- lapply(lapply(classifications, function(x) x[2]), "[[", 1)
@@ -19,7 +19,8 @@ classification_df <- data.frame(
          row.names = NULL
          )
 
-current_time <- gsub(pattern = " ", replacement = "_", x = Sys.time())
+current_time <- gsub(pattern = " ", replacement = "T", x = Sys.time())
+current_time <- gsub(pattern = ":", replacement = "", x = current_time)
 write.csv(classification_df,file = file.path(data_dir, paste("classification_", current_time, ".csv", sep = "")))
 
 #----------------------------------------------------------------------------------------

### 2016-05-13 Analysis/plot_stacked_bar.R
use taxon table for plotting stacked bar

@@ -4,7 +4,7 @@ library(ggplot2)
 library(reshape2)
 
 
-mytable <- otu_filt[,1:20]
+mytable <- taxon_table # taxon_table, otu_filt[,1:20]
 mytable_long <- melt(mytable)
 colnames(mytable_long) <- c("sample_id", "OTU", "count")
 
@@ -25,8 +25,6 @@ p + geom_bar() +
       panel.background = element_rect(fill = 'white', colour = 'white')
       )
 # dev.off()
-
-# note high value for sample "lib_B_tag_GCGCTC" in PCT-C-0500
 
 
 # head(data_full_long)

### 2016-05-13 Manuscript/v0.0/carkeek_grid.tex
add potential title
misc changes

@@
 \title{There once was a grid at ol' Carkeek}
+% Spatial dispersion of environmental DNA in a marine habitat
 
@@
 The technique is founded on the premises that:
-1) organisms shed DNA-containing material into the environment, 
-2) the DNA contained in this material breaks down over time, 
-which leads to the conclusion that:
-(1) the concentration of DNA originating from an organism should decrease with distance from its source.
+(1) organisms shed DNA-containing material into the environment, and 
+(2) the DNA contained in this material breaks down over time, 
+which leads to the conclusion that the concentration of DNA should decrease with distance from and time since the presence of its source.
 %there should be higher concentrations of DNA close to the organism that shed it
@@
 
-For an effectively immobile organism (e.g. fungal hyphae) 
-in an effectively static environmental medium (e.g. soil), 
+For an immobile organism (e.g. fungal hyphae) % effectively 
+in a static environmental medium (e.g. soil), %n effectively
 the relationship between DNA concentration and distance appears straightforward (CITE ???).
-One might expect the relationship to break down in a dynamic environment: 
+But the relationship may be complicated by a dynamic environment: %One might expect the relationship to break down in 
 An immobile tree may shed cells into the air, which may be carried by the wind for some distance.
 This has been a cause for concern for ecologists interested in surveying aquatic organisms, 
-because the robustness of eDNA surveys in aquatic environments depends on the distance that DNA travels from a source. 
+because the robustness of eDNA surveys for detecting and quantifying organisms depends on the distance that DNA travels from a source. %in aquatic environments 
 This distance will be affected by the nature of the environment: 
@@
 
+% Over two marine habitats, we may KNOW their benthic communities are different, their pelagic/planktonic communities MAY be different, and the DNA community MAY be different.
 In marine environments, this relationship is further complicated by the fact 
-that many organisms have dispersive early life history stages (gametes and larvae) 
-that can travel great distances but are difficult to track.
+that many organisms have tiny, highly dispersive early life history stages (gametes and larvae).
+%that can travel great distances but are difficult to track.
 These individuals will obscure the expected relationship between DNA concentration and adult abundance because they 
 shed genetic material into the environment and 
-they are small enough to be captured directly by environmental sampling protocols.
+are small enough to be captured directly by environmental sampling protocols.
 For example, while a traditional survey of a sessile, intertidal bivalve would 
-identify the most biomass along the shore, there may be a greater concentration 
+determine that the most biomass lies along the shore, there may be a greater concentration 
 of their genetic material at some distance away during a spawning event, or in 
@@
 
-% Over two marine habitats, we may KNOW their benthic communities are different, their pelagic/planktonic communities MAY be different, and the DNA community MAY be different.
+If DNA communities are composed mostly of species known
+
+Despite the potential for discrepancy between communities of benthic organisms and their DNA in the water column above, several studies have reported consistent patterns between the two.
 
@@
 
+% 1. Jane SF, Wilcox TM, Mckelvey KS, Young MK, Schwartz MK, Lowe WH, et al. Distance, flow and PCR inhibition: EDNA dynamics in two headwater streams. Mol Ecol Resour. 2014; 
 %-----------------------------------FIGURES-------------------------------------

### 2016-05-13 Analysis/add_names.R

@@ -3,54 +3,80 @@ the_otu_table <- otu_filt
 
 blast_file <- "query_hit_LCA.csv"
 
-taxa_df <- read.csv(file.path(data_dir, blast_file))
+blast_df <- read.csv(
+  file = file.path(data_dir, blast_file),
+  stringsAsFactors = FALSE
+)
 
-head(taxa_df)
+head(blast_df)
 
 # remove abundance annotation from query sequence names
-taxa_df$query_seq <- gsub(pattern = ";size=.*", replacement = "", taxa_df$query_seq)
-head(taxa_df)
+blast_df$query_seq <- gsub(pattern = ";size=.*", replacement = "", blast_df$query_seq)
+head(blast_df)
 
 
-taxa_vector <- match(colnames(the_otu_table), taxa_df$query_seq)
-nrow(taxa_df)
+taxa_vector <- match(colnames(the_otu_table), blast_df$query_seq)
+nrow(blast_df)
 
 # which taxa are not in the blast file?
-(not_in_blastfile <- colnames(the_otu_table)[!colnames(the_otu_table) %in% taxa_df$query_seq])
+(not_in_blastfile <- colnames(the_otu_table)[
+	!colnames(the_otu_table) %in% blast_df$query_seq
+	])
 
 # how abundant are they?
-round(colSums(the_otu_table)[!colnames(the_otu_table) %in% taxa_df$query_seq] / sum(the_otu_table), digits = 3)
+round(colSums(the_otu_table)[!colnames(the_otu_table) %in% blast_df$query_seq] / sum(the_otu_table), digits = 3)
 
 # plot
 # plot(
-  # colSums(the_otu_table), 
-  # col = as.numeric(!colnames(the_otu_table) %in% taxa_df$query_seq) + 1, 
+  # colSums(the_otu_table),
+  # col = as.numeric(!colnames(the_otu_table) %in% blast_df$query_seq) + 1,
   # log = "y", ylab = "", las = 1
 # )
 
-# taxon ID of each otu
-taxa_df$query_seq == 
-colnames(the_otu_table)
 
-taxa_by_otu <- as.character(taxa_df[taxa_vector,"class_all"])
+taxon_column <- "family_all"
+# LCA_name_all   | class_all   | order_all   | family_all
+# LCA_name_beste | class_beste | order_beste | family_beste
+table(blast_df$LCA_id_all   %in% classification_df$query)
[…]

### 2016-05-13 Analysis/life_history.R

@@ -1,0 +1,79 @@
+# add life history data
+
+life_history_file <- "life_history.csv"
+life_history <- read.csv(
+  file = file.path(data_dir, life_history_file),
+  stringsAsFactors = FALSE
+)
+
+head(life_history)
+lh_taxon <- "taxon_name"
+
+# join sequence count table and life history table, using blast results and classifications as intermediate
+
+classification_file <- "classification_2016-04-19T125449.csv"
+
+classification_df <- read.csv(
+	file = file.path(data_dir, classification_file),
+	row.names = 1,
+	stringsAsFactors = FALSE
+)
+head(classification_df)
+classification_col <- "taxname"
+
+
+# which of the taxa in the classification df have data?
+# classification_df$taxname %in% life_history$taxon_name
+# their index in the life history data
+classification_df$lifehist_row <- match(
+	classification_df$taxname, life_history$taxon_name
+)
+head(classification_df)
+
+
+for OTU (column)
+
+from OTU TABLE
+
+match on SEQ ID
+
+from TAXON DF
+
+print TAXON COLUMN
+
+from TAXON DF
+
+if TAXON
+
+in TAXON COLUMN
+
+from LIFE HISTORY TABLE
+
+else
+
+print LIFE HISTORY COLUMN
+
+from LIFE HISTORY TABLE
+
+
+
+match(
+CLASSIFICATIONS_DF$TAXON
+LIFE_HISTORY$TAXON
+)
+
+length(
+match(
+life_history$taxon_name,
+classification_df$taxname
+)
+)
+
+blast_df[taxa_vector, classification_col]
+blast_df[, classification_col]
+
+classification_df[, classification_col] %in%
+life_history[, lh_taxon]
+
+
+head(classification_df)

### 2016-05-13 Manuscript/v0.0/carkeek_grid.tex
formal incorporation of life history data

@@
 
+The vast majority (97.6\%) of sequences and OTUs (96.9\%) were matched to organisms that have high potential for dispersal at either the gamete, larval, or adult stage. 
+Of the 6 OTUs with limited dispersal, only 2 occurred in more than two samples. 
+These were assigned to \textit{Cymatogaster aggregata}, a viviparous nearshore fish with internal fertilization, and \textit{Cupolaconcha meroclista}, a sessile marine gastropod with internal fertilization and short larval dispersal (CITE???).
+Thus, we are unable to compare the patterns of dispersion for varying life history categories.
+
 \subsection*{Spatial Model Output}

### 2016-05-13 Analysis/add_names.R

@@ -1,20 +1,20 @@
 # add organism names to an otu table
 the_otu_table <- otu_filt
 
+# specify filename of blast output
 blast_file <- "query_hit_LCA.csv"
 
 blast_df <- read.csv(
   file = file.path(data_dir, blast_file),
   stringsAsFactors = FALSE
 )
-
 head(blast_df)
 
 # remove abundance annotation from query sequence names
 blast_df$query_seq <- gsub(pattern = ";size=.*", replacement = "", blast_df$query_seq)
 head(blast_df)
 
-
+# make a vector to link OTUs to their corresponding row in blast_df
 taxa_vector <- match(colnames(the_otu_table), blast_df$query_seq)
 nrow(blast_df)
 
@@ -33,111 +33,49 @@ round(colSums(the_otu_table)[!colnames(the_otu_table) %in% blast_df$query_seq] /
   # log = "y", ylab = "", las = 1
 # )
 
-
-taxon_column <- "family_all"
+taxon_column <- "LCA_name_beste"
 # LCA_name_all   | class_all   | order_all   | family_all
 # LCA_name_beste | class_beste | order_beste | family_beste
-table(blast_df$LCA_id_all   %in% classification_df$query)
-table(blast_df$LCA_id_beste %in% classification_df$query)
-
-
-table(blast_df$LCA_name_all   %in% classification_df$taxname)
-table(blast_df$LCA_name_beste %in% classification_df$taxname)
-table(is.na(blast_df$LCA_name_beste))
-table(is.na(blast_df$LCA_name_beste))
-
-# get the index where each of the blast results are found in the
-# classification hierarchy data frame
-blast_to_classif <- match(blast_df$LCA_name_beste, classification_df$taxname)
-
-head(classification_df)
-split(classification_df, classification_df$query)
-classification_df$query[blast_to_classif]
-
-
-
 
 taxa_by_otu <- blast_df[taxa_vector, taxon_column]
 source("aggregate_cols.R")
 taxon_table <- aggregate_cols(the_otu_table, col_groups = taxa_by_otu)
 
+# GO TO life_history.R
 
 stop()
-# OK pick up here
 
 # columns in the blast_df with taxon_id
-as.character(blast_df$LCA_id_all)
-as.character(blast_df$LCA_id_beste)
+# as.character(blast_df$LCA_id_all)
+# as.character(blast_df$LCA_id_beste)
 
 # column of the classification_df with taxon_id
[…]

### 2016-05-13 Analysis/life_history.R

@@ -7,7 +7,11 @@ life_history <- read.csv(
 )
 
 head(life_history)
-lh_taxon <- "taxon_name"
+# Exclude data for groups where no data is given but a higher level is
+life_history <- life_history[
+!apply(life_history, 1, function(x) all(is.na(x[5:8])))
+,]
+# lh_taxon <- "taxon_name"
 
 # join sequence count table and life history table, using blast results and classifications as intermediate
 
@@ -24,12 +28,103 @@ classification_col <- "taxname"
 
 # which of the taxa in the classification df have data?
 # classification_df$taxname %in% life_history$taxon_name
-# their index in the life history data
+
+# Add a column to the classification_df, specifiying the corresponding 
+# row in the life history data
 classification_df$lifehist_row <- match(
 	classification_df$taxname, life_history$taxon_name
 )
 head(classification_df)
 
+
+##########
+table(blast_df$LCA_id_all   %in% classification_df$query)
+table(blast_df$LCA_id_beste %in% classification_df$query)
+
+table(blast_df$LCA_name_all   %in% classification_df$taxname)
+table(blast_df$LCA_name_beste %in% classification_df$taxname)
+# table(is.na(blast_df$LCA_name_beste))
+# table(is.na(blast_df$LCA_name_beste))
+
+# get the index where each of the blast results are found in the
+# classification hierarchy data frame
+blast_to_classif <- match(blast_df$LCA_name_beste, classification_df$taxname)
+
+classification_by_query <- split(classification_df, classification_df$query)
+classification_df$query[blast_to_classif]
+
+has_numbers <- function(x){
+	any(!is.na(x[,"lifehist_row"]))
+}
+table(sapply(classification_by_query, has_numbers))
+
+# get the first occurrence of a taxon in the classification_df
+get_lh_data <- function(taxon){
+	if(class(taxon) != "character"){
+		stop("input should be a character string specifying a taxon name")
+	}
+	# show all of the rows of the query containing the taxon
+	temp <- classification_df[
+		classification_df$query == classification_df$query[
+			match(taxon, classification_df$taxname)],
+	]
+	has_data <- temp$lifehist_row[1:match(taxon, temp$taxname)]
[…]

### 2016-08-12 Analysis/life_history.R
minor change

@@ -7,6 +7,8 @@ life_history <- read.csv(
 )
 
 head(life_history)
+tail(life_history)
+
 # Exclude data for groups where no data is given but a higher level is
 life_history <- life_history[
 !apply(life_history, 1, function(x) all(is.na(x[5:8])))
@@ -38,11 +40,12 @@ head(classification_df)
 
 
 ##########
+# How many of the values from a given column of the blast_df are in classification_df
 table(blast_df$LCA_id_all   %in% classification_df$query)
 table(blast_df$LCA_id_beste %in% classification_df$query)
+table(blast_df$LCA_name_all   %in% classification_df$taxname) # all 1153
+table(blast_df$LCA_name_beste %in% classification_df$taxname) # all 1153
 
-table(blast_df$LCA_name_all   %in% classification_df$taxname)
-table(blast_df$LCA_name_beste %in% classification_df$taxname)
 # table(is.na(blast_df$LCA_name_beste))
 # table(is.na(blast_df$LCA_name_beste))
 
@@ -50,13 +53,12 @@ table(blast_df$LCA_name_beste %in% classification_df$taxname)
 # classification hierarchy data frame
 blast_to_classif <- match(blast_df$LCA_name_beste, classification_df$taxname)
 
-classification_by_query <- split(classification_df, classification_df$query)
-classification_df$query[blast_to_classif]
-
-has_numbers <- function(x){
-	any(!is.na(x[,"lifehist_row"]))
-}
-table(sapply(classification_by_query, has_numbers))
+# I don't think I need to do the split anymore.
+# classification_by_query <- split(classification_df, classification_df$query)
+# has_numbers <- function(x){
+	# any(!is.na(x[,"lifehist_row"]))
+# }
+# table(sapply(classification_by_query, has_numbers))
 
 # get the first occurrence of a taxon in the classification_df
 get_lh_data <- function(taxon){
@@ -74,12 +76,11 @@ get_lh_data <- function(taxon){
 get_lh_data(1)
 get_lh_data("Hominidae")
 
-rm(blast_to_lh)
 blast_df$lifehist_row <- sapply(blast_df$LCA_name_beste, get_lh_data)
 
 life_history[blast_df$lifehist_row,]
 otu_to_lifehist <- blast_df[taxa_vector, "lifehist_row"]
-life_history[otu_to_lifehist,range_gamete_km]
[…]

### 2016-08-12 Analysis/dissimilarity.R
add lines to plot

@@ -17,6 +17,7 @@ find_bad_PCR <- function(my_table, my_metadata,
 	# At how many standard deviations above the mean dissimilarity
 	# should a sample be excluded?
 	THRESHOLD_SD <- threshold_sd
+
 	if(!(identical(my_metadata[,sample_id_column], rownames(my_table)))){
 		stop("hold on, the otu and metadata rows are not in the same order; disaster could ensue")
 	}
@@ -78,6 +79,7 @@ find_bad_PCR <- function(my_table, my_metadata,
 	           xlab = "Pairwise Bray-Curtis Dissimilarity",
 	           bg = rgb(0,0,0,alpha = 0.2), #col = color_vector,
 	           las = 1)
+    abline(h = seq(from = 0.5, to = 60.5, by = 1), col = "gray")
 
 	# attempting to color the points red over the threshold
 	# bad_x <- as.vector(my_dis[[bad_env_sample]])

### 2016-08-12 Analysis/graph_grid.R
add code to plot network connections among sites

@@ -1,0 +1,30 @@
+
+the_edges <- data.frame(t(combn(attributes(mydist)$Labels, 2)))
+
+
+the_edges$dis <- apply(the_edges, 1, FUN = function(x) as.matrix(mydist)[x[1],x[2]])
+
+coords <- cbind(my_metadata[,colname_xcoord], log(my_metadata[,colname_ycoord] + 100))
+rownames(coords) <- my_metadata[, colname_env_sample]
+grid_graph <- graph.data.frame(the_edges, directed = FALSE)
+plot.igraph(
+  x = grid_graph, 
+  layout = coords, 
+  edge.curved = seq(-0.3, 0.3, length = ecount(grid_graph)), 
+  edge.color = hsv(0,0,0, alpha = 0.2*scale01(the_edges[,3])), 
+  edge.width = 3,
+  vertex.color = NA, 
+  vertex.frame.color = NA, 
+  # add = TRUE, 
+  vertex.label = NA
+)
+newcoords <- apply(coords, 2, scale11)
+#  points(x = 0, y = -1, col = rainbow(5), cex = 2, pch = 19)
+points(
+		x = newcoords,
+		col = mycolors[mypam$pamobject$clustering[my_metadata[,colname_env_sample]]],
+		pch = 19, #as.character(mypam$pamobject$clustering[my_metadata$env_sample_name]-1)
+		cex = 4
+	)
+
+help.search("lower triangle")

### 2016-08-12 Analysis/scale11.R
R code to scale points between -1 and 1 in order to facilitate network plot

@@ -1,0 +1,4 @@
+scale11 <- function(x) {
+  zeroone <- (x-min(x)) / (max(x)-min(x))
+  return(zeroone * 2 - 1)
+}

### 2016-08-12 Analysis/graph_on_coordinates.R
more R code associated with network plots of sites

@@ -1,0 +1,37 @@
+nsamp <- 9
+
+mat <- matrix(
+  data = rnorm(nsamp*10), 
+  nrow = nsamp
+)
+
+coords <- as.matrix(expand.grid(c(-1,0,1), c(-1,0,1)))
+
+distmat <- as.matrix(dist(mat))
+
+expand.grid(1:nsamp, 1:nsamp)
+graph_data <- with(expand.grid(1:nsamp, 1:nsamp), data.frame(v1 = Var2, v2 = Var1))
+graph_data$magnitude <- as.vector(distmat)
+
+combn(1:9, 2)
+?graph
+duplicates <- which(apply(graph_data, 1, function(x)
+  any(duplicated(x))))
+graph_data <- graph_data[-duplicates,]
+
+g <- graph.data.frame(graph_data, directed = FALSE)
+
+plot.igraph(
+  x = g, 
+  layout = coords, 
+  edge.curved = seq(-0.2, 0.2, length = ecount(g)), 
+  edge.color = hsv(0,0,0,alpha = scale01(graph_data[,3])), 
+  vertex.color = hsv(0.6,1,1), 
+  vertex.frame.color = NA, 
+  add = TRUE, 
+  vertex.label = NA
+)
+points(coords, pch = 24)
+plot(coords, pch = 24)
+?igraph.plotting
+?plot.igraph

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
push misc changes since last visited (2016-05-13)
formatting
removed irrelevant text

@@
 \section*{Introduction}
-
-Biodiversity surveillance is being revolutionized by DNA-based detection of organisms from environmental samples.
-?(specifically speed and scope of ecological studies).
-% "The surveillance of biodiversity using DNA-based detection from environmental samples"
-While this approach has been used for decades to survey micro-organisms (CITE VENTER), it has more recently become used as a tool for surveying macro-organisms (CITE EARLY EDNA).
-The technique is founded on the premises that:
-(1) organisms shed DNA-containing material into the environment, and 
-(2) the DNA contained in this material breaks down over time, 
-which leads to the conclusion that the concentration of DNA should decrease with distance from and time since the presence of its source.
-%there should be higher concentrations of DNA close to the organism that shed it
-
-%If DNA never broke down, there would be equal concentrations of DNA from two individuals that passed through the same environment many years apart.
-Similarly, there would be no relationship between DNA concentration and distance from a source.
-
-For an immobile organism (e.g. fungal hyphae) % effectively 
-in a static environmental medium (e.g. soil), %n effectively
-the relationship between DNA concentration and distance appears straightforward (CITE ???).
-But the relationship may be complicated by a dynamic environment: %One might expect the relationship to break down in 
-An immobile tree may shed cells into the air, which may be carried by the wind for some distance.
-This has been a cause for concern for ecologists interested in surveying aquatic organisms, 
-because the robustness of eDNA surveys for detecting and quantifying organisms depends on the distance that DNA travels from a source. %in aquatic environments 
-This distance will be affected by the nature of the environment: 
-Material is less likely to be transported (quickly over) great distances in an ice-covered lake 
-than along a turbulent shoreline or a fast-moving stream.
-One expects more DNA downstream of a fish in a river than upstream, and
-
-% Over two marine habitats, we may KNOW their benthic communities are different, their pelagic/planktonic communities MAY be different, and the DNA community MAY be different.
-In marine environments, this relationship is further complicated by the fact 
-that many organisms have tiny, highly dispersive early life history stages (gametes and larvae).
-%that can travel great distances but are difficult to track.
-These individuals will obscure the expected relationship between DNA concentration and adult abundance because they 
-shed genetic material into the environment and 
-are small enough to be captured directly by environmental sampling protocols.
-For example, while a traditional survey of a sessile, intertidal bivalve would 
-determine that the most biomass lies along the shore, there may be a greater concentration 
-of their genetic material at some distance away during a spawning event, or in 
-an ocean current carrying many larvae.
-
-If DNA communities are composed mostly of species known
-
-Despite the potential for discrepancy between communities of benthic organisms and their DNA in the water column above, several studies have reported consistent patterns between the two.
-
-Many researchers are justifiably cautious about the ?(adoption) of this new form of data.
-Their apprehension is rooted in the premise that traditional survey approaches are more accurate because the chain of inference between observation and ecological data is usually short: A researcher sees two swans in Lake Hopatcong and infers the lake is occupied by at least 2 swans.
-DNA based surveys, on the other hand, consist of a longer chain of inference:
-DNA sequences are reported by a sequencing machine, the machine identifies the sequence of products of a polymerase chain reaction (PCR), PCR amplifies pieces of DNA from a purified genomic DNA sample, DNA is purified (extracted) from an environmental sample, environmental samples contain DNA from organisms present, the organisms present are representative of the biological community about which we wish to make inference.
-?( reverse order?
-tie to concrete example (swans of Lake Hopatcong)).
-Clearly, this process is more complex than visual surveys, as the relationship between several steps is complex or unknown.
-But consider that the processes ?(behind | underlying) other more widely-used ecological survey techniques are similarly complex, such as bird surveys based on song, or visual identification of fungal spores.
-When alternate survey approaches are impossible or inefficient, we are more willing to accept any available survey data, regardless of the complexity or uncertainty underlying it.
-(microbiologists have enthusiastically relied on DNA-based surveys for years for this reason, (though yes, they also do not have the problem of disconnect between individual and cell)).
-
-The ability of DNA surveys to make quantitative inference about communities has been touted by some (CITE new fish quantitation paper) and doubted by others (CITE european eelgrass PLOSONE).
-For example, a study linking (blah blah blah) concluded that "metabarcoding is powerful, yet blind" (CITE european eelgrass).
-Conversely, others have reported strong quantitative and intuitive links between DNA-based and traditional survey methods (CITE Port 2016 MOLECO).
-These studies usually rely on simple statistical models to link DNA quantity to some measurable ecosystem property like biomass (but see CITE).
-When confronted with data collected in ?(complex ways/studies/whatever), simple models ?(may | often) fail to detect relationships when they exist, or vice versa ?(they are prone to inflated risk of BOTH type I and type II error) (CITE, see Woltman 2012).
-For example, (CITE, look for that Gelman paper) have demonstrated that when data are structured in a hierarchical fashion (e.g.
-test scores of students in schools belonging to districts belonging to states), a low number of replicates at the first level of hierarchy (SEE THE PAPER).
-Similarly, (describe hospital/school problems).
-
-Shelton et al. (CITE Shelton 2016) outlined an approach for structuring statistical models of DNA surveys that address these issues.
-This framework improved on alternative statistical techniques by explicitly accounting for the ?(hierarchical | nested | multilevel) structure of the study design, which allows error and uncertainty at each level to be ?(explicitly accounted for| modeled | propagated throughout the model).
-That study demonstrated an improvement in the estimate of higher-level (e.g. ecological community) quantities when the processes linking them to the data are specified.
-As an example, it was shown that incorporation of data about the mismatch between primer and template DNA sequence can improve the estimate of the relative abundance of unique DNA templates input to a PCR.
-
-Here, we apply this framework to a DNA survey of ?(nearshore | coastal) marine habitat.
-(TODO add commentary on current dogma surrounding distribution of DNA in well-mixed (marine) habitats).
-We document the variability associated with lab based ?(procedures | replication | treatment; i.e. filter+DNA+PCR+seq), and the spatial scale over which DNA communities vary in this habitat.
-We ?(show that | tested whether) a taxon's spatial distribution predicts ( the slope of the relationship between distance from shore and DNA abundance ~or~ to what degree DNA abundance is explained by distance from shore for each taxon).
-We focus partly on species with known life histories that define their spatial distribution (e.g.
-shallow water livebearing fishes or sessile intertidal organisms with ?(motile/planktonic/pelagic) larvae or gametes).
-For these taxa whose spatial distribution is well-documented and restricted, we calculate the rate of change in space and compare this rate among taxa with similar spatial distributions.
-In turn, the distribution of rate of change serves as an estimate of the spatial distribution of DNA in this habitat.
-
-We would love to estimate the minimum distance over which eDNA community differences can be detected.
-
-% We measure 'extra-(corporeal|organismal) DNA
-Some authors have cautioned against the use of DNA-based macrobial communities in marine environments because they are subject to dynamic physical forces (CITE).
-%If environmental circulation
 
@@
 families in the phylum Nemertea), we used conservative, coarse approximations at a higher taxonomic rank.
-We assessed whether or not marine invertebrate taxa are thought to be present in Puget Sound by checking for their presence in a comprehensive checklist of invertebrates of Puget Sound CITE KOZLOFF. 
+%We assessed whether or not marine invertebrate taxa are thought to be present in Puget Sound by checking for their presence in a comprehensive checklist of invertebrates of Puget Sound CITE KOZLOFF.
 These data are available as part of the REFERENCE SUPPLEMENTAL DATA.

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
add new introduction

@@
 \section*{Introduction}
+
+% While ecologists increasingly recognize the importance of ecosystem ?wc ( constituents | components ) across size classes, from viruses to whales,
+No existing biodiversity survey method completely censuses all of the organisms in a given area.
+Towed fishing nets can efficiently sample organisms larger than the mesh and slower than the boat; but ?wc( ignore | overlook | disregard ) viruses and have undesirable effects on charismatic air-breathing species.
+From a boat or aircraft, scientists can visually quantify large whales, but not the krill on which they feed.
+However, DNA-based surveys show great promise as an efficient technique for detecting a previously unthinkable breadth of organisms from a single sample.
+% , from the smallest to the largest
+% are a promising technique for doing so, because they can detect
+
+Microbiologists have used DNA sequencing to quantify the composition of microbial communities in a wide variety of habitats (CITE VENTER, ETC).
+% nucleic acid sequencing; composition (DNA) and function (RNA)
+A sample of environmental medium (e.g.
+water) containing microorganisms is collected, their DNA is isolated and sequenced, and the identity and abundance of sequences is considered a direct measure of the identity and abundance of organisms contained in the sample, which indirectly estimates the quantity of organisms in an area.
+Macroorganisms shed DNA-containing cells into the environment (environmental DNA or eDNA) that can be sampled in the same way.
+The organismal breadth of these methods make them extremely appealing for their potentially high efficiency, but the accuracy and reliability of indirect estimates of macroorganismal abundance has been debated because the entire organisms are not contained within the sample.
+Thus, ?wc( hesitation to adopt | concern over the application of | criticisms of )?
+eDNA methods are generally ?wc( rooted in | grounded in | founded on ) ?wc( concerns regarding ) the uncertainty about the scale of time and space they sample.
+Caution should be used when inferring local presence of an organisms from an environmental sample without knowing how long DNA can persist in that environment and how far it can travel.
+Ultimately, the spatial and temporal scope of an eDNA survey is determined by the rate of DNA degradation and the movement of the environmental medium.
+
+< ? Mention other indirect sampling methods like acoustic monitoring of birds; limits in terms of organismal breadth?>
+
+% Nevertheless, DNA-based surveys ?(of macrofauna?) have been criticized.
+% if citable: (Some of these criticisms are unjustifiable, motivated by reluctance to adapt to changing methodologies.
+% After all, visual methods have been the standard of ecological surveying for arguably thousands of years.
+(Indeed, even other animals use vision (and other senses) to assess the abundance of, e.g.
+prey or predators))
+% Other arguments are justified, particularly those that question the ability of environmental surveys to quantify organisms given their indirect nature.
+
+The ability for eDNA communities to reflect local macroorganism communities is most fundamentally mediated by the rate of DNA degradation.
+DNA degradation is accelerated in the environment by biological, chemical, and physical activity (CITE DEGRADATION COMPARISON). % enzymes, sodium hypochlorite, UV/heat
+These factors vary in space and thus DNA degradation rate is dependent on the environmental context:
+DNA is likely to persist in cold environments for longer than it would in hot environments (CITE TEMPERATURE STUDY).
+However, on local scales, or within a single habitat type, environmental conditions are more likely to be consistent, and thus variance among samples is less likely to be attributable to environmental conditions impacting DNA persistence.
+
+<<ASIDE: Similarly, environmental factors can affect detection efficiency even when DNA is known to be present in the environment, by inhibiting lab procedures like PCR (CITE PCR INHIBITION)>>
+
+The relationship between local organismal abundance and DNA abundance is further complicated in habitats where the environmental medium itself may transport eDNA away from its source.
+We know that genetic material moves away from its source precisely because organisms can be detected indirectly without being present in the sample.
+% A whale need not be captured in a sample to recover whale DNA from water--
+eDNA is more likely to travel farther in a highly dynamic fluid such as the open ocean or flowing river than it would through the sediment at the bottom of a stagnant pond.
+(CITE DEINER STREAM EDNA CONVEYER BELTS)
+Yet even studies of extremely dynamic habitats such as coastlines with high wave energy have found remarkable evidence that eDNA transport is limited enough that DNA methods can detect community differences separated by less than 100 meters (CITE Port).
+
+Several rigorous studies have investigated the specific effect of various environmental factors on eDNA persistence under controlled laboratory conditions (CITE ENVIRONMENTAL FACTORS ~ DEGRADATION).
+Other studies have investigated the magnitude of environmental movement on the transport of eDNA in specific contexts, such as streams (CITE DEINER STREAM CONVEYER BELT).
+In theory, given enough data from manipulative experiments, it is possible that one could measure environmental variables--temperature, UV intensity, flow, etc--at the time of eDNA sample collection in order to more precisely estimate the origin of eDNA.
+In practice, this approach is unlikely to yield appreciable gains in accuracy because even the most advanced sensors ((thinking of ADCPs here)) and analytical techniques available cannot determine the origin of particles in highly dynamic fluid environments at scales relevant to most ecological questions (CITE PITERBARG2001, more recents??).
+% http://www.rsmas.miami.edu/LAPCOD/research/prediction/
+Further, collecting and analyzing these data along with eDNA sampling efforts would severely limit the scalability of eDNA methods.
+A more practical approach is to compare the spatial distribution of communities of eDNA with expectations based on prior knowledge of the spatial distribution of communities of organisms.
+
+Species and the communities they comprise are not homogeneously distributed in space; describing and predicting the spatial heterogeneity of ecosystems is of great interest to ecologists.
+One consistently observed patterns of community spatial heterogeneity is that communities close to one another tend to be more similar than those that are farther apart (CITE NEKOLA1999, BELL 2010, ).
+This decrease in community similarity with increasing spatial separation is called distance decay and has been reported from rainforests (CITE CONDIT2002, CHUST2006), ectomycorrhizal fungi (CITE bahram2013JECO), plant communities (CITE guo2015plosone), and marine microorganisms (CITE Martiny2011PNAS,Chust2013GLOBECOBIO, Wetzel2012plosone).
+% coral reefs (CITE HUBBELL2001)
+Typically, this relationship is assessed by regressing a measure of community similarity against a measure of spatial separation for a set of sites at which a set of species' presence or abundance is quantified.
+The underlying ?wc( parameter | mechanism ) which is ultimately thought to drive the slope of this relationship is the rate of movement of individuals among sites.
+Because eDNA is shed and transported away from its source, we expect this increased movement of individual eDNA particles to have a homogenizing effect, and thus erode the distance decay relationship of eDNA communities.
+% Because the distribution of eDNA is ?(fundamentally) more diffuse than that of the ?( individuals | organisms ) from which it originated, communities of eDNA should be more homogenous than organisms.
+% no such pattern has been document from indirect sampling methods such as eDNA.
+
+Here, we explored the spatial scope of eDNA methods in a marine environment.
+We used PCR-based methods and massively parallel sequencing to quantify the presence and abundance of DNA from a suite of marine organisms in water samples collected from a grid of sites.
+% spanning _______ phyla
+We evaluated community similarity in space and tested for a distance decay relationship.
+We estimated the minimum distance over which ( community dissimilarity is / differences in community structure are ) detectable in this habitat.
+Finally, we used an unsupervised classification algorithm to determine how many unique eDNA communities were sampled and to which community was represented at each site.
+% We quantified the relationship between community similarity and distance to estimate the spatial resolution of eDNA in this context, and used unsupervised classification techniques to determine the number  of unique community types and their spatial distribution.
 

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
add environmental sampling methods

@@
 Starting from lower-intertidal patches of \textit{Zostera marina}, we collected water samples at 1 meter depth from 8 points (0, 75, 125, 250, 500, 1000, 2000, and 4000 meters) along three parallel transects separated by 1000 meters (Figure \ref{site_map}).
+To destroy residual DNA on equipment used for field sampling and filtration, we washed with a 1:10 solution of household bleach (8.25\% sodium hypochlorite; 7.25\% available chlorine) and deionized water, followed by thorough rinsing with deionized water.
+For each environmental sample, water was collected in a clean 1 liter high-density polyethylene bottle, the opening of which was covered with 500 micrometer nylon mesh to prevent entry of larger particles.
+Immediately after collecting the sample, the mesh was replaced with a clean lid and the sample was held on ice until filtering.
+1 liter from each water sample was filtered in the lab on a clean polysulfone vacuum filter holder fitted with a 47 millimeter diameter cellulose acetate membrane with 0.45 micrometer pores.
+Filter membranes were moved into 0.9 milliliters of Longmire buffer (CITE LONGMIRE) using clean forceps.
+To test for the extent of wc?(spurious results | contamination) due to laboratory procedures, we filtered three replicate 1 liter samples of deionized water.
+These samples were treated identically to the environmental samples throughout the remaining protocols.
 

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
small change

@@
 To destroy residual DNA on equipment used for field sampling and filtration, we washed with a 1:10 solution of household bleach (8.25\% sodium hypochlorite; 7.25\% available chlorine) and deionized water, followed by thorough rinsing with deionized water.
-For each environmental sample, water was collected in a clean 1 liter high-density polyethylene bottle, the opening of which was covered with 500 micrometer nylon mesh to prevent entry of larger particles.
+Each environmental sample was collected in a clean 1 liter high-density polyethylene bottle, the opening of which was covered with 500 micrometer nylon mesh to prevent entry of larger particles.
 Immediately after collecting the sample, the mesh was replaced with a clean lid and the sample was held on ice until filtering.
@@
 Filter membranes were moved into 0.9 milliliters of Longmire buffer (CITE LONGMIRE) using clean forceps.
-To test for the extent of wc?(spurious results | contamination) due to laboratory procedures, we filtered three replicate 1 liter samples of deionized water.
+To test for the extent of wc?(spurious results | contamination) attributable to laboratory procedures, we filtered three replicate 1 liter samples of deionized water.
 These samples were treated identically to the environmental samples throughout the remaining protocols.

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
add DNA purification

@@
 
-\subsection*{Laboratory Procedures}
+\subsection*{DNA Purification}
 
-Samples were randomly assigned to PCR primer and library adapter index sequences.
-The sequencing run consisted of 14 samples ('libraries') prepared using different index sequences ligated during library preparation.
-Of these libraries, ten comprised of amplicons prepared using the 16S protocol reported above, and four comprised of amplicons prepared using a 12S protocol similar to that reported by (CITE PORT 2015).
+DNA was purified from the membrane following a phenol:chloroform:isoamyl alcohol protocol similar to Renshaw (CITE RENSHAW).
+Preserved membranes were incubated at 65C for 30 minutes before adding 900 microliters of phenol:chloroform:isoamyl alcohol and shaking vigorously for 60 seconds.
+We conducted two consecutive chloroform washes by centrifuging at 14,000 rpm for 5 minutes, transferring the aqueous layer to 700 microliters chloroform, and shaking vigorously for 60 seconds.
+After a third centrifugation, 500 microliters of the aqueous layer was transferred to tubes containing 20 microliters 5 molar NaCl and 500 microliters 100\% isopropanol, and frozen at -20C for approximately 15 hours.
+Finally, all liquid was removed by centrifuging at 14000 rpm for 10 minutes, pouring off or pipetting out any remaining liquid, and drying in a vacuum centrifuge at 45C for 15 minutes.
+DNA was resuspended in 200 microliters of ultrapure water.
+Genomic DNA extracted from tissue of a species absent from the sampled environment (Struthio camelus) served as positive control for the remaining protocols.
 

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
add PCR methods

@@
 Genomic DNA extracted from tissue of a species absent from the sampled environment (Struthio camelus) served as positive control for the remaining protocols.
+
+\subsection*{PCR Amplification}
+
+From each DNA sample, we amplified an approximately 115 base pair (bp) region of the mitochondrial gene encoding 16S RNA using a two-step polymerase chain reaction (PCR) protocol described by ODONNELL (CITE ODONNELL2016).
+In the first step, primers were identical in every reaction (forward: AGTTACYYTAGGGATAACAGCG; reverse: CCGGTCTGAACTCAGATCAYGT); primers in the second step included these same sequences but with 3 variable nucleotides ('NNN') and an index sequence on the 5PRIME end (see CITE PRIMER TABLE/METADATA).
+We used the program OligoTag \cite{Coissac2012} to generate 30 unique 6 nucleotide index sequences differing by a minimum Hamming distance of 3 (\nameref{primer_table}).
+Indexed primers were assigned to samples randomly, with the identical index sequence on the forward and reverse primer to avoid errors associated with dual-indexed multiplexing (CITE Schnell2015).
+In a UV-sterilized hood, we prepared 25 microliter reactions containing 18.375 microliters ultrapure water, 2.5 microliters 10x buffer, 0.625 microliters deoxynucleotide solution (8 millimolar), 1 microliter each forward and reverse primer (10 micromolar, obtained lyophilized from Integrated DNA Technologies (Coralville, IA, USA)), 0.25 microliters Qiagen HotStar Taq polymerase, and 1.25 microliter genomic eDNA template at 1:100 dilution in ultrapure water.
+PCR thermal profiles began with an initialization step (95C; 15 min) followed by cycles (40 and 20 for the first and second reaction, respectively) of denaturation (95C; 15 sec), annealing (61C; 30 sec), and extension (72C; 30 sec).
+20 identical PCRs were conducted from each DNA extract using non-indexed primers; these were pooled into 4 groups of 5 in order to ensure ample template for the subsequent PCR with indexed primers.
+% we probably pooled >1 product from PCR1, but figuring this out from our lab notebook would be time consuming and perhaps impossible.
+In order to isolate the fragment of interest from primer dimer and other spurious fragments generated in the first PCR, we used the AxyPrep Mag FragmentSelect-I kit with solid-phase reversible immobilization (SPRI) paramagnetic beads at 2.5x the volume of PCR product (Axygen BioSciences, Corning, NY, USA).
+A 1:5 dilution in ultrapure water of the product was used as template for the second reaction.
+PCR products of the second reaction were purified using the Qiagen MinElute PCR Purification Kit (Qiagen, Hilden, Germany).
+Ultrapure water was used in place of template DNA and run along with each batch of PCRs to serve as a negative control for PCR; none of these produced visible bands on an agarose gel.
+In total, four separate replicates from each of 31 DNA samples were carried through the two-step PCR process for a total of 124 sequenced PCR products.
+These were combined with additional samples from other projects, totaling 345 samples.
 

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
add sequencing methods

@@
 These were combined with additional samples from other projects, totaling 345 samples.
+
+\subsection{DNA Sequencing}
+
+PCR products were pooled according to their primer index in equal concentration, and 150 ng was prepared for library sequencing using the KAPA high-throughput library prep kit with real-time library amplification protocol (KAPA Biosystems, Wilmington, MA, USA).
+Each of these ligated sequencing adapter including an additional 6 base pair index sequence (NEXTflex DNA barcodes; BIOO Scientific, Austin, TX, USA).
+Thus, each PCR product was identifiable via its unique combination of index sequences in the sequencing adapters and primers.
+Fragment size distribution and concentration of each library was quantified using an Agilent 2100 BioAnalyzer.
+Libraries were pooled in equal concentrations and sequenced for 150 base pairs in both directions (PE150) using an Illumina NextSeq at the Stanford Functional Genomics Facility (machine NS500615, run 115, flowcell H3LFLAFXX), where 20\% PhiX Control v3 was added to act as a sequencing control and to enhance sequencing depth.
 

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
move data availability to results, specify note about supplement

@@
 
-Pooled libraries were sequenced on the Illumina NextSeq platform at the Stanford Center for Functional Genomics (machine ID: NS50061; run ID: 115; flowcell ID: H3LFLAFXX).
-Raw sequence data in fastq format is publicly available (see Data Availability).
-
 \subsection*{Data Preparation (Bioinformatics)}
 
-Detailed bioinformatic methods are provided in the supplemental material, and scripts used from raw sequencer output onward can be found in the project directory on GitHub (see Data Availability).
+<<Consider: Detailed bioinformatic methods are provided in the supplemental material, and scripts used from raw sequencer output onward can be found in the public project directory (see Data Availability).>>
 
@@
 \section*{Results}
+
 \subsection*{Data Quality (Bioinformatics)}
+Raw sequence data in fastq format is publicly available (see Data Availability).
 All value ranges are reported as (mean $\pm$ standard deviation).\\

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
add bioinformatics

@@
 
-\subsection*{Data Preparation (Bioinformatics)}
+\subsection*{Sequence Data Handling (Bioinformatics)}
 
@@
 
-%----------------------------------------------------------------------------
+Reads passing the preliminary Illumina quality filter were demultiplexed on the basis of the adapter index sequence by the sequencing facility.
+We used fastqc to assess the fastq files output from the sequencer for low-quality indications of a problematic run.
+Forward and reverse reads were merged using PEAR v0.9.6 \cite{Zhang2014} and discarded if more than 0.01 of the bases were uncalled.
+If a read contained two consecutive base calls with quality scores less than 15 (i.e. probability of incorrect base call = 0.0316), these bases and all subsequent bases were removed from the read.
+Paired reads for which the probability of matching by chance alone exceeded 0.01 were not assembled and omitted from the analysis.
+Assembled reads were discarded if assembled sequences were not between 50 and 168 bp long, or if reads did not overlap by at least 100 bp.
+
+We used vsearch VSEARCHVERSION (CITE VSEARCH) to discard any merged reads for which the sum of the per-base error probabilities was greater than 0.5 ("expected errors") \cite{Edgar2010}.
+Sequences were demultiplexed on the basis of the primer index sequence at base positions 4-9 at both ends using the programming language AWK.
+Primer sequences were removed using cutadapt v1.7.1 \cite{Martin2011}, allowing for 2 mismatches in the primer sequence.
+Identical duplicate sequences were identified, counted, and removed in python to speed up subsequent steps by eliminating redundancy, and sequences occurring only once were removed.
+We checked for and removed any sequence likely to be a PCR artifact due to incomplete extension and subsequent mis-priming using a method described by Edgar (CITE UCHIME\cite{Edgar2010}) and implemented in vsearch VSEARCHVERSION.
+Sequences were clustered into operational taxonomic units (OTUs) using the single-linkage clustering method implemented by swarm version SWARMVERSION with a local clustering threshold (d) of 1 and fastidious processing (CITE SWARM).
+
+In order to determine the most likely taxon from which each sequence originated, the representative sequence from each OTU was then queried against the NCBI nucleotide database (GenBank; version October 2015) using the blastn command line utility (CITE BLAST).
+In order to maximize the accuracy of this computationally intensive step, we implemented a nested approach whereby each sequence was first queried using strict parameters (e-value = 5e-52), and if no match was found, the query was repeated with decreasingly strict e-values (5e-48 5e-44 5e-40 5e-36 5e-33 5e-29 5e-25 5e-21 5e-17 5e-13).
+Other parameters were unchanged among repetitions (word size: 7; maximum matches: 1000; culling limit: 100; minimum percent identity: 0).
+Each query sequence can be an equally good match to multiple taxa either because of invariability among taxa or errors in the database (e.g.
+human sequences are commonly attributed to other organisms when they in fact represent lab contamination).
+In order to guard against these spurious results, we used an algorithm to find the lowest common taxon for at least 80\% of the matched taxa, implemented in the R package taxize (CITE TAXIZE VERSION).
+Similarly, we repeated analyses using the dataset consolidated at the same taxonomic rank across all queries, for the rank of both family and order.
+
+The data for input to further analyses are thus a contingency table of counts of unique sequences, OTUs, or taxa present in each PCR.
+
 %METHODS
-%Sequences passing were demultiplexed on the basis of the ?(8-bp) sequence ligated during library preparation and read during the index read, and filtered for general quality by the sequencing facility.
 %
-%We used fastqc to assess the fastq files output from the sequencer for low-quality indications of a problematic run.
 %

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
decontamination methods

@@
 
+Reads that do not contain primer index sequences (or combinations thereof) in the expected position are known to ?(be derived from contamination | result from processes) that leave the same signature as contamination.
+For example, errors during oligonucleotide synthesis or sequencing of the indexed primers could cause reads to be erroneously assigned to samples.
+Similarly, we calculated rates of cross-library contamination by counting occurrences of primer sequences from 12S amplicons prepared separately but pooled and sequenced alongside our samples.
+These occurrences indicate an error in the preparation or sequencing procedures.
+This represents a general minimum rate at which we can expect that sequences from one environmental sample could be erroneously assigned, and so we considered for further analysis only those reads occurring with greater frequency than this across the entire dataset.
+
 In order to determine the most likely taxon from which each sequence originated, the representative sequence from each OTU was then queried against the NCBI nucleotide database (GenBank; version October 2015) using the blastn command line utility (CITE BLAST).
@@
 
-%METHODS
-%
-%
-%Reads that contain primer index sequences (or combinations thereof) that we did not actually use in laboratory procedures are known to ?(be derived from contamination) be a result of processes that leave the same signature as contamination.
-For example, errors during oligonucleotide synthesis of the indexed primers would cause reads to be erroneously assigned to samples.
-%Because this represents the minimum rate at which we can be certain that sequences from one environmental sample could be erroneously assigned, we considered for further analysis only those reads occurring with greater frequency than this.
-%
-We calculated rates of cross-library contamination by counting occurrences of primer sequences: 12S primer sequences appearing in a 16S library (and vice versa) indicate an error in the preparation or sequencing procedures.
-
 We checked for experimental error by evaluating the Bray-Curtis dissimilarity of proportional read abundance among replicate PCRs of the same DNA sample (0.033$\pm$0.063), and excluded one PCR replicate for which the dissimilarities between itself and the other replicates exceeded 1 SD (lib\_B\_tag\_GCGCTC).
 
-%To infer a general estimate of the rate of cross-sample contamination likely to be associated with lab procedures, we calculated the proportion of reads containing primer index sequences (or combinations thereof?) that we did not actually use
-%
 %RESULTS
-%Reads were of generally high-quality, with ?(96.48)? % of base calls having a 0.001 or lower probability of incorrect base call (Phred q-score of 30 or higher).
-%
-%
+%Reads were of generally high-quality, with XXX% of base calls having a 0.001 or lower probability of incorrect base call (Phred q-score of 30 or higher).
 %Total sequencing clusters passing filter per read: 371,576,190 on NEXTSEQ
@@
 % THESE BELOW ARE FROM A MISEQ RUN
-%reads, reads passing filter, % >= Q30, for each run ( 7,052,307 ; 6,752,829; 97.47 :: 7,052,307 ; 6,752,829 ; 95.50)
+% reads passing filter, % >= Q30, for each run ( 7,052,307 ; 6,752,829; 97.47 :: 7,052,307 ; 6,752,829 ; 95.50)
 %

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
add some basic sequencing results

@@
 
-We checked for experimental error by evaluating the Bray-Curtis dissimilarity of proportional read abundance among replicate PCRs of the same DNA sample (0.033$\pm$0.063), and excluded one PCR replicate for which the dissimilarities between itself and the other replicates exceeded 1 SD (lib\_B\_tag\_GCGCTC).
-
-%RESULTS
-%Reads were of generally high-quality, with XXX% of base calls having a 0.001 or lower probability of incorrect base call (Phred q-score of 30 or higher).
-%Total sequencing clusters passing filter per read: 371,576,190 on NEXTSEQ
-%
-% THESE BELOW ARE FROM A MISEQ RUN
-% reads passing filter, % >= Q30, for each run ( 7,052,307 ; 6,752,829; 97.47 :: 7,052,307 ; 6,752,829 ; 95.50)
-%
-%Cluster density of 358 \UTF{00EF}\UTF{00BF}\UTF{0153}22	clusters per mm^2
-%
-%XXX % of reads aligned to the PhiX genome
-%(31.02 \UTF{00EF}\UTF{00BF}\UTF{0153}0.59	; 29.17 \UTF{00EF}\UTF{00BF}\UTF{0153}0.89)
-%
-%and error rate as determined from alignment of PhiX reads to the PhiX genome was low (Read 1: 1.26 \UTF{00EF}\UTF{00BF}\UTF{0153}0.07; Read 2: 1.38 \UTF{00EF}\UTF{00BF}\UTF{0153}0.07)
-%----------------------------------------------------------------------------
+We checked for experimental error by evaluating the Bray-Curtis dissimilarity of proportional read abundance among replicate PCRs from the same DNA sample (0.033$\pm$0.063), and excluded one PCR replicate for which the dissimilarities between itself and the other replicates exceeded 1 SD.% (lib\_B\_tag\_GCGCTC)
 
@@
 Raw sequence data in fastq format is publicly available (see Data Availability).
-All value ranges are reported as (mean $\pm$ standard deviation).\\
+All value ranges are reported as (mean $\pm$ standard deviation).
+A total of 371,576,190 reads passing filter were generated in each direction, with XXX\% of base calls having a 0.001 or lower probability of incorrect base call (Phred q-score of 30 or higher).
+%XXX % of reads aligned to the PhiX genome
+%and error rate as determined from alignment of PhiX reads to the PhiX genome was low (Read 1: 1.26 \UTF{00EF}\UTF{00BF}\UTF{0153}0.07; Read 2: 1.38 \UTF{00EF}\UTF{00BF}\UTF{0153}0.07)
 There was a very low frequency of cross-contamination from other libraries into those reported here (5e-05$\pm$8e-05; max 0.00034)

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
add note about orphan text and clarify PCR similarity

@@
 
+<<< orphaned text but perhaps useful somewhere...>>>
 Samples collected of ecological communities may vary in dissimilarity from 0 (completely identical) to 1 (completely different).
@@
 The rate at which community dissimilarity approaches the mean gives an indication of the rate of community turnover.
+<<< end orphan text >>>
 
@@
 
-We assessed the consistency of PCR by conducting 4 replicate PCRs for each environmental sample and calculating the mean pairwise Bray-Curtis dissimilarity of the resulting communities (scaled to minimum read depth per sample).
-92 of the 96 amplicon samples had mean Bray-Curtis dissimilarity $\leq$ 0.052; 1 sample had a value of 0.341, which elevates the value of the other replicates.
-After removal of this sample, the highest mean Bray-Curtis dissimilarity among replicates within an environmental sample was 0.034.
+PCR replicates were remarkably consistent.
+92 of the 96 environmental samples had a mean Bray-Curtis dissimilarity $\leq$ 0.052 among PCR replicates.
+1 environmental sample had high dissimilarity (0.341) among PCR replicates before removing a single faulty PCR.
+After removal of this PCR replicate, the highest mean Bray-Curtis dissimilarity among replicates within an environmental sample was 0.034.
 

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
analytical methods

@@
 
+We checked for experimental error by evaluating the Bray-Curtis dissimilarity of proportional read abundance among replicate PCRs from the same DNA sample (0.033$\pm$0.063), and excluded one PCR replicate for which the dissimilarities between itself and the other replicates exceeded 1 SD.% (lib\_B\_tag\_GCGCTC)
+
+To account for variation in the number of sequencing reads (sequencing depth) recovered per sample, we multiplied the within-sample proportional abundance of each OTU by the minimum sequencing depth, and rounded to the nearest integer.
+
+Because each step in this workflow is sensitive to contamination, it is possible that some sequences are not truly derived from the environmental sample, and instead represent contamination during field sampling, filtration, DNA extraction, PCR, fragment size selection, quantitation, sequencing adapter ligation, or the sequencing process itself.
+Some authors have argued that these risks could bias sequence abundance, making those data meaningless and prohibiting quantitative estimates; instead these authors advocate for converting count data to binary presence absence data on the basis of sequence abundance greater than some arbitrary threshold.
+Recent work indicates that this binary treatment of data can overestimate taxon richness and falsely elevate the estimate of taxon turnover among samples (CITE LERAY FORTHCOMING).
+We take the view that contaminants are unlikely to manifest as sequences in the final dataset in consistent abundance across replicates; indeed, our data show that the process from PCR onward is remarkably consistent.
+Thus, we calculated from our data the maximum number of sequence counts (after scaling to correct for sequencing depth variation) for which there is turnover in presence-absence among PCR replicates within an environmental sample.
+We use this number to determine a conservative minimum threshold above which we can be confident that counts are consistent among replicates and not of (?spurious | dubious?) origin, and exclude from further analysis observations where the mean abundance across PCR replicates within samples does not reach this threshold.
+% the number is turnover_threshold=178
+
 In order to determine the most likely taxon from which each sequence originated, the representative sequence from each OTU was then queried against the NCBI nucleotide database (GenBank; version October 2015) using the blastn command line utility (CITE BLAST).
@@
 The data for input to further analyses are thus a contingency table of counts of unique sequences, OTUs, or taxa present in each PCR.
-
-We checked for experimental error by evaluating the Bray-Curtis dissimilarity of proportional read abundance among replicate PCRs from the same DNA sample (0.033$\pm$0.063), and excluded one PCR replicate for which the dissimilarities between itself and the other replicates exceeded 1 SD.% (lib\_B\_tag\_GCGCTC)
-
-To account for variation in the number of sequencing reads (sequencing depth) recovered per sample, we multiplied the within-sample proportional abundance of each OTU by the minimum sequencing depth (130402), and rounded to the nearest integer.
-
-Because each step in the massively parallel sequencing workflow is sensitive to contamination, it is possible that some sequences are the result of contamination during field sampling, filtration, DNA extraction, PCR, fragment size selection, quantitation, sequencing adapter ligation, or the sequencing process itself.
-Some authors have argued that these risks could bias sequence abundance, making those data meaningless and prohibiting quantitative estimates, yet convert count data to binary presence absence data on the basis of the sequence abundance greater than some arbitrary threshold).
-%massively parallel metabarcoding,
-Recent work has shown that this binary treatment of data can (?falsely?) overestimate taxon richness and falsely elevate the estimate of taxon turnover among samples (CITE LERAY FORTHCOMING).
-We take the view that contaminants are unlikely to manifest as sequences in the final dataset in consistent abundance across replicates; indeed, our data show that the process from PCR onward is remarkably consistent.
-Thus, we calculated from our data the maximum number of sequence counts (after scaling to correct for sequencing depth variation) for which there is turnover in presence-absence among PCR replicates within an environmental sample.
-We use this number to determine a conservative minimum threshold above which we can be confident that counts are consistent among replicates and not of (?spurious | dubious?) origin, and exclude from further analysis observations where the mean abundance across PCR replicates within samples does not reach this threshold.
-% the number is turnover_threshold=178
 
@@
 % because of the nature of our data (lots of zeros, with some very high counts), the mean is not very informative
-We simultaneously assessed the existence of distinct community types and the membership of samples to those community types using a partitioning around mediods algorithm (CITE PAM, sometimes referred to as k-medoids clustering), as implemented in the R package fpc (CITE fpc).
+We simultaneously assessed the existence of distinct community types and the membership of samples to those community types using an unsupervised classification alogrithm known as partitioning around medoids (CITE PAM, sometimes referred to as k-medoids clustering), as implemented in the R package fpc (CITE fpc).
 The classification of samples to communities was made on the basis of their pairwise Bray-Curtis dissimilarity, calculated using the function vegdist in the R package vegan (CITE VEGAN).
@@
 % Calculate Moran's I on ?(relative abundance | raw count) of ?(taxa | OTU | uniq_sequences) to report dispersion of DNA in space?
+There is a multiple comparison issue and the results are generally not informative to my eye.
 
-To estimate the maximum dissimilarity and the rate of community turnover in space, we modeled community dissimilarity as a function of distance from shore following a Michaelis-Menten relationship:
+To estimate the maximum dissimilarity and the rate of community turnover in space (distance decay), we modeled community dissimilarity as a function of distance from shore following a Michaelis-Menten relationship:
 \begin{equation}\label{MichaelisMenten}
@@
 \subsection*{Organismal Life History Data}
-We collated coarse-scale data on life history characteristic for each of the major taxonomic groups recovered,
-including dispersal range of the gametes, larvae, and adults, adult habitat type and selectivity, and adult body size.
+We collated coarse-scale data on life history characteristic for each of the major taxonomic groups recovered, including dispersal range of the gametes, larvae, and adults, adult habitat type and selectivity, and adult body size.
 Dispersal range was given as an order-of-magnitude approximation of the scale of dispersal: for example, internally fertilized species were assigned a gamete range of 0 km, while broadcast spawners were assigned a gamete range of 10 km.
@@
 %and error rate as determined from alignment of PhiX reads to the PhiX genome was low (Read 1: 1.26 \UTF{00EF}\UTF{00BF}\UTF{0153}0.07; Read 2: 1.38 \UTF{00EF}\UTF{00BF}\UTF{0153}0.07)
+Each environmental sample in the present study was represented by no less than 130402 reads.
 There was a very low frequency of cross-contamination from other libraries into those reported here (5e-05$\pm$8e-05; max 0.00034)
 
-PCR replicates were remarkably consistent.
+Sequence abundances across PCR replicates were remarkably consistent.
 92 of the 96 environmental samples had a mean Bray-Curtis dissimilarity $\leq$ 0.052 among PCR replicates.
@@
 After removal of this PCR replicate, the highest mean Bray-Curtis dissimilarity among replicates within an environmental sample was 0.034.
+Sequences with abundance greater than 178 reads experienced no turnover in presence across PCR replicates within a sample.
 

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
remove model formulation stuff

@@
 
-\subsection*{Spatial Model Formulation}
-We use the general framework outlined by Shelton et al (CITE).
-That study outlined the structure for estimation of the proportional biomass of a taxon ($B_i$) given the proportional counts of sequences recovered from a parallel sequencing run ($Z_i$).
-
-We modeled the counts of DNA sequences ($Z$) from each of a given taxon $i$, in each replicate PCR $j$, from each replicate of a given location $k$ (hence, $Z_{ijk}$), as though they are ?(proportional to/drawn from)?
-a Poisson distribution.
-A Poisson distribution is described by one and only one parameter, $\lambda$, which is equal to both the mean and variance.
-Because in this case our modeled values are discrete counts, we use the natural exponent, $e^\lambda$.
-% WHY IS IT EXPONENTIATED?
-COUNTS?
-Thus,
-\begin{equation}\label{some_cool_eqn_name}
-	Z_{ijk} \sim Poisson(e^{\lambda_{ijk}})
-\end{equation}
-
-In turn, we further assume this parameter $\lambda$ is linearly proportional to a suite of taxon-, pcr-, and site- specific parameters describing the variance associated with each sub-process linking the amount of DNA ($Y$) of a given taxon $i$ at a given location $k$ in a DNA extract (hence $Y_{ik}$):
-
-\begin{equation}\label{GLM}
-	\lambda_{ijk} = \beta_0 + \beta_i + \eta_{ijk} + \epsilon_{ijk}
-\end{equation}
-
-%(\ref{my_equation_label})
-Where $\beta_0$ is a general intercept across all taxa, $\beta_i$ is a fixed effect accounting for the variance associated with taxon $i$, and $\eta_{ijk}$ and $\epsilon_{ijk}$ are random effects of variance resulting from the processes associated with PCR and spatial location, respectively.
-
-%We constructed the following mathematical model (\ref{my_equation_label}) to better understand the concept:
-% using '\ref{something}' in the text will refer to any object (e.g. figure, equation, table) which contains the corresponding '\label{something}'
-% Note that you need to run latex a few times to get it to register numbers correctly
-
-%\begin{equation}\label{my_equation_label}
-%	Y = 2a + 2a
-%\end{equation}
-%
-%Where $a$ represents an apple.
-
 %-------------------------------------------------------------------------------
@@
 Thus, we are unable to compare the patterns of dispersion for varying life history categories.
-
-\subsection*{Spatial Model Output}
 

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
discussion

@@
 \subsection*{Community Analysis}
-Excluding spatially-invariant taxa ?(taxa which occur in only one spatial location) had no discernible effect on the outcome of the PAM analysis (number of clusters, assignment to clusters).
+Excluding spatially-invariant taxa ?(taxa which occur in only one spatial location) had no discernible effect on the outcome of the PAM analysis (number of clusters, assignment to clusters) (CITE FIGUREPAM).
 
@@
 \section*{Discussion}
-Boy those results sure are neat. Now, the pressing question becomes: How do you like them apples?
+Indirect surveys of organismal presence are a key development in ecosystem ?(surveillance | monitoring) in the face of increased anthropogenic pressure and dwindling resources for ecological research (is this citable, or even true?
+Either way, budget constraints drive interest in methods with greater bang per buck).
+Detection of organisms using environmental DNA is an especially promising method, given the ?(rapid pace of technological innovation and cost efficiency | rapid pace of advancement in technological innovation and cost efficiency) in the field of DNA sequencing and quantification.
+To date, little work has ?(specifically | explicitly )? addressed the spatial resolution of DNA in a marine system.
+
+We demonstrate that, despite some expected homogenization of signal compared to known community distributions, a distance-decay relationship exists for eDNA communities in this dynamic environment.
+The distribution of community types did/does not fit our expectations based on known distributions of benthic communities.
+In this area (as in much of Puget Sound), a narrow band of shallow, soft bottom hugs the shoreline and supports eelgrass meadows which harbor diverse epifaunal and infaunal communities, and attract transient species which use the meadows for shelter and to feed on epifaunal biota.
+This shallow habitat quickly gives way to a central depth of approximately 200 meters.
+Intertidal shoreline varies between soft, fine sediment to cobble and boulder rubble.
+Hard intertidal surfaces support a well-documented ?(biota/fauna/community) ?(including/dominated by)?
+barnacles (Sessilia), mussels (Bivalvia:Mytilidae), anemones (Actinaria), sea stars (Asteroidea), urchins (Echinoidea), Bryzoans (Ectoprocta), and crustaceans (Decapoda).
+Soft intertidal sediments are inhabited by burrowing bivalves (Bivalvia:various families), segmented worms (Annelida), acorn worms (Enteropneusta), and in the lower intertidal and high subtidal ranges by eelgrass ?(beds | meadows).
+Hard bottoms: lower intertidal and high subtidal = macroalgae such as nereocystis etc (check M. Foley's list) which provides habitat for a diverse community of ?(fish and invertebrates)
+
+Our samples were collected at the surface over depths ranging from approximately 2 to 200 meters.
+The benthic habitats directly below each sample point vary from eelgrass meadow to soft-bottom ?( sub-photic? ) habitats.
+At the sampling points over the ?(deepest water/ greatest depth), samples were closer to the bottom than they were to the next closest sampling point.
+However, we found no distinct pattern (/signal) of deep-water communities, indicating (1) the a greater biomass of nearshore (/shallow) organisms results in ?(more recoverable genetic material | greater concentrations of eDNA)?
+even over long distances.
+
+{ Material from these deep water communities is not transported to the surface in ?(quantities / amounts )?
+sufficient for detection in sequencing based approaches using general primers, as in the present study.
+
+% We found no evidence for....
+
+Contrary to our ?(perhaps naive) expectations based on well-documented patterns of benthic community distribution, we found no discernible pattern of communitiy turnover as a function of distance from shore, nor did the composition of those communities change as expected.
+The most abundant taxa at sites furthest from shore are taxa whose adult stages are restricted to intertidal hard substrates.
+However, the larvae and gametes of these taxa are pelagic and can be transported long distances by water movement (CITE).
+
+Of the ______________ taxa ?(for which we recovered data | detected by our methods), we were able to confirm only one which lacks any microscopic and potentially dispersive life stage, the Shiner Surfperch (Embiotocidae).
+Adults and juveniles of this species are strictly associated with shallow, near-shore habitats.
+Like all members of the family, they are internal fertilizers that give birth to live young, meaning there are no dispersive gametes or larvae which may travel far from juveniles and adults and be detected by our sampling methods.
+C. aggregata DNA was abundant at sites nearest to shore, (M_____\%+-SD of all sequences,), rare or absent elsewhere (CITE FIG).
+We interpret this as evidence that the chaotic spatial distribution of eDNA communities (CITE FIG PAM) results from our primers' affinity for many species which at some point exist as microscopic pelagic gametes or larvae.
+Communities of microscopic organisms can be exceptionally variable in space over small scales due to oceanographic features such as eddies.
+Such fine-scale variation would be missed by our sampling design; denser sampling in concert with current profiling would ?(_______) the role of ?(ocean movement???) on eDNA dispersal.
+
+Moreover, if entire multicellular individuals (such as larvae) were directly collected in our water samples, community profiles would be further distorted.
+Our sampling bottles excluded particles larger than 500 micrometers, but gametes and very small larvae could have gained entry.
+It is possible that even a single small individual, containing many thousand mitochondria, would overwhelm the signal of another species from which hundreds of cells had been sloughed from many, larger individuals.
+Data on larval size distribution at the time of sampling from each species in our data set would allow us to estimate the frequency of such events.
+Nevertheless, it is precisely the sensitivity to small particles that makes the eDNA approach powerful, so we are reluctant to recommend that aquatic eDNA sampling use finer pre-filtering.
+Instead, we emphasize the importance of designing and selecting primer sets that selectively amplify target organisms.
+In the case of the present study, in order to recover patterns matching our expectations, this would be non-transient, benthic marine organisms lacking any pelagic life stage.
+Likewise, our results highlight the need for curated life-history databases.
+As technological advances increase the speed and throughput of biodiversity surveys, making sense of these data in a timely manner requires that natural history data be stored in standard formats in centralized repositories.
+<< The rate at which we can make sense of high-throughput biodiversity survey methods will be limited by our ability to incoporate external data.
+Some databases like EOL, GBIF, Fishbase, that house records of taxonomy, occurrence, and other basic data types are growing in number of users, support, and maintenance, but there is no centralized, standardized repository for natural history data.
+As NCBI's nucleotide and protein sequence database (GenBank) has facilitated a multitude of transformative studies in diverse fields,
+an ecological analog would ?(be a huge boon for | revolutionize ) biodiversity science.
 

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
fix textypos and render pdf

@@
 In the first step, primers were identical in every reaction (forward: AGTTACYYTAGGGATAACAGCG; reverse: CCGGTCTGAACTCAGATCAYGT); primers in the second step included these same sequences but with 3 variable nucleotides ('NNN') and an index sequence on the 5PRIME end (see CITE PRIMER TABLE/METADATA).
-We used the program OligoTag \cite{Coissac2012} to generate 30 unique 6 nucleotide index sequences differing by a minimum Hamming distance of 3 (\nameref{primer_table}).
+We used the program OligoTag \cite{Coissac2012} to generate 30 unique 6 nucleotide index sequences differing by a minimum Hamming distance of 3 (CITE primer\_table). %\nameref{primer_table}).
 Indexed primers were assigned to samples randomly, with the identical index sequence on the forward and reverse primer to avoid errors associated with dual-indexed multiplexing (CITE Schnell2015).
@@
 
-Of the ______________ taxa ?(for which we recovered data | detected by our methods), we were able to confirm only one which lacks any microscopic and potentially dispersive life stage, the Shiner Surfperch (Embiotocidae).
+Of the XXXX taxa ?(for which we recovered data | detected by our methods), we were able to confirm only one which lacks any microscopic and potentially dispersive life stage, the Shiner Surfperch (Embiotocidae).
 Adults and juveniles of this species are strictly associated with shallow, near-shore habitats.
 Like all members of the family, they are internal fertilizers that give birth to live young, meaning there are no dispersive gametes or larvae which may travel far from juveniles and adults and be detected by our sampling methods.
-C. aggregata DNA was abundant at sites nearest to shore, (M_____\%+-SD of all sequences,), rare or absent elsewhere (CITE FIG).
+C. aggregata DNA was abundant at sites nearest to shore, (MXXXXX\%+-SD of all sequences,), rare or absent elsewhere (CITE FIG).
 We interpret this as evidence that the chaotic spatial distribution of eDNA communities (CITE FIG PAM) results from our primers' affinity for many species which at some point exist as microscopic pelagic gametes or larvae.
 Communities of microscopic organisms can be exceptionally variable in space over small scales due to oceanographic features such as eddies.
-Such fine-scale variation would be missed by our sampling design; denser sampling in concert with current profiling would ?(_______) the role of ?(ocean movement???) on eDNA dispersal.
+Such fine-scale variation would be missed by our sampling design; denser sampling in concert with current profiling would ?(XXXX) the role of ?(ocean movement???) on eDNA dispersal.
 

### 2016-11-16 Analysis/distance_decay.R
fix legend to escape character in latex

@@ -397,7 +397,7 @@ if(!exists("legend_text")){
 
 plot_name   <- "distance_decay"
 
-legend_text[plot_name] <- "Distance decay relationship of environmental DNA communities. Each point represents the Bray-Curtis similarity of a site sampled along three parallel transects comprising a 3000 by 4000 meter grid. Blue dashed line represents fit of a nonlinear least squares regression (see Methods), and shading denotes the 95% confidence interval. Boxplot is comparisons within-sample across PCR replicates, separated by a vertical line at zero, where the central line is the median, the box encompasses the interquartile range, and the lines extend to 1.5 times the interquartile range. Boxplot outliers are omitted for clarity."
+legend_text[plot_name] <- "Distance decay relationship of environmental DNA communities. Each point represents the Bray-Curtis similarity of a site sampled along three parallel transects comprising a 3000 by 4000 meter grid. Blue dashed line represents fit of a nonlinear least squares regression (see Methods), and shading denotes the 95\% confidence interval. Boxplot is comparisons within-sample across PCR replicates, separated by a vertical line at zero, where the central line is the median, the box encompasses the interquartile range, and the lines extend to 1.5 times the interquartile range. Boxplot outliers are omitted for clarity."
 
 if(EXPORT){
   pdf_file    <- file.path(fig_dir, paste(plot_name, ".pdf", sep = ""))

### 2016-11-16 Analysis/distance_decay_multi.R

@@ -366,7 +366,7 @@ OTU abundance scaled relative to within-taxon maximum (m-p),
 and exclusion of OTUs found at only one site (q-t).
 Columns indicate the similarity index used (Bray Curtis or Morisita-Horn) and
 whether the input was full abundance data or binary (0,1) transformed data.
-Lines and bands illustrate the fit and 95% confidence interval of both the
+Lines and bands illustrate the fit and 95\% confidence interval of both the
 main nonlinear model (red, dashed line) and a simple linear model (blue, solid line).
 Results using the Jaccard distance are omitted because of its similarity to Bray-Curtis."
 }

### 2016-11-16 Analysis/diversity_space.R

@@ -123,7 +123,7 @@ if(!exists("legend_text")){ legend_text <- list()}
 legend_text[plot_name] <- {
 "Aggregate diversity metrics of each site plotted against distance from shore.
 Both Simpson's Index (top) and richness (bottom) are shown for a variety of data subsets and transformations (left to right: mean, unfiltered mean, log(x + 1), transformed, scaled, spatially variable, and taxon clustered). 
-Lines and bands illustrate the fit and 95% confidence interval of a linear model.
+Lines and bands illustrate the fit and 95\% confidence interval of a linear model.
 See methods text for detailed data descriptions."
 
 }

### 2016-11-16 Manuscript/v1.0_PeerJ/carkeek_grid.tex
add first publication version

%--------------------------------------||--------------------------------------%

\documentclass[11pt,letterpaper]{article} %  font size

%-----------------------------------PACKAGES-----------------------------------%

\usepackage[T1]{fontenc} % Choose an output font encoding (T1) that has support for the accented characters used by the most widespread European languages
\usepackage[utf8]{inputenc} % Allow input of accented characters (and more...)
\usepackage{graphicx} %  figures
\graphicspath{ {../../Figures/} {../Figures/} } % include images in the directory 'Figures' at the same level
\usepackage[round]{natbib} % names in citations
\usepackage{lineno} % line numbers
\usepackage{authblk} % allows more intuitive formatting for multiple authors/affiliations
\usepackage[margin=1in]{geometry} % make the margins 1 inch on all sides of the document
%\usepackage{amsmath} % useful for formatting math stuff, especially complex equations
%\usepackage{pdflscape} % rotate table into landscape mode
%\usepackage{subfigure} % side by side figures
%\usepackage{longtable} % for tables that span multiple pages.
\usepackage{setspace} % double spacing
\usepackage{float} % figure placement [H]
\usepackage{nameref,hyperref} % named references (e.g. "Supplemental Material")

%----------------------------------FORMATTING-----------------------------------

\date{\today}
\doublespacing % initiate double spacing (package setspace)
%\linespread{2} % alternate method of double spacing

%------------------------------------TITLE--------------------------------------

\title{Spatial distribution of environmental DNA in a nearshore marine habitat}

%-----------------------------------AUTHORS-------------------------------------

% using package 'authblk':
\author[1]{James L. O'Donnell\thanks{jodonnellbio@gmail.com}}
\author[1]{Ryan P. Kelly}
\author[2]{Andrew O. Shelton}
\author[2]{Jameal F. Samhouri}
\author[1,3]{Natalie C. Lowell}
\author[4]{Gregory D. Williams}

% Make sure authors specify/clarify department, institution, and address.
\affil[1]{School of Marine and Environmental Affairs, University of Washington, 3707 Brooklyn Ave NE, Seattle, Washington 98105, USA}
\affil[2]{Northwest Fisheries Science Center, NOAA Fisheries, 2725 Montlake Blvd E, Seattle, Washington 98112, USA}
\affil[3]{School of Aquatic and Fishery Sciences, University of Washington, 1122 NE Boat St, Seattle, Washington 98105, USA}
\affil[4]{Pacific States Marine Fisheries Commission, Under contract to the Northwest Fisheries Science Center, NOAA Fisheries, 2725 Montlake Blvd E, Seattle, WA 98112}

%----------------------------------FORMATTING-----------------------------------
\begin{document}
\maketitle
\linenumbers % start line numbers
\def\linenumberfont{\normalfont\small\rmfamily} % change line number font

%-----------------------------------NOTES---------------------------------------

% Target journal: PeerJ. Backups: Frontiers in...marine science?, BMC Ecology, BMC Genetics

%----------------------------------REVIEWERS------------------------------------
% A list of recommended reviewers.
% Frédéric Mahé mahefrederic@gmail.com

% Sophie Arnaud-Haond, sophie.arnaud@ifremer.fr
% Dominique Cowart dcowart@illinois.edu dominique.cowart@gmail.com (Metabarcoding is powerful yet still blind)

% Ann Bucklin, ann.bucklin@uconn.edu (Metabarcoding of marine zooplankton)

% Naiara Rodríguez-Ezpeleta, nrodriguez@azti.es (Benchmarking DNA Metabarcoding for Biodiversity-Based Monitoring and Assessment)

% Xavier Turon, xturon@ceab.csic.es (Metabarcoding Extracellular DNA from Sediments of Marine Canyons)

% Andrew Mahon, mahon2a@cmich.edu

% Eske Willerslev ewillerslev@snm.ku.dk

% Philip Thomsen pfthomsen@snm.ku.dk

%----------------------------------KEYWORDS-------------------------------------
\section*{Keywords}

metagenomics, metabarcoding, environmental monitoring, molecular ecology, marine, estuarine

%----------------------------------ABSTRACT-------------------------------------
\section*{Abstract}
In the face of increasing threats to biodiversity, the advancement of methods for surveying biological communities is a major priority for ecologists.
Recent advances in molecular biological technologies have made it possible to detect and sequence DNA from environmental samples (environmental DNA or eDNA); however, eDNA techniques have not yet seen widespread adoption as a routine method for biological surveillance primarily due to gaps in our understanding of the dynamics of eDNA in space and time.
In order to identify the effective spatial scale of this approach in a dynamic marine environment, we collected marine surface water samples from transects ranging from the intertidal zone to 4 kilometers from shore.
Using massively parallel sequencing of 16S amplicons, we identified a diverse community of metazoans and quantified their spatial patterns using a variety of statistical tools.
We find evidence for multiple, discrete eDNA communities in this habitat, and show that these communities decrease in similarity as they become further apart.
Offshore communities tend to be richer but less even than those inshore, though diversity was not spatially autocorrelated.
Taxon-specific relative abundance coincided with our expectations of spatial distribution in taxa lacking a microscopic, pelagic life-history stage, though most of the taxa detected do not meet these criteria.
Finally, we use carefully replicated laboratory procedures to show that laboratory treatments were remarkably similar in most cases, while allowing us to detect a faulty replicate, emphasizing the importance of replication to metabarcoding studies.
While there is much work to be done before eDNA techniques can be confidently deployed as a standard method for ecological monitoring, this study serves as a first analysis of diversity at the fine spatial scales relevant to marine ecologists and confirms the promise of eDNA in dynamic environments.

%---------------------------------INTRODUCTION----------------------------------
\section*{Introduction}

No existing biodiversity survey method completely censuses all of the organisms in a given area. For example, towed fishing nets can efficiently sample organisms larger than the mesh and slower than the boat; but overlook viruses and have undesirable effects on charismatic air-breathing species. From a boat or aircraft, scientists can count whales by sight, but not the krill on which they feed. However, DNA-based surveys show great promise as an efficient technique for detecting a previously unthinkable breadth of organisms from a single sample.

Microbiologists have used nucleic acid sequencing to quantify the composition and function of microbial communities in a wide variety of habitats \citep{Handelsman1998, Tyson2004, Venter2004, Iverson2012}. To do so, a sample of environmental medium (e.g. water) containing microorganisms is collected, their DNA or RNA is isolated and sequenced, and the identity and abundance of sequences is considered to reflect the community of  organisms contained in the sample, which indirectly estimates the quantity of organisms in an area.

Macroorganisms shed DNA-containing cells into the environment (environmental DNA or eDNA) that can be sampled in the same way  \citep{Ficetola2008,Thomsen2012}. Potentially, eDNA methods allow a broad swath of macroorganisms to be surveyed from basic environmental samples. However, the accuracy and reliability of indirect estimates of macroorganismal abundance has been debated because the entire organisms are not contained within the sample \citep{Cowart2015}. Concern surrounding eDNA methods is rooted in uncertainty about the attributes of eDNA in the environment relative to actual organisms \citep{Shelton2016, Evans2016}. Basic questions such as how long DNA can persist in that environment and how far DNA can travel remain largely unknown (but see \cite{Klymus2015,Turner2015,Strickler2015,Deiner2014}) and impede inference about local organismal presence from an environmental sample. As a result, estimating the spatial and temporal resolution of eDNA studies in the field is a key step in making these methods practical.

The relationship between local organismal abundance and eDNA is further complicated in habitats where the environmental medium itself may transport eDNA away from its source. We know that genetic material moves away from its source precisely because organisms can be detected indirectly without being present in the sample \citep{Kelly2016}. One might reasonably expect eDNA to travel farther in a highly dynamic fluid such as the open ocean or flowing river than it would through the sediment at the bottom of a stagnant pond \citep{Deiner2014, Shogren2016}. Yet even studies of extremely dynamic habitats such as coastlines with high wave energy have found remarkable evidence that eDNA transport is limited enough that DNA methods can detect differences among communities separated by less than 100 meters \citep{Port2016}.

While rigorous laboratory studies have investigated the effects of some environmental factors on eDNA persistence \citep{Klymus2015, Barnes2014, Sassoubre2016} and the transport of eDNA in specific contexts \citep{Deiner2014}, we suggest that field studies comparing the spatial distribution of communities of eDNA with expectations based on prior knowledge of organisms' distributions are also critical to developing a working understanding of eDNA in the real world.

Species and the communities they comprise are not homogeneously distributed in space; describing and predicting the spatial heterogeneity of ecosystems is of great interest to ecologists.  There is a large and active literature focused on understanding the patterns and causes of community variability in space \citep{Hubbell2001, Anderson2011}. One consistently observed pattern of community spatial heterogeneity is that communities close to one another tend to be more similar than those that are farther apart \citep{Nekola1999}. This decrease in community similarity with increasing spatial separation is called distance decay and has been reported from communities of tropical trees \citep{Condit2002, Chust2006}, ectomycorrhizal fungi \citep{Bahram2013}, salt marsh plants \citep{Guo2015}, and microorganisms \citep{Martiny2011, Chust2013, Wetzel2012, Bell2010}. Typically, this relationship is assessed by regressing a measure of community similarity against a measure of spatial separation for a set of sites at which a set of species' abundances (or presences) is calculated.

We can apply these methods derived from community ecology to understand spatial patterns and patchiness of eDNA. The underlying mechanism thought to drive the slope of this relationship is the rate of movement of individuals among sites, which may be driven by underlying processes such as habitat suitability. Because eDNA is shed and transported away from its source, the increased movement of eDNA particles should homogenize community similarity, and thus erode the distance decay relationship of eDNA communities.

Puget Sound is a deep, narrow fjord in Washington, USA, where a narrow band of shallow bottom hugs the shoreline and abruptly gives way to a central depth of up to 300 meters. This form allows the juxtaposition of communities associated with distinctly different habitats: shallow, intertidal benthos, and euphotic pelagic \citep{Burns1985}. At the upper reaches of the intertidal, the shoreline substrate varies from soft, fine sediment to cobble and boulder rubble. Soft intertidal sediments are inhabited by burrowing bivalves (Bivalvia), segmented worms (Annelida), and acorn worms (Enteropneusta), and in some lower intertidal and high subtidal ranges by eelgrass (\textit{Zostera marina}) \citep{Kozloff1973, Dethier2010} . Eelgrass meadows harbor epifaunal and infaunal biota, and attract transient species which use the meadows for shelter and to feed on resident organisms. Hard intertidal surfaces support a well-documented biota including barnacles (Sessilia), mussels (Bivalvia:Mytilidae), anemones (Actinaria), sea stars (Asteroidea), urchins (Echinoidea), Bryzoans (Ectoprocta), crustaceans (Decapoda), and a variety of algae \citep{Dethier2010}. Hard bottoms of the lower intertidal and high subtidal are home to macroalgae such as Laminariales and Desmarestiales which provides habitat for a distinct community of fish and invertebrates. The upper pelagic is home to a diverse assemblage of microscopic plankton including diatoms and larvae \citep{Strickland1983}, as well as transitory fish and marine mammals.

We took advantage of this setting to explore the spatial variation and distribution of marine eDNA communities. Using PCR-based methods and massively parallel sequencing, we surveyed 16S mtDNA from a suite of marine animals in water samples collected over a grid of sites extending from the shoreline out to 4 kilometers offshore in Puget Sound, Washington, USA. We leverage this sampling design to perform the first explicitly spatial analysis of eDNA-derived community similarity. We investigate two primary objectives. First we examine the spatial patterning of eDNA and determine the degree to which eDNA community similarity can be predicted by physical proximity. We expect that physical proximity will be a strong predictor of community similarity, and that community differences can be detected over small distances. Second, we examine the distribution of diversity from eDNA data, and compare it to our expectations based on distributions of macrobial communities. We expect that distinct eDNA communities exist in this setting, and that their spatial distribution coincides with that of adult macrobial organisms. Because of the vastly different communities of benthic macrobial metazoans as a function of distance from shore, we expect that more than one eDNA community is present across our 4 kilometer sampling grid, and that communities change as a function of distance from shore. For this reason, we examine two diversity measures of eDNA communities that have been widely used to reveal broad scale patterns based on macrobiota in many ecological systems. Finally, we identify the taxa represented in the eDNA communities, which span a range of life-history characteristics, and we expect that the spatial distribution of eDNA will most closely resemble the distribution of adults in taxa with low dispersal potential.

%-----------------------------------METHODS-------------------------------------
\section*{Methods}
There are seven discrete steps to our methodology: (1) Environmental sample collection, (2) isolation of particulates from water via filtration, (3) isolation of DNA from filter membrane, (4) amplification of target locus via PCR, (5) sequencing of amplicons, (6) bioinformatic translation of raw sequence data into tables of sequence abundance among samples, and (7) community ecological analyses of eDNA. We provide brief overviews of these steps here, and encourage the reader to review the fully detailed methods presented in the supplementary material (\nameref{supplement}).

\subsection*{Environmental Sampling}
Starting from lower-intertidal patches of \textit{Zostera marina}, we collected water samples at 1 meter depth from 8 points (0, 75, 125, 250, 500, 1000, 2000, and 4000 meters) along three parallel transects separated by 1000 meters (24 sample locations total; Figure \ref{site_map}). To destroy residual DNA on equipment used for field sampling and filtration, we washed with a 1:10 solution of household bleach (8.25\% sodium hypochlorite; 7.25\% available chlorine) and deionized water, followed by thorough rinsing with deionized water. Each environmental sample was collected in a clean 1 liter high-density polyethylene bottle, the opening of which was covered with 500 micrometer nylon mesh to prevent entry of larger particles. Immediately after collecting the sample, the mesh was replaced with a clean lid and the sample was held on ice until filtering.

\subsection*{Filtration}
One liter from each water sample was filtered in the lab on a clean polysulfone vacuum filter holder fitted with a 47 millimeter diameter cellulose acetate membrane with 0.45 micrometer pores. Filter membranes were moved into 900 microliters of Longmire buffer \citep{Longmire1997} using clean forceps and stored at room temperature \citep{Renshaw2014}. To test for the extent of contamination attributable to laboratory procedures, we filtered three replicate 1 liter samples of deionized water. These samples were treated identically to the environmental samples throughout the remaining protocols.

\subsection*{DNA Purification}
DNA was purified from the membrane following a phenol:chloroform:isoamyl alcohol protocol following Renshaw \citep{Renshaw2014}. Preserved membranes were incubated at 65C for 30 minutes before adding 900 microliters of phenol:chloroform:isoamyl alcohol and shaking vigorously for 60 seconds. We conducted two consecutive chloroform washes by centrifuging at 14,000 rpm for 5 minutes, transferring the aqueous layer to 700 microliters chloroform, and shaking vigorously for 60 seconds. After a third centrifugation, 500 microliters of the aqueous layer was transferred to tubes containing 20 microliters 5 molar NaCl and 500 microliters 100\% isopropanol, and frozen at -20C for approximately 15 hours. Finally, all liquid was removed by centrifuging at 14000 rpm for 10 minutes, pouring off or pipetting out any remaining liquid, and drying in a vacuum centrifuge at 45C for 15 minutes. DNA was resuspended in 200 microliters of ultrapure water. Genomic DNA extracted from tissue of a species absent from the sampled environment (\textit{Oreochromis niloticus}) served as positive control for the remaining protocols.

\subsection*{PCR Amplification}
From each DNA sample, we amplified an approximately 115 base pair (bp) region of the mitochondrial gene encoding 16S RNA using a two-step polymerase chain reaction (PCR) protocol described by \citet{ODonnell2016}. In the first set of reactions, primers were identical in every reaction (forward: AGTTACYYTAGGGATAACAGCG; reverse: CCGGTCTGAACTCAGATCAYGT); primers in the second set of reactions included these same sequences but with 3 variable nucleotides (NNN) and an index sequence on the 5$'$ end (see \nameref{sequencing_metadata}). We used the program OligoTag \citep{Coissac2012} to generate 30 unique 6-nucleotide index sequences differing by a minimum Hamming distance of 3 (see \nameref{sequencing_metadata}). Indexed primers were assigned to samples randomly, with the identical index sequence on the forward and reverse primer to avoid errors associated with dual-indexed multiplexing \citep{Schnell2015}. In a UV-sterilized hood, we prepared 25 microliter reactions containing 18.375 microliters ultrapure water, 2.5 microliters 10x buffer, 0.625 microliters deoxynucleotide solution (8 millimolar), 1 microliter each forward and reverse primer (10 micromolar, obtained lyophilized from Integrated DNA Technologies (Coralville, IA, USA)), 0.25 microliters Qiagen HotStar Taq polymerase, and 1.25 microliter genomic eDNA template at 1:100 dilution in ultrapure water. PCR thermal profiles began with an initialization step (95C; 15 min) followed by cycles (40 and 20 for the first and second reaction, respectively) of denaturation (95C; 15 sec), annealing (61C; 30 sec), and extension (72C; 30 sec). 20 identical PCRs were conducted from each DNA extract using non-indexed primers; these were pooled into 4 groups of 5 in order to ensure ample template for the subsequent PCR with indexed primers. In order to isolate the fragment of interest from primer dimer and other spurious fragments generated in the first PCR, we used the AxyPrep Mag FragmentSelect-I kit with solid-phase reversible immobilization (SPRI) paramagnetic beads at 2.5x the volume of PCR product (Axygen BioSciences, Corning, NY, USA). A 1:5 dilution in ultrapure water of the product was used as template for the second reaction. PCR products of the second reaction were purified using the Qiagen MinElute PCR Purification Kit (Qiagen, Hilden, Germany). Ultrapure water was used in place of template DNA and run along with each batch of PCRs to serve as a negative control for PCR; none of these produced visible bands on an agarose gel. In total, four separate replicates from each of 31 DNA samples were carried through the two-step PCR process for a total of 124 sequenced PCR products. These were combined with additional samples from other projects, totaling 345 samples for sequencing.

\subsection*{DNA Sequencing}
PCR products were pooled according to their primer index in equal concentration, and 150 ng was prepared for library sequencing using the KAPA high-throughput library prep kit with real-time library amplification protocol (KAPA Biosystems, Wilmington, MA, USA). Each of these ligated sequencing adapter including an additional 6 base pair index sequence (NEXTflex DNA barcodes; BIOO Scientific, Austin, TX, USA). Thus, each PCR product was identifiable via its unique combination of index sequences in the sequencing adapters and primers. Fragment size distribution and concentration of each library was quantified using an Agilent 2100 BioAnalyzer. Libraries were pooled in equal concentrations and sequenced for 150 base pairs in both directions (PE150) using an Illumina NextSeq at the Stanford Functional Genomics Facility (machine NS500615, run 115, flowcell H3LFLAFXX), where 20\% PhiX Control v3 was added to act as a sequencing control and to enhance sequencing depth. Raw sequence data in fastq format is publicly available (see Data Availability).

\subsection*{Sequence Data Processing (Bioinformatics)}
Detailed bioinformatic methods are provided in the supplemental material, and analysis scripts used from raw sequencer output onward can be found in the public project directory (see \nameref{analysis_scripts}). Briefly, we performed five steps to process the sequence data: (1) Merge paired-end reads, (2) eliminate low-quality reads, (3) eliminate PCR artifacts (chimeras), (4) cluster reads by similarity into operational taxonomic units (OTUs), and (5) match observed sequences to taxon names. Additionally, we checked for consistency among PCR replicates, excluded extremely rare sequences, and rescaled (rarefied) the data to account for differences in sequencing depth. The data for input to further analyses are a contingency table of the mean count of unique sequences, OTUs, or taxa present in each environmental sample.

\subsection*{Ecological Analyses}
After gathering the data, we use the eDNA community observed at each location to make inferences about the spatial patterning of eDNA communities. We use statistical tools from community ecology to assess the spatial structure of eDNA communities. We report similarity (1- dissimilarity) rather than dissimilarity in all cases for ease of interpretation.

\subsection*{Objective 1: Community similarity as a function of distance}
\subsubsection*{Distance Decay}
To address our first objective and determine whether or not nearby samples are more similar than distant ones, we fit a nonlinear model to represent decreasing community similarity with distance. We calculated the pairwise Bray-Curtis similarity (1 - Bray-Curtis dissimilarity) between eDNA communities using the R package vegan \citep{vegan} and the great circle distance between sampling points using the Haversine method as implemented by the R package geosphere \citep{geosphere}. This model is similar to the Michaelis-Menten function, but with an asymptote fixed at 0:

\begin{equation}\label{MM_asy0}
	y_{ij} = \frac{AB}{B + x_{ij}}
\end{equation}

Where the relationship between community similarity ($y_{ij}$) and spatial distance ($x_{ij}$) between observations $i$ and $j$ is determined by the similarity of samples at distance 0 ($A$), and the distance at which half the total change in similarity is achieved ($B$). This allows for a samples collected very close together (near 0) to have similarity significantly less than one. We assessed model fit using the R function nls \citep{R}, using the nl2sol algorithm from the Port library to solve separable nonlinear least squares using analytically computed derivatives (http://netlib.org/port/nsg.f). We set bounds of 0 and 1 for the intercept parameter and a lower bound of 0 for the distance at half similarity; starting values of these parameters were 0.5 and $x_{max}/2$, respectively. We calculated a 95\% confidence interval for the parameters and the predicted values using a first-order Taylor expansion approach implemented by the function predictNLS in the R package propagate \citep{propagate}.

There are other conceptually reasonable forms to expect the space-by-similarlity relationship to take; we present these in the supplemental material along with alternative data subsets and similarity indices (see \nameref{supplement}).

\subsection*{Objective 2: Spatial distribution of diversity}
\subsubsection*{Community Classification}
To determine the spatial distribution and variation of eDNA communities (objective 2), we used multivariate classification algorithms. We simultaneously assessed the existence of distinct community types and the membership of samples to those community types using an unsupervised classification algorithm known as partitioning around medoids (PAM; sometimes referred to as k-medoids clustering) \citep{Kaufman1990}, as implemented in the R package cluster \citep{cluster}. The classification of samples to communities was made on the basis of their pairwise Bray-Curtis similarity, calculated using the function vegdist in the R package vegan \citep{vegan}. Other distance metrics were evaluated but had no appreciable effect on the outcome of the analysis (Figure \ref{distance_decay_multi}). In order to chose an optimal number of clusters ($K$), we evaluated the distribution of silhouette widths, a measure of the similarity between each sample and its cluster compared to its similarity to other clusters. We repeated the analysis using fuzzy clustering (FANNY, \citep{Kaufman1990}; however, the results were qualitatively similar to the results using PAM so we omit them here.

\subsubsection*{Aggregate Measures of Diversity}
We calculated two measures of diversity (richness and 1-Simpson's Index) to ask if aggregate metrics of the eDNA community showed evidence of spatial patterning. Richness is a measure of the number of distinct types of organisms present and so ranges from 1 (only one taxon observed) to R, the number of taxa observed across all samples. 1-Simpson's is a measure of the evenness of the distribution of abundance of taxa in a sample and ranges from 0 to 1, with the value interpreted as the probability that two sequences randomly selected from the sample will belong to different taxa. Thus larger values of the index indicate more evenly divided communities. We calculated Moran's I for both diversity metrics to test for spatial autocorrelation. We also tested for a linear effect of log-transformed distance from shore on each measure of diversity to ask how diversity changes over this strong environmental gradient.

\subsubsection*{Taxon and Life History Patterns}
After assigning taxon names to the abundance data, we plotted the distribution in space of a selection of taxa to compare with our expectations on the basis of adult distributions (objective 2). Our aim was to understand where each taxon occurred in the greatest proportional abundance, and its distribution in space relative to that maximum. Thus, we rescaled each sample to proportional abundance, extracted the data from a single taxon, and scaled those values between 0 and 1. We collated  life history characteristics for each of the major taxonomic groups recovered, including dispersal range of the gametes, larvae, and adults, adult habitat type and selectivity, and adult body size. Dispersal range was given as an order-of-magnitude approximation of the scale of dispersal: for example, internally fertilized species were assigned a gamete range of 0 km, while broadcast spawners were assigned a gamete range of 10 km. Similarly, adult range size was approximated as 0 km (sessile), 1 km (motile but not pelagic), or 10 km (highly mobile, pelagic). Variables were specified as 'multiple' for groups known to span more than 1 magnitude of range size. For groups to which sequences were annotated with high confidence, but for which life history strategy is diverse or poorly known (e.g. families in the phylum Nemertea), we used conservative, coarse approximations at a higher taxonomic rank (see \nameref{life_history_data}).

%-----------------------------------RESULTS-------------------------------------
\section*{Results}
\subsection*{Sequence Data Processing (Bioinformatics)}
Preliminary sequence analysis strongly suggested that the observed variation among environmental samples reflects true variation in the environment, rather than variability due to lab protocols, for the following reasons (note that all value ranges are reported as mean plus and minus one standard deviation). First, all libraries passed the FastQC per-base sequence quality filter, generating a total of 371,576,190 reads passing filter generated in each direction. Second, samples in this study were represented by an adequate number of reads (333,537.9 $\pm$ 112,200.5), with no individual sample receiving fewer than 130,402 reads. Third, there was a very low frequency of cross-contamination from other libraries into those reported here (5e-05$\pm$8e-05; max proportion 0.00034). Fourth, after scaling all samples to the same sequencing depth, OTUs with abundance greater than 178 reads (0.14\% of a sample's reads) experienced no turnover among PCR replicates within a sample. Fifth, sequence abundances among PCR replicates within water samples were remarkably consistent. A single sample had low similarity among PCR replicates (0.659) after removing this outlier, the lowest mean similarity among replicates within a sample was 0.966. Overall similarities among PCR replicates within a sample were extremely high (0.976 $\pm$ 0.013), and far higher than that of than among samples (0.3 $\pm$ 0.16).

\subsection*{Ecological Analyses}
\subsubsection*{Distance Decay}
Physical proximity is a good predictor of eDNA community similarity: Similarity decreased from 0.40 (95\%CI = 0.36, 0.45) to half that amount at 4500 meters (95\%CI = 2900, 7500) (Figure \ref{distance_decay}).

\subsubsection*{Community Classification}
Despite a clear trend in community similarity as a function of spatial separation, the results from our classification analysis are difficult to interpret. The silhouette analysis indicated the presence of 8 distinct communities; however, the gain in mean silhouette width from 2 was small (0.1), and lacked a distinctive peak (Figure \ref{pam_sil}), indicating substantial uncertainty in the clustering algorithm. Thus, we present the results of cluster assignment for both $K$ = 2 and $K$ = 8 to illustrate the range of results (Figure \ref{pam_in_space}). Excluding taxa which occur in only one site had no discernible effect on the outcome of the PAM analysis (number of clusters, assignment to clusters). While there was no distinct spatial divide indicating the presence of an inshore versus an offshore community, one of the two communities (at $K$ = 2) occurred in only 2 out of 18 samples inside 1000 meters from shore, and never occurred within 125 meters of shore, suggesting the presence of an inshore and offshore community.

\subsubsection*{Diversity in Space}
Sites offshore tend to be less diverse (richness) and more even (1-Simpson's D) than those inshore (Figure \ref{diversity_distance}). Mean OTU richness declined by 1.42 per 1000 meters from a mean of 17.6 taxa (95\%CI = 2.15) inshore to 11.9 taxa (95\%CI = 4.31) at offshore locations (p = 0.0415; Fig. figures/diversity.pdf). 1-Simpson's diversity, the probability that two reads chosen at random from a sample belong to different species, increased by .0666 per 1000 meters from 0.225 (95\%CI = 0.0558) to 0.491 (95\%CI = $\pm$ 0.112), indicating that sequence reads were less evenly distributed among taxa in offshore samples (p <<< 0.05; Figure \ref{diversity_distance}). There was no evidence for spatial autocorrelation for any of the diversity metrics (Moran's I, p > 0.05; Figure \ref{diversity}).

\subsubsection*{Taxon and Life History Patterns}
We were able to assign a taxon name with confidence to 136 of 146 OTU sequences. The vast majority of sequences (97.6\%) and OTUs (96.9\%) were matched to organisms that have high potential for dispersal at either the gamete, larval, or adult stage, making it impossible to determine whether the source of that DNA was adults with well-documented spatial patterns (e.g. sessile nearshore specialists) or highly mobile early life history stages. Of the 6 OTUs for which dispersal is limited during all life history stages, only 2 occurred in more than two samples, precluding a quantitative comparison of spatial dispersion based on life history characteristics. These were assigned to \textit{Cymatogaster aggregata}, a viviparous nearshore fish with internal fertilization, and \textit{Cupolaconcha meroclista}, a sessile Vermetid gastropod with presumed internal fertilization and short larval dispersal \citep{Strathmann2006, Phillips2010, Calvo2004}. \textit{Cymatogaster aggregata} was distinctly more abundant close to shore, with no sequences occurring in any sample beyond 250 meters (Figure \ref{otu_in_space_select}). \textit{Cupolaconcha meroclista} showed no such distinct spatial trend, occurring in nearly equal abundance at three sites, 75, 500, and 2000 meters from shore. An additional species that was highly abundant in the sequence data, the krill \textit{Thysanoessa raschii}, has pelagic adults, highly seasonal reproduction, and sinking eggs; their distribution was consistent with our expectations based on a tendency of adults to aggregate offshore. Finally, the two most abundant taxa in the dataset were the mussel genus \textit{Mytilus} and the Barnacle order Sessilia; the adults of both taxa are sessile and occur exclusively on hard intertidal substrata but have highly motile larvae.

%----------------------------------DISCUSSION-----------------------------------
\section*{Discussion}
Indirect surveys of organismal presence are a key development in ecosystem monitoring in the face of increased anthropogenic pressure and dwindling resources for ecological research. Monitoring of organisms using environmental DNA is an especially promising method, given the rapid pace of advancement in technological innovation and cost efficiency in the field of DNA sequencing and quantification. For the first time in a marine environment, we document four key patterns: (1) communities far from one another tend to be less similar than those that are nearby, (2) distinct eDNA communities exist and are distributed in a non-random fashion, (3) diversity declines with distance from shore, and (4) spatial patterning of eDNA is associated with taxon-specific life history characteristics.

\subsubsection*{(1) Communities far from one another tend to be less similar than those that are nearby}
We demonstrate that more distant locations have less similar eDNA communities than more proximate locations in Puget Sound, a dynamic marine environment. Our finding is in line with observations based on traditional surveys of terrestrial plants and fungi \citep{Nekola1999, Bahram2013, Condit2002, Chust2006} and of microorganisms in freshwater \citep{Wetzel2012}, marine \citep{Chust2013}, and estuarine \citep{Martiny2011} environments. To our knowledge, it is the first to report such a pattern using massively parallel sequencing of environmental DNA in the marine environment, and the first using any technique to describe this pattern from macrobial metazoans. We note that the theoretical expectation is that samples at very close distance be nearly completely similar, while our samples separated by the 50 meters were only 40\% similar. We interpret this to reflect the highly dynamic nature of this environment, which could cause DNA to be distributed quickly from its source, eroding the rise in similarity at small distances. At the same time, community similarity decreased to very low levels at larger scales, indicating that DNA distribution is not completely unpredictable. This finding implies that the effectively sampled area of individual water samples for eDNA analysis is likely to be quite small (<100m) in this nearshore environment. Our estimated distance-decay relationship does indicate that proximate samples are more similar than distant samples, but we suggest this pattern is partially obscured by other factors, including signal from mobile, microscopic life-stages.

\subsubsection*{(2) Distinct eDNA communities exist and are distributed in a non-random fashion}
 We demonstrate strong evidence for distinct community types and the non-random spatial patterning of those communities. While the spatial distributions of communities is surprising if one were concerned only with the macroscopic life stages of metazoans, it indeed does align with the broader view that even offshore pelagic communities are comprised of and influenced by nearshore organisms. While there was no distinct break in community types between onshore and offshore sites, there was some clustering of community types that may be explained by oceanographic features such as nearshore eddies generated by strong tidal exchange in a steep bathymetric setting \citep{Yang2010}. It would be useful to better understand such features during the period of sampling, by way of oceanographic monitoring devices.

 \subsubsection*{(3) Richness declines and evenness increases with distance from shore}
We detected a general pattern of declining richness and increasing evenness with increasing distance offshore. Such a pattern is consistent with many other ecosystems which show strong clines in diversity metrics over environmental gradients. The coastal ocean is a highly productive and diverse ecosystem \citep{Ray1988}. However, our study is novel in that it corroborates a cline well-known on macroscales for macrobiota on a much smaller spatial scale for microscopic animals, suggesting that there may be a self-similarity across scales in diversity patterning \citep{Levin1992}. Intriguingly, the cline in diversity from inshore to offshore was not determined by shared changes in communities as one moved offshore; the classification analysis suggested a fair amount of differences among communities at a given offshore distance (Figure \ref{pam_in_space}). Furthermore, the uncertainty in identification of the number of distinct clusters to best characterize the community underlines the difficulty of identifying community patterns with the number of taxonomic groups considered here. We suspect that the signature of eDNA from microscopic life-stages may explain our inability to easily detect spatial community level patterns that align with our initial expectations.

\subsubsection*{(4) Spatial patterning of eDNA is associated with taxon-specific life history characteristics.}
In contrast to our expectations, other taxa including species with sessile adult stages restricted to benthic hard substrates (e.g. barnacles, mussels) are among the most abundant taxa at sites furthest from shore. However, the larvae and gametes of these taxa are abundant, pelagic, and can be transported long distances by water movement \citep{Strathmann1987}. This indicates that we likely detected DNA of their pelagic phase gametes and larvae. It is always possible that DNA of adults was advected over long distances and detected offshore but in light of our results with krill and surfperch, we view this as unlikely. We interpret our results as evidence that the chaotic spatial distribution of eDNA communities (Figure \ref{pam_in_space}) results from our primers' affinity for many species which at some point exist as microscopic pelagic gametes or larvae. Our results emphasize that expected results based on easily visually observed individuals or detectable with traditional sampling gear such as nets may be very different from results using eDNA. This does caution that eDNA surveys may have different purposes and may not be directly comparable to existing surveys \citep{Shelton2016}.

We acknowledge that sampling artifacts may have affected our results. For example if entire multicellular individuals were captured in our samples, their DNA could be in much greater density than eDNA, affecting the observed community. Our sampling bottles excluded particles larger than 500 micrometers, but gametes and very small larvae could have gained entry. It is possible that even a single small individual, containing many thousand mitochondria, would overwhelm the signal of another species from which hundreds of cells had been sloughed from many, larger individuals. Data on larval size distribution at the time of sampling from each species in our data set would allow us to estimate the frequency of such events. Nevertheless, it is precisely the sensitivity to small particles that makes the eDNA approach powerful, so we are reluctant to recommend that aquatic eDNA sampling use finer pre-filtering. Instead, we emphasize the importance of designing and selecting primer sets that selectively amplify target organisms. In the case of the present study, in order to recover patterns matching our expectations, this would be non-transient, benthic marine organisms lacking any pelagic life stage.

Our results also highlight the need for curated life-history databases. As technological advances increase the speed and throughput of DNA sequencing and sequence processing, making sense of these data in a timely manner requires that natural history data be stored in standard formats in centralized repositories. The rate at which we can make sense of high-throughput survey methods will be limited by our ability to collate auxiliary data. Databases such as Global Biodiversity Information Facility (GBIF), Encyclopedia of Life (EOL), and FishBase \citep{eol,  fishbase} contain records of taxonomy, occurrence, and other rudimentary data types, but there is no centralized, standardized repository for even basic natural history data such as body size. As NCBI's nucleotide and protein sequence database (GenBank) has facilitated a multitude of transformative studies in diverse fields, an ecological analog would be a huge boon for biodiversity science.

Surveys based on eDNA are intensely scrutinized because of the danger that the final data are subject to complicated laboratory and bioinformatic procedures. Finding virtually no variability among lab and bioinformatic treatments from the point of PCR onward, we were confident our results represented actual field-based differences among samples. However, we note that one PCR replicate had a clear signal of contamination in that the sequence community was extremely similar to those from a different environmental sample. The source of this error is difficult to identify, but seems most likely to be an error during PCR preparation, either in assignment or pipetting during preparation of indexed primers. While the remainder of our results would be largely unchanged had we sequenced a single replicate per environmental sample, we believe the sequencing of PCR replicates is critical for ensuring data quality in eDNA sequencing studies.

While there is much work to be done before eDNA techniques can be confidently deployed as a standard method for ecological monitoring, this study serves as a first analysis of diversity at the fine spatial scales that are likely to be relevant to eDNA work in the field across a range of study systems.

%-------------------------------ACKNOWLEDGEMENTS--------------------------------
\section*{Acknowledgements}
We wish to thank Robert Morris, E. Virginia Armbrust, and James Kralj.

%-----------------------------------FUNDING-------------------------------------
\section*{Funding}
This work was supported by a grant from the David and Lucile Packard Foundation to RPK (grant 2014-39827). The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.

%--------------------------------CONTRIBUTIONS----------------------------------
\section*{Author Contributions}
Conceived and designed the experiments:
JL O'Donnell, RP Kelly, AO Shelton;
Collected the data:
JL O'Donnell, NC Lowell, GD Williams, RP Kelly, AO Shelton, JF Samhouri;
Conducted the analyses:
JL O'Donnell;
Wrote the first draft:
JL O'Donnell;
Edited the manuscript:
JL O'Donnell, AO Shelton, RP Kelly, JF Samhouri, GD Williams, NC Lowell

%-----------------------------------ETHICS--------------------------------------
\section*{Ethics Statement}
The authors declare no conflict of interest. No permits were required to do any of the research described here.

%-------------------------------------DATA-------------------------------------%
\section*{Data Availablity}
\label{data}

\subsection{Sequence Data}
\label{sequence_data}
All sequence files and metadata are available from EMBL:\\ \verb!http://www.ebi.ac.uk/ena/data/view/TODO! \\
(TODO SUBMIT DATA AND METADATA; ADD URL)

\subsection{Project Repository}
The following components are available from the project repository on GitHub: \\
\verb!https://github.com/jimmyodonnell/Carkeek_eDNA_grid!\\
(TODO INSERT DOI FROM FINAL COMMIT BEFORE PUBLICATION)

\subsubsection{Sequencing Metadata}
\label{sequencing_metadata}
Sequencing metadata is available in: \verb!Data/metadata_spatial.csv!

\subsubsection{Life History Data}
\label{life_history_data}
Life history data is available in: \verb!Data/life_history.csv!

\subsubsection{Analysis Scripts}
\label{analysis_scripts}
All analyses were performed using scripts available in the Analysis subdirectory.

%----------------------------------REFERENCES-----------------------------------
%\section*{References} % commented out because the section title is automatically inserted if using an automatically-generated bibliography

\bibliographystyle{apalike} % or: plain,unsrt,alpha,abbrv,acm,apalike,ieeetr
\bibliography{carkeek_grid} % path to your .bib file excluding .bib extension

%-----------------------------------FIGURES-------------------------------------
\pagebreak
\section*{Figures}

\begin{figure}[H] % [h!] forces the figure to be placed roughly here
  \centering
    \includegraphics[width=1\textwidth]{site_map.pdf}
    \caption{\protect\input{"../../Figures/site_map_legend.txt"}}
  \label{site_map}
\end{figure}

\begin{figure}[H] % [h!] forces the figure to be placed roughly here
  \centering
    \includegraphics[width=1\textwidth]{distance_decay.pdf}
    \caption{\protect\input{"../../Figures/distance_decay_legend.txt"}}
  \label{distance_decay}
\end{figure}

\begin{figure}[H] % [h!] forces the figure to be placed roughly here
  \centering
    \includegraphics[width=1\textwidth]{pam_in_space.pdf}
    \caption{\protect\input{"../../Figures/pam_in_space_legend.txt"}}
  \label{pam_in_space}
\end{figure}

\begin{figure}[H] % [h!] forces the figure to be placed roughly here
  \centering
    \includegraphics[width=1\textwidth]{pam_sil.pdf}
    \caption{\protect\input{"../../Figures/pam_sil_legend.txt"}}
  \label{pam_sil}
\end{figure}

\begin{figure}[H] % [h!] forces the figure to be placed roughly here
  \centering
    \includegraphics[width=1\textwidth]{diversity.pdf}
    \caption{\protect\input{"../../Figures/diversity_legend.txt"}}
  \label{diversity}
\end{figure}

\begin{figure}[H] % [h!] forces the figure to be placed roughly here
  \centering
    \includegraphics[width=1\textwidth]{diversity_distance.pdf}
    \caption{\protect\input{"../../Figures/diversity_distance_legend.txt"}}
  \label{diversity_distance}
\end{figure}
\pagebreak

\begin{figure}[H] % [h!] forces the figure to be placed roughly here
  \centering
    \includegraphics[width=1\textwidth]{otu_in_space_select.pdf}
    \caption{\protect\input{"../../Figures/otu_in_space_select_legend.txt"}}
  \label{otu_in_space_select}
\end{figure}

%----------------------------------SUPPLEMENT-----------------------------------
\pagebreak
\section*{Supplemental Material}
\label{supplement}

\subsection*{Methods}
\subsubsection*{Bioinformatics}
Reads passing the preliminary Illumina quality filter were demultiplexed on the basis of the adapter index sequence by the sequencing facility. We used fastqc to assess the fastq files output from the sequencer for low-quality indications of a problematic run. Forward and reverse reads were merged using PEAR v0.9.6 \cite{Zhang2014} and discarded if more than 0.01 of the bases were uncalled. If a read contained two consecutive base calls with quality scores less than 15 (i.e. probability of incorrect base call = 0.0316), these bases and all subsequent bases were removed from the read. Paired reads for which the probability of matching by chance alone exceeded 0.01 were not assembled and omitted from the analysis. Assembled reads were discarded if assembled sequences were not between 50 and 168 bp long, or if reads did not overlap by at least 100 bp.

We used vsearch v2.1.1 \citep{vsearch} to discard any merged reads for which the sum of the per-base error probabilities was greater than 0.5 (``expected errors'') \cite{Edgar2010}. Sequences were demultiplexed on the basis of the primer index sequence at base positions 4-9 at both ends using the programming language AWK. Primer sequences were removed using cutadapt v1.7.1 \cite{Martin2011}, allowing for 2 mismatches in the primer sequence. Identical duplicate sequences were identified, counted, and removed in python to speed up subsequent steps by eliminating redundancy, and sequences occurring only once were removed. We checked for and removed any sequence likely to be a PCR artifact due to incomplete extension and subsequent mis-priming using a method described by \citet{Edgar2010} and implemented in vsearch v2.0.2. Sequences were clustered into operational taxonomic units (OTUs) using the single-linkage clustering method implemented by swarm version 2.1.1 with a local clustering threshold (d) of 1 and fastidious processing \citep{swarm}.

Cross-contamination of environmental, DNA, or PCR samples can result in erroneous inference about the presence of a given DNA sequence in a sample. However, other processes can contribute to the same signature of contamination. For example, errors during oligonucleotide synthesis or sequencing of the indexes could cause reads to be erroneously assigned to samples. The frequency of such errors can be estimated by counting the occurrence of sequences known to be absent from a given sample, and of reads that do not contain primer index sequences in the expected position or combinations. These occurrences indicate an error in the preparation or sequencing procedures. We estimated a rate of incorrect sample assignment by calculating the maximum rate of occurrence of index sequences combinations we did not actually use, as well as the rates of cross-library contamination by counting occurrences of primer sequences from 12S amplicons prepared in a lab more than 1000 kilometers away, but pooled and sequenced alongside our samples. This represents a general minimum rate at which we can expect that sequences from one environmental sample could be erroneously assigned to another, and so we considered for further analysis only those reads occurring with greater frequency than this across the entire dataset.

We checked for experimental error by evaluating the Bray-Curtis similarity (1 - Bray-Curtis dissimilarity) among replicate PCRs from the same DNA sample. We calculated the mean and standard deviation across the dataset, and excluded any PCR replicates for which the similarity between itself and the other replicates was less than 1.5 standard deviations from the mean.

To account for variation in the number of sequencing reads (sequencing depth) recovered per sample, we rarefied the within-sample abundance of each OTU by the minimum sequencing depth \citep{vegan}.

Because each step in this workflow is sensitive to contamination, it is possible that some sequences are not truly derived from the environmental sample, and instead represent contamination during field sampling, filtration, DNA extraction, PCR, fragment size selection, quantitation, sequencing adapter ligation, or the sequencing process itself. We take the view that contaminants are unlikely to manifest as sequences in the final dataset in consistent abundance across replicates; indeed, our data show that the process from PCR onward is remarkably consistent. Thus, after scaling to correct for sequencing depth variation, we calculated from our data the maximum number of sequence counts for which there is turnover in presence-absence among PCR replicates within an environmental sample. We use this number to determine a conservative minimum threshold above which we can be confident that counts are consistent among replicates and not of spurious origin, and exclude from further analysis observations where the mean abundance across PCR replicates within samples does not reach this threshold. For further analyses we use the mean abundance across PCR replicates for each of the 24 environmental samples.

In order to determine the most likely taxon from which each sequence originated, the representative sequence from each OTU was then queried against the NCBI nucleotide collection (GenBank; version October 7, 2015; 32,827,936 sequences) using the blastn command line utility \citep{Camacho2009}. In order to maximize the accuracy of this computationally intensive step, we implemented a nested approach whereby each sequence was first queried using strict parameters (e-value = 5e-52), and if no match was found, the query was repeated with decreasingly strict e-values (5e-48 5e-44 5e-40 5e-36 5e-33 5e-29 5e-25 5e-21 5e-17 5e-13). Other parameters were unchanged among repetitions (word size: 7; maximum matches: 1000; culling limit: 100; minimum percent identity: 0). Each query sequence can be an equally good match to multiple taxa either because of invariability among taxa or errors in the database (e.g. human sequences are commonly attributed to other organisms when they in fact represent lab contamination). In order to guard against these spurious results, we used an algorithm to find the lowest common taxon for at least 80\% of the matched taxa, implemented in the R package taxize 0.7.8 \citep{Chamberlain2013, Chamberlain2016}. Similarly, we repeated analyses using the dataset consolidated at the same taxonomic rank across all queries, for the rank of both family and order.

\subsubsection*{Alternative distance decay model formulations}

\paragraph{Linear:} We fit a straight line through the points after log-transforming the spatial distances to estimate the intercept and slope. This model ignores the bounds of our response variable of community similarity.

\paragraph{Michaelis-Menten:} We fit a Michaelis-Menten-like curve to our data.
Our formulation can be thought of as a modification of the Michaelis-Menten equation, but with the addition of a parameter in the numerator which modifies the intercept.

\begin{equation}\label{MM_full}
	y = \frac{AB + Cx}{B+x}
\end{equation}

Where $C$ is the asymptote of minimum similarity. This formulation allows us to estimate the maximum similarity in the system, and the rate at which it is achieved. If the value of the parameter ($AB$) is 0 (e.g. if the intercept is 0), the form is identical to the Michaelis-Menten equation:

\begin{equation}\label{MichaelisMenten}
	y = \frac{Cx}{B + x}
\end{equation}

This is conceptually satisfying in that a fit through [0,1] reflects the theoretical expectation that samples at zero distance from one another are necessarily identical. Given an efficient sampling technique, replicate samples taken at the same position in space should be identical, and thus the intercept of the regression of similarity against distance should be 1, and deviation from 1 is an indicator of the efficiency of the sampling method. % NOTE: We did not take multiple environmental samples from the same position in space, but conducted multiple PCRs from each environmental sample.

Finally, we considered a model which estimates an asymptote as the total change in similarity ($D$):

\begin{equation}\label{Harold}
	y = \frac{A + Dx}{B + x}
\end{equation}

However, this model failed to converge and produced uninformative estimates of all parameters.

\subsection*{Supplemental Figures}

\begin{figure}[H] % [h!] forces the figure to be placed roughly here
  \centering
    \includegraphics[height=0.6\textheight]{distance_decay_multi.pdf}
    \caption{\protect\input{"../../Figures/distance_decay_multi_legend.txt"}}
  \label{distance_decay_multi}
\end{figure}

\begin{figure}[H] % [h!] forces the figure to be placed roughly here
  \centering
    \includegraphics[width=1\textwidth]{diversity_distance_all.pdf}
    \caption{\protect\input{"../../Figures/diversity_distance_all_legend.txt"}}
  \label{diversity_distance_all}
\end{figure}

% \pagebreak
% \section*{Orphaned Text}
% Samples collected of ecological communities may vary in dissimilarity from 0 (completely identical) to 1 (completely different). For samples collected from multiple locations, the relationship between their spatial distance and community dissimilarity is of interest because it reflects the amount of community heterogeneity---changes in abundance and composition---over the spatial scale sampled.
% In general, the relationship between community dissimilarity (0 = identical; 1 = completely different) and spatial distance is expected to be asymptotic, because communities nearer to each other tend to be more similar than those farther apart. The intercept is expected to be 0, because only within-sample comparisons can have 0 spatial separation, and communities have no dissimilarity within a sample if sampling method is repeatable.
% (technically impossible for samples to be taken at exactly 0; they are actually immediately adjacent.)
% Deviation from 0 indicates heterogeneous community composition/structure over fine scales. Likewise, dissimilarity cannot exceed 1, and reaches 1 only when multiple discrete community types are sampled. The trend is expected to be asymptotic if communities within a habitat are spatially structured, where the value of the asymptote and the rate of increase provide insight about community turnover. A flat relationship indicates that heterogeneity is not assorted spatially, and can be interpreted in different ways, depending on the asymptote. % ?or no relationship both
% If the asymptote is close to 1, there is high spatial heterogeneity over the spatial scale of sampling. If the asymptote is close 0, all samples are similar, and we infer there is complete community homogeneity over the scale sampled. The rate at which community dissimilarity approaches the mean gives an indication of the rate of community turnover.
%
% In theory, given enough data from manipulative experiments, it is possible that one could measure environmental variables--temperature, UV intensity, flow, etc--at the time of eDNA sample collection in order to more precisely estimate the origin of eDNA. In practice, this approach is unlikely to yield appreciable gains in accuracy because even the most advanced sensors (ADCPs) and analytical techniques available cannot determine the origin of particles in highly dynamic fluid environments at scales relevant to most ecological questions \citep{Piterbarg2001}. %http://www.rsmas.miami.edu/LAPCOD/research/prediction/
% Further, collecting and analyzing these data would severely limit the scalability of eDNA methods. A more practical approach is to compare the spatial distribution of communities of eDNA with expectations based on prior knowledge of the spatial distribution of communities of organisms.

\end{document}

### 2016-11-16 Manuscript/v1.0_PeerJ/carkeek_grid.tex
comment out TODOs, fix much less than symbol

@@
+%-----------------------------------NOTES---------------------------------------
+% (TODO SUBMIT DATA AND METADATA; ADD URL)
+% (TODO INSERT DOI FROM FINAL COMMIT BEFORE PUBLICATION)
+
+% Target journal: PeerJ. Backups: Frontiers in...marine science?, BMC Ecology, BMC Genetics
+
 %--------------------------------------||--------------------------------------%
@@
 \def\linenumberfont{\normalfont\small\rmfamily} % change line number font
-
-%-----------------------------------NOTES---------------------------------------
-
-% Target journal: PeerJ. Backups: Frontiers in...marine science?, BMC Ecology, BMC Genetics
 
@@
 figures/diversity.pdf).
-1-Simpson's diversity, the probability that two reads chosen at random from a sample belong to different species, increased by .0666 per 1000 meters from 0.225 (95\%CI = 0.0558) to 0.491 (95\%CI = $\pm$ 0.112), indicating that sequence reads were less evenly distributed among taxa in offshore samples (p <<< 0.05; Figure \ref{diversity_distance}).
+1-Simpson's diversity, the probability that two reads chosen at random from a sample belong to different species, increased by .0666 per 1000 meters from 0.225 (95\%CI = 0.0558) to 0.491 (95\%CI = $\pm$ 0.112), indicating that sequence reads were less evenly distributed among taxa in offshore samples (p $\ll$ 0.05; Figure \ref{diversity_distance}).
 There was no evidence for spatial autocorrelation for any of the diversity metrics (Moran's I, p > 0.05; Figure \ref{diversity}).
@@
 \label{sequence_data}
-All sequence files and metadata are available from EMBL:\\ \verb!http://www.ebi.ac.uk/ena/data/view/TODO! \\
-(TODO SUBMIT DATA AND METADATA; ADD URL)
+All sequence files and metadata are available from EMBL:\\ \verb!http://www.ebi.ac.uk/ena/data/view/FIXME! \\
 
@@
 \verb!https://github.com/jimmyodonnell/Carkeek_eDNA_grid!\\
-(TODO INSERT DOI FROM FINAL COMMIT BEFORE PUBLICATION)
+\verb!http://dx.doi.org/FIXME!\\
 

### 2016-11-18 Analysis/diversity.R
refer to Simpsons index as evenness

@@ -7,7 +7,7 @@ EXPORT <- TRUE
 div_metrics <- list()
 
 # div_metrics[["Shannon"]] <- lapply(otu_table, diversity, index = "shannon") 
-div_metrics[["Simpson"]] <- lapply(otu_table, function(x){
+div_metrics[["Evenness"]] <- lapply(otu_table, function(x){
   1 - diversity(x, index = "simpson")})
 div_metrics[["Richness"]] <- lapply(otu_table, function(x) {rowSums(x > 0)})
 

### 2016-11-18 Manuscript/v1.0_PeerJ/carkeek_grid.tex
incorporate Ole's edits and some of Jameal's

@@
 
+% \pagebreak
 %----------------------------------ABSTRACT-------------------------------------
@@
 \subsubsection*{Aggregate Measures of Diversity}
-We calculated two measures of diversity (richness and 1-Simpson's Index) to ask if aggregate metrics of the eDNA community showed evidence of spatial patterning.
-Richness is a measure of the number of distinct types of organisms present and so ranges from 1 (only one taxon observed) to R, the number of taxa observed across all samples.
-1-Simpson's is a measure of the evenness of the distribution of abundance of taxa in a sample and ranges from 0 to 1, with the value interpreted as the probability that two sequences randomly selected from the sample will belong to different taxa.
-Thus larger values of the index indicate more evenly divided communities.
+We calculated two measures of diversity, richness and evenness, to ask if aggregate metrics of the eDNA community showed evidence of spatial patterning.
+Richness is a measure of the number of distinct types of organisms present and so ranges from 1 (only one taxon observed) to $S$, the number of taxa observed across all samples.
+To calculate the evenness of the distribution of abundance of taxa in a sample, we used the complement of the \citet{Simpson1949} index ($1 - \Sigma{p_{i}^2}$, where $p_i$ is the proportional abundance of taxon $i$).
+The values of this index ranges from 0 to 1, with the value interpreted as the probability that two sequences randomly selected from the sample will belong to different taxa; thus, larger values of the index indicate more evenly divided communities \citep{Magurran2004}.
 We calculated Moran's I for both diversity metrics to test for spatial autocorrelation.
@@
 \subsubsection*{Diversity in Space}
-Sites offshore tend to be less diverse (richness) and more even (1-Simpson's D) than those inshore (Figure \ref{diversity_distance}).
-Mean OTU richness declined by 1.42 per 1000 meters from a mean of 17.6 taxa (95\%CI = 2.15) inshore to 11.9 taxa (95\%CI = 4.31) at offshore locations (p = 0.0415; Fig.
-figures/diversity.pdf).
-1-Simpson's diversity, the probability that two reads chosen at random from a sample belong to different species, increased by .0666 per 1000 meters from 0.225 (95\%CI = 0.0558) to 0.491 (95\%CI = $\pm$ 0.112), indicating that sequence reads were less evenly distributed among taxa in offshore samples (p $\ll$ 0.05; Figure \ref{diversity_distance}).
+Sites offshore tend to be less rich and more even than those inshore (Figure \ref{diversity_distance}).
+Mean OTU richness declined by 1.42 per 1000 meters from a mean of 17.6 taxa (95\%CI = 2.15) inshore to 11.9 taxa (95\%CI = 4.31) at offshore locations (p = 0.0415; Figure \ref{diversity_distance}).
+Evenness, the probability that two reads chosen at random from a sample belong to different species, increased by .0666 per 1000 meters from 0.225 (95\%CI = 0.0558) to 0.491 (95\%CI = $\pm$ 0.112), indicating that sequence reads were less evenly distributed among taxa in offshore samples (p $\ll$ 0.05; Figure \ref{diversity_distance}).
 There was no evidence for spatial autocorrelation for any of the diversity metrics (Moran's I, p > 0.05; Figure \ref{diversity}).
@@
 Monitoring of organisms using environmental DNA is an especially promising method, given the rapid pace of advancement in technological innovation and cost efficiency in the field of DNA sequencing and quantification.
-For the first time in a marine environment, we document four key patterns: (1) communities far from one another tend to be less similar than those that are nearby, (2) distinct eDNA communities exist and are distributed in a non-random fashion, (3) diversity declines with distance from shore, and (4) spatial patterning of eDNA is associated with taxon-specific life history characteristics.
+For the first time in a marine environment, we document four key patterns: (1) eDNA communities far from one another tend to be less similar than those that are nearby, (2) distinct eDNA communities exist and are distributed in a non-random fashion, (3) diversity declines with distance from shore, and (4) spatial patterning of eDNA is associated with taxon-specific life history characteristics.
 
@@
 While the spatial distributions of communities is surprising if one were concerned only with the macroscopic life stages of metazoans, it indeed does align with the broader view that even offshore pelagic communities are comprised of and influenced by nearshore organisms.
+This result underscores the idea that areas immediately offshore act as ecotones, a mixing zone of taxa characteristic of benthic and pelagic environments.
 While there was no distinct break in community types between onshore and offshore sites, there was some clustering of community types that may be explained by oceanographic features such as nearshore eddies generated by strong tidal exchange in a steep bathymetric setting \citep{Yang2010}.

### 2016-11-18 Manuscript/v1.0_PeerJ/carkeek_grid.tex
finish Jameal's edits

@@
 
-% Make sure authors specify/clarify department, institution, and address.
+% Make sure authors specify department, institution, and address.
 \affil[1]{School of Marine and Environmental Affairs, University of Washington, 3707 Brooklyn Ave NE, Seattle, Washington 98105, USA}
@@
 
-% \pagebreak
+\pagebreak
 %----------------------------------ABSTRACT-------------------------------------
@@
 
-No existing biodiversity survey method completely censuses all of the organisms in a given area.
+The patterns and causes of variability in ecological communities across space are both seminal and contentious areas of study in ecology \citep{Hubbell2001, Anderson2011}.
+One consistently observed pattern of community spatial heterogeneity is that communities close to one another tend to be more similar than those that are farther apart \citep{Nekola1999}.
+This decrease in community similarity with increasing spatial separation is called distance decay and has been reported from communities of tropical trees \citep{Condit2002, Chust2006}, ectomycorrhizal fungi \citep{Bahram2013}, salt marsh plants \citep{Guo2015}, and microorganisms \citep{Martiny2011, Chust2013, Wetzel2012, Bell2010}.
+Typically, this relationship is assessed by regressing a measure of community similarity against a measure of spatial separation for a set of sites at which a set of species' abundances (or presences) is calculated.
+Yet no existing biodiversity survey method completely censuses all of the organisms in a given area.
+The lack of a single `silver bullet' method of sampling contributes inconclusiveness to the study of spatial patterning in ecology \citep{Levin1992}, and leaves open the possibility of new and more comprehensive methods.
+
+From a boat or aircraft, scientists can count whales by sight, but not the krill on which they feed.
 For example, towed fishing nets can efficiently sample organisms larger than the mesh and slower than the boat; but overlook viruses and have undesirable effects on charismatic air-breathing species.
-From a boat or aircraft, scientists can count whales by sight, but not the krill on which they feed.
 However, DNA-based surveys show great promise as an efficient technique for detecting a previously unthinkable breadth of organisms from a single sample.
@@
 
-Species and the communities they comprise are not homogeneously distributed in space; describing and predicting the spatial heterogeneity of ecosystems is of great interest to ecologists.
-There is a large and active literature focused on understanding the patterns and causes of community variability in space \citep{Hubbell2001, Anderson2011}.
-One consistently observed pattern of community spatial heterogeneity is that communities close to one another tend to be more similar than those that are farther apart \citep{Nekola1999}.
-This decrease in community similarity with increasing spatial separation is called distance decay and has been reported from communities of tropical trees \citep{Condit2002, Chust2006}, ectomycorrhizal fungi \citep{Bahram2013}, salt marsh plants \citep{Guo2015}, and microorganisms \citep{Martiny2011, Chust2013, Wetzel2012, Bell2010}.
-Typically, this relationship is assessed by regressing a measure of community similarity against a measure of spatial separation for a set of sites at which a set of species' abundances (or presences) is calculated.
-
-We can apply these methods derived from community ecology to understand spatial patterns and patchiness of eDNA.
-The underlying mechanism thought to drive the slope of this relationship is the rate of movement of individuals among sites, which may be driven by underlying processes such as habitat suitability.
+We apply methods derived from community ecology to understand spatial patterns and patchiness of eDNA.
+The underlying mechanism thought to drive the slope of the distance decay relationship in ecological communities is the rate of movement of individuals among sites, which may be driven by underlying processes such as habitat suitability.
 Because eDNA is shed and transported away from its source, the increased movement of eDNA particles should homogenize community similarity, and thus erode the distance decay relationship of eDNA communities.
@@
 We took advantage of this setting to explore the spatial variation and distribution of marine eDNA communities.
-Using PCR-based methods and massively parallel sequencing, we surveyed 16S mtDNA from a suite of marine animals in water samples collected over a grid of sites extending from the shoreline out to 4 kilometers offshore in Puget Sound, Washington, USA.
+Using PCR-based methods and massively parallel sequencing, we surveyed mitochondrial 16S sequences from a suite of marine animals in water samples collected over a grid of sites extending from the shoreline out to 4 kilometers offshore in Puget Sound, Washington, USA.
 We leverage this sampling design to perform the first explicitly spatial analysis of eDNA-derived community similarity.

### 2016-11-22 Manuscript/v1.0_PeerJ/carkeek_grid.tex
final formatting of Ole and Jameal's edits
fix typo
incorporate Linda Park's feedback

@@
 % Target journal: PeerJ. Backups: Frontiers in...marine science?, BMC Ecology, BMC Genetics
+%----------------------------------REVIEWERS------------------------------------
+% A list of recommended reviewers.
+% Frédéric Mahé mahefrederic@gmail.com
+% Sophie Arnaud-Haond, sophie.arnaud@ifremer.fr
+% Dominique Cowart dcowart@illinois.edu dominique.cowart@gmail.com (Metabarcoding is powerful yet still blind)
+% Ann Bucklin, ann.bucklin@uconn.edu (Metabarcoding of marine zooplankton)
+% Naiara Rodríguez-Ezpeleta, nrodriguez@azti.es (Benchmarking DNA Metabarcoding for Biodiversity-Based Monitoring and Assessment)
+% Xavier Turon, xturon@ceab.csic.es (Metabarcoding Extracellular DNA from Sediments of Marine Canyons)
+% Andrew Mahon, mahon2a@cmich.edu
+% Eske Willerslev ewillerslev@snm.ku.dk
+% Philip Thomsen pfthomsen@snm.ku.dk
 
@@
 \maketitle
-\linenumbers % start line numbers
-\def\linenumberfont{\normalfont\small\rmfamily} % change line number font
-
-%----------------------------------REVIEWERS------------------------------------
-% A list of recommended reviewers.
-% Frédéric Mahé mahefrederic@gmail.com
-
-% Sophie Arnaud-Haond, sophie.arnaud@ifremer.fr
-% Dominique Cowart dcowart@illinois.edu dominique.cowart@gmail.com (Metabarcoding is powerful yet still blind)
-
-% Ann Bucklin, ann.bucklin@uconn.edu (Metabarcoding of marine zooplankton)
-
-% Naiara Rodríguez-Ezpeleta, nrodriguez@azti.es (Benchmarking DNA Metabarcoding for Biodiversity-Based Monitoring and Assessment)
-
-% Xavier Turon, xturon@ceab.csic.es (Metabarcoding Extracellular DNA from Sediments of Marine Canyons)
-
-% Andrew Mahon, mahon2a@cmich.edu
-
-% Eske Willerslev ewillerslev@snm.ku.dk
-
-% Philip Thomsen pfthomsen@snm.ku.dk
 
@@
 \pagebreak
+\linenumbers % start line numbers
+\def\linenumberfont{\normalfont\small\rmfamily} % change line number font
 %----------------------------------ABSTRACT-------------------------------------
@@
 From a boat or aircraft, scientists can count whales by sight, but not the krill on which they feed.
-For example, towed fishing nets can efficiently sample organisms larger than the mesh and slower than the boat; but overlook viruses and have undesirable effects on charismatic air-breathing species.
+For example, towed fishing nets can efficiently sample organisms larger than the mesh and slower than the boat, but overlook viruses and have undesirable effects on charismatic air-breathing species.
 However, DNA-based surveys show great promise as an efficient technique for detecting a previously unthinkable breadth of organisms from a single sample.
@@
 Microbiologists have used nucleic acid sequencing to quantify the composition and function of microbial communities in a wide variety of habitats \citep{Handelsman1998, Tyson2004, Venter2004, Iverson2012}.
-To do so, a sample of environmental medium (e.g.
-water) containing microorganisms is collected, their DNA or RNA is isolated and sequenced, and the identity and abundance of sequences is considered to reflect the community of  organisms contained in the sample, which indirectly estimates the quantity of organisms in an area.
+To do so, microorganisms are collected in a sample of environmental medium (e.g.
+water), their DNA or RNA is isolated and sequenced, and the identity and abundance of sequences is considered to reflect the community of  organisms contained in the sample, which indirectly estimates the quantity of organisms in an area.
 
@@
 The relationship between local organismal abundance and eDNA is further complicated in habitats where the environmental medium itself may transport eDNA away from its source.
-We know that genetic material moves away from its source precisely because organisms can be detected indirectly without being present in the sample \citep{Kelly2016}.
+We know that genetic material can move away from its source precisely because organisms can be detected indirectly without being present in the sample \citep{Kelly2016}.
 One might reasonably expect eDNA to travel farther in a highly dynamic fluid such as the open ocean or flowing river than it would through the sediment at the bottom of a stagnant pond \citep{Deiner2014, Shogren2016}.
@@
 Starting from lower-intertidal patches of \textit{Zostera marina}, we collected water samples at 1 meter depth from 8 points (0, 75, 125, 250, 500, 1000, 2000, and 4000 meters) along three parallel transects separated by 1000 meters (24 sample locations total; Figure \ref{site_map}).
+Samples were collected by attaching bottles to a PVC pole and lowering it over the side of a boat over the span of one hour on 27 June 2014.
 To destroy residual DNA on equipment used for field sampling and filtration, we washed with a 1:10 solution of household bleach (8.25\% sodium hypochlorite; 7.25\% available chlorine) and deionized water, followed by thorough rinsing with deionized water.
@@
 DNA was resuspended in 200 microliters of ultrapure water.
-Genomic DNA extracted from tissue of a species absent from the sampled environment (\textit{Oreochromis niloticus}) served as positive control for the remaining protocols.
+Four replicates of genomic DNA extracted from tissue of a species absent from the sampled environment (\textit{Oreochromis niloticus}) served as positive control for the remaining protocols.
 
@@
 \subsection*{DNA Sequencing}
-PCR products were pooled according to their primer index in equal concentration, and 150 ng was prepared for library sequencing using the KAPA high-throughput library prep kit with real-time library amplification protocol (KAPA Biosystems, Wilmington, MA, USA).
-Each of these ligated sequencing adapter including an additional 6 base pair index sequence (NEXTflex DNA barcodes; BIOO Scientific, Austin, TX, USA).
+Up to 30 PCR products were combined according to their primer index in equal concentration into one of 14 pools, and 150 nanograms from each were prepared for library sequencing using the KAPA high-throughput library prep kit with real-time library amplification protocol (KAPA Biosystems, Wilmington, MA, USA).
+Each of these ligated sequencing adapters included an additional 6 base pair index sequence (NEXTflex DNA barcodes; BIOO Scientific, Austin, TX, USA).
 Thus, each PCR product was identifiable via its unique combination of index sequences in the sequencing adapters and primers.
@@
 Databases such as Global Biodiversity Information Facility (GBIF), Encyclopedia of Life (EOL), and FishBase \citep{eol,  fishbase} contain records of taxonomy, occurrence, and other rudimentary data types, but there is no centralized, standardized repository for even basic natural history data such as body size.
-As NCBI's nucleotide and protein sequence database (GenBank) has facilitated a multitude of transformative studies in diverse fields, an ecological analog would be a huge boon for biodiversity science.
+As NCBI's nucleotide and protein sequence database (GenBank) has facilitated transformative studies in diverse fields, an ecological analog would be a boon for biodiversity science.
 
@@
 \label{sequence_data}
-All sequence files and metadata are available from EMBL:\\ \verb!http://www.ebi.ac.uk/ena/data/view/FIXME! \\
+All sequence files and metadata are available from EMBL:\\ \verb!http://www.ebi.ac.uk/ena/data/view/FIXME!
 
@@
 \verb!https://github.com/jimmyodonnell/Carkeek_eDNA_grid!\\
-\verb!http://dx.doi.org/FIXME!\\
+\verb!http://dx.doi.org/FIXME!
 
@@
 This formulation allows us to estimate the maximum similarity in the system, and the rate at which it is achieved.
-If the value of the parameter ($AB$) is 0 (e.g.
+If the value of the parameter ($AB$) is 0 (i.e.
 if the intercept is 0), the form is identical to the Michaelis-Menten equation:

### 2016-11-23 Analysis/multilevel_model/Ole/eDNA_grid_model.R
cleanup old files

@@ -1,236 +1,0 @@
-#!/usr/bin/env Rscript
-
-# setwd("")
-library(R2jags)
-
-# This is code to estimate some model parameters for proportional abundance of DNA in a sample given reads in the file from a sequencer
-
-
-# Let:
-# i = taxon i (1:I); Carkeek grid dataset I = ? (limit to 10 for tests)
-# j = PCR replicate j (1:J); Carkeek grid dataset J = 4
-# k = location k (1:K); Carkeek grid dataset K = 24
-
-
-# Requires as input:
-# 1. otu file
-# 2. metadata file
-
-# a table of counts of sequences (Z, length = ...)
-counts_table <- read.csv("OTUs_top10_4000m.csv", row.names = 1)
-# rows/rownames = samples, columns/colnames = taxa, cells = integer counts
-# if table is incorrectly oriented, transpose it:
-# counts_table <- as.data.frame(t(counts_table))
-
-# load metadata
-metadata <- read.csv("metadata_spatial.csv", stringsAsFactors = FALSE)
-
-# what is the name of the column containing the sample id (SEQUENCING samples)
-colname_sampleid <- "sample_id"
-
-# what is the name of the column containing PCR replicate levels
-colname_pcr <- "PCR_replicate"
-
-# what is the name of the column containing relevant site replicate levels
-colname_site <- "transect"
-
-# of each taxon (length = I)
-taxa <- colnames(counts_table)
-
-# from each PCR replicate (length = J)
-pcr <- unique(metadata[, colname_pcr])
-
-# at each site (length = K)
-sites <- unique(metadata[, colname_site])
-
-
-
-
-
-
-
-
-# JAGS won't do some basic R functions, so we need to give it the lengths explicitly
-N_taxa <- length(taxa)
-N_pcr <- length(pcr)
-N_sites <- length(sites)
-
-
-# make a vector of the site for each sample name in the counts table
-site_by_rowname <- metadata[ , colname_site][
-  match(rownames(counts_table), metadata[ , colname_sampleid])
-]
-
-# unused for now
-# sampleid_by_site <- split(x = metadata[, colname_sampleid], f = metadata[, colname_site], drop = TRUE)
-
-counts_by_site <- split(x = counts_table, f = site_by_rowname)
-
-
-
-
-
-
-# create an array for JAGS called "counts"
[…]

### 2016-11-23 Analysis/multilevel_model/eDNA_model_grid.R

@@ -1,405 +1,0 @@
-#!/usr/bin/env Rscript
-
-# This is code to estimate some model parameters for proportional abundance of DNA in a sample given reads in the file from a sequencer
-
-library(R2jags)
-library(reshape2) # acast: turn long-form dataframe into array
-
-setwd("~/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Analysis")
-setwd("/Users/jimmy.odonnell/Projects/Carkeek_eDNA_grid/Analysis")
-
-data_dir <- file.path("..", "Data")
-fig_dir <- file.path("..", "Figures")
-
-# Let:
-# i = taxon i (1:I); Carkeek grid dataset I = ? (limit to 10 for tests)
-# j = PCR replicate j (1:J); Carkeek grid dataset J = 4
-# k = position k from shore (1:K); Carkeek grid dataset K = 8
-# l = position l along shore (1:L); L = 3
-
-# Requires as input:
-# 1. otu file
-# 2. metadata file
-
-# a table of counts of sequences (Z, length = ...)
-# rows/rownames = samples, columns/colnames = taxa, cells = integer counts
-counts_file_path <- file.path(data_dir, "OTUs_top10_4000m.csv")
-counts_file_path <- file.path(data_dir, "otu_table_filtered_per_samp.csv")
-counts_table <- read.csv(file = counts_file_path, row.names = 1)
-
-# REMEMBER TO REMOVE CONTROL TAXON IF IT'S STILL THERE!
-counts_table <- counts_table[,! names(counts_table) %in% "DUP_3"]
-
-
-counts_table <- counts_table[,1:10] # use only the top ten for now
-
-# if table is incorrectly oriented, transpose it:
-# counts_table <- as.data.frame(t(counts_table))
-
-
-# load metadata
-metadata_file_path <- file.path(data_dir, "metadata_spatial.csv")
-metadata <- read.csv(file = metadata_file_path, stringsAsFactors = FALSE)
-
-# what is the name of the column containing the sample id (SEQUENCING samples)
-colname_sampleid <- "sample_id"
-
-# what is the name of the column containing PCR replicate levels
-colname_pcr <- "PCR_replicate"
-
-# what is the name of the column containing position in x direction (along shore) replicate levels
-colname_posx <- "transect"
-
[…]

### 2016-11-23 Analysis/multilevel_model/eDNA_model_single.R

@@ -1,235 +1,0 @@
-#!/usr/bin/env Rscript
-
-# This is code to estimate some model parameters for proportional abundance of DNA in a sample given reads in the file from a sequencer
-
-library(R2jags)
-
-setwd("~/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Analysis")
-
-data_dir <- file.path("..", "Data")
-fig_dir <- file.path("..", "Figures")
-
-# Let:
-# i = taxon i (1:I); Carkeek grid dataset I = ? (limit to 10 for tests)
-# j = PCR replicate j (1:J); Carkeek grid dataset J = 4
-# k = location k (1:K); Carkeek grid dataset K = 24
-
-
-# Requires as input:
-# 1. otu file
-# 2. metadata file
-
-# a table of counts of sequences (Z, length = ...)
-counts_file_path <- file.path(data_dir, "OTUs_top10_4000m.csv")
-counts_table <- read.csv(file = counts_file_path, row.names = 1)
-# rows/rownames = samples, columns/colnames = taxa, cells = integer counts
-# if table is incorrectly oriented, transpose it:
-# counts_table <- as.data.frame(t(counts_table))
-
-# load metadata
-metadata_file_path <- file.path(data_dir, "metadata_spatial.csv")
-metadata <- read.csv(file = metadata_file_path, stringsAsFactors = FALSE)
-
-# what is the name of the column containing the sample id (SEQUENCING samples)
-colname_sampleid <- "sample_id"
-
-# what is the name of the column containing PCR replicate levels
-colname_pcr <- "PCR_replicate"
-
-# what is the name of the column containing relevant site replicate levels
-colname_site <- "transect"
-
-# of each taxon (length = I)
-taxa <- colnames(counts_table)
-
-# from each PCR replicate (length = J)
-pcr <- unique(metadata[, colname_pcr])
-
-# at each site (length = K)
-sites <- unique(metadata[, colname_site])
-
-
-
-
-
-
-
-
-# JAGS won't do some basic R functions, so we need to give it the lengths explicitly
-N_taxa <- length(taxa)
-N_pcr <- length(pcr)
-N_sites <- length(sites)
-
-
[…]

### 2016-11-23 Analysis/multilevel_model/general_framework/Analysis/BHM_eDNA.R

@@ -1,196 +1,0 @@
-# Estimation of some underlying/unobserved quantity from some other observed quantity
-# (e.g. relative abundance from eDNA counts)
-
-# Set this to the analysis directory of this project
-analysis_dir <- "~/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Analysis"
-
-
-
-setwd(analysis_dir)
-data_dir <- file.path("..", "Data")
-fig_dir <- file.path("..", "Figures")
-
-
-############################################################################
-# The following requires as input:
-# 1. A matrix of counts (integers) of DNA sequences, where:
-# 		rows = samples
-# 		columns = taxa whose names are given in column names
-# eg: dput(mydata) 
-# structure(c(18094L, 18136L, 21198L, 13643L, 14198L, 12219L, 12703L, 
-# 13873L, 11773L, 12022L, 11865L, 13554L, 13840L, 9871L, 10999L, 
-# 10070L, 10827L, 8920L, 9708L, 10033L, 9303L, 6802L, 7853L, 7927L, 
-# 3118L, 3344L, 4107L), .Dim = c(3L, 9L), .Dimnames = list(NULL, 
-    # c("taxon_A", "taxon_B", "taxon_C", "taxon_D", "taxon_E", 
-    # "taxon_F", "taxon_G", "taxon_H", "taxon_I")))
-
-
-# If loading from a file
-# counts <- as.matrix(read.csv(file = file.path(data_dir, "counts.csv")))
-# taxon_names <- colnames(counts)
-
-# If reading from a currently existing object (e.g. you simulated some eDNA data)
-# counts <- as.matrix(mydata)
-
-# make into a vector
-counts_vec <- as.vector(counts)
-names(counts_vec) <- rep(x = colnames(counts), each = nrow(counts))
-
-
-
-
-############
-# Create Design Matrix for the categorical effects
-############
-
-# Should have nrow = length of the input 
-
-cat_mat <- matrix(data = 0, nrow = length(counts_vec), ncol = ncol(counts))
-
-# Not sure why, but you set the first species to equal 1...
-# I think this might be because everything else is expressed relative to this?
-cat_mat[,1]	<- 1
-
-# Changed Ole's original version because it will break if you don't use the "right" taxon names (i.e. with an integer corresponding to their index in the third position == A.3)
-for(i in 1:length(taxon_names)){
[…]

### 2016-11-23 Analysis/multilevel_model/general_framework/Analysis/original OS - eDNA Pipers v4_JO.r

@@ -1,1801 +1,0 @@
-rm(list=ls())
-### GO GET THE NECESSARY LIBRARIES
-
-##### TILAPIA IS OTU 4 ####
-
-library(gtools)
-library(R2jags)
-
-# Set this to the subdirectory Analysis within the project folder (which should also contain subdirectories called Data and Figures)
-# analysis_dir <- "/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Analysis"
-
-
-# data_dir <- file.path("..", "Data")
-# fig_dir <- file.path("..", "Figures")
-
-setwd("/Users/threeprime/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Analysis/general_framework/Data")
-
-
-
-
-
-
-
-#------------------------------------------
-## Read in the clusters data + sort to make long form
-# this file is a matrix of counts (integers) where
-# rows = taxa
-# columns = samples
-# AND
-# columns = distance measure from taxon's primer binding site and each of the tags (primer indexes)
-# for original modeling paper, dim =  9,151; only top 10 OTUs
-dat		<-	read.csv("abundance_distance_matrix+OLE.csv")
-dat		<-	dat[dat$X!='OTU_4',] # remove Tilapia # no need once tables are separated appropriately
-
-# OTU_table.csv
-# primer_mismatch_both.csv
-# primer_mismatch_F.csv
-# primer_mismatch_R.csv
-#------------------------------------------
-
-
-
-
-
-
-
-
-
-#------------------------------------------
-# Read in the "duplicate table":
-# this file is a matrix of counts (integers)
-# rows = unique DNA sequences ("DUP")
-# columns = samples
-# dim = (27973,75); 75 samples, 27973 unique sequences
-dat.total.DNA	<-	read.csv("all_clusters.csv", header=TRUE, row.names = 1)
-
-# duplicate_table.csv
-#------------------------------------------
-
-# calculate the total number of sequence counts per sample
-dat.DNA			<-	colSums(dat.total.DNA)
-
-
-
-
-
-
-
-#------------------------------------------
-# Read in the sequencing metadata
-# contains primer index data, environmental sample name, etc
-dat.surv	<-	read.csv("sample_data.csv")
-# sequencing_metadata.csv
-#------------------------------------------
-
-
-# isolate the primer index sequence ("tag") and the sample name
[…]

### 2016-11-23 Analysis/multilevel_model/long_form.R

@@ -1,51 +1,0 @@
-# Prep data and manipulate for multilevel modeling
-#-------------------------------------------------------------------------------
-# COMBINE UP DATA
-if(
-  identical(
-    rownames(otu_table_in),
-    metadata[,colname_sampleid]
-  )
-){
-  data_full <- cbind.data.frame(metadata, otu_table_in)
-  if(identical(rownames(data_full), data_full[,colname_sampleid])){
-    rownames(data_full) <- NULL
-  } else {
-    "something went wrong, check thyself before you wreck thyself"
-  }
-} else {
-  warning("something does not match up, and you could screw up the data big time by proceeding.")
-}
-
-# data_full_path <- file.path(data_dir, "data_full_filt.csv") # "data_full_top10.csv"
-# write.csv(x = data_full, file = data_full_path, row.names = FALSE)
-#-------------------------------------------------------------------------------
-
-
-#-------------------------------------------------------------------------------
-# LONG FORMAT:
-library(reshape2) # melt()
-otu_long <- melt(otu_clean)
-colnames(otu_long) <- c("sample_id", "OTU", "count")
-otu_long$OTU <- as.character(otu_long$OTU)
-otu_long$sample_id <- as.character(otu_long$sample_id)
-
-row_match_long <- match(otu_long$sample_id , metadata[, colname_sampleid])
-
-data_full_long <- cbind.data.frame(metadata[row_match_long,], otu_long[,c("OTU", "count")])
-
-data_full_long_path <- file.path(data_dir, "data_full_long.csv")
-# write.csv(x = data_full_long, file = data_full_long_path, row.names = FALSE)
-# I think I may have been crazy when I wrote the next few lines:
-library(reshape2)
-data_for_cast <- data.frame(
-  sample_id = data_full_long$sample_id,
-  OTU = data_full_long$OTU,
-  count = data_full_long$count)
-
-otu_long_mean <- melt(dcast(data = data_for_cast, formula = sample_id ~ OTU, fun.aggregate = mean, na.rm = TRUE))
-colnames(otu_long_mean) <- c("sample_id", "OTU", "count")
-row_match_mean <- match(otu_long_mean$sample_id , metadata[, colname_sampleid])
-otu_long_mean_full <- cbind.data.frame(metadata[row_match_long,], otu_long[,c("OTU", "count")])
-split(data_for_cast$count, f = c(data_for_cast$sample_id, data_for_cast$OTU))
[…]

### 2016-11-23 Analysis/multilevel_model/mcmc_diagnostics.R

@@ -1,194 +1,0 @@
-# important things:
-# Obs v predicted 
-# Acf 
-# Prior against marginal posterior 
-# Also traces 
-
-# presumes your JAGS output is called "jags_out"
-
-# and if you input:
-# jags_params <- c(
-                # "beta",
-                # "sigma2",
-                # "tau2",
-                # "phi",
-                # "P",
-                # "beta_0"
-# )
-
-jags_params[6]
-# check out the structure of the output
-summary(jags_out)
-str(jags_out)
-names(jags_out) # "model" "BUGSoutput" "parameters.to.save" "model.file" "n.iter" "DIC"
-
-
-jags_out$parameters.to.save[! jags_out$parameters.to.save %in% "deviance" ]
-
-str(jags_out$BUGSoutput)
-names(jags_out$BUGSoutput)
-
-
-# simulations list
-mcmc_list <- jags_out$BUGSoutput$sims.list
-names(mcmc_list)
-(mcmc_list[jags_params[6]])
-mcmc_list[2]
-
-
-# simulations array
-mcmc_array <- jags_out$BUGSoutput$sims.array
-str(mcmc_array)
-dim(mcmc_array)
-# mcmc_array[ iterations , chains , parameter ]
-# N_iter
-# N_chain
-
-jags_params[]
-names(mcmc_array[1,1,1])
-
-
-
-param_current <- "sigma2"
-
-xdim <- c(min(density(mcmc_array[,, param_current])$x), max(density(mcmc_array[,, param_current])$x))
-ydim <- c(min(density(mcmc_array[,, param_current])$y), max(density(mcmc_array[,, param_current])$y))
-plot(density(mcmc_array[,,param_current]), xlim = xdim, ylim = ydim, type = "l")
-points(density(mcmc_array[,3,1]), type = "l")
-xrange <- c(0,0.3)
-plot("Density", xlim = xrange, ylim = ydim, type = "n", main = param_current)
-points(density(mcmc_array[,,param_current]), type='l')
-pred_x <- seq(xrange[1], xrange[2], 0.01)
-points(pred_x, dgamma(pred_x, 0.01, 0.01), type='l', lty = 2)
-legend("topright", legend = c("Marginal Posterior", "Prior"), lty = c(1,2), bty = "n")
-
[…]

### 2016-11-23 Analysis/multilevel_model/reshape_otu_table.R

@@ -1,161 +1,0 @@
-# put OTU data in long format:
-
-library(reshape2)
-
-setwd("~/Documents/GoogleDrive/Kelly_Lab/Projects/Carkeek_eDNA_grid/Analysis")
-setwd("/Users/jimmy.odonnell/Projects/Carkeek_eDNA_grid/Analysis")
-
-data_dir <- file.path("..", "Data")
-fig_dir <- file.path("..", "Figures")
-
-
-# a table of counts of sequences (Z, length = ...)
-counts_file_path <- file.path(data_dir, "OTUs_top10_4000m.csv")
-
-counts_table <- read.csv(file = counts_file_path, row.names = 1)
-counts_table <- counts_table[,1:10] # use only the top ten for now
-
-# rows/rownames = samples, columns/colnames = taxa, cells = integer counts
-# if table is incorrectly oriented, transpose it:
-# counts_table <- as.data.frame(t(counts_table))
-
-# REMEMBER TO REMOVE CONTROL TAXON IF IT'S STILL THERE!
-# counts_table <- counts_table[,! names(counts_table) %in% "DUP_3"]
-
-
-# load metadata
-metadata_file_path <- file.path(data_dir, "metadata_spatial.csv")
-metadata <- read.csv(file = metadata_file_path, stringsAsFactors = FALSE)
-
-# what is the name of the column containing the sample id (SEQUENCING samples)
-colname_sampleid <- "sample_id"
-
-# what is the name of the column containing PCR replicate levels
-colname_pcr <- "PCR_replicate"
-
-# what is the name of the column containing position in x direction (along shore) replicate levels
-colname_posx <- "transect"
-
-# what is the name of the column containing position in y direction (from shore) replicate levels
-colname_posy <- "dist_from_shore"
-
-# give a name to call the variable describing the things that were counted
-varname_taxa <- "taxon"
-
-# give a name to call the variable describing what the counts are
-varname_counts <- "reads"
-
-
-# of each taxon (length = I)
-taxa <- colnames(counts_table)
-
-# from each PCR replicate (length = J)
-pcr <- as.character(unique(metadata[, colname_pcr]))
-
-# at each position in x direction (length = K)
-posx <- unique(metadata[, colname_posx])
-
-# at each position in y direction (length = L)
-posy <- as.character(unique(metadata[, colname_posy]))
-
-# *** NOTE *** #
[…]

### 2016-11-23 Analysis/multilevel_model/simulate_eDNA_counts.R

@@ -1,158 +1,0 @@
-############################################################################
-# This script simulates some data that resembles eDNA sequence data
-############################################################################
-
-# Original code by Ole Shelton
-# Hacked up by Jimmy O'Donnell
-# Comments mostly reflect Jimmy's imperfect understanding of the underlying concepts.
-
-library(gtools) # gtools::rdirichlet
-library(R2jags) # R2jags::jags
-
-
-
-# Suppose we have an aquarium which contains N_tax taxa
-N_tax <- 9
-
-
-
-# Give them names
-taxon_names <- paste("taxon_", LETTERS[1:N_tax], sep = "")
-
-
-# The distribution of DNA molecules derived from each of the taxa might look like the following
-# Ole originally called these alpha because they eventually become the alpha parameter of the Dirichlet distribution, which sorta determines how likely each component is to be drawn
-# I call these "true" because they represent the "true" abundance of DNA from each taxon in the sample.
-# Set seed to be able to reproduce draws from a probability distribution (gets reset each time one of these functions is used)
-# Instead of runif, you might try using rexp or random from the Pareto distribution to more accurately resemble "real" eDNA results
-set.seed(407744)
-DNA_truth		<- sort(round(runif(N_tax,1,100)), decreasing = TRUE)
-names(DNA_truth) <- taxon_names
-
-# those same values expressed as proportions
-DNA_truth_prop <- DNA_truth/sum(DNA_truth)
-
-
-
-
-
-############################################################################
-# The actual simulation 
-############################################################################
-
-# This simulates data that might be expected from high throughput sequencing of PCR amplicons generated from environmental samples: proportional abundance of sequences from each of the taxa/OTUs/dups
-# draw samples from the Dirichlet distribution, using alpha given by DNA_truth to approximate the sampling of DNA molecules from the environment
-# Set seed to be able to reproduce draws from a probability distribution (gets reset each time one of these functions is used)
-
-
-# If you were to take N_samples samples, what kind of answer would you expect?
-N_samples_small	<-	3
-set.seed(407744)
-eDNA_counts_small <- rdirichlet(N_samples_small, DNA_truth)
-colnames(eDNA_counts_small) <- taxon_names
-
-
-############
-# OR...
[…]

### 2016-11-23 Analysis/multilevel_model/stan_3.R

@@ -1,30 +1,0 @@
-data_dir <- file.path("..", "Data")
-
-data_file <- file.path(data_dir, "data_full_top10.csv")
-
-data_full <- read.csv(data_file)
-
-mydata <- data_full
-
-colname_level1 <- "sequenced_sample"
-colname_level2 <- "transect_line"
-colname_level3 <- "transect_position"
-colname_predictor <- "dist_from_shore"
-colname_outcome <- "DUP_8"
-
-lookup_3in2 <- unique(mydata[,c(colname_level2, colname_level3)])[,colname_level3]
-
-plot(mydata[,colname_predictor], mydata[,colname_outcome])
-
-stan_data <- list(Ni          = length(unique(mydata[,colname_level1])), 
-                  Nj          = length(unique(mydata[,colname_level2])), 
-                  Nk          = length(unique(mydata[,colname_level3])), 
-                  level2      = mydata[,colname_level2],
-                  level3      = mydata[,colname_level3],
-                  lev3ForLev2 = lookup_3in2,
-                  Y_ijk       = mydata[,colname_outcome],
-                  X_1ijk      = mydata[,colname_predictor]
-                  )
-
-lm_out <- lm(mydata[,colname_outcome] ~ mydata[,colname_predictor] )
-summary(lm_out)

### 2016-11-23 Analysis/multilevel_model/stan_4.R

@@ -1,172 +1,0 @@
-data_dir <- file.path("..", "Data")
-
-data_file <- file.path(data_dir, "data_full_long_20.csv")
-
-data_full <- read.csv(data_file)
-
-mydata <- data_full
-
-colname_level1 <- "sequenced_sample"
-colname_level2 <- "transect_line"
-colname_level3 <- "transect_position"
-colname_level4 <- "OTU"
-colname_predictor <- "dist_from_shore"
-colname_outcome <- "count"
-
-lookup_3in2 <- unique(mydata[,c(colname_level2, colname_level3)])[,colname_level3]
-lookup_4in3 <- as.character(unique(mydata[,c(colname_level3, colname_level4)])[,colname_level4])
-
-# plot(mydata[,colname_predictor], mydata[,colname_outcome], xlab = "distance from shore", ylab = "number of reads", las = 1)
-
-stan_data <- list(Ni          = length(unique(mydata[,colname_level1])), 
-                  Nj          = length(unique(mydata[,colname_level2])), 
-                  Nk          = length(unique(mydata[,colname_level3])), 
-                  Nl          = length(unique(mydata[,colname_level4])), 
-                  level2      = mydata[,colname_level2],
-                  level3      = mydata[,colname_level3],
-                  level4      = mydata[,colname_level4],
-                  lev3ForLev2 = lookup_3in2,
-                  lev4ForLev3 = lookup_4in3,
-                  Y_ijkl      = mydata[,colname_outcome],
-                  X_1ijkl     = mydata[,colname_predictor]
-                  )
-
-lm_out <- lm(mydata[,colname_outcome] ~ mydata[,colname_predictor] )
-summary(lm_out)
-
-#-------------------------------------------------------------------------------
-# MODEL IN STAN CODE
-#-------------------------------------------------------------------------------
-stan_code <- '
-data {
-  // Define variables in data
-  // Number of level-1 observations (an integer)
-  int<lower=0> Ni;
-  // Number of level-2 clusters
-  int<lower=0> Nj;
[…]

### 2017-01-04 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex
work on response to reviews

\documentclass{article}
\usepackage{changepage}% \adjustwidth{<leftskip>}{<rightskip>} http://ctan.org/pkg/changepage
\usepackage{lipsum}% http://ctan.org/pkg/lipsum
\usepackage[margin=1in]{geometry} % make the margins 1 inch on all sides of the document

% The following four lines prevent indentation unless it is invoked with \indent
\newlength\tindent
\setlength{\tindent}{\parindent}
\setlength{\parindent}{0pt}
\renewcommand{\indent}{\hspace*{\tindent}}

\title{Response to Reviews}

% You can make your own environment
\newenvironment{response}
	{
	\begin{adjustwidth}{1cm}{0cm}
	\itshape % make italic
	}
	{
	\end{adjustwidth}
	}

\begin{document}

\maketitle

\noindent
{\large
Manuscript Title:
Spatial distribution of environmental DNA in a nearshore marine habitat
\\
Journal:
PeerJ
\\
Manuscript ID:
14649
\\
Authors:
James L. O'Donnell,
Ryan P. Kelly,
Andrew O. Shelton,
Jameal F. Samhouri,
Natalie C. Lowell,
Gregory D. Williams
\\
}

We are grateful to the editor and reviewers for their insightful feedback.
Reviewer feedback is below in normal font, with our responses in indented italics.
Additional changes to the manuscript independent of the reviews are presented at the end of this document.

%------------------------------------EDITOR-------------------------------------
\subsection*{Editor}
Both reviewers commented favourably on this manuscript and have made some useful and constructive comments which you should consider and respond to. I would also like you to respond to the comment by Bernd Haenfling on your pre-print as he is in agreement with reviewer #1 in relation to the degree of novelty this work represents (line 100 and your concluding sentence). Personally, I found your discussion of the results in the light of the advantages and limitations of this methodology interesting and highly informative. I wondered what the community analyses results would look like if you compartmentalised by e.g. life history or type (e.g. fish)?

\begin{response}
  Thank you for the positive feedback.
	Indeed, we agree with the comment made on the preprint. Our omission was not intended as a slight, merely an oversight given the rapid pace of publication in this field.
	TODO: We have added the appropriate citations and corrected our wording.
	TODO: We were also curious about the analysis you mention. Unfortunately, we found no meaningful patterns, and have made an attempt to clarify this in the text.
	\\
\end{response}

%------------------------------------REVIEWER1----------------------------------
\subsection*{Reviewer 1 (Owen Wangensteen)}

\subsubsection*{Basic reporting}
The article is well written. The structure is correct. The introduction and the material and methods sections are a little bit long, but can be easily improved by deleting some sentences (detailed below in this review). The references used are relevant and updated. The raw data and the analysis scripts are all publicly available.

\subsubsection*{Experimental design}
The objectives are clearly established and the experimental design is perfect for addressing the proposed questions. The laboratory methods are rigorous and the analysis pipeline includes steps for assuring excellent quality of the resulting data set.

\subsubsection*{Validity of the findings}
The findings are very robust. The high level of replication in the PCR analyses and the high-standards bioinformatic filtering steps used for removing any sequence attributable to possible contamination assure an excellent resulting data set to work with. The statistical analysis of the data are very clearly explained and the results are highly trustworthy.

\subsubsection*{Comments for the Author}
The manuscript is, in general, very well written and the robust results obtained from the enhanced laboratory procedures and bioinformatic pipelines used are presented in a clear manner. I think that the introduction and material & methods sections are a little too long, but they can be enhanced by deleting some sentences (detailed below).

I think that it would be informative to add a paragraph in the discussion section commenting the caveats of using 16S as metabarcoding marker for taxonomic assignment of the MOTUs (i.e. the issues with the assignment of many sequences which can be assigned only to the family or lower levels), compared to other more variable markers with denser reference databases, such as COI, which, in most cases, allow assignments to the genus or species level.

\begin{response}
	We agree with your point that a locus with only coarse-scale resolution (16S) may obfuscate patterns of diversity compared to a locus with fine-scale resolution (CO1). A subset of us has recently published an in-depth comparison of multiple loci for metabarcoding, and TODO: we have added a more substantial comment on this and referenced to this now-published work.
\\
\end{response}

\subsubsection*{Specific Corrections}
Some specific corrections are detailed below:

L 36 to 40. This whole paragraph may be deleted to shorten the introduction, without any significant information loss.
\begin{response}
  TODO: Check with group on this.\\
\end{response}

L90. Correct “Actiniaria” and “Bryozoans”.
\begin{response}
  Thanks for catching these errors. We made these changes.\\
\end{response}

L91. Delete “(Decapoda)”. These habitats hold many other Crustacea other than Decapoda as well. Actually, I would write “barnacles (Sessilia) and other crustaceans” in line 89.
\begin{response}
  Excellent suggestion; we made this change.\\
\end{response}

L93: “which provide”
\begin{response}
  \\
\end{response}

L94: The microscopic plankton is not composed just by “diatoms and larvae”. Many important holoplanktonic organisms (e.g. copepods) could be mentioned here.
\begin{response}
  \\
\end{response}

L100: I am not sure that this is “the first explicitly spatial analysis of eDNA-derived community similarity”, since some other papers have been published which reported patterns of spatial variability in marine systems (although using less quantitative approaches); e.g. De Vargas et al. 2015. I think that the authors should change this to “an explicitly spatial analysis of eDNA-derived community similarity”.
\begin{response}
  \\
\end{response}

L117 to 122. This whole paragraph can be deleted. This methodology and the steps involved are now commonplace among aquatic biodiversity researchers and need not be enumerated.
\begin{response}
  \\
\end{response}

L166 to 170. Why not use “µl” instead of “microliter”?
\begin{response}
  \\
\end{response}

L169. I guess the authors mean “1.25 microliter genomic or eDNA template”.
\begin{response}
  \\
\end{response}

L171 to 172. Use “95 ºC” instead of “95C”.
\begin{response}
  \\
\end{response}

L195. “(machine NS500615, run 115, flowcell H3LFLAFXX)”. I don't think we need to know these details!
\begin{response}
  \\
\end{response}

L196. “to enhance sequencing depth”. PhiX is added “to increase sequence diversity” and not to enhance sequencing depth. Actually, the more PhiX is spiked-in, the less amplicon reads will be obtained, thus decreasing (not enhancing) sequencing depth.
\begin{response}
  \\
\end{response}

L223. Delete “a” in “This allows for a samples”.
\begin{response}
  \\
\end{response}

L270-272. I think that these sentences may be confusing due to the use of some terms related to “dispersal range”, “type of fertilization” and “type of larval development”. Some “internally fertilized species” may have larvae with planktonic development and long dispersal capacity (e.g. some barnacles), whereas some broadcast spawners (e.g. some gorgonians) may be surface brooders with low dispersal ability. I think that the main traits affecting “dispersal range” is the type of larval development (planktonic or benthic) and the total planktonic larval duration (PLD), and not the type of fertilization or the strategy for gamete releasing, per se. I think that “internally fertilized species” in L271 could be changed to “benthic brooding species” and “broadcast spawners” could be changed to “species with long planktonic larval development”. Anyway, this semi-quantitative life-history information recorded by the authors in their data tables does not seem to have been used in their further analyses (since only qualitative considerations regarding two brooding species are discussed below).
\begin{response}
  \\
\end{response}

L285. “333,537.9 ± 112,200.5”. It looks somehow awkward to use decimal values for numbers of reads here. You could round the values to “333,538 ± 112,200”
\begin{response}
  \\
\end{response}

L314-315. You can delete “the probability that two reads chosen at random from a sample belong to different species”, since it has been already explained in the Materials & Methods section.
\begin{response}
  \\
\end{response}

L344. “For the first time in a marine environment, we document four key patterns”. The authors should definitely use a more modest sentence here. At least the first two patterns they claim to report “for the first time” have been already reported previously in several works in the marine environment. E.g.: Fonseca et al. 2014, Guardiola et al. 2015, Lallias et a. 2015, De Vargas et al. 2015, etc., although possibly never with a sampling design specifically designed to get a quantitative assessment of distance decay.
\begin{response}
  \\
\end{response}

%------------------------------------REVIEWER2----------------------------------
\subsection*{Reviewer 2 (Stephen Moss)}
\subsubsection*{Basic reporting}
There were a couple of minor grammatical errors in the manuscript. 1) The final sentence (L292-294) of the “Sequence Data Processing (Bioinformatics)” paragraph on page 12 in the Results section. 2) The first sentence (L359) of the “Communities far from one another tend to be less similar than those that are nearby” paragraph on page 15 of the Discussion section.
\begin{response}
  \\
\end{response}

I was a bit confused initially over discussion of massively parallel sequencing of 16S amplicons in relation to metazoan diversity in the abstract, and it wasn’t until into the introduction (L98) that I realised it was referring to 16S sequences from the mtDNA. Perhaps some clarity on this earlier on would be helpful? Perhaps also discussing the reasoning behind the decision to use 16S rDNA from the mtDNA rather than 18S rDNA would be helpful (e.g. Tang et. al., 2012; http://www.pnas.org/content/109/40/16208)?
\begin{response}
  \\
\end{response}

I couldn’t find any reference to the details of the assigned clusters in Figure 3. I thought it might be useful to include a reference to figures or tables, perhaps as supplementary information, outlining the taxonomic composition of the clusters?
\begin{response}
  \\
\end{response}

I felt the label for Figure 5 needed expanding to provide more detailed information on what it represents, making it more meaningful/understandable as a stand-alone figure.
\begin{response}
  \\
\end{response}

The raw data is listed as available under Data Availability, though I assume this is still under embargo? The link on page 18 (L463) will need updating upon publication.
\begin{response}
  \\
\end{response}

A GitHub DOI can be provided by Zenodo (https://zenodo.org) and will need updating upon publication.
\begin{response}
  \\
\end{response}

\subsubsection*{Experimental design}
When detailing how samples were collected (L126-128) under the Environmental Sampling header in the Methods section, I felt that information on how sampling artefacts were mediated (e.g. from L409-414) could have been provided sooner.
\begin{response}
  \\
\end{response}

I felt a little more effort could have been made to facilitate the ability to replicate the analyses in this study. The methods section provided good information, but the analyses as detailed on L473 needs more clarity. An outline of the workflow steps and the input/output at each stage would be very helpful here.
\begin{response}
  \\
\end{response}

\subsubsection*{Validity of the findings}
In response to the “(3) Richness declines and evenness increases with distance from shore” heading in the Discussion section, I felt that decline in richness and increase in evenness might be expected due to dilution and osmotic factors in the ocean. Perhaps some reference to tidal flushing and dilution might be pertinent here?
\begin{response}
  \\
\end{response}

I felt that under the remit of the Ecological Analyses headings throughout the study, there could have been some focus on reporting the alpha diversity and rarefaction analysis in relation to whether effective sequencing depth had been achieved for the samples in question.
\begin{response}
  \\
\end{response}

\subsubsection{Comments for the Author}
I thought the paper was very interesting. The study was well designed, implemented, and written. It was useful to clarify the expectations and hypotheses outlined within, however, I felt more effort could have been made to provide recommendations on how some of the limitations in eDNA sampling could be overcome.
\begin{response}
  \\
\end{response}

The bioinformatics/statistical methods seemed to be very well thought out, and consistent with current best practices in the analyses of this type of sequencing data.
\begin{response}
  \\
\end{response}

\end{document}

### 2017-01-04 Manuscript/v1.1_PeerJ/carkeek_grid.tex
added reviews and prepared for revision
work on response to reviews

--- 2016-11-22 Manuscript/v1.0_PeerJ/carkeek_grid.tex
@@
 \author[1]{James L. O'Donnell\thanks{jodonnellbio@gmail.com}}
-\author[1]{Ryan P. Kelly}
-\author[2]{Andrew O. Shelton}
-\author[2]{Jameal F. Samhouri}
-\author[1,3]{Natalie C. Lowell}
-\author[4]{Gregory D. Williams}
+\author[1]{Ryan P. Kelly} % rpkelly@uw.edu
+\author[2]{Andrew O. Shelton} % ole.shelton@noaa.gov
+\author[2]{Jameal F. Samhouri} % jameal.samhouri@noaa.gov
+\author[1,3]{Natalie C. Lowell} % nclowell@uw.edu
+\author[4]{Gregory D. Williams} % greg.williams@noaa.gov
 
@@
 Eelgrass meadows harbor epifaunal and infaunal biota, and attract transient species which use the meadows for shelter and to feed on resident organisms.
-Hard intertidal surfaces support a well-documented biota including barnacles (Sessilia), mussels (Bivalvia:Mytilidae), anemones (Actinaria), sea stars (Asteroidea), urchins (Echinoidea), Bryzoans (Ectoprocta), crustaceans (Decapoda), and a variety of algae \citep{Dethier2010}.
+Hard intertidal surfaces support a well-documented biota including barnacles (Sessilia) and other crustaceans, mussels (Bivalvia:Mytilidae), anemones (Actiniaria), sea stars (Asteroidea), urchins (Echinoidea), Bryozoans (Ectoprocta), crustaceans (Decapoda), and a variety of algae \citep{Dethier2010}.
 Hard bottoms of the lower intertidal and high subtidal are home to macroalgae such as Laminariales and Desmarestiales which provides habitat for a distinct community of fish and invertebrates.
@@
 \section*{Ethics Statement}
-The authors declare no conflict of interest. No permits were required to do any of the research described here.
+The authors declare no conflict of interest.
+Consistent with the public trust doctrine, waters of the US are public, and therefore no permit was necessary to conduct this research (see Illinois Central Railroad v. Illinois, 146 U.S. 387 (1892)).
 

### 2017-01-05 Manuscript/v1.1_PeerJ/carkeek_grid.tex
typo

@@
 Hard intertidal surfaces support a well-documented biota including barnacles (Sessilia) and other crustaceans, mussels (Bivalvia:Mytilidae), anemones (Actiniaria), sea stars (Asteroidea), urchins (Echinoidea), Bryozoans (Ectoprocta), crustaceans (Decapoda), and a variety of algae \citep{Dethier2010}.
-Hard bottoms of the lower intertidal and high subtidal are home to macroalgae such as Laminariales and Desmarestiales which provides habitat for a distinct community of fish and invertebrates.
+Hard bottoms of the lower intertidal and high subtidal are home to macroalgae such as Laminariales and Desmarestiales which provide habitat for a distinct community of fish and invertebrates.
 The upper pelagic is home to a diverse assemblage of microscopic plankton including diatoms and larvae \citep{Strickland1983}, as well as transitory fish and marine mammals.

### 2017-01-05 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex
added some responsed

@@
 \begin{response}
-  TODO: Check with group on this.\\
+  We can understand the reviewer's perspective, but feel this point is a stylistic one and would prefer to leave the manuscript as it is currently worded to increase the appeal of our work to a broader audience.\\
 \end{response}
@@
 \begin{response}
-  \\
+  We made this change.\\
 \end{response}
@@
 \begin{response}
-  \\
+  We did not state that the microscopic plankton is composed solely of diatoms and larve, we made a  statement and qualified it with two examples to give meaning and context: `diverse assemblage of microscopic plankton including diatoms and larvae'.
+We mention diatoms because they are a widely-known, numerically abundant, and ecologically important group that falls outside the scope of this study on metazoans.
+We mention larvae because it subtly hints at our major finding that our samples captured many organisms best known for their sessile, benthic stages (e.g.
+barnacles).
+Yes, we could add copepods to this list if you would like, but what qualifies a group for inclusion?
+Numerical abundance?
+Biomass?\\
 \end{response}
@@
 
+%------------------------------------REVIEWER3----------------------------------
+\subsection*{Comment on the preprint (Bernd Hänfling)}
+This is well designed study and we would like to congratulate the authors to this data set and its excellent analysis.
+However, we feel that the authors have made an important omission in the discussion.
+
+Given that the main focus of this paper is to explore the spatial distribution of community eDNA we feel that the authors should consider the results of our study on lake fish communities (Hänfling et al.
+2016).
+In our paper we have used extensive spatial sampling of three transects across our main study lake with a total of 66 individual samples and to our knowledge there is no other data set on macrobial eDNA metabarcoding with a comparable extent of explicit sampling.
+Our paper confirms that the spatial distribution of individual species reflects their abundance but also shows that the pattern of distribution is not random and ecologically meaningful.
+Although there might be important differences between freshwater and marine habitats we feel that the results from our study are extremely relevant here and should not be omitted especially when claiming novelty of discussion point 2.
+
+Bernd Hänfling, Lori Lawson Handley, Dan Read and Ian Winfield
+
+Hänfling B, Lawson Handley L, Read DS, Hahn C, Li J, Nichols P, Blackman RC, Oliver A, Winfield IJ (2016) Environmental DNA metabarcoding of lake fish communities reflects long-term data from established survey methods.
+Molecular Ecology 25, 3101-3119.
+\begin{response}
+  \\
+\end{response}
+
 \end{document}

### 2017-01-05 Manuscript/v1.1_PeerJ/carkeek_grid.tex
correct affiliations to NMFS standards

@@
 \author[2]{Andrew O. Shelton} % ole.shelton@noaa.gov
-\author[2]{Jameal F. Samhouri} % jameal.samhouri@noaa.gov
-\author[1,3]{Natalie C. Lowell} % nclowell@uw.edu
-\author[4]{Gregory D. Williams} % greg.williams@noaa.gov
+\author[3]{Jameal F. Samhouri} % jameal.samhouri@noaa.gov
+\author[1,4]{Natalie C. Lowell} % nclowell@uw.edu
+\author[5]{Gregory D. Williams} % greg.williams@noaa.gov
 
@@
 \affil[1]{School of Marine and Environmental Affairs, University of Washington, 3707 Brooklyn Ave NE, Seattle, Washington 98105, USA}
-\affil[2]{Northwest Fisheries Science Center, NOAA Fisheries, 2725 Montlake Blvd E, Seattle, Washington 98112, USA}
-\affil[3]{School of Aquatic and Fishery Sciences, University of Washington, 1122 NE Boat St, Seattle, Washington 98105, USA}
-\affil[4]{Pacific States Marine Fisheries Commission, Under contract to the Northwest Fisheries Science Center, NOAA Fisheries, 2725 Montlake Blvd E, Seattle, WA 98112}
+\affil[2]{Earth Resource Technology, Inc., Under contract to the Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, 2725 Montlake Blvd E, Seattle, WA 98112, USA}
+\affil[3]{Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, 2725 Montlake Blvd E, Seattle, Washington 98112, USA}
+\affil[4]{School of Aquatic and Fishery Sciences, University of Washington, 1122 NE Boat St, Seattle, Washington 98105, USA}
+\affil[5]{Pacific States Marine Fisheries Commission, Under contract to the Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, 2725 Montlake Blvd E, Seattle, WA 98112, USA}
 

### 2017-01-05 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex
responding to reviews

@@
 \noindent
-{\large
+{
 Manuscript Title:
@@
 Both reviewers commented favourably on this manuscript and have made some useful and constructive comments which you should consider and respond to.
-I would also like you to respond to the comment by Bernd Haenfling on your pre-print as he is in agreement with reviewer #1 in relation to the degree of novelty this work represents (line 100 and your concluding sentence).
+I would also like you to respond to the comment by Bernd Haenfling on your pre-print as he is in agreement with reviewer 1 in relation to the degree of novelty this work represents (line 100 and your concluding sentence).
 Personally, I found your discussion of the results in the light of the advantages and limitations of this methodology interesting and highly informative.
@@
 	Indeed, we agree with the comment made on the preprint. Our omission was not intended as a slight, merely an oversight given the rapid pace of publication in this field.
-	TODO: We have added the appropriate citations and corrected our wording.
-	TODO: We were also curious about the analysis you mention. Unfortunately, we found no meaningful patterns, and have made an attempt to clarify this in the text.
+	We have added the appropriate citations and corrected our wording.
+	% TODO: add hanfling citation, respond to comment, update context in paper.
+	We were also curious about the effect of the data subset you mention, and conducted that analysis.
+Unfortunately, we found no meaningful patterns, and have made an attempt to clarify this in the text.
+	% TODO: comment on effect of subsetting based on life-history strategy or taxon
 	\\
@@
 The manuscript is, in general, very well written and the robust results obtained from the enhanced laboratory procedures and bioinformatic pipelines used are presented in a clear manner.
-I think that the introduction and material & methods sections are a little too long, but they can be enhanced by deleting some sentences (detailed below).
+I think that the introduction and material and methods sections are a little too long, but they can be enhanced by deleting some sentences (detailed below).
 
@@
 
-L90. Correct “Actiniaria” and “Bryozoans”.
+L90. Correct ``Actiniaria'' and ``Bryozoans''.
 \begin{response}
@@
 
-L91. Delete “(Decapoda)”. These habitats hold many other Crustacea other than Decapoda as well. Actually, I would write “barnacles (Sessilia) and other crustaceans” in line 89.
+L91. Delete ``(Decapoda)''. These habitats hold many other Crustacea other than Decapoda as well. Actually, I would write ``barnacles (Sessilia) and other crustaceans'' in line 89.
 \begin{response}
@@
 
-L93: “which provide”
+L93: ``which provide''
 \begin{response}
@@
 
-L94: The microscopic plankton is not composed just by “diatoms and larvae”. Many important holoplanktonic organisms (e.g. copepods) could be mentioned here.
+L94: The microscopic plankton is not composed just by ``diatoms and larvae''. Many important holoplanktonic organisms (e.g. copepods) could be mentioned here.
 \begin{response}
-  We did not state that the microscopic plankton is composed solely of diatoms and larve, we made a  statement and qualified it with two examples to give meaning and context: `diverse assemblage of microscopic plankton including diatoms and larvae'.
+  We did not state that the microscopic plankton is composed solely of diatoms and larve, we made a  statement and qualified it with two examples to give meaning and context: `a diverse assemblage of microscopic plankton including diatoms and larvae'.
 We mention diatoms because they are a widely-known, numerically abundant, and ecologically important group that falls outside the scope of this study on metazoans.
@@
 
-L100: I am not sure that this is “the first explicitly spatial analysis of eDNA-derived community similarity”, since some other papers have been published which reported patterns of spatial variability in marine systems (although using less quantitative approaches); e.g.
+L100: I am not sure that this is ``the first explicitly spatial analysis of eDNA-derived community similarity'', since some other papers have been published which reported patterns of spatial variability in marine systems (although using less quantitative approaches); e.g.
 De Vargas et al.
 2015.
-I think that the authors should change this to “an explicitly spatial analysis of eDNA-derived community similarity”.
+I think that the authors should change this to ``an explicitly spatial analysis of eDNA-derived community similarity''.
 \begin{response}
-  \\
+  Indeed, our wording did not clearly convey our meaning.
+Our intent was to indicate the novelty of the experimental design which was intended to capture variability in transport of eDNA.
+We have made the suggested change.\\
+	% TODO: Comment on the novelty of this study in the context of refs: Fonseca et al.
+2014, Guardiola et al.
+2015 (PLOSONE), Lallias et a.
+2015, De Vargas et al.
+2015, Hanfling 2016, Guardiola 2016 PeerJ e.g.
+`while other studies have documented the spatial diversity of meiofauna (Fonseca 2014), '
 \end{response}
@@
 \begin{response}
-  \\
+  We appreciate this perspective, but disagree that the method is commonplace to the extent that it can be omitted.
+Our perspective is that few ecologists understand eDNA methodology, and will likely skip the details we outline in the methods.
+We include this very brief overview for those readers in order to give perspective to the complicated methodology.\\
 \end{response}
 
-L166 to 170. Why not use “µl” instead of “microliter”?
+L166 to 170. Why not use ``µl'' instead of ``microliter''?
 \begin{response}
-  \\
+  Our experience is that such characters are often mangled at some point the publication process. The journal is welcome to make this change if they so desire.\\
 \end{response}
 
-L169. I guess the authors mean “1.25 microliter genomic or eDNA template”.
+L169. I guess the authors mean ``1.25 microliter genomic or eDNA template''.
 \begin{response}
-  \\
+  We intended `genomic eDNA' to mean the full DNA extracted from samples, but your suggestion is a good one and we have implemented it.\\
 \end{response}
 
-L171 to 172. Use “95 ºC” instead of “95C”.
+L171 to 172. Use ``95 ºC'' instead of ``95C''.
 \begin{response}
-  \\
+  We made this change.\\
 \end{response}
 
-L195. “(machine NS500615, run 115, flowcell H3LFLAFXX)”. I don't think we need to know these details!
+L195. ``(machine NS500615, run 115, flowcell H3LFLAFXX)''. I don't think we need to know these details!
 \begin{response}
-  \\
+  While this does seem like overkill, it is not unheard of for machines or sequencing facilities to develop problems that can bias data, and we included this with the intent of being as transparent as possible.
+Nevertheless, we made this change.\\
 \end{response}
@@
 L196.
-“to enhance sequencing depth”.
-PhiX is added “to increase sequence diversity” and not to enhance sequencing depth.
+``to enhance sequencing depth''.
+PhiX is added ``to increase sequence diversity'' and not to enhance sequencing depth.
 Actually, the more PhiX is spiked-in, the less amplicon reads will be obtained, thus decreasing (not enhancing) sequencing depth.
 \begin{response}
-  \\
+  We are grateful to have a reviewer reading with such a careful eye!
+We believe both statements are correct, though subtly so.
+Because PhiX increases sequence diversity, its effect is to increase the number of clusters which can be correctly differentiated from one another on the flowcell, thus increasing sequencing depth.
+Similar to your statement, the \textbf{less} PhiX is spiked-in, the fewer clusters will be correctly differentiated from one another, and the fewer reads passing filter will be obtained, thus also decreasing sequencing depth.\\
 \end{response}
 
-L223. Delete “a” in “This allows for a samples”.
+L223. Delete ``a'' in ``This allows for a samples''.
 \begin{response}
-  \\
+  We made this change.\\
 \end{response}
@@
 L270-272.
-I think that these sentences may be confusing due to the use of some terms related to “dispersal range”, “type of fertilization” and “type of larval development”.
-Some “internally fertilized species” may have larvae with planktonic development and long dispersal capacity (e.g.
+I think that these sentences may be confusing due to the use of some terms related to ``dispersal range'', ``type of fertilization'' and ``type of larval development''.
+Some ``internally fertilized species'' may have larvae with planktonic development and long dispersal capacity (e.g.
 some barnacles), whereas some broadcast spawners (e.g.
 some gorgonians) may be surface brooders with low dispersal ability.
-I think that the main traits affecting “dispersal range” is the type of larval development (planktonic or benthic) and the total planktonic larval duration (PLD), and not the type of fertilization or the strategy for gamete releasing, per se.
-I think that “internally fertilized species” in L271 could be changed to “benthic brooding species” and “broadcast spawners” could be changed to “species with long planktonic larval development”.
+I think that the main traits affecting ``dispersal range'' is the type of larval development (planktonic or benthic) and the total planktonic larval duration (PLD), and not the type of fertilization or the strategy for gamete releasing, per se.
+I think that ``internally fertilized species'' in L271 could be changed to ``benthic brooding species'' and ``broadcast spawners'' could be changed to ``species with long planktonic larval development''.
 Anyway, this semi-quantitative life-history information recorded by the authors in their data tables does not seem to have been used in their further analyses (since only qualitative considerations regarding two brooding species are discussed below).
 \begin{response}
-  \\
+  You have made an excellent point that we were not clear in our wording. \\
 \end{response}
 
-L285. “333,537.9 ± 112,200.5”. It looks somehow awkward to use decimal values for numbers of reads here. You could round the values to “333,538 ± 112,200”
+L285. ``333,537.9 ± 112,200.5''. It looks somehow awkward to use decimal values for numbers of reads here. You could round the values to ``333,538 ± 112,200''
 \begin{response}
-  \\
+  Because these are summary statistics (mean and standard deviation), we believe the use of decimals is appropriate.\\
 \end{response}
 
-L314-315. You can delete “the probability that two reads chosen at random from a sample belong to different species”, since it has been already explained in the Materials & Methods section.
+L314-315. You can delete ``the probability that two reads chosen at random from a sample belong to different species'', since it has been already explained in the Materials and Methods section.
 \begin{response}
-  \\
+  We made this change.\\
 \end{response}
@@
 L344.
-“For the first time in a marine environment, we document four key patterns”.
+``For the first time in a marine environment, we document four key patterns''.
 The authors should definitely use a more modest sentence here.
-At least the first two patterns they claim to report “for the first time” have been already reported previously in several works in the marine environment.
+At least the first two patterns they claim to report ``for the first time'' have been already reported previously in several works in the marine environment.
 E.g.: Fonseca et al.
@@
 \begin{response}
-  \\
+  Our intent was to convey the novelty of the sampling design. We have clarified the novelty of our study in the context of recently published work in the introduction.\\
+	% TODO: Comment on the novelty of this study in the context of refs: Fonseca et al.
+2014, Guardiola et al.
+2015 (PLOSONE), Lallias et a.
+2015, De Vargas et al.
+2015, Hanfling 2016, Guardiola 2016 PeerJ
 \end{response}
@@
 There were a couple of minor grammatical errors in the manuscript.
-1) The final sentence (L292-294) of the “Sequence Data Processing (Bioinformatics)” paragraph on page 12 in the Results section.
-2) The first sentence (L359) of the “Communities far from one another tend to be less similar than those that are nearby” paragraph on page 15 of the Discussion section.
+1) The final sentence (L292-294) of the ``Sequence Data Processing (Bioinformatics)'' paragraph on page 12 in the Results section.
+2) The first sentence (L359) of the ``Communities far from one another tend to be less similar than those that are nearby'' paragraph on page 15 of the Discussion section.
 \begin{response}
-  \\
+  OurResponse\\
 \end{response}
@@
 \begin{response}
-  \\
+  OurResponse\\
 \end{response}
@@
 \begin{response}
-  \\
+  Yes; these links are a placeholder until the manuscript is accepted.\\
+	% TODO: include EMBL/EBI url
 \end{response}
@@
 \begin{response}
-  \\
+  We are aware of this and will create a GitHub version and DOI of the respository once the final version is obtained.\\
+	% TODO: upon acceptance, get DOI and update
 \end{response}
@@
 \subsubsection*{Validity of the findings}
-In response to the “(3) Richness declines and evenness increases with distance from shore” heading in the Discussion section, I felt that decline in richness and increase in evenness might be expected due to dilution and osmotic factors in the ocean.
+In response to the ``(3) Richness declines and evenness increases with distance from shore'' heading in the Discussion section, I felt that decline in richness and increase in evenness might be expected due to dilution and osmotic factors in the ocean.
 Perhaps some reference to tidal flushing and dilution might be pertinent here?
 \begin{response}
-  \\
+  OurResponse\\
 \end{response}

### 2017-01-05 Manuscript/v1.1_PeerJ/carkeek_grid.tex
responding to reviews

@@
 Using PCR-based methods and massively parallel sequencing, we surveyed mitochondrial 16S sequences from a suite of marine animals in water samples collected over a grid of sites extending from the shoreline out to 4 kilometers offshore in Puget Sound, Washington, USA.
-We leverage this sampling design to perform the first explicitly spatial analysis of eDNA-derived community similarity.
+We leverage this sampling design to perform an explicitly spatial analysis of eDNA-derived community similarity.
 We investigate two primary objectives.
@@
 DNA was purified from the membrane following a phenol:chloroform:isoamyl alcohol protocol following Renshaw \citep{Renshaw2014}.
-Preserved membranes were incubated at 65C for 30 minutes before adding 900 microliters of phenol:chloroform:isoamyl alcohol and shaking vigorously for 60 seconds.
+Preserved membranes were incubated at 65 ºC for 30 minutes before adding 900 microliters of phenol:chloroform:isoamyl alcohol and shaking vigorously for 60 seconds.
 We conducted two consecutive chloroform washes by centrifuging at 14,000 rpm for 5 minutes, transferring the aqueous layer to 700 microliters chloroform, and shaking vigorously for 60 seconds.
-After a third centrifugation, 500 microliters of the aqueous layer was transferred to tubes containing 20 microliters 5 molar NaCl and 500 microliters 100\% isopropanol, and frozen at -20C for approximately 15 hours.
-Finally, all liquid was removed by centrifuging at 14000 rpm for 10 minutes, pouring off or pipetting out any remaining liquid, and drying in a vacuum centrifuge at 45C for 15 minutes.
+After a third centrifugation, 500 microliters of the aqueous layer was transferred to tubes containing 20 microliters 5 molar NaCl and 500 microliters 100\% isopropanol, and frozen at -20 ºC for approximately 15 hours.
+Finally, all liquid was removed by centrifuging at 14000 rpm for 10 minutes, pouring off or pipetting out any remaining liquid, and drying in a vacuum centrifuge at 45 ºC for 15 minutes.
 DNA was resuspended in 200 microliters of ultrapure water.
@@
 In a UV-sterilized hood, we prepared 25 microliter reactions containing 18.375 microliters ultrapure water, 2.5 microliters 10x buffer, 0.625 microliters deoxynucleotide solution (8 millimolar), 1 microliter each forward and reverse primer (10 micromolar, obtained lyophilized from Integrated DNA Technologies (Coralville, IA, USA)), 0.25 microliters Qiagen HotStar Taq polymerase, and 1.25 microliter genomic eDNA template at 1:100 dilution in ultrapure water.
-PCR thermal profiles began with an initialization step (95C; 15 min) followed by cycles (40 and 20 for the first and second reaction, respectively) of denaturation (95C; 15 sec), annealing (61C; 30 sec), and extension (72C; 30 sec).
+PCR thermal profiles began with an initialization step (95 ºC; 15 min) followed by cycles (40 and 20 for the first and second reaction, respectively) of denaturation (95 ºC; 15 sec), annealing (61 ºC; 30 sec), and extension (72 ºC; 30 sec).
 20 identical PCRs were conducted from each DNA extract using non-indexed primers; these were pooled into 4 groups of 5 in order to ensure ample template for the subsequent PCR with indexed primers.
@@
 Fragment size distribution and concentration of each library was quantified using an Agilent 2100 BioAnalyzer.
-Libraries were pooled in equal concentrations and sequenced for 150 base pairs in both directions (PE150) using an Illumina NextSeq at the Stanford Functional Genomics Facility (machine NS500615, run 115, flowcell H3LFLAFXX), where 20\% PhiX Control v3 was added to act as a sequencing control and to enhance sequencing depth.
+Libraries were pooled in equal concentrations and sequenced for 150 base pairs in both directions (PE150) using an Illumina NextSeq at the Stanford Functional Genomics Facility, where 20\% PhiX Control v3 was added to act as a sequencing control and to enhance sequencing depth by increasing sequence diversity.
 Raw sequence data in fastq format is publicly available (see Data Availability).
@@
 Where the relationship between community similarity ($y_{ij}$) and spatial distance ($x_{ij}$) between observations $i$ and $j$ is determined by the similarity of samples at distance 0 ($A$), and the distance at which half the total change in similarity is achieved ($B$).
-This allows for a samples collected very close together (near 0) to have similarity significantly less than one.
+This allows for samples collected very close together (near 0) to have similarity significantly less than one.
 We assessed model fit using the R function nls \citep{R}, using the nl2sol algorithm from the Port library to solve separable nonlinear least squares using analytically computed derivatives (http://netlib.org/port/nsg.f).
@@
 We collated  life history characteristics for each of the major taxonomic groups recovered, including dispersal range of the gametes, larvae, and adults, adult habitat type and selectivity, and adult body size.
-Dispersal range was given as an order-of-magnitude approximation of the scale of dispersal: for example, internally fertilized species were assigned a gamete range of 0 km, while broadcast spawners were assigned a gamete range of 10 km.
+For each life history stage, we made an order-of-magnitude approximation of the scale of dispersal: for example, internally fertilized species were assigned a gamete range of 0 km, while broadcast spawners were assigned a gamete range of 10 km.
 Similarly, adult range size was approximated as 0 km (sessile), 1 km (motile but not pelagic), or 10 km (highly mobile, pelagic).
@@
 Mean OTU richness declined by 1.42 per 1000 meters from a mean of 17.6 taxa (95\%CI = 2.15) inshore to 11.9 taxa (95\%CI = 4.31) at offshore locations (p = 0.0415; Figure \ref{diversity_distance}).
-Evenness, the probability that two reads chosen at random from a sample belong to different species, increased by .0666 per 1000 meters from 0.225 (95\%CI = 0.0558) to 0.491 (95\%CI = $\pm$ 0.112), indicating that sequence reads were less evenly distributed among taxa in offshore samples (p $\ll$ 0.05; Figure \ref{diversity_distance}).
+Evenness increased by .0666 per 1000 meters from 0.225 (95\%CI = 0.0558) to 0.491 (95\%CI = $\pm$ 0.112), indicating that sequence reads were less evenly distributed among taxa in offshore samples (p $\ll$ 0.05; Figure \ref{diversity_distance}).
 There was no evidence for spatial autocorrelation for any of the diversity metrics (Moran's I, p > 0.05; Figure \ref{diversity}).
@@
 Monitoring of organisms using environmental DNA is an especially promising method, given the rapid pace of advancement in technological innovation and cost efficiency in the field of DNA sequencing and quantification.
-For the first time in a marine environment, we document four key patterns: (1) eDNA communities far from one another tend to be less similar than those that are nearby, (2) distinct eDNA communities exist and are distributed in a non-random fashion, (3) diversity declines with distance from shore, and (4) spatial patterning of eDNA is associated with taxon-specific life history characteristics.
+We document four key patterns: (1) eDNA communities far from one another tend to be less similar than those that are nearby, (2) distinct eDNA communities exist and are distributed in a non-random fashion, (3) diversity declines with distance from shore, and (4) spatial patterning of eDNA is associated with taxon-specific life history characteristics.
 

### 2017-01-06 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex
respond to reviews

@@
 \usepackage{changepage}% \adjustwidth{<leftskip>}{<rightskip>} http://ctan.org/pkg/changepage
-\usepackage{lipsum}% http://ctan.org/pkg/lipsum
+\usepackage[T1]{fontenc} % Choose an output font encoding (T1) that has support for the accented characters used by the most widespread European languages
+\usepackage[utf8x]{inputenc} % Allow input of accented characters (and more...)
+\usepackage[usenames, dvipsnames]{xcolor} % to use color
 \usepackage[margin=1in]{geometry} % make the margins 1 inch on all sides of the document
+
+% font:
+% Chivo
+% \usepackage[familydefault,light]{Chivo}
+
+% Carlito:
+% \usepackage[sfdefault,lf]{carlito}
+%% The 'lf' option for lining figures
+%% The 'sfdefault' option to make the base font sans serif
+% \renewcommand*\oldstylenums[1]{\carlitoOsF #1}
+
+% roboto
+% \usepackage[sfdefault]{roboto}  %% Option 'sfdefault' only if the base font of the document is to be sans serif
+
+% Nimbus sans
+\usepackage{nimbussans}
+\renewcommand*\familydefault{\sfdefault} %% Only if the base font of the document is to be sans serif
 
@@
 
-\title{Response to Reviews}
+% define a new color
+\definecolor{peerjBlue}{RGB}{0,144,255} % also try G=167
 
-% You can make your own environment
+% Make your own environment for responses
 \newenvironment{response}
@@
 	\begin{adjustwidth}{1cm}{0cm}
-	\itshape % make italic
+	\color{peerjBlue}
+	% \itshape % make italic
 	}
@@
 
+\title{Response to Reviews}
+
 \maketitle
 
-\noindent
-{
-Manuscript Title:
+% \noindent
+% {
+\begin{tabular}{ l l }
+Manuscript Title: &
 Spatial distribution of environmental DNA in a nearshore marine habitat
 \\
-Journal:
+Journal: &
 PeerJ
 \\
-Manuscript ID:
+Manuscript ID: &
 14649
 \\
-Authors:
-James L. O'Donnell,
-Ryan P. Kelly,
-Andrew O. Shelton,
-Jameal F. Samhouri,
-Natalie C. Lowell,
-Gregory D. Williams
+Authors: &
+O'Donnell JL,
+Kelly RP,
+Shelton AO,
+Samhouri JF,
+Lowell NC,
+Williams GD
 \\
-}
+\end{tabular}
+
+\bigskip
+% }
 
 We are grateful to the editor and reviewers for their insightful feedback.
-Reviewer feedback is below in normal font, with our responses in indented italics.
-Additional changes to the manuscript independent of the reviews are presented at the end of this document.
+Reviewer feedback is below in regular typeface, with our responses indented and colored.
+% Additional changes to the manuscript independent of the reviews are presented at the end of this document.
 
@@
 The raw data and the analysis scripts are all publicly available.
+\begin{response}
+	Thank you for the positive feedback.
+	We feel the introduction is appropriately detailed to put our study in the context of both the field of metabarcoding and the field of spatial ecology.
+	We strove for excruciating detail in the methods section as a counter to the chronic under-reporting of details in the metabarcoding field, which hinders reproducible research.
+	We took your later suggestions into account, but please note that the other reviewer requested more analytical detail in the methods.
+	We would happily move extraneous details to the supplement if the editor feels that manuscript length is an issue as written (e.g. molecular laboratory details such as PCR parameters).\\
+\end{response}
 
@@
 The laboratory methods are rigorous and the analysis pipeline includes steps for assuring excellent quality of the resulting data set.
+\begin{response}
+	Thank you; we agree.\\
+\end{response}
 
@@
 The statistical analysis of the data are very clearly explained and the results are highly trustworthy.
+\begin{response}
+	Thank you; we agree.\\
+\end{response}
 
@@
 I think that the introduction and material and methods sections are a little too long, but they can be enhanced by deleting some sentences (detailed below).
+\begin{response}
+	Thank you; please see our previous comment on manuscript length, as well as our responses to your specific comments on extraneous details.\\
+\end{response}
 
@@
 the issues with the assignment of many sequences which can be assigned only to the family or lower levels), compared to other more variable markers with denser reference databases, such as COI, which, in most cases, allow assignments to the genus or species level.
-
 \begin{response}
-	We agree with your point that a locus with only coarse-scale resolution (16S) may obfuscate patterns of diversity compared to a locus with fine-scale resolution (CO1).
-A subset of us has recently published an in-depth comparison of multiple loci for metabarcoding, and TODO: we have added a more substantial comment on this and referenced to this now-published work.
+	We agree with your point that a locus with only coarse-scale resolution (16S) may obfuscate patterns of diversity compared to a locus with fine-scale resolution.
+A subset of us has recently published an in-depth comparison of multiple loci for metabarcoding, and we have added a more substantial comment on this and referenced to this now-published work.
 \\
@@
 \begin{response}
-  We did not state that the microscopic plankton is composed solely of diatoms and larve, we made a  statement and qualified it with two examples to give meaning and context: `a diverse assemblage of microscopic plankton including diatoms and larvae'.
+  We did not mean to imply that the microscopic plankton is composed solely of diatoms and larve, we made a  statement and qualified it with two examples to give meaning and context: `a diverse assemblage of microscopic plankton including diatoms and larvae'.
 	We mention diatoms because they are a widely-known, numerically abundant, and ecologically important group that falls outside the scope of this study on metazoans.
-We mention larvae because it subtly hints at our major finding that our samples captured many organisms best known for their sessile, benthic stages (e.g.
-barnacles).
-Yes, we could add copepods to this list if you would like, but what qualifies a group for inclusion?
-Numerical abundance?
-Biomass?\\
+	We mention larvae because it subtly hints at our major finding that our samples captured many organisms best known for their sessile, benthic stages (e.g. barnacles).
+	Yes, we could add copepods to this list if you would like, but what qualifies a group for inclusion? Numerical abundance? Biomass?
+	Still, if you think the statement is somehow misleading as written, we prefer to omit any examples.\\
 \end{response}
@@
 2015, Hanfling 2016, Guardiola 2016 PeerJ e.g.
-`while other studies have documented the spatial diversity of meiofauna (Fonseca 2014), '
+`While other studies have documented the spatial diversity of meiofauna in sediments (Fonseca 2014), '
 \end{response}
@@
   While this does seem like overkill, it is not unheard of for machines or sequencing facilities to develop problems that can bias data, and we included this with the intent of being as transparent as possible.
-Nevertheless, we made this change.\\
+Nevertheless, we removed this information.\\
 \end{response}
@@
 \begin{response}
-  We are grateful to have a reviewer reading with such a careful eye!
-We believe both statements are correct, though subtly so.
+  We are grateful that you have read the manuscript with such a careful eye!
+	Both statements are correct for subtle reasons, and we have incorporated your phrasing to our explanation.
 	Because PhiX increases sequence diversity, its effect is to increase the number of clusters which can be correctly differentiated from one another on the flowcell, thus increasing sequencing depth.
@@
 \begin{response}
-  You have made an excellent point that we were not clear in our wording. \\
+  You make an excellent point that we were not clear in our wording.
+	We did not mean to suggest that we made only one general estimate of dispersal per taxon; rather, we made separate estimates for each stage.
+	That is, even if a species (e.g.
+Balanus glandula) is internally fertilized and has sessile adults in the intertidal, if it has a larval stage with high potential for dispersal, we cannot rule out an expectation that it will be most abundant in the sequence data from samples far from shore due to larval abundance.
+	We have attempted to clarify this in the text.
+	That said, you are also correct that this data is almost unused except to single out Cymatogaster aggregata and Cupolaconcha meroclista as taxa for which an a priori expectation based on adult distribution should be met.
+	We added a concluding sentence to the paragraph to explain the reason for these data.
+	We chose to keep the description here to show that we have been diligent in investigating potential factors obscuring taxon-specific patterns.
+	\\
 \end{response}
@@
 \begin{response}
-  Our intent was to convey the novelty of the sampling design. We have clarified the novelty of our study in the context of recently published work in the introduction.\\
+  Our intent was to convey the novelty of the sampling design.
+	We removed `For the first time in a marine environment', and we have clarified the novelty of our study in the context of recently published work in the introduction.\\
 	% TODO: Comment on the novelty of this study in the context of refs: Fonseca et al.
@@
 \subsubsection*{Basic reporting}
-There were a couple of minor grammatical errors in the manuscript.
-1) The final sentence (L292-294) of the ``Sequence Data Processing (Bioinformatics)'' paragraph on page 12 in the Results section.
+There were a couple of minor grammatical errors in the manuscript. 1) The final sentence (L292-294) of the ``Sequence Data Processing (Bioinformatics)'' paragraph on page 12 in the Results section.
+\begin{response}
+  We corrected this.\\
+\end{response}
+
 2) The first sentence (L359) of the ``Communities far from one another tend to be less similar than those that are nearby'' paragraph on page 15 of the Discussion section.
 \begin{response}
-  OurResponse\\
+	%TODO email Steve about this
+  It is unclear what the specific error is, but we acknowledge this sentence is poorly worded.
+	We changed this to read `We demonstrate that distant locations have less-similar eDNA communities than proximate locations in Puget Sound, a dynamic marine environment. '\\
 \end{response}
@@
 \begin{response}
+	% TODO: edit legend for figure in `Analysis/diversity.R`
   OurResponse\\
@@
 \begin{response}
-  Yes; these links are a placeholder until the manuscript is accepted.\\
+  Yes; these links were a placeholder until the manuscript is accepted.\\
 	% TODO: include EMBL/EBI url
@@
 \begin{response}
-  OurResponse\\
+  Perhaps we misunderstand your comment, but in the relevant part of the methods section, we state: `Each environmental sample was collected in a clean 1 liter high-density polyethylene bottle, the opening of which was covered with 500 micrometer nylon mesh to prevent entry of larger particles.'\\
 \end{response}
@@
 \begin{response}
+	%TODO email Steve about this
+	% TODO: Add outline of the workflow steps of banzai
   OurResponse\\
@@
 \begin{response}
+	% TODO: Add reference to tidal flushing and dilution as a possible mechanism for diversity from shore
   OurResponse\\
@@
 \begin{response}
+	% TODO: report alpha diversity and rarefaction analysis to demonstrate effective sequence depth had been achieved
   OurResponse\\
@@
 
-\subsubsection{Comments for the Author}
+\subsubsection*{Comments for the Author}
 I thought the paper was very interesting.
@@
 \begin{response}
+	% TODO: provide recommendations on how some of the limitations in eDNA sampling could be overcome
   OurResponse\\
@@
 \begin{response}
-  OurResponse\\
+	Thank you; we agree.\\
 \end{response}
@@
 %------------------------------------REVIEWER3----------------------------------
-\subsection*{Comment on the preprint (Bernd Hänfling)}
+\subsection*{Comment on the preprint (Bernd Hänfling, Lori Lawson Handley, Dan Read and Ian Winfield)}
 This is well designed study and we would like to congratulate the authors to this data set and its excellent analysis.
-However, we feel that the authors have made an important omission in the discussion.
+However, we feel that the authors have made an important omission in the discussion.\\
 
@@
 Our paper confirms that the spatial distribution of individual species reflects their abundance but also shows that the pattern of distribution is not random and ecologically meaningful.
-Although there might be important differences between freshwater and marine habitats we feel that the results from our study are extremely relevant here and should not be omitted especially when claiming novelty of discussion point 2.
-
-Bernd Hänfling, Lori Lawson Handley, Dan Read and Ian Winfield
+Although there might be important differences between freshwater and marine habitats we feel that the results from our study are extremely relevant here and should not be omitted especially when claiming novelty of discussion point 2.\\
 
@@
 \begin{response}
-  OurResponse\\
+	% http://onlinelibrary.wiley.com/doi/10.1111/mec.13660/full
+  Thank you for your comment and for drawing our attention to your paper. As you note therein, the rate of publication in this field can make it difficult to keep up.
+	% We acknowledge that your study took place in a spatial context becuase you analyzed samples collected from transects in a lake.
+	% Likewise, we note that you also found a non-
+	% (Note that in the site map, references to Derwent and Windermere are mistakenly switched.)
+	% Despite your admirably extensive sampling, we contest that your study explicity addresses spatial patterns in community eDNA.
+	% First,
+	% In your paper, you note that `the results regarding vertical sampling should be regarded as preliminary' and `Our site occupancy estimates should therefore be treated as preliminary'
+	% Yet these site occupancy estimates are the data on which the
+	% Note that even for the mock community, we found no formal analysis of whether read count or relative abundance was correlated with input DNA mass.
+	% Sampling design complicates the relationship between distance and community similarity because depth and distance are confounded.
+	% no true replication of PCR because PCR replicates were pooled prior to sequencing
+	% In the relevant section of this paper (`Spatial distribution of eDNA records within Windermere'), we could find no formal analysis of spatial distribution, only references to Figure 5 and S14,
+	% In figure `Fig. S15: Spatial Distribution of species recorded in Windermere North and South Basins at more than 2 sites (Site Occupancy) using 12S data',
+	% relevant section: Site occupancy data based on 12S sequences were used to investigate the spatial distribution of each species recorded at more than two sites around Windermere (Fig.
+S15, Supporting information).
+The general pattern emerging from this analysis is that species-specific eDNA was not evenly distributed around the lake.
+Although some species such as P.
+fluviatilis, R.
+rutilus, E.
+lucius and S.
+trutta are recorded almost ubiquitously within the lake, eDNA from other species is predominantly found in one of the two basins.
+S.
+alpinus, P.
+phoxinus and G.
+aculeatus eDNA was common in the North Basin but very rare in the South Basin, whereas A.
+brama and A.
+anguilla eDNA was more common in South Basin (Fig.
+S15, Supporting information).
+Overall, the relative proportion of sequence read counts for different species across sample sites was significantly different between Windermere North and South Basins (χ2 = 47 817; df = 13; P < 0.001 and χ2 = 134750; df = 11; P < 0.001 for 12S and CytB, respectively, Fig.
+6a, b).
+A similar pattern was observed for the relative proportion of sites occupied (χ2 = 61.43; df = 13; P < 0.001 and χ2 = 48.65; df = 11; P < 0.001 for 12S and CytB, respectively, Fig.
+6c, d).
+Distribution of eDNA in the two Windermere basins reflected the expected association between species and ecological condition.
+eDNA from species associated with eutrophic conditions (R.
+rutilus, T.
+tinca, S.
+erythrophthalmus, A.
+brama and A.
+anguilla) was more abundant in the South than North Basin, while eDNA from species that prefer less eutrophic conditions (S.
+salar, S.
+trutta, S.
+alpinus, P.
+phoxinus and C.
+gobio) was more abundant in the North than South Basin (Fig.
+6).
+	\\
 \end{response}

### 2017-01-06 Manuscript/v1.1_PeerJ/carkeek_grid.tex
respond to reviews

@@
 \affil[2]{Earth Resource Technology, Inc., Under contract to the Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, 2725 Montlake Blvd E, Seattle, WA 98112, USA}
-\affil[3]{Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, 2725 Montlake Blvd E, Seattle, Washington 98112, USA}
+\affil[3]{Conservation Biology Division, Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, 2725 Montlake Blvd E, Seattle, Washington 98112, USA}
 \affil[4]{School of Aquatic and Fishery Sciences, University of Washington, 1122 NE Boat St, Seattle, Washington 98105, USA}
@@
 We collated life history characteristics for each of the major taxonomic groups recovered, including dispersal range of the gametes, larvae, and adults, adult habitat type and selectivity, and adult body size.
-For each life history stage, we made an order-of-magnitude approximation of the scale of dispersal: for example, internally fertilized species were assigned a gamete range of 0 km, while broadcast spawners were assigned a gamete range of 10 km.
+For each life history stage of each taxon group, we made an order-of-magnitude approximation of the scale of dispersal.
+For example, internally fertilized species were assigned a gamete range of 0 km, while broadcast spawners were assigned a gamete range of 10 km.
 Similarly, adult range size was approximated as 0 km (sessile), 1 km (motile but not pelagic), or 10 km (highly mobile, pelagic).
-Variables were specified as 'multiple' for groups known to span more than 1 magnitude of range size.
+Variables were specified as 'multiple' for life history stages known to span more than 1 magnitude of range size.
 For groups to which sequences were annotated with high confidence, but for which life history strategy is diverse or poorly known (e.g.
 families in the phylum Nemertea), we used conservative, coarse approximations at a higher taxonomic rank (see \nameref{life_history_data}).
+These data were used to contextualize group-specific spatial distributions and inform expectations based on known adult distributions.
 
@@
 A single sample had low similarity among PCR replicates (0.659) after removing this outlier, the lowest mean similarity among replicates within a sample was 0.966.
-Overall similarities among PCR replicates within a sample were extremely high (0.976 $\pm$ 0.013), and far higher than that of than among samples (0.3 $\pm$ 0.16).
+Overall similarities among PCR replicates within a sample were extremely high (0.976 $\pm$ 0.013), and far higher than those among samples (0.3 $\pm$ 0.16).
 
@@
 \subsubsection*{(1) Communities far from one another tend to be less similar than those that are nearby}
-We demonstrate that more distant locations have less similar eDNA communities than more proximate locations in Puget Sound, a dynamic marine environment.
+We demonstrate that distant locations have less-similar eDNA communities than proximate locations in Puget Sound, a dynamic marine environment.
 Our finding is in line with observations based on traditional surveys of terrestrial plants and fungi \citep{Nekola1999, Bahram2013, Condit2002, Chust2006} and of microorganisms in freshwater \citep{Wetzel2012}, marine \citep{Chust2013}, and estuarine \citep{Martiny2011} environments.
@@
 
+The marker we chose for this study detects a wide variety of metazoans while excluding other more common taxa; however, it does not effectively discriminate among species within a higher group in all cases.
+Other markers, such as mitochondrial cytochrome c oxidase subunit 1 (COX1, CO1, or COI) provide adequate species-level resolution in some metazoan groups, but have other shortcomings including taxon dropout (Deagle2014) and amplification of more abundant non-metazoans, as we discovered in an accompanying study (Kelly2016FMS).
+Both have undesirable effects of biasing estimates of diversity: in our case, it is possible that the lumping of multiple species into one group underestimates the true richness of the group and of the entire sample.
+In turn, this could obscure true underlying patterns of diversity.
+In the case of COX1, well-documented primer biases cause failure to amplify some taxa, particularly in mixed samples, with the same result.
+In fact, even surveys relying on traditional capture techniques (e.g.
+seine nets) and morphological characteristics are subject to biases imposed by the sampling gear (e.g.
+mesh size), the observer (e.g.
+taxonomic expertise), and organisms (e.g.
+morphologically cryptic species).
+Until whole-genome sequencing of individual cells is a reality, the tradeoffs between taxonomic breadth and resolution will continue to be problematic for metabarcoding studies, just as they are for more traditional ecological survey methods (Kelly2016FMS).
+
 Our results also highlight the need for curated life-history databases.
@@
 We wish to thank Robert Morris, E. Virginia Armbrust, and James Kralj.
+The manuscript was improved by suggestions from Magnus Johnson, Owen Wangensteen, and Stephen Moss.
 

### 2017-01-07 Manuscript/v1.1_PeerJ/carkeek_grid.tex
add paragraph on marker choice to discussion

@@
 The marker we chose for this study detects a wide variety of metazoans while excluding other more common taxa; however, it does not effectively discriminate among species within a higher group in all cases.
-Other markers, such as mitochondrial cytochrome c oxidase subunit 1 (COX1, CO1, or COI) provide adequate species-level resolution in some metazoan groups, but have other shortcomings including taxon dropout (Deagle2014) and amplification of more abundant non-metazoans, as we discovered in an accompanying study (Kelly2016FMS).
+Other markers, such as mitochondrial cytochrome c oxidase subunit 1 (COX1, CO1, or COI) provide adequate species-level resolution in some metazoan groups, but have other shortcomings including taxon dropout \citep{Deagle2014} and amplification of more abundant non-metazoans, as we discovered in an accompanying study \citep{Kelly2016FMS}.
 Both have undesirable effects of biasing estimates of diversity: in our case, it is possible that the lumping of multiple species into one group underestimates the true richness of the group and of the entire sample.
@@
 morphologically cryptic species).
-Until whole-genome sequencing of individual cells is a reality, the tradeoffs between taxonomic breadth and resolution will continue to be problematic for metabarcoding studies, just as they are for more traditional ecological survey methods (Kelly2016FMS).
+Until whole-genome sequencing of individual cells is a reality, the tradeoffs between taxonomic breadth and resolution will continue to be problematic for metabarcoding studies, just as they are for more traditional ecological survey methods \citep{Kelly2016FMS}.
 

### 2017-01-07 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex
strip comments
respond to preprint comment

@@
 \usepackage[margin=1in]{geometry} % make the margins 1 inch on all sides of the document
-
-% font:
-% Chivo
-% \usepackage[familydefault,light]{Chivo}
-
-% Carlito:
-% \usepackage[sfdefault,lf]{carlito}
-%% The 'lf' option for lining figures
-%% The 'sfdefault' option to make the base font sans serif
-% \renewcommand*\oldstylenums[1]{\carlitoOsF #1}
-
-% roboto
-% \usepackage[sfdefault]{roboto}  %% Option 'sfdefault' only if the base font of the document is to be sans serif
 
@@
   Thank you for the positive feedback.
-	Indeed, we agree with the comment made on the preprint. Our omission was not intended as a slight, merely an oversight given the rapid pace of publication in this field.
-	We have added the appropriate citations and corrected our wording.
-	% TODO: add hanfling citation, respond to comment, update context in paper.
-	We were also curious about the effect of the data subset you mention, and conducted that analysis.
+	Indeed, we agree with the comment made on the preprint.
+	Our omission was not intended as a slight, merely an oversight given the rapid pace of publication in this field.
+	We have serious qualms about the quality of the data in their paper that we believe preclude us from recommending it as a study of spatial variation in eDNA.
+	However, the suggestions of reviewer 1 (Wangensteen) were justified and we have added the appropriate citations and corrected our wording.
+	% TODO:  respond to comment, update context in paper.
+	\\
+
+	Like you, we were curious about the effect of the data subset you mention, and conducted that analysis using the life history data.
 Unfortunately, we found no meaningful patterns, and have made an attempt to clarify this in the text.
@@
 \begin{response}
-  Our intent was to convey the novelty of the sampling design.
+  Our intent was precisely to convey the novelty of the sampling design; we failed to do that clearly.
 	We removed `For the first time in a marine environment', and we have clarified the novelty of our study in the context of recently published work in the introduction.\\
@@
 %------------------------------------REVIEWER3----------------------------------
-\subsection*{Comment on the preprint (Bernd Hänfling, Lori Lawson Handley, Dan Read and Ian Winfield)}
+\subsection*{Comment on the preprint}
+
+\subsubsection*{Bernd Hänfling, Lori Lawson Handley, Dan Read and Ian Winfield}
+
 This is well designed study and we would like to congratulate the authors to this data set and its excellent analysis.
@@
 	% http://onlinelibrary.wiley.com/doi/10.1111/mec.13660/full
-  Thank you for your comment and for drawing our attention to your paper. As you note therein, the rate of publication in this field can make it difficult to keep up.
+  Thank you for your comment and for drawing our attention to your paper.
+	As you note therein, the rate of publication in this field can make it difficult to keep up.
+	Our omission was not intended as a slight, merely an oversight. % given the rapid pace of publication in this field.
 	% We acknowledge that your study took place in a spatial context becuase you analyzed samples collected from transects in a lake.
-	% Likewise, we note that you also found a non-
-	% (Note that in the site map, references to Derwent and Windermere are mistakenly switched.)
-	% Despite your admirably extensive sampling, we contest that your study explicity addresses spatial patterns in community eDNA.
-	% First,
+	We commend your extensive environmental sampling; however, upon reading your paper, we have reservations about your interpretations of the results given the molecular protocols. % are reluctant to
+  \\
+
+	We note that you conducted replicate PCRs of each environmental sample, and pooled them prior to sequencing.
+	While this practice is not uncommon in the field, it precludes the ability to test for any effects of PCR on the communities.
+	We believe this is particularly problematic in the case of your study because, as far as we understand your methods, you used indexed ('tagged') primers in a single PCR reaction.
+	We have shown previously that this can bias datasets in such a way that completely obscures any underlying pattern (see: http://dx.doi.org/10.1371/journal.pone.0148698).
+	As we report in that study, indexed primers can bias relative read abundance by as much as 70\%.
+	That is, in one environmental sample, a species may comprise 70\% of the reads in PCR replicates amplified with one indexed primer set, while being completely absent from replicates amplified by another indexed primer set.
+	This would most certainly result in a significant difference among most any comparison, such as the Chi-squared test you conducted to compare lake basins.
+	If we have misunderstood your methodology, or if you have results of replication tests on the indexed primers in community samples, we would be happy to consider further the results of your study.
+  \\
+
+	Regardless, we have neutralized our statements on the novelty of our work on the basis of points brought up by one of the reviewers.
+
 	% In your paper, you note that `the results regarding vertical sampling should be regarded as preliminary' and `Our site occupancy estimates should therefore be treated as preliminary'
@@
 	% Sampling design complicates the relationship between distance and community similarity because depth and distance are confounded.
-	% no true replication of PCR because PCR replicates were pooled prior to sequencing
-	% In the relevant section of this paper (`Spatial distribution of eDNA records within Windermere'), we could find no formal analysis of spatial distribution, only references to Figure 5 and S14,
-	% In figure `Fig. S15: Spatial Distribution of species recorded in Windermere North and South Basins at more than 2 sites (Site Occupancy) using 12S data',
-	% relevant section: Site occupancy data based on 12S sequences were used to investigate the spatial distribution of each species recorded at more than two sites around Windermere (Fig.
-S15, Supporting information).
-The general pattern emerging from this analysis is that species-specific eDNA was not evenly distributed around the lake.
-Although some species such as P.
-fluviatilis, R.
-rutilus, E.
-lucius and S.
-trutta are recorded almost ubiquitously within the lake, eDNA from other species is predominantly found in one of the two basins.
-S.
-alpinus, P.
-phoxinus and G.
-aculeatus eDNA was common in the North Basin but very rare in the South Basin, whereas A.
-brama and A.
-anguilla eDNA was more common in South Basin (Fig.
-S15, Supporting information).
-Overall, the relative proportion of sequence read counts for different species across sample sites was significantly different between Windermere North and South Basins (χ2 = 47 817; df = 13; P < 0.001 and χ2 = 134750; df = 11; P < 0.001 for 12S and CytB, respectively, Fig.
-6a, b).
-A similar pattern was observed for the relative proportion of sites occupied (χ2 = 61.43; df = 13; P < 0.001 and χ2 = 48.65; df = 11; P < 0.001 for 12S and CytB, respectively, Fig.
-6c, d).
-Distribution of eDNA in the two Windermere basins reflected the expected association between species and ecological condition.
-eDNA from species associated with eutrophic conditions (R.
-rutilus, T.
-tinca, S.
-erythrophthalmus, A.
-brama and A.
-anguilla) was more abundant in the South than North Basin, while eDNA from species that prefer less eutrophic conditions (S.
-salar, S.
-trutta, S.
-alpinus, P.
-phoxinus and C.
-gobio) was more abundant in the North than South Basin (Fig.
-6).
-	\\
 \end{response}

### 2017-01-09 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex
typo
respond to reviews

@@
 \usepackage[usenames, dvipsnames]{xcolor} % to use color
-\usepackage[margin=1in]{geometry} % make the margins 1 inch on all sides of the document
+\usepackage[margin=1in]{geometry} % make the margins 1 inch on all sides
 
@@
   Thank you for the positive feedback.
-	Indeed, we agree with the comment made on the preprint.
-	Our omission was not intended as a slight, merely an oversight given the rapid pace of publication in this field.
+	Indeed, we agree with the comment made on the preprint that we overlooked some recently published works.
 	We have serious qualms about the quality of the data in their paper that we believe preclude us from recommending it as a study of spatial variation in eDNA.

### 2017-01-09 Manuscript/v1.1_PeerJ/carkeek_grid.tex
respond to reviews
more responding to reviews

@@
 While rigorous laboratory studies have investigated the effects of some environmental factors on eDNA persistence \citep{Klymus2015, Barnes2014, Sassoubre2016} and the transport of eDNA in specific contexts \citep{Deiner2014}, we suggest that field studies comparing the spatial distribution of communities of eDNA with expectations based on prior knowledge of organisms' distributions are also critical to developing a working understanding of eDNA in the real world.
+Research to date has documented the non-random spatial distribution of % spatial variation of
+meiofaunal \citep{Fonseca2014,Guardiola2016}, microbial \citep{Lallias2015}, and extracellular \citep{Guardiola2015} eDNA of marine and estuarine sediments, and of microscopic plankton in open ocean waters \citep{DeVargas2015}.
+These studies conducted targeted sampling at intermediate (tens of kilometers) to global (thousands of kilometers) scales.
+Here, we use a grid-based environmental sampling strategy to assess spatial variability of eDNA in a coastal marine environment at a fine scale (tens to thousands of meters), using molecular methods that focus on macrobial metazoans.
 
@@
 The marker we chose for this study detects a wide variety of metazoans while excluding other more common taxa; however, it does not effectively discriminate among species within a higher group in all cases.
-Other markers, such as mitochondrial cytochrome c oxidase subunit 1 (COX1, CO1, or COI) provide adequate species-level resolution in some metazoan groups, but have other shortcomings including taxon dropout \citep{Deagle2014} and amplification of more abundant non-metazoans, as we discovered in an accompanying study \citep{Kelly2016FMS}.
-Both have undesirable effects of biasing estimates of diversity: in our case, it is possible that the lumping of multiple species into one group underestimates the true richness of the group and of the entire sample.
-In turn, this could obscure true underlying patterns of diversity.
-In the case of COX1, well-documented primer biases cause failure to amplify some taxa, particularly in mixed samples, with the same result.
+Other markers, such as mitochondrial cytochrome c oxidase subunit 1 (COX1, CO1, or COI) may provide adequate species-level resolution in some metazoan groups, but have other shortcomings including taxon dropout \citep{Deagle2014} and amplification of more abundant non-metazoans, as we discovered in an accompanying study \citep{Kelly2016FMS}.
+Both have undesirable effects of biasing estimates of diversity.
+In our case, it is possible that the lumping of multiple species into one group underestimates the true richness of the group and of the entire sample, in turn obscuring true underlying patterns of diversity.
+In the case of COX1, well-documented primer biases cause failure to amplify some taxa, particularly in mixed samples, with the same result \citep{Deagle2014}.
 In fact, even surveys relying on traditional capture techniques (e.g.
@@
 morphologically cryptic species).
+Similarly, no single molecular marker adequately and effectively samples all taxa without bias  \citep{Drummond2015}, and thus the choice of marker is an important and context-dependent one.
 Until whole-genome sequencing of individual cells is a reality, the tradeoffs between taxonomic breadth and resolution will continue to be problematic for metabarcoding studies, just as they are for more traditional ecological survey methods \citep{Kelly2016FMS}.

### 2017-01-10 Manuscript/v1.1_PeerJ/carkeek_grid.tex
fix references

@@
 meiofaunal \citep{Fonseca2014,Guardiola2016}, microbial \citep{Lallias2015}, and extracellular \citep{Guardiola2015} eDNA of marine and estuarine sediments, and of microscopic plankton in open ocean waters \citep{DeVargas2015}.
-These studies conducted targeted sampling at intermediate (tens of kilometers) to global (thousands of kilometers) scales.
+These studies conducted targeted sampling at intermediate (thousands of meters) to global (thousands of kilometers) scales.
 Here, we use a grid-based environmental sampling strategy to assess spatial variability of eDNA in a coastal marine environment at a fine scale (tens to thousands of meters), using molecular methods that focus on macrobial metazoans.
@@
 One liter from each water sample was filtered in the lab on a clean polysulfone vacuum filter holder fitted with a 47 millimeter diameter cellulose acetate membrane with 0.45 micrometer pores.
-Filter membranes were moved into 900 microliters of Longmire buffer \citep{Longmire1997} using clean forceps and stored at room temperature \citep{Renshaw2014}.
+Filter membranes were moved into 900 microliters of Longmire buffer \citep{Longmire1997} using clean forceps and stored at room temperature \citep{Renshaw2015}.
 To test for the extent of contamination attributable to laboratory procedures, we filtered three replicate 1 liter samples of deionized water.
@@
 \subsection*{DNA Purification}
-DNA was purified from the membrane following a phenol:chloroform:isoamyl alcohol protocol following Renshaw \citep{Renshaw2014}.
+DNA was purified from the membrane following a phenol:chloroform:isoamyl alcohol protocol following Renshaw \citep{Renshaw2015}.
 Preserved membranes were incubated at 65 ºC for 30 minutes before adding 900 microliters of phenol:chloroform:isoamyl alcohol and shaking vigorously for 60 seconds.

### 2017-01-10 Analysis/diversity.R
fix figure legend

@@ -15,7 +15,11 @@ which_data <- "mean"
 
 plot_name <- "diversity"
 if(!exists("legend_text")){legend_text <- list()}
-legend_text[plot_name] <- {"Aggregate measures of diversity at each sample site."}
+legend_text[plot_name] <- {
+"Aggregate measures of diversity at each sample site.
+Data are rarefied counts of mitochondrial 16S sequences collected from 3 parallel transects in Puget Sound, Washington, USA.
+Evenness (left) is the probability that two sequences drawn at random are identical by chance alone; richness (right) represents the total number of unique sequences from that location."
+}
 
 if(EXPORT){
   pdf_file    <- file.path(fig_dir, paste(plot_name, ".pdf", sep = ""))

### 2017-01-10 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex
more responding to reviews
add data link
respond to reviews

@@
   Indeed, our wording did not clearly convey our meaning.
-Our intent was to indicate the novelty of the experimental design which was intended to capture variability in transport of eDNA.
+Our intent was to indicate the novelty of the experimental design which was intended to capture relatively fine-scale spatial distribution of eDNA in marine waters.
 We have made the suggested change.\\
-	% TODO: Comment on the novelty of this study in the context of refs: Fonseca et al.
-2014, Guardiola et al.
-2015 (PLOSONE), Lallias et a.
-2015, De Vargas et al.
-2015, Hanfling 2016, Guardiola 2016 PeerJ e.g.
-`While other studies have documented the spatial diversity of meiofauna in sediments (Fonseca 2014), '
 \end{response}
@@
 	We removed `For the first time in a marine environment', and we have clarified the novelty of our study in the context of recently published work in the introduction.\\
-	% TODO: Comment on the novelty of this study in the context of refs: Fonseca et al.
-2014, Guardiola et al.
-2015 (PLOSONE), Lallias et a.
-2015, De Vargas et al.
-2015, Hanfling 2016, Guardiola 2016 PeerJ
 \end{response}
@@
 \begin{response}
-	% TODO: edit legend for figure in `Analysis/diversity.R`
-  OurResponse\\
+  Thank you for noticing this; we have corrected the legend to be meaningful as a standalone figure.\\
 \end{response}
@@
   Yes; these links were a placeholder until the manuscript is accepted.\\
-	% TODO: include EMBL/EBI url
 \end{response}

### 2017-01-11 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex
more responding to reviews

@@
 \usepackage[margin=1in]{geometry} % make the margins 1 inch on all sides
+\usepackage{hyperref}
 
@@
 We are grateful to the editor and reviewers for their insightful feedback.
-Reviewer feedback is below in regular typeface, with our responses indented and colored.
+Reviewer feedback is below in regular typeface, with our responses indented and colored, including references to the line numbers in the track changes version of the manuscript.
 % Additional changes to the manuscript independent of the reviews are presented at the end of this document.
@@
 	However, the suggestions of reviewer 1 (Wangensteen) were justified and we have added the appropriate citations and corrected our wording.
-	% TODO:  respond to comment, update context in paper.
 	\\
@@
 Perhaps some clarity on this earlier on would be helpful?
-Perhaps also discussing the reasoning behind the decision to use 16S rDNA from the mtDNA rather than 18S rDNA would be helpful (e.g.
-Tang et.
-al., 2012; http://www.pnas.org/content/109/40/16208)?
 \begin{response}
-  OurResponse\\
+  This is a good suggestion; we have modified the abstract to clarify this as early as possible.\\
+\end{response}
+
+Perhaps also discussing the reasoning behind the decision to use 16S rDNA from the mtDNA rather than 18S rDNA would be helpful (e.g. Tang et. al., 2012; http://www.pnas.org/content/109/40/16208)?
+\begin{response}
+  We added a brief explanation of our marker choice to the section Methods:PCR Amplification, and a more general discussion of marker choice to the discussion.\\
 \end{response}
@@
 \begin{response}
-  OurResponse\\
+	We had a similar first reaction to the results of the classification analysis: \textit{Do these correspond to known ecological communities?}
+  The utility of clustering methods is that they simplify the complexity of highly dimensional data; in our case, hundreds to thousands of OTUs, depending on the data subset.
+	While additional analyses, such as PCA on a reduced-dimensionality dataset, may help identify which OTUs driving community differentiation, this result will not necessarily correspond to the clusters reported using PAM, especially considering the uncertainty regarding the appropriate number of clusters in our data.
+	In short, we are unaware of an analytical or visual method for further simplification of the classification results.
+	If you feel this is imperative information, we could potentially modify our \href{https://github.com/jimmyodonnell/Carkeek_eDNA_grid/blob/master/Figures/stacked_bar_community.pdf}{stacked bar plot} in order to indicate the cluster to which each sample was assigned, though making sense of this result in a spatial context is not straightforward.
+	\\
 \end{response}
@@
 \begin{response}
-  Yes; these links were a placeholder until the manuscript is accepted.\\
+  Yes; these links were a placeholder until the manuscript is accepted and have now been updated.\\
 \end{response}
@@
 	%TODO email Steve about this
-	% TODO: Add outline of the workflow steps of banzai
-  OurResponse\\
+  In our initial draft, we provided an outline of the bioinformatics procedures in the section Methods:Sequence Data Processing (Bioinformatics), and document the procedures in painstaking detail in the Supplemental Material section Methods:Bioinformatics.
+	However, we overlooked the inclusion of a direct link to the scripts used to process the raw sequence data, and have now added those.
+	We certainly appreciate any encouragement to move towards more reproducible research, though we are not yet where we would like to be.
+	Any specific recommendations are welcome.
+	\\
 \end{response}
@@
 \begin{response}
-	% TODO: Add reference to tidal flushing and dilution as a possible mechanism for diversity from shore
-  OurResponse\\
+  This is an excellent point. We have modified the relevant section to incorporate this potential mechanism.
+	\\
 \end{response}

### 2017-01-11 Manuscript/v1.1_PeerJ/carkeek_grid.tex
add data link
more responding to reviews

@@
 In order to identify the effective spatial scale of this approach in a dynamic marine environment, we collected marine surface water samples from transects ranging from the intertidal zone to 4 kilometers from shore.
-Using massively parallel sequencing of 16S amplicons, we identified a diverse community of metazoans and quantified their spatial patterns using a variety of statistical tools.
+Using PCR primers that target a diverse assemblage of metazoans, we amplified a region of mitochondrial 16S rDNA from the samples and sequenced the products on an Illumina platform in order to detect communities and quantify their spatial patterns using a variety of statistical tools.
 We find evidence for multiple, discrete eDNA communities in this habitat, and show that these communities decrease in similarity as they become further apart.
@@
 \subsection*{PCR Amplification}
-From each DNA sample, we amplified an approximately 115 base pair (bp) region of the mitochondrial gene encoding 16S RNA using a two-step polymerase chain reaction (PCR) protocol described by \citet{ODonnell2016}.
+We chose a primer set that amplifies an approximately 115 base pair (bp) region of the mitochondrial 16S rRNA gene in at least 10 metazoan phyla from this habitat, excludes non-metazoans, and resolves taxonomy to the family level in most cases using a public sequence database \citep{Kelly2016FMS}.
+We used a two-step polymerase chain reaction (PCR) protocol described by \citet{ODonnell2016} to generate 4 replicate products from each DNA sample.
 In the first set of reactions, primers were identical in every reaction (forward: AGTTACYYTAGGGATAACAGCG; reverse: CCGGTCTGAACTCAGATCAYGT); primers in the second set of reactions included these same sequences but with 3 variable nucleotides (NNN) and an index sequence on the 5$'$ end (see \nameref{sequencing_metadata}).
@@
 It would be useful to better understand such features during the period of sampling, by way of oceanographic monitoring devices.
+Finally, the uncertainty in identification of the number of distinct clusters to best characterize the community underlines the difficulty of identifying community patterns with the number of taxonomic groups considered here.
+We suspect that the signature of eDNA from microscopic life-stages may explain our inability to easily detect spatial community level patterns that align with our initial expectations.
 
  \subsubsection*{(3) Richness declines and evenness increases with distance from shore}
-We detected a general pattern of declining richness and increasing evenness with increasing distance offshore.
+We found that richness declined while evenness increased with distance from shore.
 Such a pattern is consistent with many other ecosystems which show strong clines in diversity metrics over environmental gradients.
-The coastal ocean is a highly productive and diverse ecosystem \citep{Ray1988}.
 However, our study is novel in that it corroborates a cline well-known on macroscales for macrobiota on a much smaller spatial scale for microscopic animals, suggesting that there may be a self-similarity across scales in diversity patterning \citep{Levin1992}.
+The coastal ocean is a highly productive and diverse ecosystem, where biomass is concentrated most heavily along the bottom and shoreline \citep{Ray1988}.
+Our
 Intriguingly, the cline in diversity from inshore to offshore was not determined by shared changes in communities as one moved offshore; the classification analysis suggested a fair amount of differences among communities at a given offshore distance (Figure \ref{pam_in_space}).
-Furthermore, the uncertainty in identification of the number of distinct clusters to best characterize the community underlines the difficulty of identifying community patterns with the number of taxonomic groups considered here.
-We suspect that the signature of eDNA from microscopic life-stages may explain our inability to easily detect spatial community level patterns that align with our initial expectations.
 
@@
 \section*{Acknowledgements}
-We wish to thank Robert Morris, E. Virginia Armbrust, and James Kralj.
-The manuscript was improved by suggestions from Magnus Johnson, Owen Wangensteen, and Stephen Moss.
+We wish to thank Linda Park, Robert Morris, E. Virginia Armbrust, and James Kralj.
+The manuscript was improved by suggestions from editor Magnus Johnson, and reviewers Owen Wangensteen and Stephen Moss.
 
@@
 
-\subsection{Sequence Data}
+\subsection*{Sequence Data}
 \label{sequence_data}
-All sequence files and metadata are available from EMBL:\\ \verb!http://www.ebi.ac.uk/ena/data/view/FIXME!
+All sequence data are available from NCBI under BioProject PRJNA338801.\\
+Scripts to process raw sequence data into the contingency tables used for ecological analyses can be found at: \\
+\verb!https://github.com/jimmyodonnell/banzai!\\
 
-\subsection{Project Repository}
+\subsection*{Project Repository}
 The following components are available from the project repository on GitHub: \\
@@
 
-\subsubsection{Sequencing Metadata}
+\subsubsection*{Sequencing Metadata}
 \label{sequencing_metadata}
@@
 
-\subsubsection{Life History Data}
+\subsubsection*{Life History Data}
 \label{life_history_data}
@@
 
-\subsubsection{Analysis Scripts}
+\subsubsection*{Analysis Scripts}
 \label{analysis_scripts}
-All analyses were performed using scripts available in the Analysis subdirectory.
+All analyses were performed using scripts available in the Analysis subdirectory of the project's repository on GitHub.
 

### 2017-01-11 Manuscript/v1.1_PeerJ/carkeek_grid.tex
respond to reviews

@@
 The coastal ocean is a highly productive and diverse ecosystem, where biomass is concentrated most heavily along the bottom and shoreline \citep{Ray1988}.
-Our
+This differential in biomass concentration from the shoreline to open waters may contribute to the opposing trends we detected.
+Where particles (organisms, tissues, and cells) are sparse, fewer would be collected per sample of constant volume, thus decreasing the probability of drawing as many types (richness) and increasing the probability that any two particles originate from the same type (evenness).
 Intriguingly, the cline in diversity from inshore to offshore was not determined by shared changes in communities as one moved offshore; the classification analysis suggested a fair amount of differences among communities at a given offshore distance (Figure \ref{pam_in_space}).

### 2017-01-11 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex
add rarefaction

@@
 \begin{response}
-	% TODO: report alpha diversity and rarefaction analysis to demonstrate effective sequence depth had been achieved
-  OurResponse\\
+  We believe we adequately addressed alpha diversity by reporting richness and evenness both in a 2-dimensional and linear spatial context, with accompanying figures.
+	However, we have now also added an OTU accumulation figure to the supplement and commented on it in the results.\\
 \end{response}
@@
 	% TODO: provide recommendations on how some of the limitations in eDNA sampling could be overcome
-  OurResponse\\
+  Thank you for the kind words.	\\
 \end{response}

### 2017-01-11 Analysis/rarefaction.R

@@ -1,0 +1,52 @@
+# Calculate alpha diversity (species richness) and plot species accumulation curve
+library(vegan)
+
+EXPORT <- FALSE
+
+dat      <- otu_table[["clean"]]
+metadat  <- metadata[["clean"]]
+
+# sum the counts of each OTU across PCR replicates within a sample
+dat_reduced <- do.call(rbind, 
+  lapply(
+    split(as.data.frame(dat), metadat[,colname_env_sample]), 
+    colSums)
+)
+
+dat <- dat_reduced
+
+# change rownames for plotting
+rownames(dat) <- gsub("PCT-", "", rownames(dat))
+
+min_reads <- min(rowSums(dat))
+
+alpha_div <- rarefy(dat, sample = min_reads)
+
+rarefied  <- rrarefy(dat, sample = min_reads)
+
+plot_name <- "rarefaction"
+
+if(!exists("legend_text")){legend_text <- list()}
+legend_text[plot_name] <- {
+"Accumulation of OTUs from 24 environmental samples using randomized rarefaction.
+Four replicate PCRs were conducted using DNA each environmental sample and independently sequenced, but these are collapsed here to illustrate a single representation of richness.
+Sample names indicate the position in the sampling grid: south (S), central (C), or north (N), followed by the distance along the transect, in meters (0, 75, 125, 250, 500, 1000, 2000, 4000).
+Vertical line indicates the minimum combined number of sequence reads per sample.
+Horizontal lines indicate OTU richness for each sample at the minimum combined number of sequence reads."
+}
+
+if(EXPORT){
+  pdf_file    <- file.path(fig_dir, paste(plot_name, ".pdf", sep = ""))
+  legend_file <- file.path(fig_dir, paste(plot_name, "_legend.txt", sep = ""))
+  writeLines(legend_text[[plot_name]], con = legend_file)
+  pdf(file = pdf_file, width = 5, height = 4) #, width = 8, height = 3
+}
+
+par(mar = c(4,4,1,1))
+rarecurve(dat, step = 10000, sample = min_reads, 
+  xlab = "Number of sequences", ylab = "OTUs", col = "orchid")
+
+if(EXPORT){
+  dev.off()
+}
+

### 2017-01-12 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex
respond to reviews

@@
   Thank you for the positive feedback.
-	Indeed, we agree with the comment made on the preprint that we overlooked some recently published works.
-	We have serious qualms about the quality of the data in their paper that we believe preclude us from recommending it as a study of spatial variation in eDNA.
-	However, the suggestions of reviewer 1 (Wangensteen) were justified and we have added the appropriate citations and corrected our wording.
+	We thank the commenters for pointing out recent literature that we’d missed when developing the original manuscript.
+	We have now altered the language in our paper to avoid suggesting that our study is the first to do any kind of explicit spatial analysis, but (as noted below in the response to the commenter), because we have reservations about the way the referenced study was conducted, we have avoided citing it as an example.
+	However, the suggestions of reviewer 1 (Wangensteen) on this matter are well taken and we have added the appropriate citations and corrected our wording.
 	\\
@@
 Unfortunately, we found no meaningful patterns, and have made an attempt to clarify this in the text.
-	% TODO: comment on effect of subsetting based on life-history strategy or taxon
 	\\
@@
 \begin{response}
-	%TODO email Steve about this
   It is unclear what the specific error is, but we acknowledge this sentence is poorly worded.
@@
 \begin{response}
-	%TODO email Steve about this
   In our initial draft, we provided an outline of the bioinformatics procedures in the section Methods:Sequence Data Processing (Bioinformatics), and document the procedures in painstaking detail in the Supplemental Material section Methods:Bioinformatics.
@@
   We believe we adequately addressed alpha diversity by reporting richness and evenness both in a 2-dimensional and linear spatial context, with accompanying figures.
-	However, we have now also added an OTU accumulation figure to the supplement and commented on it in the results.\\
+	However, we have now also added an OTU accumulation figure to the supplement and commented on it in the results.
+	We note that our goal is not to recover an absolute measure of alpha diversity, but to make comparisons among sites for standardized sampling effort (i.e. sequencing depth).
+	\\
 \end{response}
@@
 \begin{response}
-	% TODO: provide recommendations on how some of the limitations in eDNA sampling could be overcome
-  Thank you for the kind words.	\\
+  Thank you for the kind words.
+	We acknowledge there are many limitations to eDNA sampling; however, we believe that making more general recommendations than those we provide in the dicussion is beyond the scope of this study.
+	The more general limitations and recommendations can and do receive regular, detailed attention in the many reviews published on this subject.
+	Of course, if there are specific limitations you feel we have erroneously overlooked and which are within the purview of our results, we are happy to revisit this issue.
+	\\
 \end{response}
@@
 	As you note therein, the rate of publication in this field can make it difficult to keep up.
-	Our omission was not intended as a slight, merely an oversight. % given the rapid pace of publication in this field.
-	% We acknowledge that your study took place in a spatial context becuase you analyzed samples collected from transects in a lake.
-	We commend your extensive environmental sampling; however, upon reading your paper, we have reservations about your interpretations of the results given the molecular protocols. % are reluctant to
+	% Our omission was not intended as a slight, merely an oversight. % given the rapid pace of publication in this field.
+	We commend your extensive environmental sampling; however, upon reading your paper, we have reservations about your interpretations of the results given the molecular protocols.
   \\
@@
 	While this practice is not uncommon in the field, it precludes the ability to test for any effects of PCR on the communities.
-	We believe this is particularly problematic in the case of your study because, as far as we understand your methods, you used indexed ('tagged') primers in a single PCR reaction.
+	We believe this is particularly problematic in the case of your study because, as far as we understand your methods, you used indexed (`tagged') primers in a single PCR reaction.
 	We have shown previously that this can bias datasets in such a way that completely obscures any underlying pattern (see: http://dx.doi.org/10.1371/journal.pone.0148698).
@@
 	This would most certainly result in a significant difference among most any comparison, such as the Chi-squared test you conducted to compare lake basins.
-	If we have misunderstood your methodology, or if you have results of replication tests on the indexed primers in community samples, we would be happy to consider further the results of your study.
+	% If we have misunderstood your methodology, or if you have results of replication tests on the indexed primers in community samples, we would be happy to consider further the results of your study.
   \\
 
-	Regardless, we have neutralized our statements on the novelty of our work on the basis of points brought up by one of the reviewers.
+	Regardless, we have neutralized our statements on the novelty of our work on the basis of the points raised by your comment and those of the reviewers.
 

### 2017-01-12 Manuscript/v1.1_PeerJ/carkeek_grid.tex
add rarefaction
respond to reviews

@@
 %-----------------------------------NOTES---------------------------------------
-% (TODO SUBMIT DATA AND METADATA; ADD URL)
 % (TODO INSERT DOI FROM FINAL COMMIT BEFORE PUBLICATION)
@@
 Overall similarities among PCR replicates within a sample were extremely high (0.976 $\pm$ 0.013), and far higher than those among samples (0.3 $\pm$ 0.16).
+Across PCR replicates, each sample was represented by at least 781425 reads in the raw data and contained between 111 and 443 rarefied OTUs (Supplemental Figure \ref{rarefaction}).
 
@@
 Finally, the two most abundant taxa in the dataset were the mussel genus \textit{Mytilus} and the Barnacle order Sessilia; the adults of both taxa are sessile and occur exclusively on hard intertidal substrata but have highly motile larvae.
+Because large-scale dispersal could not be ruled out for the vast majority of taxa, subsetting the community data by taxonomic group had no qualitative effect on the spatial patterning or diversity metrics, and we omit those results here.
 
@@
 
-\subsubsection*{(4) Spatial patterning of eDNA is associated with taxon-specific life history characteristics.}
+\subsubsection*{(4) Spatial patterning of eDNA is associated with taxon-specific life history}
 In contrast to our expectations, other taxa including species with sessile adult stages restricted to benthic hard substrates (e.g.
@@
 
+\begin{figure}[H] % [h!] forces the figure to be placed roughly here
+  \centering
+    \includegraphics[height=0.6\textheight]{rarefaction.pdf}
+    \caption{\protect\input{"../../Figures/rarefaction_legend.txt"}}
+  \label{rarefaction}
+\end{figure}
+
 % \pagebreak

### 2017-01-13 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex
add Jameal's edits, render track changes version

@@
 	We mention larvae because it subtly hints at our major finding that our samples captured many organisms best known for their sessile, benthic stages (e.g. barnacles).
-	Yes, we could add copepods to this list if you would like, but what qualifies a group for inclusion? Numerical abundance? Biomass?
-	Still, if you think the statement is somehow misleading as written, we prefer to omit any examples.\\
+	We added copepods to this list, though we recognize that it is necessarily incomplete.\\
 \end{response}
@@
 	That is, in one environmental sample, a species may comprise 70\% of the reads in PCR replicates amplified with one indexed primer set, while being completely absent from replicates amplified by another indexed primer set.
-	This would most certainly result in a significant difference among most any comparison, such as the Chi-squared test you conducted to compare lake basins.
+	This would very likely result in the detection of a significant difference by various statistical tests, such as the Chi-squared test you conducted to compare lake basins.
 	% If we have misunderstood your methodology, or if you have results of replication tests on the indexed primers in community samples, we would be happy to consider further the results of your study.

### 2017-01-13 Manuscript/v1.1_PeerJ/carkeek_grid.tex
add Jameal's edits, render track changes version

@@
 Hard bottoms of the lower intertidal and high subtidal are home to macroalgae such as Laminariales and Desmarestiales which provide habitat for a distinct community of fish and invertebrates.
-The upper pelagic is home to a diverse assemblage of microscopic plankton including diatoms and larvae \citep{Strickland1983}, as well as transitory fish and marine mammals.
+The upper pelagic is home to a diverse assemblage of microscopic plankton including diatoms, copepods, and larvae \citep{Strickland1983}, as well as transitory fish and marine mammals.
 

### 2017-01-13 Manuscript/v1.1_PeerJ/track_changes.sh

@@ -1,0 +1,4 @@
+#!/usr/bin/env bash
+
+latexdiff ../v1.0_PeerJ/carkeek_grid.tex carkeek_grid.tex > changes_tracked.tex
+

### 2017-01-13 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex
add line numbers to response to reviews

@@
 We are grateful to the editor and reviewers for their insightful feedback.
-Reviewer feedback is below in regular typeface, with our responses indented and colored, including references to the line numbers in the track changes version of the manuscript.
+Reviewer feedback is below in regular typeface, with our responses indented and colored.
+\textbf{Line numbers refer to the track changes version of the manuscript}.
 % Additional changes to the manuscript independent of the reviews are presented at the end of this document.
@@
 	Like you, we were curious about the effect of the data subset you mention, and conducted that analysis using the life history data.
-Unfortunately, we found no meaningful patterns, and have made an attempt to clarify this in the text.
+Unfortunately, we found no meaningful patterns, and have made an attempt to clarify this in the text (L356).
 	\\
@@
 	We agree with your point that a locus with only coarse-scale resolution (16S) may obfuscate patterns of diversity compared to a locus with fine-scale resolution.
-A subset of us has recently published an in-depth comparison of multiple loci for metabarcoding, and we have added a more substantial comment on this and referenced to this now-published work.
+A subset of us has recently published an in-depth comparison of multiple loci for metabarcoding, and we have added a more substantial comment on this and referenced to this now-published work (L452).
 \\
@@
 \begin{response}
-  Thanks for catching these errors. We made these changes.\\
+  Thanks for catching these errors. We made these changes (L99, L100).\\
 \end{response}
@@
 \begin{response}
-  Excellent suggestion; we made this change.\\
+  Excellent suggestion; we made this change (L98).\\
 \end{response}
@@
 \begin{response}
-  We made this change.\\
+  We made this change (L102).\\
 \end{response}
@@
 	We mention larvae because it subtly hints at our major finding that our samples captured many organisms best known for their sessile, benthic stages (e.g. barnacles).
-	We added copepods to this list, though we recognize that it is necessarily incomplete.\\
+	We added copepods to this list, though we recognize that it is necessarily incomplete (L104).\\
 \end{response}
@@
 Our intent was to indicate the novelty of the experimental design which was intended to capture relatively fine-scale spatial distribution of eDNA in marine waters.
-We have made the suggested change.\\
+We have made the suggested change (L110).\\
 \end{response}
@@
 \begin{response}
-  We intended `genomic eDNA' to mean the full DNA extracted from samples, but your suggestion is a good one and we have implemented it.\\
+  We intended `genomic eDNA' to mean the full DNA extracted from samples, but your suggestion is a good one and we have implemented it (L182).\\
 \end{response}
@@
 \begin{response}
-  We made this change.\\
+  We made this change (L154, L159, L162, L184, L185).\\
 \end{response}
@@
   While this does seem like overkill, it is not unheard of for machines or sequencing facilities to develop problems that can bias data, and we included this with the intent of being as transparent as possible.
-Nevertheless, we removed this information.\\
+Nevertheless, we removed this information (L209).\\
 \end{response}
@@
 	Because PhiX increases sequence diversity, its effect is to increase the number of clusters which can be correctly differentiated from one another on the flowcell, thus increasing sequencing depth.
-	Similar to your statement, the \textbf{less} PhiX is spiked-in, the fewer clusters will be correctly differentiated from one another, and the fewer reads passing filter will be obtained, thus also decreasing sequencing depth.\\
+	Similar to your statement, the \textbf{less} PhiX is spiked-in, the fewer clusters will be correctly differentiated from one another, and the fewer reads passing filter will be obtained, thus also decreasing sequencing depth (L210).\\
 \end{response}
@@
 \begin{response}
-  We made this change.\\
+  We made this change (L237).\\
 \end{response}
@@
 Balanus glandula) is internally fertilized and has sessile adults in the intertidal, if it has a larval stage with high potential for dispersal, we cannot rule out an expectation that it will be most abundant in the sequence data from samples far from shore due to larval abundance.
-	We have attempted to clarify this in the text.
+	We have attempted to clarify this in the text (L284).
 	That said, you are also correct that this data is almost unused except to single out Cymatogaster aggregata and Cupolaconcha meroclista as taxa for which an a priori expectation based on adult distribution should be met.
-	We added a concluding sentence to the paragraph to explain the reason for these data.
+	We added a concluding sentence to the paragraph to explain the reason for these data (L292).
 	We chose to keep the description here to show that we have been diligent in investigating potential factors obscuring taxon-specific patterns.
@@
 \begin{response}
-  We made this change.\\
+  We made this change (L332).\\
 \end{response}
@@
   Our intent was precisely to convey the novelty of the sampling design; we failed to do that clearly.
-	We removed `For the first time in a marine environment', and we have clarified the novelty of our study in the context of recently published work in the introduction.\\
+	We removed `For the first time in a marine environment' (L264), and we have clarified the novelty of our study in the context of recently published work in the introduction (L76).\\
 \end{response}
@@
 \begin{response}
-  We corrected this.\\
+  We corrected this (L309).\\
 \end{response}
@@
   It is unclear what the specific error is, but we acknowledge this sentence is poorly worded.
-	We changed this to read `We demonstrate that distant locations have less-similar eDNA communities than proximate locations in Puget Sound, a dynamic marine environment. '\\
+	We changed this to read `We demonstrate that distant locations have less-similar eDNA communities than proximate locations in Puget Sound, a dynamic marine environment' (L371).\\
 \end{response}
@@
 \begin{response}
-  This is a good suggestion; we have modified the abstract to clarify this as early as possible.\\
+  This is a good suggestion; we have modified the abstract to clarify this as early as possible (L9).\\
 \end{response}
@@
 \begin{response}
-  We added a brief explanation of our marker choice to the section Methods:PCR Amplification, and a more general discussion of marker choice to the discussion.\\
+  We added a brief explanation of our marker choice to the section Methods:PCR Amplification (L166), and a more general discussion of marker choice to the discussion (L452).\\
 \end{response}
@@
 \begin{response}
-  Thank you for noticing this; we have corrected the legend to be meaningful as a standalone figure.\\
+  Thank you for noticing this; we have corrected the legend to be meaningful as a standalone figure (Figure 5 legend).\\
 \end{response}
@@
 \begin{response}
-  Yes; these links were a placeholder until the manuscript is accepted and have now been updated.\\
+  Yes; these links were a placeholder until the manuscript is accepted and have now been updated (L514).\\
 \end{response}
@@
 \begin{response}
-  Perhaps we misunderstand your comment, but in the relevant part of the methods section, we state: `Each environmental sample was collected in a clean 1 liter high-density polyethylene bottle, the opening of which was covered with 500 micrometer nylon mesh to prevent entry of larger particles.'\\
+  Perhaps we misunderstand your comment, but in the relevant part of the methods section, we state: `Each environmental sample was collected in a clean 1 liter high-density polyethylene bottle, the opening of which was covered with 500 micrometer nylon mesh to prevent entry of larger particles' (L140).\\
 \end{response}
@@
 \begin{response}
-  In our initial draft, we provided an outline of the bioinformatics procedures in the section Methods:Sequence Data Processing (Bioinformatics), and document the procedures in painstaking detail in the Supplemental Material section Methods:Bioinformatics.
-	However, we overlooked the inclusion of a direct link to the scripts used to process the raw sequence data, and have now added those.
+  In our initial draft, we provided an outline of the bioinformatics procedures in the section Methods:Sequence Data Processing (Bioinformatics) (L213), and document the procedures in painstaking detail in the Supplemental Material section Methods:Bioinformatics (L725).
+	However, we overlooked the inclusion of a direct link to the scripts used to process the raw sequence data, and have now added those (L515).
 	We certainly appreciate any encouragement to move towards more reproducible research, though we are not yet where we would like to be.
@@
 \begin{response}
-  This is an excellent point. We have modified the relevant section to incorporate this potential mechanism.
+  This is an excellent point. We have modified the relevant section to incorporate this potential mechanism (L411).
 	\\
@@
   We believe we adequately addressed alpha diversity by reporting richness and evenness both in a 2-dimensional and linear spatial context, with accompanying figures.
-	However, we have now also added an OTU accumulation figure to the supplement and commented on it in the results.
+	However, we have now also added an OTU accumulation figure to the supplement (Figure 10) and commented on it in the results (L310).
 	We note that our goal is not to recover an absolute measure of alpha diversity, but to make comparisons among sites for standardized sampling effort (i.e. sequencing depth).

### 2017-01-13 Manuscript/v1.1_PeerJ/carkeek_grid.tex
add line numbers to response to reviews

@@
 \usepackage{float} % figure placement [H]
-\usepackage{nameref,hyperref} % named references (e.g. "Supplemental Material")
+\usepackage{nameref} % named references (e.g. "Supplemental Material")
+\usepackage{breakcites} % add line breaks in citations where necessary
+\usepackage[breaklinks=true]{hyperref}
 
@@
 Research to date has documented the non-random spatial distribution of % spatial variation of
-meiofaunal \citep{Fonseca2014,Guardiola2016}, microbial \citep{Lallias2015}, and extracellular \citep{Guardiola2015} eDNA of marine and estuarine sediments, and of microscopic plankton in open ocean waters \citep{DeVargas2015}.
+meiofaunal \citep{Fonseca2014,Guardiola2016}, microbial \citep{Lallias2015},
+and extracellular \citep{Guardiola2015} eDNA of marine and estuarine sediments, and of microscopic plankton in open ocean waters \citep{DeVargas2015}.
 These studies conducted targeted sampling at intermediate (thousands of meters) to global (thousands of kilometers) scales.
@@
 Indexed primers were assigned to samples randomly, with the identical index sequence on the forward and reverse primer to avoid errors associated with dual-indexed multiplexing \citep{Schnell2015}.
-In a UV-sterilized hood, we prepared 25 microliter reactions containing 18.375 microliters ultrapure water, 2.5 microliters 10x buffer, 0.625 microliters deoxynucleotide solution (8 millimolar), 1 microliter each forward and reverse primer (10 micromolar, obtained lyophilized from Integrated DNA Technologies (Coralville, IA, USA)), 0.25 microliters Qiagen HotStar Taq polymerase, and 1.25 microliter genomic eDNA template at 1:100 dilution in ultrapure water.
+In a UV-sterilized hood, we prepared 25 microliter reactions containing 18.375 microliters ultrapure water, 2.5 microliters 10x buffer, 0.625 microliters deoxynucleotide solution (8 millimolar), 1 microliter each forward and reverse primer (10 micromolar, obtained lyophilized from Integrated DNA Technologies (Coralville, IA, USA)), 0.25 microliters Qiagen HotStar Taq polymerase, and 1.25 microliter genomic or eDNA template at 1:100 dilution in ultrapure water.
 PCR thermal profiles began with an initialization step (95 ºC; 15 min) followed by cycles (40 and 20 for the first and second reaction, respectively) of denaturation (95 ºC; 15 sec), annealing (61 ºC; 30 sec), and extension (72 ºC; 30 sec).

### 2017-01-18 Analysis/diversity.R
remove richness/evenness, add letter labels

@@ -2,7 +2,7 @@ library(vegan)
 
 # requires a list of OTU tables and a list of corresponding metadata
 
-EXPORT <- FALSE
+EXPORT <- TRUE
 
 div_metrics <- list()
 
@@ -18,7 +18,7 @@ if(!exists("legend_text")){legend_text <- list()}
 legend_text[plot_name] <- {
 "Aggregate measures of diversity at each sample site.
 Data are rarefied counts of mitochondrial 16S sequences collected from 3 parallel transects in Puget Sound, Washington, USA.
-Evenness (left) is the probability that two sequences drawn at random are different; richness (right) represents the total number of unique sequences from that location."
+Evenness (A) is the probability that two sequences drawn at random are different; richness (B) represents the total number of unique sequences from that location."
 }
 
 if(EXPORT){
@@ -69,7 +69,15 @@ if(i == 1){
   title(ylab = "Distance from 0 (meters)")
 }
 
-title(main = names(div_metrics[i]))
+# add letter for legend
+mtext(
+  text = LETTERS[i], 
+  side = 1, 
+  adj = 1.3, 
+  # at = c(max(metadata[[which_data]][,colname_xcoord])*1.2, 0), 
+  line = 1, 
+  cex = 2 )
+
 title(xlab = "Transect", line = 1)
 
 # add color legend

### 2017-01-18 Analysis/diversity.R
decrease margins and move letters out

@@ -2,7 +2,7 @@ library(vegan)
 
 # requires a list of OTU tables and a list of corresponding metadata
 
-EXPORT <- TRUE
+EXPORT <- FALSE
 
 div_metrics <- list()
 
@@ -30,7 +30,7 @@ if(EXPORT){
 
 the_layout <- layout(mat = matrix(c(1,2), nrow = 1), widths = c(11,6))
 
-margins_subplot <- list(c(4,5,3,5), c(4,0,3,3), c(4,0,3,3))
+margins_subplot <- list(c(3,5,1,5), c(3,0,1,3))
 
 for(i in 1:length(div_metrics)){
 
@@ -73,7 +73,7 @@ if(i == 1){
 mtext(
   text = LETTERS[i], 
   side = 1, 
-  adj = 1.3, 
+  adj = 1.4, 
   # at = c(max(metadata[[which_data]][,colname_xcoord])*1.2, 0), 
   line = 1, 
   cex = 2 )

### 2017-01-18 Analysis/site_map.R
add label to z scale (depth)

@@ -5,6 +5,7 @@
 tif_file <- file.path(data_dir, "ngdc_pug_snd_dm_subset.tif")
 
 library(raster)
+library(grid) #grid.text()
 data_raster <- raster(tif_file)
 
 ignore_above_0 <- function(x){
@@ -66,7 +67,7 @@ if(EXPORT){
   writeLines(legend_text[[plot_name]], con = legend_file)
   pdf(file = pdf_file, width = 6, height = 5)
 }
-par(mar = c(5,6,1,4))
+par(mar = c(5,6,1,5))
 
 library(rasterVis)
 myTheme <- rasterTheme(region = col_depth)
@@ -77,6 +78,11 @@ levelplot(
   par.settings = myTheme
 ) + 
 layer(sp.points(mypoints, col = "orangered", pch = 4))
+
+# add label to depth color (zcale) legend
+trellis.focus("legend", side="right", clipp.off=TRUE, highlight=FALSE)
+grid.text(label = "Depth (m)", x = 0.2, y = -0.025, hjust = 0.5, vjust = 1)
+trellis.unfocus()
 
 if(EXPORT){
   dev.off()

### 2017-01-18 Analysis/cluster.R
add letter labels

@@ -188,9 +188,9 @@ if(EXPORT){
 }
 
 #-------------------------------------------------------------------------------
-## Plot the sites in space, colored by cluster, for ALL levels of K
+## Plot the sites in space, colored by cluster, for two levels of K
 plot_name <- "pam_in_space"
-legend_text[plot_name] <- "Cluster membership of sampled sites. Distance from onshore starting point is log scaled. Sites are colored and labeled by their assignment to a cluster by PAM analysis for number of clusters (K) chosen based on a priori expectations (2) and mean silhouette width (8)."
+legend_text[plot_name] <- "Cluster membership of sampled sites. Distance from onshore starting point is log scaled. Sites are colored and labeled by their assignment to a cluster by PAM analysis for number of clusters (K) chosen based on a priori expectations (A; 2) and mean silhouette width (B; 8)."
 
 if(EXPORT){
   pdf_file    <- file.path(fig_dir, paste(plot_name, ".pdf", sep = ""))
@@ -246,7 +246,18 @@ axis(side = 1,
 	at = unique(my_metadata[,colname_xcoord]), 
 	labels = unique(my_metadata[,colname_xcoord]), 
 	line = 0, lwd = 0, lwd.ticks = 1, las = 1)
+
+# add letter label
+mtext(
+  LETTERS[which(K_plot == i)],
+  side = 1, 
+  line = 1, 
+  adj = 0, 
+  cex = 3
+)
+
 }
+
 if(EXPORT){
   dev.off()
 }

### 2017-01-19 Analysis/diversity_space.R
add letter labels

@@ -1,7 +1,7 @@
 #-------------------------------------------------------------------------------
 # Diversity in Space
 #-------------------------------------------------------------------------------
-EXPORT <- TRUE
+EXPORT <- FALSE
 
 dataset <- "mean"
 
@@ -164,7 +164,7 @@ plot_name <- "diversity_distance"
 if(!exists("legend_text")){ legend_text <- list()}
 legend_text[plot_name] <- {
 "Aggregate diversity metrics of each site plotted against distance from shore.
-Both Simpson's Index (left) and richness (right) are shown, and have been computed from the mean abundance of unique DNA sequences found across 4 PCR replicates at each of 24 sites.
+Both Simpson's Index of evenness (A) and richness (B) are shown, and have been computed from the mean abundance of unique DNA sequences found across 4 PCR replicates at each of 24 sites.
 Lines and bands illustrate the fit and 95% confidence interval of a linear model.
 "
 }
@@ -185,7 +185,12 @@ for(i in 1:length(col_sub)){
 	
 	# ylims <- range(c(range(div_dist[[metric_name]]$conf), range(temp_dat$y)))
 
-	plot(x = div_dist[[col_sub[i]]]$dat$x, y = div_dist[[col_sub[i]]]$dat$y,
+	plot_dat <- data.frame(
+		x = div_dist[[col_sub[i]]]$dat$x, 
+		y = div_dist[[col_sub[i]]]$dat$y
+	)
+
+	plot(x = plot_dat$x, y = plot_dat$y,
 	  # log = "x", 
 	  # axes = FALSE,
 	  xlab = "Distance from shore (meters)", 
@@ -194,14 +199,24 @@ for(i in 1:length(col_sub)){
 
 	# tick_x <- c(0,50,250,500,1000,2000,4000)
 	# axis(1, at = log(tick_x + 1), labels = tick_x)
-	
 	# axis(2)
-	
 	# box()
 
 	plot_model(div_dist[[col_sub[i]]]$conf, x_pred, line_type = 2)
+	
+	legend_x <- ifelse(i == 1, min(plot_dat$x), max(plot_dat$x))
+    
+	legend_adj <- ifelse(i == 1, c(0, -1), c(1, 0.5) )
+	
+	text(
+		x = legend_x, 
+		y = max(plot_dat$y)*0.95, 
+		adj = legend_adj,
+		label = LETTERS[i], 
+		cex = 3)
 
 }
+
 if(EXPORT){
   dev.off()
 }

### 2017-01-19 Analysis/otu_in_space.R
remove taxa labels, add letter labels

@@ -96,11 +96,19 @@ y_for_plotting <- as.numeric(as.factor(my_metadata[ , colname_ycoord]))
 
 if(!exists("legend_text")){legend_text <- list()}
 
-legend_text[plot_name] <- {
-"Distribution of eDNA from select taxa.
-Circles are colored and scaled by the proportion of that taxon's maximum proportional abundance.
-That is, the largest circle is the same size in each of the panels, and occurs where that taxon contributed the greatest proportional abundance of reads to that sample."
+select_taxa_legend <- vector()
+for(i in 1:length(select_taxa)){
+  select_taxa_legend[i] <- paste0(select_taxa[i], " (", LETTERS[i], ")" )
 }
+select_taxa_legend0 <- paste(select_taxa_legend, collapse = ", ")
+
+select_taxa[i]
+legend_text[plot_name] <- {paste0(
+"Distribution of eDNA from select taxa.
+Taxa represented are: ", select_taxa_legend0,  
+". Circles are colored and scaled by the proportion of that taxon's maximum proportional abundance.
+That is, the largest circle is the same size in each of the panels, and occurs where that taxon contributed the greatest proportional abundance of reads to that sample."
+)}
 
 if(EXPORT){
   pdf_file    <- file.path(fig_dir, paste(plot_name, ".pdf", sep = ""))
@@ -109,7 +117,7 @@ if(EXPORT){
   pdf(file = pdf_file, width = 10, height = 4)
 }
 
-par(mfrow = c(1,length(select_taxa)))
+par(mfrow = c(1,length(select_taxa)), mar = c(4, 4, 1, 1))
 
 for(i in 1:length(select_taxa)){
 	point_data  <- scale01(my_table[, select_taxa[i]])
@@ -125,7 +133,7 @@ for(i in 1:length(select_taxa)){
 		pch = 21,
 		bg = point_color,
 		cex = point_size,
-		main = select_taxa[i],
+		# main = select_taxa[i],
 		las = 1,
 		xlab = "Distance along shore (meters)",
 		ylab = "Distance from 0 (meters)"
@@ -139,10 +147,14 @@ for(i in 1:length(select_taxa)){
 			point_data == 0,
 			4, NA_integer_)
 	)
+	
+	# add letter label
+	mtext(LETTERS[i], side = 1, line = 2, adj = -0.2, cex = 2)
     
[…]

### 2017-01-19 Analysis/diversity_space.R
add letter labels

@@ -123,7 +123,7 @@ if(!exists("legend_text")){ legend_text <- list()}
 legend_text[plot_name] <- {
 "Aggregate diversity metrics of each site plotted against distance from shore.
 Both Simpson's Index (top) and richness (bottom) are shown for a variety of data subsets and transformations (left to right: mean, unfiltered mean, log(x + 1), transformed, scaled, spatially variable, and taxon clustered). 
-Lines and bands illustrate the fit and 95\% confidence interval of a linear model.
+Lines and bands illustrate the fit and 95% confidence interval of a linear model.
 See methods text for detailed data descriptions."
 
 }
@@ -143,6 +143,7 @@ for(i in 1:length(div_dist)){
 	)
 	plot_model(div_dist[[i]]$conf, x_pred, line_type = 2)
 	# abline(div_dist[[i]]$out, col = "red")
+	mtext(LETTERS[i], side = 1, line = 2, adj = -0.2, cex = 2)
 }
 if(EXPORT){
   dev.off()

### 2017-01-19 Manuscript/v1.2_PeerJ/carkeek_grid.tex
make editorial changes

--- 2017-01-13 Manuscript/v1.1_PeerJ/carkeek_grid.tex
@@
 %-----------------------------------NOTES---------------------------------------
-% (TODO INSERT DOI FROM FINAL COMMIT BEFORE PUBLICATION)
 
@@
 The classification of samples to communities was made on the basis of their pairwise Bray-Curtis similarity, calculated using the function vegdist in the R package vegan \citep{vegan}.
-Other distance metrics were evaluated but had no appreciable effect on the outcome of the analysis (Figure \ref{distance_decay_multi}).
+Other distance metrics were evaluated but had no appreciable effect on the outcome of the analysis (Supplemental Figure \ref{distance_decay_multi}).
 In order to chose an optimal number of clusters ($K$), we evaluated the distribution of silhouette widths, a measure of the similarity between each sample and its cluster compared to its similarity to other clusters.
@@
 %-----------------------------------FUNDING-------------------------------------
-\section*{Funding}
-This work was supported by a grant from the David and Lucile Packard Foundation to RPK (grant 2014-39827).
+% \section*{Funding}
+% This work was supported by a grant from the David and Lucile Packard Foundation to RPK (grant 2014-39827).
 The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
-
+%
 %--------------------------------CONTRIBUTIONS----------------------------------
-\section*{Author Contributions}
-Conceived and designed the experiments:
-JL O'Donnell, RP Kelly, AO Shelton;
-Collected the data:
-JL O'Donnell, NC Lowell, GD Williams, RP Kelly, AO Shelton, JF Samhouri;
-Conducted the analyses:
-JL O'Donnell;
-Wrote the first draft:
-JL O'Donnell;
-Edited the manuscript:
-JL O'Donnell, AO Shelton, RP Kelly, JF Samhouri, GD Williams, NC Lowell
+% \section*{Author Contributions}
+% Conceived and designed the experiments:
+% JL O'Donnell, RP Kelly, AO Shelton;
+% Collected the data:
+% JL O'Donnell, NC Lowell, GD Williams, RP Kelly, AO Shelton, JF Samhouri;
+% Conducted the analyses:
+% JL O'Donnell;
+% Wrote the first draft:
+% JL O'Donnell;
+% Edited the manuscript:
+% JL O'Donnell, AO Shelton, RP Kelly, JF Samhouri, GD Williams, NC Lowell
 
 %-----------------------------------ETHICS--------------------------------------
-\section*{Ethics Statement}
-The authors declare no conflict of interest.
-Consistent with the public trust doctrine, waters of the US are public, and therefore no permit was necessary to conduct this research (see Illinois Central Railroad v. Illinois, 146 U.S. 387 (1892)).
+% \section*{Ethics Statement}
+% The authors declare no conflict of interest.
+% Consistent with the public trust doctrine, waters of the US are public, and therefore no permit was necessary to conduct this research (see Illinois Central Railroad v.
+Illinois, 146 U.S.
+387 (1892)).
 
@@
 \verb!https://github.com/jimmyodonnell/Carkeek_eDNA_grid!\\
-\verb!http://dx.doi.org/FIXME!
+\verb!https://doi.org/10.5281/zenodo.242976!
 
@@
   \centering
-    \includegraphics[width=1\textwidth]{pam_in_space.pdf}
-    \caption{\protect\input{"../../Figures/pam_in_space_legend.txt"}}
-  \label{pam_in_space}
-\end{figure}
-
-\begin{figure}[H] % [h!] forces the figure to be placed roughly here
-  \centering
     \includegraphics[width=1\textwidth]{pam_sil.pdf}
@@
   \centering
-    \includegraphics[width=1\textwidth]{diversity.pdf}
-    \caption{\protect\input{"../../Figures/diversity_legend.txt"}}
-  \label{diversity}
+    \includegraphics[width=1\textwidth]{pam_in_space.pdf}
+    \caption{\protect\input{"../../Figures/pam_in_space_legend.txt"}}
+  \label{pam_in_space}
 \end{figure}
@@
 \pagebreak
+
+\begin{figure}[H] % [h!] forces the figure to be placed roughly here
+  \centering
+    \includegraphics[width=1\textwidth]{diversity.pdf}
+    \caption{\protect\input{"../../Figures/diversity_legend.txt"}}
+  \label{diversity}
+\end{figure}
 

### 2017-01-19 Analysis/diversity_space.R
fix letter labels and legend

@@ -122,7 +122,7 @@ plot_name <- "diversity_distance_all"
 if(!exists("legend_text")){ legend_text <- list()}
 legend_text[plot_name] <- {
 "Aggregate diversity metrics of each site plotted against distance from shore.
-Both Simpson's Index (top) and richness (bottom) are shown for a variety of data subsets and transformations (left to right: mean, unfiltered mean, log(x + 1), transformed, scaled, spatially variable, and taxon clustered). 
+Both Simpson's Index (top; A-F) and richness (bottom; G-L) are shown for a variety of data subsets and transformations (left to right: mean (A,G), unfiltered mean (B,H), log(x + 1) transformed (C,I), scaled (D,J), spatially variable (E,K), and taxon clustered (F,L)). 
 Lines and bands illustrate the fit and 95% confidence interval of a linear model.
 See methods text for detailed data descriptions."
 

### 2017-01-19 Analysis/rarefaction.R
fix y axis labels

@@ -43,7 +43,7 @@ if(EXPORT){
 }
 
 par(mar = c(4,4,1,1))
-rarecurve(dat, step = 10000, sample = min_reads, 
+rarecurve(dat, step = 10000, sample = min_reads, las = 1, 
   xlab = "Number of sequences", ylab = "OTUs", col = "orchid")
 
 if(EXPORT){

### 2017-01-19 Manuscript/v1.2_PeerJ/carkeek_grid.tex
move supplement to separate file

@@
 Detailed bioinformatic methods are provided in the supplemental material, and analysis scripts used from raw sequencer output onward can be found in the public project directory (see \nameref{analysis_scripts}).
-Briefly, we performed five steps to process the sequence data: (1) Merge paired-end reads, (2) eliminate low-quality reads, (3) eliminate PCR artifacts (chimeras), (4) cluster reads by similarity into operational taxonomic units (OTUs), and (5) match observed sequences to taxon names.
+Briefly, we performed five steps to process the sequence data: (1) Merge paired-end reads \citep{Zhang2014}, (2) eliminate low-quality reads \citep{Edgar2010,vsearch}, (3) eliminate PCR artifacts (chimeras) \citep{Edgar2010,vsearch,Martin2011}, (4) cluster reads by similarity into operational taxonomic units (OTUs) \citep{swarm}, and (5) match observed sequences to taxon names \citep{Camacho2009, Chamberlain2013, Chamberlain2016}.
 Additionally, we checked for consistency among PCR replicates, excluded extremely rare sequences, and rescaled (rarefied) the data to account for differences in sequencing depth.
@@
 
-There are other conceptually reasonable forms to expect the space-by-similarlity relationship to take; we present these in the supplemental material along with alternative data subsets and similarity indices (see \nameref{supplement}).
+There are other conceptually reasonable forms to expect the space-by-similarlity relationship to take; we present these in the supplemental material along with alternative data subsets and similarity indices.
 
@@
 The classification of samples to communities was made on the basis of their pairwise Bray-Curtis similarity, calculated using the function vegdist in the R package vegan \citep{vegan}.
-Other distance metrics were evaluated but had no appreciable effect on the outcome of the analysis (Supplemental Figure \ref{distance_decay_multi}).
+Other distance metrics were evaluated but had no appreciable effect on the outcome of the analysis (Supplemental Figure 1).
 In order to chose an optimal number of clusters ($K$), we evaluated the distribution of silhouette widths, a measure of the similarity between each sample and its cluster compared to its similarity to other clusters.
@@
 Overall similarities among PCR replicates within a sample were extremely high (0.976 $\pm$ 0.013), and far higher than those among samples (0.3 $\pm$ 0.16).
-Across PCR replicates, each sample was represented by at least 781425 reads in the raw data and contained between 111 and 443 rarefied OTUs (Supplemental Figure \ref{rarefaction}).
+Across PCR replicates, each sample was represented by at least 781425 reads in the raw data and contained between 111 and 443 rarefied OTUs (Supplemental Figure 2).
 
@@
 Evenness increased by .0666 per 1000 meters from 0.225 (95\%CI = 0.0558) to 0.491 (95\%CI = $\pm$ 0.112), indicating that sequence reads were less evenly distributed among taxa in offshore samples (p $\ll$ 0.05; Figure \ref{diversity_distance}).
+The subset of data used for this analysis had no qualitative effect on the outcome of this analysis (Supplemental Figure 3).
 There was no evidence for spatial autocorrelation for any of the diversity metrics (Moran's I, p > 0.05; Figure \ref{diversity}).
@@
 
-%----------------------------------SUPPLEMENT-----------------------------------
-\pagebreak
-\section*{Supplemental Material}
-\label{supplement}
-
-\subsection*{Methods}
-\subsubsection*{Bioinformatics}
-Reads passing the preliminary Illumina quality filter were demultiplexed on the basis of the adapter index sequence by the sequencing facility.
-We used fastqc to assess the fastq files output from the sequencer for low-quality indications of a problematic run.
-Forward and reverse reads were merged using PEAR v0.9.6 \cite{Zhang2014} and discarded if more than 0.01 of the bases were uncalled.
-If a read contained two consecutive base calls with quality scores less than 15 (i.e.
-probability of incorrect base call = 0.0316), these bases and all subsequent bases were removed from the read.
-Paired reads for which the probability of matching by chance alone exceeded 0.01 were not assembled and omitted from the analysis.
-Assembled reads were discarded if assembled sequences were not between 50 and 168 bp long, or if reads did not overlap by at least 100 bp.
-
-We used vsearch v2.1.1 \citep{vsearch} to discard any merged reads for which the sum of the per-base error probabilities was greater than 0.5 (``expected errors'') \cite{Edgar2010}.
-Sequences were demultiplexed on the basis of the primer index sequence at base positions 4-9 at both ends using the programming language AWK.
-Primer sequences were removed using cutadapt v1.7.1 \cite{Martin2011}, allowing for 2 mismatches in the primer sequence.
-Identical duplicate sequences were identified, counted, and removed in python to speed up subsequent steps by eliminating redundancy, and sequences occurring only once were removed.
-We checked for and removed any sequence likely to be a PCR artifact due to incomplete extension and subsequent mis-priming using a method described by \citet{Edgar2010} and implemented in vsearch v2.0.2.
-Sequences were clustered into operational taxonomic units (OTUs) using the single-linkage clustering method implemented by swarm version 2.1.1 with a local clustering threshold (d) of 1 and fastidious processing \citep{swarm}.
-
-Cross-contamination of environmental, DNA, or PCR samples can result in erroneous inference about the presence of a given DNA sequence in a sample.
-However, other processes can contribute to the same signature of contamination.
-For example, errors during oligonucleotide synthesis or sequencing of the indexes could cause reads to be erroneously assigned to samples.
-The frequency of such errors can be estimated by counting the occurrence of sequences known to be absent from a given sample, and of reads that do not contain primer index sequences in the expected position or combinations.
-These occurrences indicate an error in the preparation or sequencing procedures.
-We estimated a rate of incorrect sample assignment by calculating the maximum rate of occurrence of index sequences combinations we did not actually use, as well as the rates of cross-library contamination by counting occurrences of primer sequences from 12S amplicons prepared in a lab more than 1000 kilometers away, but pooled and sequenced alongside our samples.
-This represents a general minimum rate at which we can expect that sequences from one environmental sample could be erroneously assigned to another, and so we considered for further analysis only those reads occurring with greater frequency than this across the entire dataset.
-
-We checked for experimental error by evaluating the Bray-Curtis similarity (1 - Bray-Curtis dissimilarity) among replicate PCRs from the same DNA sample.
-We calculated the mean and standard deviation across the dataset, and excluded any PCR replicates for which the similarity between itself and the other replicates was less than 1.5 standard deviations from the mean.
-
-To account for variation in the number of sequencing reads (sequencing depth) recovered per sample, we rarefied the within-sample abundance of each OTU by the minimum sequencing depth \citep{vegan}.
-
-Because each step in this workflow is sensitive to contamination, it is possible that some sequences are not truly derived from the environmental sample, and instead represent contamination during field sampling, filtration, DNA extraction, PCR, fragment size selection, quantitation, sequencing adapter ligation, or the sequencing process itself.
-We take the view that contaminants are unlikely to manifest as sequences in the final dataset in consistent abundance across replicates; indeed, our data show that the process from PCR onward is remarkably consistent.
-Thus, after scaling to correct for sequencing depth variation, we calculated from our data the maximum number of sequence counts for which there is turnover in presence-absence among PCR replicates within an environmental sample.
-We use this number to determine a conservative minimum threshold above which we can be confident that counts are consistent among replicates and not of spurious origin, and exclude from further analysis observations where the mean abundance across PCR replicates within samples does not reach this threshold.
-For further analyses we use the mean abundance across PCR replicates for each of the 24 environmental samples.
-
-In order to determine the most likely taxon from which each sequence originated, the representative sequence from each OTU was then queried against the NCBI nucleotide collection (GenBank; version October 7, 2015; 32,827,936 sequences) using the blastn command line utility \citep{Camacho2009}.
-In order to maximize the accuracy of this computationally intensive step, we implemented a nested approach whereby each sequence was first queried using strict parameters (e-value = 5e-52), and if no match was found, the query was repeated with decreasingly strict e-values (5e-48 5e-44 5e-40 5e-36 5e-33 5e-29 5e-25 5e-21 5e-17 5e-13).
-Other parameters were unchanged among repetitions (word size: 7; maximum matches: 1000; culling limit: 100; minimum percent identity: 0).
-Each query sequence can be an equally good match to multiple taxa either because of invariability among taxa or errors in the database (e.g.
-human sequences are commonly attributed to other organisms when they in fact represent lab contamination).
-In order to guard against these spurious results, we used an algorithm to find the lowest common taxon for at least 80\% of the matched taxa, implemented in the R package taxize 0.7.8 \citep{Chamberlain2013, Chamberlain2016}.
-Similarly, we repeated analyses using the dataset consolidated at the same taxonomic rank across all queries, for the rank of both family and order.
-
-\subsubsection*{Alternative distance decay model formulations}
-
-\paragraph{Linear:} We fit a straight line through the points after log-transforming the spatial distances to estimate the intercept and slope.
-This model ignores the bounds of our response variable of community similarity.
-
-\paragraph{Michaelis-Menten:} We fit a Michaelis-Menten-like curve to our data.
-Our formulation can be thought of as a modification of the Michaelis-Menten equation, but with the addition of a parameter in the numerator which modifies the intercept.
-
-\begin{equation}\label{MM_full}
-	y = \frac{AB + Cx}{B+x}
-\end{equation}
-
-Where $C$ is the asymptote of minimum similarity.
-This formulation allows us to estimate the maximum similarity in the system, and the rate at which it is achieved.
-If the value of the parameter ($AB$) is 0 (i.e.
-if the intercept is 0), the form is identical to the Michaelis-Menten equation:
-
-\begin{equation}\label{MichaelisMenten}
-	y = \frac{Cx}{B + x}
-\end{equation}
-
-This is conceptually satisfying in that a fit through [0,1] reflects the theoretical expectation that samples at zero distance from one another are necessarily identical.
-Given an efficient sampling technique, replicate samples taken at the same position in space should be identical, and thus the intercept of the regression of similarity against distance should be 1, and deviation from 1 is an indicator of the efficiency of the sampling method.
-% NOTE: We did not take multiple environmental samples from the same position in space, but conducted multiple PCRs from each environmental sample.
-
-Finally, we considered a model which estimates an asymptote as the total change in similarity ($D$):
-
-\begin{equation}\label{Harold}
-	y = \frac{A + Dx}{B + x}
-\end{equation}
-
-However, this model failed to converge and produced uninformative estimates of all parameters.
-
-\subsection*{Supplemental Figures}
-
-\begin{figure}[H] % [h!] forces the figure to be placed roughly here
-  \centering
-    \includegraphics[height=0.6\textheight]{distance_decay_multi.pdf}
-    \caption{\protect\input{"../../Figures/distance_decay_multi_legend.txt"}}
-  \label{distance_decay_multi}
-\end{figure}
-
-\begin{figure}[H] % [h!] forces the figure to be placed roughly here
-  \centering
-    \includegraphics[width=1\textwidth]{diversity_distance_all.pdf}
-    \caption{\protect\input{"../../Figures/diversity_distance_all_legend.txt"}}
-  \label{diversity_distance_all}
-\end{figure}
-
-\begin{figure}[H] % [h!] forces the figure to be placed roughly here
-  \centering
-    \includegraphics[height=0.6\textheight]{rarefaction.pdf}
-    \caption{\protect\input{"../../Figures/rarefaction_legend.txt"}}
-  \label{rarefaction}
-\end{figure}
-
-% \pagebreak
-% \section*{Orphaned Text}
-% Samples collected of ecological communities may vary in dissimilarity from 0 (completely identical) to 1 (completely different).
-For samples collected from multiple locations, the relationship between their spatial distance and community dissimilarity is of interest because it reflects the amount of community heterogeneity---changes in abundance and composition---over the spatial scale sampled.
-% In general, the relationship between community dissimilarity (0 = identical; 1 = completely different) and spatial distance is expected to be asymptotic, because communities nearer to each other tend to be more similar than those farther apart.
-The intercept is expected to be 0, because only within-sample comparisons can have 0 spatial separation, and communities have no dissimilarity within a sample if sampling method is repeatable.
-% (technically impossible for samples to be taken at exactly 0; they are actually immediately adjacent.)
-% Deviation from 0 indicates heterogeneous community composition/structure over fine scales.
-Likewise, dissimilarity cannot exceed 1, and reaches 1 only when multiple discrete community types are sampled.
-The trend is expected to be asymptotic if communities within a habitat are spatially structured, where the value of the asymptote and the rate of increase provide insight about community turnover.
-A flat relationship indicates that heterogeneity is not assorted spatially, and can be interpreted in different ways, depending on the asymptote.
-% ?or no relationship both
-% If the asymptote is close to 1, there is high spatial heterogeneity over the spatial scale of sampling.
-If the asymptote is close 0, all samples are similar, and we infer there is complete community homogeneity over the scale sampled.
-The rate at which community dissimilarity approaches the mean gives an indication of the rate of community turnover.
-%
-% In theory, given enough data from manipulative experiments, it is possible that one could measure environmental variables--temperature, UV intensity, flow, etc--at the time of eDNA sample collection in order to more precisely estimate the origin of eDNA.
-In practice, this approach is unlikely to yield appreciable gains in accuracy because even the most advanced sensors (ADCPs) and analytical techniques available cannot determine the origin of particles in highly dynamic fluid environments at scales relevant to most ecological questions \citep{Piterbarg2001}.
-%http://www.rsmas.miami.edu/LAPCOD/research/prediction/
-% Further, collecting and analyzing these data would severely limit the scalability of eDNA methods.
-A more practical approach is to compare the spatial distribution of communities of eDNA with expectations based on prior knowledge of the spatial distribution of communities of organisms.
-
 \end{document}

### 2017-01-19 Manuscript/v1.2_PeerJ/supplement.tex
move supplement to separate file

--- 2017-01-13 Manuscript/v1.1_PeerJ/carkeek_grid.tex
@@
-%-----------------------------------NOTES---------------------------------------
-% (TODO INSERT DOI FROM FINAL COMMIT BEFORE PUBLICATION)
-
-% Target journal: PeerJ. Backups: Frontiers in...marine science?, BMC Ecology, BMC Genetics
-%----------------------------------REVIEWERS------------------------------------
-% A list of recommended reviewers.
-% Frédéric Mahé mahefrederic@gmail.com
-% Sophie Arnaud-Haond, sophie.arnaud@ifremer.fr
-% Dominique Cowart dcowart@illinois.edu dominique.cowart@gmail.com (Metabarcoding is powerful yet still blind)
-% Ann Bucklin, ann.bucklin@uconn.edu (Metabarcoding of marine zooplankton)
-% Naiara Rodríguez-Ezpeleta, nrodriguez@azti.es (Benchmarking DNA Metabarcoding for Biodiversity-Based Monitoring and Assessment)
-% Xavier Turon, xturon@ceab.csic.es (Metabarcoding Extracellular DNA from Sediments of Marine Canyons)
-% Andrew Mahon, mahon2a@cmich.edu
-% Eske Willerslev ewillerslev@snm.ku.dk
-% Philip Thomsen pfthomsen@snm.ku.dk
-
-%--------------------------------------||--------------------------------------%
-
 \documentclass[11pt,letterpaper]{article} %  font size
@@
 
-\title{Spatial distribution of environmental DNA in a nearshore marine habitat}
+\title{Supplemental Material for \\ \textit{Spatial distribution of environmental DNA in a nearshore marine habitat}}
 
@@
 
-%----------------------------------KEYWORDS-------------------------------------
-\section*{Keywords}
-
-metagenomics, metabarcoding, environmental monitoring, molecular ecology, marine, estuarine
-
 \pagebreak
@@
 \def\linenumberfont{\normalfont\small\rmfamily} % change line number font
-%----------------------------------ABSTRACT-------------------------------------
-\section*{Abstract}
-In the face of increasing threats to biodiversity, the advancement of methods for surveying biological communities is a major priority for ecologists.
-Recent advances in molecular biological technologies have made it possible to detect and sequence DNA from environmental samples (environmental DNA or eDNA); however, eDNA techniques have not yet seen widespread adoption as a routine method for biological surveillance primarily due to gaps in our understanding of the dynamics of eDNA in space and time.
-In order to identify the effective spatial scale of this approach in a dynamic marine environment, we collected marine surface water samples from transects ranging from the intertidal zone to 4 kilometers from shore.
-Using PCR primers that target a diverse assemblage of metazoans, we amplified a region of mitochondrial 16S rDNA from the samples and sequenced the products on an Illumina platform in order to detect communities and quantify their spatial patterns using a variety of statistical tools.
-We find evidence for multiple, discrete eDNA communities in this habitat, and show that these communities decrease in similarity as they become further apart.
-Offshore communities tend to be richer but less even than those inshore, though diversity was not spatially autocorrelated.
-Taxon-specific relative abundance coincided with our expectations of spatial distribution in taxa lacking a microscopic, pelagic life-history stage, though most of the taxa detected do not meet these criteria.
-Finally, we use carefully replicated laboratory procedures to show that laboratory treatments were remarkably similar in most cases, while allowing us to detect a faulty replicate, emphasizing the importance of replication to metabarcoding studies.
-While there is much work to be done before eDNA techniques can be confidently deployed as a standard method for ecological monitoring, this study serves as a first analysis of diversity at the fine spatial scales relevant to marine ecologists and confirms the promise of eDNA in dynamic environments.
-
-%---------------------------------INTRODUCTION----------------------------------
-\section*{Introduction}
-
-The patterns and causes of variability in ecological communities across space are both seminal and contentious areas of study in ecology \citep{Hubbell2001, Anderson2011}.
-One consistently observed pattern of community spatial heterogeneity is that communities close to one another tend to be more similar than those that are farther apart \citep{Nekola1999}.
-This decrease in community similarity with increasing spatial separation is called distance decay and has been reported from communities of tropical trees \citep{Condit2002, Chust2006}, ectomycorrhizal fungi \citep{Bahram2013}, salt marsh plants \citep{Guo2015}, and microorganisms \citep{Martiny2011, Chust2013, Wetzel2012, Bell2010}.
-Typically, this relationship is assessed by regressing a measure of community similarity against a measure of spatial separation for a set of sites at which a set of species' abundances (or presences) is calculated.
-Yet no existing biodiversity survey method completely censuses all of the organisms in a given area.
-The lack of a single `silver bullet' method of sampling contributes inconclusiveness to the study of spatial patterning in ecology \citep{Levin1992}, and leaves open the possibility of new and more comprehensive methods.
-
-From a boat or aircraft, scientists can count whales by sight, but not the krill on which they feed.
-For example, towed fishing nets can efficiently sample organisms larger than the mesh and slower than the boat, but overlook viruses and have undesirable effects on charismatic air-breathing species.
-However, DNA-based surveys show great promise as an efficient technique for detecting a previously unthinkable breadth of organisms from a single sample.
-
-Microbiologists have used nucleic acid sequencing to quantify the composition and function of microbial communities in a wide variety of habitats \citep{Handelsman1998, Tyson2004, Venter2004, Iverson2012}.
-To do so, microorganisms are collected in a sample of environmental medium (e.g.
-water), their DNA or RNA is isolated and sequenced, and the identity and abundance of sequences is considered to reflect the community of  organisms contained in the sample, which indirectly estimates the quantity of organisms in an area.
-
-Macroorganisms shed DNA-containing cells into the environment (environmental DNA or eDNA) that can be sampled in the same way  \citep{Ficetola2008,Thomsen2012}.
-Potentially, eDNA methods allow a broad swath of macroorganisms to be surveyed from basic environmental samples.
-However, the accuracy and reliability of indirect estimates of macroorganismal abundance has been debated because the entire organisms are not contained within the sample \citep{Cowart2015}.
-Concern surrounding eDNA methods is rooted in uncertainty about the attributes of eDNA in the environment relative to actual organisms \citep{Shelton2016, Evans2016}.
-Basic questions such as how long DNA can persist in that environment and how far DNA can travel remain largely unknown (but see \cite{Klymus2015,Turner2015,Strickler2015,Deiner2014}) and impede inference about local organismal presence from an environmental sample.
-As a result, estimating the spatial and temporal resolution of eDNA studies in the field is a key step in making these methods practical.
-
-The relationship between local organismal abundance and eDNA is further complicated in habitats where the environmental medium itself may transport eDNA away from its source.
-We know that genetic material can move away from its source precisely because organisms can be detected indirectly without being present in the sample \citep{Kelly2016}.
-One might reasonably expect eDNA to travel farther in a highly dynamic fluid such as the open ocean or flowing river than it would through the sediment at the bottom of a stagnant pond \citep{Deiner2014, Shogren2016}.
-Yet even studies of extremely dynamic habitats such as coastlines with high wave energy have found remarkable evidence that eDNA transport is limited enough that DNA methods can detect differences among communities separated by less than 100 meters \citep{Port2016}.
-
-While rigorous laboratory studies have investigated the effects of some environmental factors on eDNA persistence \citep{Klymus2015, Barnes2014, Sassoubre2016} and the transport of eDNA in specific contexts \citep{Deiner2014}, we suggest that field studies comparing the spatial distribution of communities of eDNA with expectations based on prior knowledge of organisms' distributions are also critical to developing a working understanding of eDNA in the real world.
-Research to date has documented the non-random spatial distribution of % spatial variation of
-meiofaunal \citep{Fonseca2014,Guardiola2016}, microbial \citep{Lallias2015},
-and extracellular \citep{Guardiola2015} eDNA of marine and estuarine sediments, and of microscopic plankton in open ocean waters \citep{DeVargas2015}.
-These studies conducted targeted sampling at intermediate (thousands of meters) to global (thousands of kilometers) scales.
-Here, we use a grid-based environmental sampling strategy to assess spatial variability of eDNA in a coastal marine environment at a fine scale (tens to thousands of meters), using molecular methods that focus on macrobial metazoans.
-
-We apply methods derived from community ecology to understand spatial patterns and patchiness of eDNA.
-The underlying mechanism thought to drive the slope of the distance decay relationship in ecological communities is the rate of movement of individuals among sites, which may be driven by underlying processes such as habitat suitability.
-Because eDNA is shed and transported away from its source, the increased movement of eDNA particles should homogenize community similarity, and thus erode the distance decay relationship of eDNA communities.
-
-Puget Sound is a deep, narrow fjord in Washington, USA, where a narrow band of shallow bottom hugs the shoreline and abruptly gives way to a central depth of up to 300 meters.
-This form allows the juxtaposition of communities associated with distinctly different habitats: shallow, intertidal benthos, and euphotic pelagic \citep{Burns1985}.
-At the upper reaches of the intertidal, the shoreline substrate varies from soft, fine sediment to cobble and boulder rubble.
-Soft intertidal sediments are inhabited by burrowing bivalves (Bivalvia), segmented worms (Annelida), and acorn worms (Enteropneusta), and in some lower intertidal and high subtidal ranges by eelgrass (\textit{Zostera marina}) \citep{Kozloff1973, Dethier2010} .
-Eelgrass meadows harbor epifaunal and infaunal biota, and attract transient species which use the meadows for shelter and to feed on resident organisms.
-Hard intertidal surfaces support a well-documented biota including barnacles (Sessilia) and other crustaceans, mussels (Bivalvia:Mytilidae), anemones (Actiniaria), sea stars (Asteroidea), urchins (Echinoidea), Bryozoans (Ectoprocta), crustaceans (Decapoda), and a variety of algae \citep{Dethier2010}.
-Hard bottoms of the lower intertidal and high subtidal are home to macroalgae such as Laminariales and Desmarestiales which provide habitat for a distinct community of fish and invertebrates.
-The upper pelagic is home to a diverse assemblage of microscopic plankton including diatoms, copepods, and larvae \citep{Strickland1983}, as well as transitory fish and marine mammals.
-
-We took advantage of this setting to explore the spatial variation and distribution of marine eDNA communities.
-Using PCR-based methods and massively parallel sequencing, we surveyed mitochondrial 16S sequences from a suite of marine animals in water samples collected over a grid of sites extending from the shoreline out to 4 kilometers offshore in Puget Sound, Washington, USA.
-We leverage this sampling design to perform an explicitly spatial analysis of eDNA-derived community similarity.
-We investigate two primary objectives.
-First we examine the spatial patterning of eDNA and determine the degree to which eDNA community similarity can be predicted by physical proximity.
-We expect that physical proximity will be a strong predictor of community similarity, and that community differences can be detected over small distances.
-Second, we examine the distribution of diversity from eDNA data, and compare it to our expectations based on distributions of macrobial communities.
-We expect that distinct eDNA communities exist in this setting, and that their spatial distribution coincides with that of adult macrobial organisms.
-Because of the vastly different communities of benthic macrobial metazoans as a function of distance from shore, we expect that more than one eDNA community is present across our 4 kilometer sampling grid, and that communities change as a function of distance from shore.
-For this reason, we examine two diversity measures of eDNA communities that have been widely used to reveal broad scale patterns based on macrobiota in many ecological systems.
-Finally, we identify the taxa represented in the eDNA communities, which span a range of life-history characteristics, and we expect that the spatial distribution of eDNA will most closely resemble the distribution of adults in taxa with low dispersal potential.
-
-%-----------------------------------METHODS-------------------------------------
-\section*{Methods}
-There are seven discrete steps to our methodology: (1) Environmental sample collection, (2) isolation of particulates from water via filtration, (3) isolation of DNA from filter membrane, (4) amplification of target locus via PCR, (5) sequencing of amplicons, (6) bioinformatic translation of raw sequence data into tables of sequence abundance among samples, and (7) community ecological analyses of eDNA.
-We provide brief overviews of these steps here, and encourage the reader to review the fully detailed methods presented in the supplementary material (\nameref{supplement}).
-
-\subsection*{Environmental Sampling}
-Starting from lower-intertidal patches of \textit{Zostera marina}, we collected water samples at 1 meter depth from 8 points (0, 75, 125, 250, 500, 1000, 2000, and 4000 meters) along three parallel transects separated by 1000 meters (24 sample locations total; Figure \ref{site_map}).
-Samples were collected by attaching bottles to a PVC pole and lowering it over the side of a boat over the span of one hour on 27 June 2014.
-To destroy residual DNA on equipment used for field sampling and filtration, we washed with a 1:10 solution of household bleach (8.25\% sodium hypochlorite; 7.25\% available chlorine) and deionized water, followed by thorough rinsing with deionized water.
-Each environmental sample was collected in a clean 1 liter high-density polyethylene bottle, the opening of which was covered with 500 micrometer nylon mesh to prevent entry of larger particles.
-Immediately after collecting the sample, the mesh was replaced with a clean lid and the sample was held on ice until filtering.
-
-\subsection*{Filtration}
-One liter from each water sample was filtered in the lab on a clean polysulfone vacuum filter holder fitted with a 47 millimeter diameter cellulose acetate membrane with 0.45 micrometer pores.
-Filter membranes were moved into 900 microliters of Longmire buffer \citep{Longmire1997} using clean forceps and stored at room temperature \citep{Renshaw2015}.
-To test for the extent of contamination attributable to laboratory procedures, we filtered three replicate 1 liter samples of deionized water.
-These samples were treated identically to the environmental samples throughout the remaining protocols.
-
-\subsection*{DNA Purification}
-DNA was purified from the membrane following a phenol:chloroform:isoamyl alcohol protocol following Renshaw \citep{Renshaw2015}.
-Preserved membranes were incubated at 65 ºC for 30 minutes before adding 900 microliters of phenol:chloroform:isoamyl alcohol and shaking vigorously for 60 seconds.
-We conducted two consecutive chloroform washes by centrifuging at 14,000 rpm for 5 minutes, transferring the aqueous layer to 700 microliters chloroform, and shaking vigorously for 60 seconds.
-After a third centrifugation, 500 microliters of the aqueous layer was transferred to tubes containing 20 microliters 5 molar NaCl and 500 microliters 100\% isopropanol, and frozen at -20 ºC for approximately 15 hours.
-Finally, all liquid was removed by centrifuging at 14000 rpm for 10 minutes, pouring off or pipetting out any remaining liquid, and drying in a vacuum centrifuge at 45 ºC for 15 minutes.
-DNA was resuspended in 200 microliters of ultrapure water.
-Four replicates of genomic DNA extracted from tissue of a species absent from the sampled environment (\textit{Oreochromis niloticus}) served as positive control for the remaining protocols.
-
-\subsection*{PCR Amplification}
-We chose a primer set that amplifies an approximately 115 base pair (bp) region of the mitochondrial 16S rRNA gene in at least 10 metazoan phyla from this habitat, excludes non-metazoans, and resolves taxonomy to the family level in most cases using a public sequence database \citep{Kelly2016FMS}.
-We used a two-step polymerase chain reaction (PCR) protocol described by \citet{ODonnell2016} to generate 4 replicate products from each DNA sample.
-In the first set of reactions, primers were identical in every reaction (forward: AGTTACYYTAGGGATAACAGCG; reverse: CCGGTCTGAACTCAGATCAYGT); primers in the second set of reactions included these same sequences but with 3 variable nucleotides (NNN) and an index sequence on the 5$'$ end (see \nameref{sequencing_metadata}).
-We used the program OligoTag \citep{Coissac2012} to generate 30 unique 6-nucleotide index sequences differing by a minimum Hamming distance of 3 (see \nameref{sequencing_metadata}).
-Indexed primers were assigned to samples randomly, with the identical index sequence on the forward and reverse primer to avoid errors associated with dual-indexed multiplexing \citep{Schnell2015}.
-In a UV-sterilized hood, we prepared 25 microliter reactions containing 18.375 microliters ultrapure water, 2.5 microliters 10x buffer, 0.625 microliters deoxynucleotide solution (8 millimolar), 1 microliter each forward and reverse primer (10 micromolar, obtained lyophilized from Integrated DNA Technologies (Coralville, IA, USA)), 0.25 microliters Qiagen HotStar Taq polymerase, and 1.25 microliter genomic or eDNA template at 1:100 dilution in ultrapure water.
-PCR thermal profiles began with an initialization step (95 ºC; 15 min) followed by cycles (40 and 20 for the first and second reaction, respectively) of denaturation (95 ºC; 15 sec), annealing (61 ºC; 30 sec), and extension (72 ºC; 30 sec).
-20 identical PCRs were conducted from each DNA extract using non-indexed primers; these were pooled into 4 groups of 5 in order to ensure ample template for the subsequent PCR with indexed primers.
-In order to isolate the fragment of interest from primer dimer and other spurious fragments generated in the first PCR, we used the AxyPrep Mag FragmentSelect-I kit with solid-phase reversible immobilization (SPRI) paramagnetic beads at 2.5x the volume of PCR product (Axygen BioSciences, Corning, NY, USA).
-A 1:5 dilution in ultrapure water of the product was used as template for the second reaction.
-PCR products of the second reaction were purified using the Qiagen MinElute PCR Purification Kit (Qiagen, Hilden, Germany).
-Ultrapure water was used in place of template DNA and run along with each batch of PCRs to serve as a negative control for PCR; none of these produced visible bands on an agarose gel.
-In total, four separate replicates from each of 31 DNA samples were carried through the two-step PCR process for a total of 124 sequenced PCR products.
-These were combined with additional samples from other projects, totaling 345 samples for sequencing.
-
-\subsection*{DNA Sequencing}
-Up to 30 PCR products were combined according to their primer index in equal concentration into one of 14 pools, and 150 nanograms from each were prepared for library sequencing using the KAPA high-throughput library prep kit with real-time library amplification protocol (KAPA Biosystems, Wilmington, MA, USA).
-Each of these ligated sequencing adapters included an additional 6 base pair index sequence (NEXTflex DNA barcodes; BIOO Scientific, Austin, TX, USA).
-Thus, each PCR product was identifiable via its unique combination of index sequences in the sequencing adapters and primers.
-Fragment size distribution and concentration of each library was quantified using an Agilent 2100 BioAnalyzer.
-Libraries were pooled in equal concentrations and sequenced for 150 base pairs in both directions (PE150) using an Illumina NextSeq at the Stanford Functional Genomics Facility, where 20\% PhiX Control v3 was added to act as a sequencing control and to enhance sequencing depth by increasing sequence diversity.
-Raw sequence data in fastq format is publicly available (see Data Availability).
-
-\subsection*{Sequence Data Processing (Bioinformatics)}
-Detailed bioinformatic methods are provided in the supplemental material, and analysis scripts used from raw sequencer output onward can be found in the public project directory (see \nameref{analysis_scripts}).
-Briefly, we performed five steps to process the sequence data: (1) Merge paired-end reads, (2) eliminate low-quality reads, (3) eliminate PCR artifacts (chimeras), (4) cluster reads by similarity into operational taxonomic units (OTUs), and (5) match observed sequences to taxon names.
-Additionally, we checked for consistency among PCR replicates, excluded extremely rare sequences, and rescaled (rarefied) the data to account for differences in sequencing depth.
-The data for input to further analyses are a contingency table of the mean count of unique sequences, OTUs, or taxa present in each environmental sample.
-
-\subsection*{Ecological Analyses}
-After gathering the data, we use the eDNA community observed at each location to make inferences about the spatial patterning of eDNA communities.
-We use statistical tools from community ecology to assess the spatial structure of eDNA communities.
-We report similarity (1- dissimilarity) rather than dissimilarity in all cases for ease of interpretation.
-
-\subsection*{Objective 1: Community similarity as a function of distance}
-\subsubsection*{Distance Decay}
-To address our first objective and determine whether or not nearby samples are more similar than distant ones, we fit a nonlinear model to represent decreasing community similarity with distance.
-We calculated the pairwise Bray-Curtis similarity (1 - Bray-Curtis dissimilarity) between eDNA communities using the R package vegan \citep{vegan} and the great circle distance between sampling points using the Haversine method as implemented by the R package geosphere \citep{geosphere}.
-This model is similar to the Michaelis-Menten function, but with an asymptote fixed at 0:
-
-\begin{equation}\label{MM_asy0}
-	y_{ij} = \frac{AB}{B + x_{ij}}
-\end{equation}
-
-Where the relationship between community similarity ($y_{ij}$) and spatial distance ($x_{ij}$) between observations $i$ and $j$ is determined by the similarity of samples at distance 0 ($A$), and the distance at which half the total change in similarity is achieved ($B$).
-This allows for samples collected very close together (near 0) to have similarity significantly less than one.
-We assessed model fit using the R function nls \citep{R}, using the nl2sol algorithm from the Port library to solve separable nonlinear least squares using analytically computed derivatives (http://netlib.org/port/nsg.f).
-We set bounds of 0 and 1 for the intercept parameter and a lower bound of 0 for the distance at half similarity; starting values of these parameters were 0.5 and $x_{max}/2$, respectively.
-We calculated a 95\% confidence interval for the parameters and the predicted values using a first-order Taylor expansion approach implemented by the function predictNLS in the R package propagate \citep{propagate}.
-
-There are other conceptually reasonable forms to expect the space-by-similarlity relationship to take; we present these in the supplemental material along with alternative data subsets and similarity indices (see \nameref{supplement}).
-
-\subsection*{Objective 2: Spatial distribution of diversity}
-\subsubsection*{Community Classification}
-To determine the spatial distribution and variation of eDNA communities (objective 2), we used multivariate classification algorithms.
-We simultaneously assessed the existence of distinct community types and the membership of samples to those community types using an unsupervised classification algorithm known as partitioning around medoids (PAM; sometimes referred to as k-medoids clustering) \citep{Kaufman1990}, as implemented in the R package cluster \citep{cluster}.
-The classification of samples to communities was made on the basis of their pairwise Bray-Curtis similarity, calculated using the function vegdist in the R package vegan \citep{vegan}.
-Other distance metrics were evaluated but had no appreciable effect on the outcome of the analysis (Figure \ref{distance_decay_multi}).
-In order to chose an optimal number of clusters ($K$), we evaluated the distribution of silhouette widths, a measure of the similarity between each sample and its cluster compared to its similarity to other clusters.
-We repeated the analysis using fuzzy clustering (FANNY, \citep{Kaufman1990}; however, the results were qualitatively similar to the results using PAM so we omit them here.
-
-\subsubsection*{Aggregate Measures of Diversity}
-We calculated two measures of diversity, richness and evenness, to ask if aggregate metrics of the eDNA community showed evidence of spatial patterning.
-Richness is a measure of the number of distinct types of organisms present and so ranges from 1 (only one taxon observed) to $S$, the number of taxa observed across all samples.
-To calculate the evenness of the distribution of abundance of taxa in a sample, we used the complement of the \citet{Simpson1949} index ($1 - \Sigma{p_{i}^2}$, where $p_i$ is the proportional abundance of taxon $i$).
-The values of this index ranges from 0 to 1, with the value interpreted as the probability that two sequences randomly selected from the sample will belong to different taxa; thus, larger values of the index indicate more evenly divided communities \citep{Magurran2004}.
-We calculated Moran's I for both diversity metrics to test for spatial autocorrelation.
-We also tested for a linear effect of log-transformed distance from shore on each measure of diversity to ask how diversity changes over this strong environmental gradient.
-
-\subsubsection*{Taxon and Life History Patterns}
-After assigning taxon names to the abundance data, we plotted the distribution in space of a selection of taxa to compare with our expectations on the basis of adult distributions (objective 2).
-Our aim was to understand where each taxon occurred in the greatest proportional abundance, and its distribution in space relative to that maximum.
-Thus, we rescaled each sample to proportional abundance, extracted the data from a single taxon, and scaled those values between 0 and 1.
-We collated life history characteristics for each of the major taxonomic groups recovered, including dispersal range of the gametes, larvae, and adults, adult habitat type and selectivity, and adult body size.
-For each life history stage of each taxon group, we made an order-of-magnitude approximation of the scale of dispersal.
-For example, internally fertilized species were assigned a gamete range of 0 km, while broadcast spawners were assigned a gamete range of 10 km.
-Similarly, adult range size was approximated as 0 km (sessile), 1 km (motile but not pelagic), or 10 km (highly mobile, pelagic).
-Variables were specified as 'multiple' for life history stages known to span more than 1 magnitude of range size.
-For groups to which sequences were annotated with high confidence, but for which life history strategy is diverse or poorly known (e.g.
-families in the phylum Nemertea), we used conservative, coarse approximations at a higher taxonomic rank (see \nameref{life_history_data}).
-These data were used to contextualize group-specific spatial distributions and inform expectations based on known adult distributions.
-
-%-----------------------------------RESULTS-------------------------------------
-\section*{Results}
-\subsection*{Sequence Data Processing (Bioinformatics)}
-Preliminary sequence analysis strongly suggested that the observed variation among environmental samples reflects true variation in the environment, rather than variability due to lab protocols, for the following reasons (note that all value ranges are reported as mean plus and minus one standard deviation).
-First, all libraries passed the FastQC per-base sequence quality filter, generating a total of 371,576,190 reads passing filter generated in each direction.
-Second, samples in this study were represented by an adequate number of reads (333,537.9 $\pm$ 112,200.5), with no individual sample receiving fewer than 130,402 reads.
-Third, there was a very low frequency of cross-contamination from other libraries into those reported here (5e-05$\pm$8e-05; max proportion 0.00034).
-Fourth, after scaling all samples to the same sequencing depth, OTUs with abundance greater than 178 reads (0.14\% of a sample's reads) experienced no turnover among PCR replicates within a sample.
-Fifth, sequence abundances among PCR replicates within water samples were remarkably consistent.
-A single sample had low similarity among PCR replicates (0.659) after removing this outlier, the lowest mean similarity among replicates within a sample was 0.966.
-Overall similarities among PCR replicates within a sample were extremely high (0.976 $\pm$ 0.013), and far higher than those among samples (0.3 $\pm$ 0.16).
-Across PCR replicates, each sample was represented by at least 781425 reads in the raw data and contained between 111 and 443 rarefied OTUs (Supplemental Figure \ref{rarefaction}).
-
-\subsection*{Ecological Analyses}
-\subsubsection*{Distance Decay}
-Physical proximity is a good predictor of eDNA community similarity: Similarity decreased from 0.40 (95\%CI = 0.36, 0.45) to half that amount at 4500 meters (95\%CI = 2900, 7500) (Figure \ref{distance_decay}).
-
-\subsubsection*{Community Classification}
-Despite a clear trend in community similarity as a function of spatial separation, the results from our classification analysis are difficult to interpret.
-The silhouette analysis indicated the presence of 8 distinct communities; however, the gain in mean silhouette width from 2 was small (0.1), and lacked a distinctive peak (Figure \ref{pam_sil}), indicating substantial uncertainty in the clustering algorithm.
-Thus, we present the results of cluster assignment for both $K$ = 2 and $K$ = 8 to illustrate the range of results (Figure \ref{pam_in_space}).
-Excluding taxa which occur in only one site had no discernible effect on the outcome of the PAM analysis (number of clusters, assignment to clusters).
-While there was no distinct spatial divide indicating the presence of an inshore versus an offshore community, one of the two communities (at $K$ = 2) occurred in only 2 out of 18 samples inside 1000 meters from shore, and never occurred within 125 meters of shore, suggesting the presence of an inshore and offshore community.
-
-\subsubsection*{Diversity in Space}
-Sites offshore tend to be less rich and more even than those inshore (Figure \ref{diversity_distance}).
-Mean OTU richness declined by 1.42 per 1000 meters from a mean of 17.6 taxa (95\%CI = 2.15) inshore to 11.9 taxa (95\%CI = 4.31) at offshore locations (p = 0.0415; Figure \ref{diversity_distance}).
-Evenness increased by .0666 per 1000 meters from 0.225 (95\%CI = 0.0558) to 0.491 (95\%CI = $\pm$ 0.112), indicating that sequence reads were less evenly distributed among taxa in offshore samples (p $\ll$ 0.05; Figure \ref{diversity_distance}).
-There was no evidence for spatial autocorrelation for any of the diversity metrics (Moran's I, p > 0.05; Figure \ref{diversity}).
-
-\subsubsection*{Taxon and Life History Patterns}
-We were able to assign a taxon name with confidence to 136 of 146 OTU sequences.
-The vast majority of sequences (97.6\%) and OTUs (96.9\%) were matched to organisms that have high potential for dispersal at either the gamete, larval, or adult stage, making it impossible to determine whether the source of that DNA was adults with well-documented spatial patterns (e.g.
-sessile nearshore specialists) or highly mobile early life history stages.
-Of the 6 OTUs for which dispersal is limited during all life history stages, only 2 occurred in more than two samples, precluding a quantitative comparison of spatial dispersion based on life history characteristics.
-These were assigned to \textit{Cymatogaster aggregata}, a viviparous nearshore fish with internal fertilization, and \textit{Cupolaconcha meroclista}, a sessile Vermetid gastropod with presumed internal fertilization and short larval dispersal \citep{Strathmann2006, Phillips2010, Calvo2004}.
-\textit{Cymatogaster aggregata} was distinctly more abundant close to shore, with no sequences occurring in any sample beyond 250 meters (Figure \ref{otu_in_space_select}).
-\textit{Cupolaconcha meroclista} showed no such distinct spatial trend, occurring in nearly equal abundance at three sites, 75, 500, and 2000 meters from shore.
-An additional species that was highly abundant in the sequence data, the krill \textit{Thysanoessa raschii}, has pelagic adults, highly seasonal reproduction, and sinking eggs; their distribution was consistent with our expectations based on a tendency of adults to aggregate offshore.
-Finally, the two most abundant taxa in the dataset were the mussel genus \textit{Mytilus} and the Barnacle order Sessilia; the adults of both taxa are sessile and occur exclusively on hard intertidal substrata but have highly motile larvae.
-Because large-scale dispersal could not be ruled out for the vast majority of taxa, subsetting the community data by taxonomic group had no qualitative effect on the spatial patterning or diversity metrics, and we omit those results here.
-
-%----------------------------------DISCUSSION-----------------------------------
-\section*{Discussion}
-Indirect surveys of organismal presence are a key development in ecosystem monitoring in the face of increased anthropogenic pressure and dwindling resources for ecological research.
-Monitoring of organisms using environmental DNA is an especially promising method, given the rapid pace of advancement in technological innovation and cost efficiency in the field of DNA sequencing and quantification.
-We document four key patterns: (1) eDNA communities far from one another tend to be less similar than those that are nearby, (2) distinct eDNA communities exist and are distributed in a non-random fashion, (3) diversity declines with distance from shore, and (4) spatial patterning of eDNA is associated with taxon-specific life history characteristics.
-
-\subsubsection*{(1) Communities far from one another tend to be less similar than those that are nearby}
-We demonstrate that distant locations have less-similar eDNA communities than proximate locations in Puget Sound, a dynamic marine environment.
-Our finding is in line with observations based on traditional surveys of terrestrial plants and fungi \citep{Nekola1999, Bahram2013, Condit2002, Chust2006} and of microorganisms in freshwater \citep{Wetzel2012}, marine \citep{Chust2013}, and estuarine \citep{Martiny2011} environments.
-To our knowledge, it is the first to report such a pattern using massively parallel sequencing of environmental DNA in the marine environment, and the first using any technique to describe this pattern from macrobial metazoans.
-We note that the theoretical expectation is that samples at very close distance be nearly completely similar, while our samples separated by the 50 meters were only 40\% similar.
-We interpret this to reflect the highly dynamic nature of this environment, which could cause DNA to be distributed quickly from its source, eroding the rise in similarity at small distances.
-At the same time, community similarity decreased to very low levels at larger scales, indicating that DNA distribution is not completely unpredictable.
-This finding implies that the effectively sampled area of individual water samples for eDNA analysis is likely to be quite small (<100m) in this nearshore environment.
-Our estimated distance-decay relationship does indicate that proximate samples are more similar than distant samples, but we suggest this pattern is partially obscured by other factors, including signal from mobile, microscopic life-stages.
-
-\subsubsection*{(2) Distinct eDNA communities exist and are distributed in a non-random fashion}
- We demonstrate strong evidence for distinct community types and the non-random spatial patterning of those communities.
-While the spatial distributions of communities is surprising if one were concerned only with the macroscopic life stages of metazoans, it indeed does align with the broader view that even offshore pelagic communities are comprised of and influenced by nearshore organisms.
-This result underscores the idea that areas immediately offshore act as ecotones, a mixing zone of taxa characteristic of benthic and pelagic environments.
-While there was no distinct break in community types between onshore and offshore sites, there was some clustering of community types that may be explained by oceanographic features such as nearshore eddies generated by strong tidal exchange in a steep bathymetric setting \citep{Yang2010}.
-It would be useful to better understand such features during the period of sampling, by way of oceanographic monitoring devices.
-Finally, the uncertainty in identification of the number of distinct clusters to best characterize the community underlines the difficulty of identifying community patterns with the number of taxonomic groups considered here.
-We suspect that the signature of eDNA from microscopic life-stages may explain our inability to easily detect spatial community level patterns that align with our initial expectations.
-
- \subsubsection*{(3) Richness declines and evenness increases with distance from shore}
-We found that richness declined while evenness increased with distance from shore.
-Such a pattern is consistent with many other ecosystems which show strong clines in diversity metrics over environmental gradients.
-However, our study is novel in that it corroborates a cline well-known on macroscales for macrobiota on a much smaller spatial scale for microscopic animals, suggesting that there may be a self-similarity across scales in diversity patterning \citep{Levin1992}.
-The coastal ocean is a highly productive and diverse ecosystem, where biomass is concentrated most heavily along the bottom and shoreline \citep{Ray1988}.
-This differential in biomass concentration from the shoreline to open waters may contribute to the opposing trends we detected.
-Where particles (organisms, tissues, and cells) are sparse, fewer would be collected per sample of constant volume, thus decreasing the probability of drawing as many types (richness) and increasing the probability that any two particles originate from the same type (evenness).
-Intriguingly, the cline in diversity from inshore to offshore was not determined by shared changes in communities as one moved offshore; the classification analysis suggested a fair amount of differences among communities at a given offshore distance (Figure \ref{pam_in_space}).
-
-\subsubsection*{(4) Spatial patterning of eDNA is associated with taxon-specific life history}
-In contrast to our expectations, other taxa including species with sessile adult stages restricted to benthic hard substrates (e.g.
-barnacles, mussels) are among the most abundant taxa at sites furthest from shore.
-However, the larvae and gametes of these taxa are abundant, pelagic, and can be transported long distances by water movement \citep{Strathmann1987}.
-This indicates that we likely detected DNA of their pelagic phase gametes and larvae.
-It is always possible that DNA of adults was advected over long distances and detected offshore but in light of our results with krill and surfperch, we view this as unlikely.
-We interpret our results as evidence that the chaotic spatial distribution of eDNA communities (Figure \ref{pam_in_space}) results from our primers' affinity for many species which at some point exist as microscopic pelagic gametes or larvae.
-Our results emphasize that expected results based on easily visually observed individuals or detectable with traditional sampling gear such as nets may be very different from results using eDNA.
-This does caution that eDNA surveys may have different purposes and may not be directly comparable to existing surveys \citep{Shelton2016}.
-
-We acknowledge that sampling artifacts may have affected our results.
-For example if entire multicellular individuals were captured in our samples, their DNA could be in much greater density than eDNA, affecting the observed community.
-Our sampling bottles excluded particles larger than 500 micrometers, but gametes and very small larvae could have gained entry.
-It is possible that even a single small individual, containing many thousand mitochondria, would overwhelm the signal of another species from which hundreds of cells had been sloughed from many, larger individuals.
-Data on larval size distribution at the time of sampling from each species in our data set would allow us to estimate the frequency of such events.
-Nevertheless, it is precisely the sensitivity to small particles that makes the eDNA approach powerful, so we are reluctant to recommend that aquatic eDNA sampling use finer pre-filtering.
-Instead, we emphasize the importance of designing and selecting primer sets that selectively amplify target organisms.
-In the case of the present study, in order to recover patterns matching our expectations, this would be non-transient, benthic marine organisms lacking any pelagic life stage.
-
-The marker we chose for this study detects a wide variety of metazoans while excluding other more common taxa; however, it does not effectively discriminate among species within a higher group in all cases.
-Other markers, such as mitochondrial cytochrome c oxidase subunit 1 (COX1, CO1, or COI) may provide adequate species-level resolution in some metazoan groups, but have other shortcomings including taxon dropout \citep{Deagle2014} and amplification of more abundant non-metazoans, as we discovered in an accompanying study \citep{Kelly2016FMS}.
-Both have undesirable effects of biasing estimates of diversity.
-In our case, it is possible that the lumping of multiple species into one group underestimates the true richness of the group and of the entire sample, in turn obscuring true underlying patterns of diversity.
-In the case of COX1, well-documented primer biases cause failure to amplify some taxa, particularly in mixed samples, with the same result \citep{Deagle2014}.
-In fact, even surveys relying on traditional capture techniques (e.g.
-seine nets) and morphological characteristics are subject to biases imposed by the sampling gear (e.g.
-mesh size), the observer (e.g.
-taxonomic expertise), and organisms (e.g.
-morphologically cryptic species).
-Similarly, no single molecular marker adequately and effectively samples all taxa without bias  \citep{Drummond2015}, and thus the choice of marker is an important and context-dependent one.
-Until whole-genome sequencing of individual cells is a reality, the tradeoffs between taxonomic breadth and resolution will continue to be problematic for metabarcoding studies, just as they are for more traditional ecological survey methods \citep{Kelly2016FMS}.
-
-Our results also highlight the need for curated life-history databases.
-As technological advances increase the speed and throughput of DNA sequencing and sequence processing, making sense of these data in a timely manner requires that natural history data be stored in standard formats in centralized repositories.
-The rate at which we can make sense of high-throughput survey methods will be limited by our ability to collate auxiliary data.
-Databases such as Global Biodiversity Information Facility (GBIF), Encyclopedia of Life (EOL), and FishBase \citep{eol,  fishbase} contain records of taxonomy, occurrence, and other rudimentary data types, but there is no centralized, standardized repository for even basic natural history data such as body size.
-As NCBI's nucleotide and protein sequence database (GenBank) has facilitated transformative studies in diverse fields, an ecological analog would be a boon for biodiversity science.
-
-Surveys based on eDNA are intensely scrutinized because of the danger that the final data are subject to complicated laboratory and bioinformatic procedures.
-Finding virtually no variability among lab and bioinformatic treatments from the point of PCR onward, we were confident our results represented actual field-based differences among samples.
-However, we note that one PCR replicate had a clear signal of contamination in that the sequence community was extremely similar to those from a different environmental sample.
-The source of this error is difficult to identify, but seems most likely to be an error during PCR preparation, either in assignment or pipetting during preparation of indexed primers.
-While the remainder of our results would be largely unchanged had we sequenced a single replicate per environmental sample, we believe the sequencing of PCR replicates is critical for ensuring data quality in eDNA sequencing studies.
-
-While there is much work to be done before eDNA techniques can be confidently deployed as a standard method for ecological monitoring, this study serves as a first analysis of diversity at the fine spatial scales that are likely to be relevant to eDNA work in the field across a range of study systems.
-
-%-------------------------------ACKNOWLEDGEMENTS--------------------------------
-\section*{Acknowledgements}
-We wish to thank Linda Park, Robert Morris, E. Virginia Armbrust, and James Kralj.
-The manuscript was improved by suggestions from editor Magnus Johnson, and reviewers Owen Wangensteen and Stephen Moss.
-
-%-----------------------------------FUNDING-------------------------------------
-\section*{Funding}
-This work was supported by a grant from the David and Lucile Packard Foundation to RPK (grant 2014-39827).
-The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
-
-%--------------------------------CONTRIBUTIONS----------------------------------
-\section*{Author Contributions}
-Conceived and designed the experiments:
-JL O'Donnell, RP Kelly, AO Shelton;
-Collected the data:
-JL O'Donnell, NC Lowell, GD Williams, RP Kelly, AO Shelton, JF Samhouri;
-Conducted the analyses:
-JL O'Donnell;
-Wrote the first draft:
-JL O'Donnell;
-Edited the manuscript:
-JL O'Donnell, AO Shelton, RP Kelly, JF Samhouri, GD Williams, NC Lowell
-
-%-----------------------------------ETHICS--------------------------------------
-\section*{Ethics Statement}
-The authors declare no conflict of interest.
-Consistent with the public trust doctrine, waters of the US are public, and therefore no permit was necessary to conduct this research (see Illinois Central Railroad v. Illinois, 146 U.S. 387 (1892)).
-
-%-------------------------------------DATA-------------------------------------%
-\section*{Data Availablity}
-\label{data}
-
-\subsection*{Sequence Data}
-\label{sequence_data}
-All sequence data are available from NCBI under BioProject PRJNA338801.\\
-Scripts to process raw sequence data into the contingency tables used for ecological analyses can be found at: \\
-\verb!https://github.com/jimmyodonnell/banzai!\\
-
-\subsection*{Project Repository}
-The following components are available from the project repository on GitHub: \\
-\verb!https://github.com/jimmyodonnell/Carkeek_eDNA_grid!\\
-\verb!http://dx.doi.org/FIXME!
-
-\subsubsection*{Sequencing Metadata}
-\label{sequencing_metadata}
-Sequencing metadata is available in: \verb!Data/metadata_spatial.csv!
-
-\subsubsection*{Life History Data}
-\label{life_history_data}
-Life history data is available in: \verb!Data/life_history.csv!
-
-\subsubsection*{Analysis Scripts}
-\label{analysis_scripts}
-All analyses were performed using scripts available in the Analysis subdirectory of the project's repository on GitHub.
-
-%----------------------------------REFERENCES-----------------------------------
-%\section*{References} % commented out because the section title is automatically inserted if using an automatically-generated bibliography
-
-\bibliographystyle{apalike} % or: plain,unsrt,alpha,abbrv,acm,apalike,ieeetr
-\bibliography{carkeek_grid} % path to your .bib file excluding .bib extension
-
-%-----------------------------------FIGURES-------------------------------------
-\pagebreak
-\section*{Figures}
-
-\begin{figure}[H] % [h!] forces the figure to be placed roughly here
-  \centering
-    \includegraphics[width=1\textwidth]{site_map.pdf}
-    \caption{\protect\input{"../../Figures/site_map_legend.txt"}}
-  \label{site_map}
-\end{figure}
-
-\begin{figure}[H] % [h!] forces the figure to be placed roughly here
-  \centering
-    \includegraphics[width=1\textwidth]{distance_decay.pdf}
-    \caption{\protect\input{"../../Figures/distance_decay_legend.txt"}}
-  \label{distance_decay}
-\end{figure}
-
-\begin{figure}[H] % [h!] forces the figure to be placed roughly here
-  \centering
-    \includegraphics[width=1\textwidth]{pam_in_space.pdf}
-    \caption{\protect\input{"../../Figures/pam_in_space_legend.txt"}}
-  \label{pam_in_space}
-\end{figure}
-
-\begin{figure}[H] % [h!] forces the figure to be placed roughly here
-  \centering
-    \includegraphics[width=1\textwidth]{pam_sil.pdf}
-    \caption{\protect\input{"../../Figures/pam_sil_legend.txt"}}
-  \label{pam_sil}
-\end{figure}
-
-\begin{figure}[H] % [h!] forces the figure to be placed roughly here
-  \centering
-    \includegraphics[width=1\textwidth]{diversity.pdf}
-    \caption{\protect\input{"../../Figures/diversity_legend.txt"}}
-  \label{diversity}
-\end{figure}
-
-\begin{figure}[H] % [h!] forces the figure to be placed roughly here
-  \centering
-    \includegraphics[width=1\textwidth]{diversity_distance.pdf}
-    \caption{\protect\input{"../../Figures/diversity_distance_legend.txt"}}
-  \label{diversity_distance}
-\end{figure}
-\pagebreak
-
-\begin{figure}[H] % [h!] forces the figure to be placed roughly here
-  \centering
-    \includegraphics[width=1\textwidth]{otu_in_space_select.pdf}
-    \caption{\protect\input{"../../Figures/otu_in_space_select_legend.txt"}}
-  \label{otu_in_space_select}
-\end{figure}
 
@@
 \pagebreak
-\section*{Supplemental Material}
-\label{supplement}
+% \section*{Supplemental Material}
+% \label{supplement}
 
-\subsection*{Methods}
-\subsubsection*{Bioinformatics}
+\section*{Methods}
+\subsection*{Bioinformatics}
 Reads passing the preliminary Illumina quality filter were demultiplexed on the basis of the adapter index sequence by the sequencing facility.
 We used fastqc to assess the fastq files output from the sequencer for low-quality indications of a problematic run.
-Forward and reverse reads were merged using PEAR v0.9.6 \cite{Zhang2014} and discarded if more than 0.01 of the bases were uncalled.
+Forward and reverse reads were merged using PEAR v0.9.6 \citep{Zhang2014} and discarded if more than 0.01 of the bases were uncalled.
 If a read contained two consecutive base calls with quality scores less than 15 (i.e.
@@
 
-We used vsearch v2.1.1 \citep{vsearch} to discard any merged reads for which the sum of the per-base error probabilities was greater than 0.5 (``expected errors'') \cite{Edgar2010}.
+We used vsearch v2.1.1 \citep{vsearch} to discard any merged reads for which the sum of the per-base error probabilities was greater than 0.5 (``expected errors'') \citep{Edgar2010}.
 Sequences were demultiplexed on the basis of the primer index sequence at base positions 4-9 at both ends using the programming language AWK.
-Primer sequences were removed using cutadapt v1.7.1 \cite{Martin2011}, allowing for 2 mismatches in the primer sequence.
+Primer sequences were removed using cutadapt v1.7.1 \citep{Martin2011}, allowing for 2 mismatches in the primer sequence.
 Identical duplicate sequences were identified, counted, and removed in python to speed up subsequent steps by eliminating redundancy, and sequences occurring only once were removed.
-We checked for and removed any sequence likely to be a PCR artifact due to incomplete extension and subsequent mis-priming using a method described by \citet{Edgar2010} and implemented in vsearch v2.0.2.
+We checked for and removed any sequence likely to be a PCR artifact due to incomplete extension and subsequent mis-priming using a method described by \citet{Edgar2010} and implemented in vsearch v2.1.1 \citep{vsearch}.
 Sequences were clustered into operational taxonomic units (OTUs) using the single-linkage clustering method implemented by swarm version 2.1.1 with a local clustering threshold (d) of 1 and fastidious processing \citep{swarm}.
@@
 
-\subsubsection*{Alternative distance decay model formulations}
+\subsection*{Alternative distance decay model formulations}
 
@@
 
-\subsection*{Supplemental Figures}
+%----------------------------------REFERENCES-----------------------------------
+%\section*{References} % commented out because the section title is automatically inserted if using an automatically-generated bibliography
+
+\bibliographystyle{apalike} % or: plain,unsrt,alpha,abbrv,acm,apalike,ieeetr
+\bibliography{carkeek_grid} % path to your .bib file excluding .bib extension
+
+\section*{Supplemental Figures}
 
@@
   \centering
+    \includegraphics[height=0.6\textheight]{rarefaction.pdf}
+    \caption{\protect\input{"../../Figures/rarefaction_legend.txt"}}
+  \label{rarefaction}
+\end{figure}
+
+\begin{figure}[H] % [h!] forces the figure to be placed roughly here
+  \centering
     \includegraphics[width=1\textwidth]{diversity_distance_all.pdf}
@@
 
-\begin{figure}[H] % [h!] forces the figure to be placed roughly here
-  \centering
-    \includegraphics[height=0.6\textheight]{rarefaction.pdf}
-    \caption{\protect\input{"../../Figures/rarefaction_legend.txt"}}
-  \label{rarefaction}
-\end{figure}
-
-% \pagebreak
-% \section*{Orphaned Text}
-% Samples collected of ecological communities may vary in dissimilarity from 0 (completely identical) to 1 (completely different).
-For samples collected from multiple locations, the relationship between their spatial distance and community dissimilarity is of interest because it reflects the amount of community heterogeneity---changes in abundance and composition---over the spatial scale sampled.
-% In general, the relationship between community dissimilarity (0 = identical; 1 = completely different) and spatial distance is expected to be asymptotic, because communities nearer to each other tend to be more similar than those farther apart.
-The intercept is expected to be 0, because only within-sample comparisons can have 0 spatial separation, and communities have no dissimilarity within a sample if sampling method is repeatable.
-% (technically impossible for samples to be taken at exactly 0; they are actually immediately adjacent.)
-% Deviation from 0 indicates heterogeneous community composition/structure over fine scales.
-Likewise, dissimilarity cannot exceed 1, and reaches 1 only when multiple discrete community types are sampled.
-The trend is expected to be asymptotic if communities within a habitat are spatially structured, where the value of the asymptote and the rate of increase provide insight about community turnover.
-A flat relationship indicates that heterogeneity is not assorted spatially, and can be interpreted in different ways, depending on the asymptote.
-% ?or no relationship both
-% If the asymptote is close to 1, there is high spatial heterogeneity over the spatial scale of sampling.
-If the asymptote is close 0, all samples are similar, and we infer there is complete community homogeneity over the scale sampled.
-The rate at which community dissimilarity approaches the mean gives an indication of the rate of community turnover.
-%
-% In theory, given enough data from manipulative experiments, it is possible that one could measure environmental variables--temperature, UV intensity, flow, etc--at the time of eDNA sample collection in order to more precisely estimate the origin of eDNA.
-In practice, this approach is unlikely to yield appreciable gains in accuracy because even the most advanced sensors (ADCPs) and analytical techniques available cannot determine the origin of particles in highly dynamic fluid environments at scales relevant to most ecological questions \citep{Piterbarg2001}.
-%http://www.rsmas.miami.edu/LAPCOD/research/prediction/
-% Further, collecting and analyzing these data would severely limit the scalability of eDNA methods.
-A more practical approach is to compare the spatial distribution of communities of eDNA with expectations based on prior knowledge of the spatial distribution of communities of organisms.
-
 \end{document}

### 2017-01-20 Manuscript/v1.2_PeerJ/carkeek_grid.tex
fix caption problems

@@
 \usepackage{breakcites} % add line breaks in citations where necessary
-\usepackage[breaklinks=true]{hyperref}
+% \usepackage[breaklinks=true]{hyperref}
 
@@
 There are seven discrete steps to our methodology: (1) Environmental sample collection, (2) isolation of particulates from water via filtration, (3) isolation of DNA from filter membrane, (4) amplification of target locus via PCR, (5) sequencing of amplicons, (6) bioinformatic translation of raw sequence data into tables of sequence abundance among samples, and (7) community ecological analyses of eDNA.
-We provide brief overviews of these steps here, and encourage the reader to review the fully detailed methods presented in the supplementary material (\nameref{supplement}).
+We provide brief overviews of these steps here, and encourage the reader to review the fully detailed methods presented in the supplementary material.
 
@@
     \includegraphics[width=1\textwidth]{site_map.pdf}
-    \caption{\protect\input{"../../Figures/site_map_legend.txt"}}
+    \caption{
+		Map of study area.
+		Depth in meters below sea level is indicated by shading and 25 meter contours. Sampled locations are indicated by red points.
+		}
   \label{site_map}
@@
 
+\newpage
 \begin{figure}[H] % [h!] forces the figure to be placed roughly here
@@
     \includegraphics[width=1\textwidth]{distance_decay.pdf}
-    \caption{\protect\input{"../../Figures/distance_decay_legend.txt"}}
+    \caption{
+		Distance decay relationship of environmental DNA communities.
+		Each point represents the Bray-Curtis similarity of a site sampled along three parallel transects comprising a 3000 by 4000 meter grid.
+Blue dashed line represents fit of a nonlinear least squares regression (see Methods), and shading denotes the 95\% confidence interval.
+Boxplot is comparisons within-sample across PCR replicates, separated by a vertical line at zero, where the central line is the median, the box encompasses the interquartile range, and the lines extend to 1.5 times the interquartile range.
+Boxplot outliers are omitted for clarity.
+		}
   \label{distance_decay}
@@
     \includegraphics[width=1\textwidth]{pam_sil.pdf}
-    \caption{\protect\input{"../../Figures/pam_sil_legend.txt"}}
+    \caption{
+		Silhouette widths from PAM analysis.
+		Points are the width of the PAM silhouette of each sample at each number of clusters (K).
+Red line is the mean, blue line is the median.
+Boxes encompass the interquartile range with a line at the median, and the whiskers extend to 1.5 times the interquartile range.
+Boxplot outliers are omitted for clarity.
+		}
 		\label{pam_sil}
@@
     \includegraphics[width=1\textwidth]{pam_in_space.pdf}
-    \caption{\protect\input{"../../Figures/pam_in_space_legend.txt"}}
+    \caption{
+		Cluster membership of sampled sites.
+		Distance from onshore starting point is log scaled.
+Sites are colored and labeled by their assignment to a cluster by PAM analysis for number of clusters (K) chosen based on a priori expectations (A; 2) and mean silhouette width (B; 8).
+		}
   \label{pam_in_space}
@@
     \includegraphics[width=1\textwidth]{diversity_distance.pdf}
-    \caption{\protect\input{"../../Figures/diversity_distance_legend.txt"}}
+    \caption{
+		Aggregate diversity metrics of each site plotted against distance from shore.
+		Both Simpson's Index of evenness (A) and richness (B) are shown, and have been computed from the mean abundance of unique DNA sequences found across 4 PCR replicates at each of 24 sites.
+		Lines and bands illustrate the fit and 95\% confidence interval of a linear model.
+		}
   \label{diversity_distance}
@@
     \includegraphics[width=1\textwidth]{diversity.pdf}
-    \caption{\protect\input{"../../Figures/diversity_legend.txt"}}
+    \caption{
+		Aggregate measures of diversity at each sample site.
+		Data are rarefied counts of mitochondrial 16S sequences collected from 3 parallel transects in Puget Sound, Washington, USA.
+		Evenness (A) is the probability that two sequences drawn at random are different; richness (B) represents the total number of unique sequences from that location.
+		}
   \label{diversity}
@@
     \includegraphics[width=1\textwidth]{otu_in_space_select.pdf}
-    \caption{\protect\input{"../../Figures/otu_in_space_select_legend.txt"}}
+    \caption{
+		Distribution of eDNA from select taxa.
+		Taxa represented are: Embiotocidae (A), Cupolaconcha meroclista (B), Thysanoessa raschii (C), Mytilus (D), Sessilia (E).
+Circles are colored and scaled by the proportion of that taxon's maximum proportional abundance.
+		That is, the largest circle is the same size in each of the panels, and occurs where that taxon contributed the greatest proportional abundance of reads to that sample.
+		}
   \label{otu_in_space_select}

### 2017-01-20 Manuscript/v1.2_PeerJ/supplement.tex
fix caption problems

@@
     \includegraphics[height=0.6\textheight]{distance_decay_multi.pdf}
-    \caption{\protect\input{"../../Figures/distance_decay_multi_legend.txt"}}
+    \caption{
+		Distance decay relationship of environmental DNA communities using a variety of models, metrics, and data subsets.
+		Each point represents the similarity of a site sampled along three parallel transects comprising a 3000 by 4000 meter grid.
+		Each row of plots represents a different data subset indicated in the right margin, including	the final filtered data reported in the main text (a-d), the unfiltered data including all rare OTUs (e-h), log-transformed (log(x+1)) data (i-l),	OTU abundance scaled relative to within-taxon maximum (m-p), and exclusion of OTUs found at only one site (q-t).
+		Columns indicate the similarity index used (Bray Curtis or Morisita-Horn) and whether the input was full abundance data or binary (0,1) transformed data.
+		Lines and bands illustrate the fit and 95\% confidence interval of both the	main nonlinear model (red, dashed line) and a simple linear model (blue, solid line).
+		Results using the Jaccard distance are omitted because of its similarity to Bray-Curtis.
+		}
 		\label{distance_decay_multi}
@@
     \includegraphics[height=0.6\textheight]{rarefaction.pdf}
-    \caption{\protect\input{"../../Figures/rarefaction_legend.txt"}}
+    \caption{
+		Accumulation of OTUs from 24 environmental samples using randomized rarefaction.
+		Four replicate PCRs were conducted using DNA each environmental sample and independently sequenced, but these are collapsed here to illustrate a single representation of richness.
+		Sample names indicate the position in the sampling grid: south (S), central (C), or north (N), followed by the distance along the transect, in meters (0, 75, 125, 250, 500, 1000, 2000, 4000).
+		Vertical line indicates the minimum combined number of sequence reads per sample.
+		Horizontal lines indicate OTU richness for each sample at the minimum combined number of sequence reads.
+		}
   \label{rarefaction}
@@
     \includegraphics[width=1\textwidth]{diversity_distance_all.pdf}
-    \caption{\protect\input{"../../Figures/diversity_distance_all_legend.txt"}}
+    \caption[]{
+		Aggregate diversity metrics of each site plotted against distance from shore.
+		Both Simpson's Index (top; A-F) and richness (bottom; G-L) are shown for a variety of data subsets and transformations (left to right: mean (A,G), unfiltered mean (B,H), log(x + 1) transformed (C,I), scaled (D,J), spatially variable (E,K), and taxon clustered (F,L)).
+		Lines and bands illustrate the fit and 95\% confidence interval of a linear model.
+		See methods text for detailed data descriptions.
+		}
   \label{diversity_distance_all}

### 2017-02-07 Manuscript/v1.3_PeerJ_proofs/carkeek_grid.tex
start fixing proofs

--- 2017-01-20 Manuscript/v1.2_PeerJ/carkeek_grid.tex
@@
 In order to chose an optimal number of clusters ($K$), we evaluated the distribution of silhouette widths, a measure of the similarity between each sample and its cluster compared to its similarity to other clusters.
-We repeated the analysis using fuzzy clustering (FANNY, \citep{Kaufman1990}; however, the results were qualitatively similar to the results using PAM so we omit them here.
+We repeated the analysis using fuzzy clustering (FANNY, \citep{Kaufman1990}); however, the results were qualitatively similar to the results using PAM so we omit them here.
 
